### 2009-06-04 Documentation/Tex/ants.tex
tex and documentation
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@5 fd2da0a4-d968-48df-9cb5-d8c496287cbf

%
% Complete documentation on the extended LaTeX markup used for Insight
% documentation is available in ``Documenting Insight'', which is part
% of the standard documentation for Insight.  It may be found online
% at:
%
%     http://www.itk.org/

\documentclass{InsightArticle}

\usepackage[dvips]{graphicx}
%\usepackage{listing}						
\usepackage{listings}	
\usepackage{wrapfig}
\usepackage{amssymb,amsmath}
\usepackage{multirow,booktabs,ctable,array}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  hyperref should be the last package to be loaded.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[dvips,
bookmarks,
bookmarksopen,
backref,
colorlinks,linkcolor={blue},citecolor={blue},urlcolor={blue},
]{hyperref}

\graphicspath{{./Figures/}}
\newcommand{\fv}{\mbox{\boldmath $f$}}
\newcommand{\Iv}{\mbox{\boldmath $I$}}
\newcommand{\xv}{\mbox{\boldmath $x$}}
\newcommand{\velocity}{\mbox{\boldmath $v$}}
\newcommand{\velocityinv}{\mbox{\boldmath $v$}^{-1}}
\newcommand{\Velocityinv}{\mbox{\boldmath $V$}^{-1}}
\newcommand{\Velocity}{\mbox{\boldmath $V$}}
\newcommand{\welocity}{\mbox{\boldmath $w$}}
\newcommand{\spatialimagegradient}{\mbox{\boldmath $I$}\mbox{$_{x}$}}
\newcommand{\perpspatialimagegradient}{\mbox{\boldmath $I$}\mbox{$^{\perp}_{x}$}}
\newcommand{\temporalimagederivative}{\mbox{$I_t$}}
\newcommand{\finitestrain}{\mbox{\boldmath $\varepsilon^{\ast}$}}
\newcommand{\smallstrain}{\mbox{\boldmath $\varepsilon$}}
\newcommand{\dispgradient}{{\bf \nabla \displace}}
\newcommand{\displace}{{\bf u}} 
\newcommand{\Id}{\text{\bf Id}}
\newcommand{\ode}{{\em O.D.E.}}
\newcommand{\odes}{{\em O.D.E.}s}
\newcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\phiinv}{\phi^{-1}}
\newcommand{\psiinv}{\psi^{-1}}
\newcommand{\dM}{DM}
\newcommand{\DM}{Diffeomorphometry}
\newcommand{\diff}{diffeomorphism}
\newcommand{\Diff}{Diffeomorphism}
\newcommand{\orbit}{\mathcal{O}}
\newcommand{\avg}{\mathcal{A}}
\newcommand{\avgn}{\mathcal{A}_n}
\newcommand{\avgna}{\mathcal{A}^a_{n}}
\newcommand{\nset}{\{ J_i \}_n}
\newcommand{\bari}{\bar{I}}
\newcommand{\bart}{\bar{t}}
\newcommand{\jac}{\mathcal{J}}
\newcommand{\pert}{\mbox{\boldmath $w$}}
\newcommand{\barpert}{\bar{\mbox{\boldmath $w$}}}
\newcommand{\half}{0.5}

\newcommand{\X}{{\bf X}}
\newcommand{\x}{{\bf x}}
\newcommand{\Z}{{\bf Z}}
\newcommand{\z}{{\bf z}}
\newcommand{\p}{{\bf p}}
\newcommand{\Y}{{\bf Y}}
\newcommand{\y}{{\bf y}}
\newcommand{\disp}{{\bf u}}
\newcommand{\ytild}{\tilde{{\bf y}}}
\newcommand{\q}{{\bf q}}
\newcommand{\surf}{\mathcal{S}}
\newcommand{\phij}{\phi_{ij}}
\newcommand{\aphij}{\bar{\phi}_{j}}
\newcommand{\apsij}{\bar{\psi}_{j}}
\newcommand{\aphi}{\bar{\phi}}
\newcommand{\aphiinv}{\bar{\phi}^{-1}}

\newcommand{\domain}{\Omega}
\newcommand{\meanshape}{\bf \bar{x}}
\newcommand{\g}{\mbox{\boldmath $g$}}
\newcommand{\h}{\mbox{\boldmath $h$}}
\newcommand{\yv}{\mbox{\boldmath $y$}}

\newcommand{\group}{Diff}

\newcommand{\evol}{{E_{\text{Vol}}}}
\newcommand{\epair}{{E_{\text{Pair}}}}
\newcommand{\esec}{{E_{S}}}
\newcommand{\inv}{^{-1}}
\newcommand{\vinit}{ \velocity^0_{ij}(\x,\tau=0) }
\newcommand{\avinit}{ \bar{\velocity}^0_{j}(\x,t) }
\newcommand{\avinitz}{ \bar{\velocity}^0_{j}(\x,t_a) }
\newcommand{\gz}{\nabla_{\z}}

%  This is a template for Papers to the Insight Journal. 
%  It is comparable to a technical report format.

% The title should be descriptive enough for people to be able to find
% the relevant document. 
\title{Advanced Normalization Tools (ANTS)}

% 
% NOTE: This is the last number of the "handle" URL that 
% The Insight Journal assigns to your paper as part of the
% submission process. Please replace the number "1338" with
% the actual handle number that you get assigned.
%
%\newcommand{\IJhandlerIDnumber}{}

% Increment the release number whenever significant changes are made.
% The author and/or editor can define 'significant' however they like.
\release{1.00}

% At minimum, give your name and an email address.  You can include a
% snail-mail address if you like.
\author{Brian B. Avants, Nick Tustison and Gang Song}
\authoraddress{Penn Image Computing And Science Laboratory\\
University of Pennsylvania}

\begin{document}

%
% Add hyperlink to the web location and license of the paper.
% The argument of this command is the handler identifier given
% by the Insight Journal to this paper.
% 
%\IJhandlefooter{\IJhandlerIDnumber}

\ifpdf
\else
   %
   % Commands for including Graphics when using latex
   % 
   \DeclareGraphicsExtensions{.eps,.jpg,.gif,.tiff,.bmp,.png}
   \DeclareGraphicsRule{.jpg}{eps}{.jpg.bb}{`convert #1 eps:-}
   \DeclareGraphicsRule{.gif}{eps}{.gif.bb}{`convert #1 eps:-}
   \DeclareGraphicsRule{.tiff}{eps}{.tiff.bb}{`convert #1 eps:-}
   \DeclareGraphicsRule{.bmp}{eps}{.bmp.bb}{`convert #1 eps:-}
   \DeclareGraphicsRule{.png}{eps}{.png.bb}{`convert #1 eps:-}
\fi

\maketitle

\ifhtml
\chapter*{Front Matter\label{front}}
\fi

%Setup listing environment
\definecolor{listcomment}{rgb}{0.0,0.5,0.0}
\definecolor{listkeyword}{rgb}{0.0,0.0,0.5}
\definecolor{listbackground}{gray}{0.965}

\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listbackground}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = left,
        tabsize = 2,
        language=[ISO]C++,
        floatplacement=!h
        }

% The abstract should be a paragraph or two long, and describe the
% scope of the document.
\begin{abstract}
\noindent We provide examples and highlights of Advanced Normalization Tools (ANTS) that 
address practical problems in real data.  
\end{abstract}

%\IJhandlenote{\IJhandlerIDnumber}

\tableofcontents

\section{Introduction}
The ANTS framework provides open-source, encapsulated,
software-engineered functionality for deformable normalization and
segmentation.  Independent evaluation of ANTS normalization tools, 
applied to ``control'' data, 
placed the toolkit as a top performer amongst 14
methods \cite{Klein2009}.  Developer evaluation showed stronger differences 
with other methodology in neurodegenerative neuroimaging data, 
where large deformation is required \cite{Avants2008b}. 
ANTS serves as both a base for further
algorithm development and also as an application-oriented toolkit.
ANTS enables diffeomorphic normalization with a variety of
transformation models, optimal template construction, multiple types
of diffeomorphisms, multivariate similarity metrics, diffusion tensor
processing and warping, image segmentation with and without priors and
measurement of cortical thickness from probabilistic segmentations.
The normalization tools, alone, provide a near limitless range of
functionality and allow the user to develop customized objective
functions.  Objective functions in ANTS are of the form: $$
\text{Deformation Cost} + \text{Data Terms},$$ and the command
line reflects this balance of two terms.  As mentioned above, the
data term may combine multiple different measures of
similarity that are optimized in parallel, for instance, image similarity 
and landmark terms.  This document seeks to
provide a practical overview of basic functionality and some of the common use cases that
users seek.
Additional information is available online -- see 
\href{www.picsl.upenn.edu/ANTS}{ANTS Homepage : www.picsl.upenn.edu/ANTS}.
For compilation details, see: \href{http://picsl.upenn.edu/ANTS/download.php}{Compile and download: http://picsl.upenn.edu/ANTS/download.php}.
The most important ANTS programs are described below. 
\footnote{This document is a work in progress. Please check for updates with each release.}

\subsection{The ANTS Executable}
The ANTS program itself is the central program encapsulating
normalization/registration functionality.  Its main output is an
affine transform file and a deformation field, potentially with
inverse.  Options to ANTS allows the user to navigate the similarity
and transformation options that are available.  ANTS takes multiple
similarity and optimization criteria as options.  The program is
wrapped in \verb ants.sh for normalization with ``out of the box''
parameters and in \verb buildtemplateparallel.sh for 
computationally distributed optimal template construction.

\subsection{The WarpImageMultiTransform Executable}
The WarpImageMultiTransform program applies ANTS mappings to images.  One may apply an 
arbitrarily long series of transformations to an image through this program.  Thus, it enables one 
to compose a series of affine and deformable mappings -- and/or their inverses -- such that one 
may avoid multiple interpolations of a single image.  This program is used internally in ANTS 
optimization to avoid ever using more than one interpolation.  The use-NN option applies nearest 
neighbor interpolation, otherwise linear interpolation is used.

\subsection{The ImageMath Executable} This is a multi-purpose program that has the following syntax:  ``ImageMath ImageDimension outputfilename   Operation  InputFileName  parameters ``.  Most basic scalar image operations -- and some tensor operations -- may be performed with this program.  Some operations output text files, some output images and some output only to the terminal.   ImageMath allows one to multiply images together (m), to negate images (Neg), to take an image to a power (pow), to test the invertibility of transformations (InvId), to compute the fractional anisotropy of an image (TensorFA) and to perform Markov Random Field segmentation with our without priors (Segment).   Many other operations are available.  Like all other ANTS programs, one may call ImageMath from the command line to see all of its options.   ImageMath is used heavily in ANTS scripts.

\subsection{ANTS Scripts} 
The ANTS/Scripts directory contains (hopefully) user-friendly wrappings of ANTS tools 
that enable higher-level error checking and combinations of basic ANTS functions. 
These scripts are called as \verb sh  ~ \verb antsscriptname.sh ~ and provide usage when 
called from the command line.  For instance, try \verb sh ~ \verb ants.sh .

\section{ImageRegistration with ANTS}
\begin{wrapfigure}{r}{0.5\textwidth}
\framebox[0.5\textwidth]{
\begin{minipage}[t]{0.45\textwidth}
{\bf Quick Start:} {\color{blue}call ants.sh (from ANTS/Scripts/) to get usage and apply a normalization to some of your existing data.  It is instructional to read the script, modify some of the parameters and re-run to witness the effect of your changes -- image registration is an art as well as science.  Many other ``ready to go'' scripts are available in ANTS/Scripts. The user must set his/her ANTSPATH environment variable  -- which points to the location of ANTS binaries -- within these scripts or in their environment.  Note: All ANTS programs provide usage when called from the command line.  Most require the image dimension to be specified as the first parameter.  E.g. 
  ImageMath ImageDimension 
where ImageDimenion is 2 or 3.}
\end{minipage}}
\end{wrapfigure}
There are two general applications of image registration. The first application is transforming labeled data from a template image space into an individual space. This strategy is important when appearance alone is not enough to locate a structure as, for example, in the case of hippocampus segmentation. The template provides a prediction of the hippocampus-amygdala boundary.
The second application operates in the "inverse" direction of the first: instead of mapping template to individual, we map individual(s) to the template. Voxel-based population studies of either functional or structural variables depend on mapping to a template space. The common coordinate system enables a statistical evaluation of the likelihood of consistent activation across a group or, in other contexts, the differences in anatomy between two groups.

The ANTS toolkit enables both types of mapping. The main challenge in image and brain mapping is defining the way in which images/anatomy are compared. There are two components to the comparison.
The shape transformation space defines the range of shape variation that will be allowed in the optimization.
The appearance similarity space defines the statistical assumptions that determine when one image is considered to appear similar to another.

These two components interact in a weighted optimization within a multiple resolution gradient descent framework. Each component may use either "strict" or "flexible" assumptions about the shape or appearance similarity. The selection of these models should be done in a principled way that fits the problem at hand. No single choice is appropriate for all scenarios (see the "no free lunch" theorem).

Thus, ANTS enables many operating points from both the transformation and appearance domains such that users may make choices appropriate for their problems. The ANTS command-line syntax, shown in figure~\ref{fig:cmd}, reflects these operating points and the various components that interact in the optimization.
\begin{figure}
\label{fig:cmd}
\includegraphics[width=0.9\textwidth]{Figures/ANTSSyntax.pdf} 
\itkcaption[ANTS Command Line]{The relationship between the variational 
form that defines the optimization goals and the ANTS command line.}
\vspace{-0.1in}
\end{figure}
ANTS may be used to navigate the transformation and similarity metric space. 
\subsection{Use Your Header!} ANTS uses the direction/orientation, spacing
and origin in its definitions of the mapping between images.  That is,
ANTS uses the nifti standard defintions of an image space.  So, the
image orientation (direction), origin and spacing are important!  You
may ``check'' for consistency between two image header definitions by
using: ImageMath 3 Image2repaired.nii.gz CompareHeadersAndImages
Image1.nii.gz Image2.nii.gz \newline This will compare Image2 to
Image1 and ``fix'' the header of Image2, writing out to
Image2repaired.

\subsection{ANTS Transformation Models}
The ANTS toolkit provides a hierarchy of transformations with
adjustable levels of complexity, regularization, degrees of freedom
and behavior as optimizers.  The simplest transformation model is the
rigid and/or affine transform.  The most complex -- and most flexible
-- is a symmetric diffeomorphic transformation based on optimizing and
integrating a time-varying velocity field.  Computation time also
increases with transformation model complexity, but most normalization
needs may be met with under an hour of computation.  
We provide an overview of some of the ANTS models below and 
try to communicate intuition on what one gains/loses with each choice.  
An overview of available models and similarity terms is in Table~\ref{table:chart}.
\begin{table*}
  \centering
    \begin{tabular}{c c c c}
    {\bf Category} & {\bf Transformation, $\phi$} & {\bf Similarity Measures} & {\bf Brief Description} \\
    \toprule     
    \multirow{2}*{\bf Linear}
           & Rigid$^\dagger$ & MI, MSQ & Rigid registration. \\
       {} & Affine$^\dagger$ & MI, MSQ & Affine registration. \\
       \cmidrule(l){2-4}
    \multirow{2}*{\bf Elastic}
           & Deformable & CC, PR, MI, MSQ, PSE & Demons-like algorithm. \\
       {} & DMFFD & CC, PR, MI, MSQ, PSE  & FFD variant. \\
       \cmidrule(l){2-4}
    \multirow{2}*{\bf Diffeo.}
           & Exponential$^\dagger$ &  CC,PR, MI, MSQ, PSE  & $~\min~\velocity(\x)$ \\
       {} & Greedy SyN$^\dagger$ &  CC, PR, MI, MSQ, PSE  & ~~locally in time~$\min~\velocity(\x,t)$\\
       {} & Geodesic SyN$^\dagger$ &  CC, PR, MI, MSQ, PSE  &  $~\min~\velocity(\x,t)$ over all time \\
    \bottomrule
    \end{tabular}
  \caption{Transformations and a subset of the similarity metrics available in ANTS.  Similarity metric acronyms:  CC = fast cross correlation, PR = pure cross correlation (the preferred metric), MSQ = mean squared difference, MI = mutual information, PSE = point-set expectation.  ANTS also provides the inverse of those transformations denoted by the `$\dagger$' symbol.  The brief descriptions of the diffeomorphic algorithms contrast the way in which the velocity field is optimized and used to parameterize $\phi$, the mapping.  All 
ANTS Diff algorithms generate $\phi(\x,t)$ over $t \in [ 0, 1]$ through gradient descent.}
  \label{table:chart}
\end{table*}    
\vspace{0in}

\subsubsection{Linear Registration}
\begin{figure}
\label{fig:aff}
\includegraphics[width=0.9\textwidth]{./Figures/ANTSanatomy.pdf} 
\itkcaption[ANTS Introduction]{The anatomy of an ANTS optimization and application of the resulting warping.}
\vspace{-0.1in}
\end{figure}
The most basic mapping between images is an affine mapping. The ANTS 
affine mapping syntax is shown in the affine mapping figure~\ref{fig:aff}. The
parameter breakdown of the ANTS call in the affine figure is:
\begin{itemize}
\item  ANTS -- the normalization program
\item 2 -- the expected image dimension
\item -m MI[...,...,1,32] -- the similarity metric "mutual information" with a 32-bin square joint histogram and weight of 1.  
Metric details will be discussed in a later section. 
\item -i 0 -- no deformable iterations (affine only).
\item -o ab -- the output prefix.  A specific extension may also be specified, e.g.   -o ab.mhd  .
\end{itemize}
The parameter breakdown of the WarpImageMultiTransform call in the affine figure is:
\begin{itemize}
\item WarpImageMultiTransform -- to apply the mapping output from ANTS to an image.
\item 2 -- the expected image dimension.
\item the image to be deformed is input next.
\item the deformed image output file name is next.
\item the transform itself is next -- if is affine and preceded by " -i ", we apply the inverse affine map.
\item the " -R " option dictates the domain you want to warp into -- usually the ``fixed'' image, unless using the inverse mapping, in which case one switches the role of fixed and moving images. 
\end{itemize}
Note that -i 0 (in the ANTS call) means that no deformable mapping is performed. The call "ANTS 2" means to treat the images as 2-dimensional. "ANTS 3" would be called for 3D images. The " -m " term defines the similarity metric. 
Note that no more than one image interpolation is used during optimization.  
That is, ANTS always references the original image before applying a transformation or series of transformations.

Affine mapping may become nontrivial when the region of interest 
occupies only a small portion of the image data.  In such cases, 
ANTS enables the user to define a mask to focus the optimization 
effort on the region of interest.  
Example code for (affine) registration with a mask: 
\begin{verbatim}
ANTS 3 -m MI[AtlasHead.nii,crAnatomical.nii,1,32] 
-o TEST -i 10x10x0  -r Gauss[1.5,0] -t Exp[0.5] -x mask.nii.gz
\end{verbatim}
The difference between this command 
and a regular ANTS call is the \verb"-x" option, which specifies the mask, defined in the 
template space (here, AtlasHead.nii).   The mask option also affects the deformable optimization. 
Other affine registration options include  \verb"--number-of-affine-iterations 5000x5000x5000" which specifies that the affine registration uses a 3 level image pyramid with each level 5000 iterations at most. \verb"--MI-option 48x6000" means to use Mutual Information as similarity metric with 48 bins and 6000 samples. \verb"--affine-gradient-descent-option" defines the options for the gradient descent procedure; the 4 numbers are maximum step length, relaxation factor, minimum step length and translation scale.

\subsubsection{Deformable Registration}
Affine mapping is adequate when the
difference between images involves only rotation, scaling or shearing.
Other data requires more deformable mappings to capture shape
differences and find a good alignment of image features.
Figure~\ref{fig:antshier} shows how deformable mapping may improve the
correspondence of the deformed beetle to the ford image.
\begin{figure}
\label{fig:antshier}
\includegraphics[width=0.9\textwidth]{./Figures/ANTSElastToDiff.pdf} 
\itkcaption[ANTS Transformation Examples]{This example shows the degree to which the beetle (b.img) may be deformed to the ford (a.img) under different transformation models.  Left to right increases the degrees of freedom in the mapping and thus the registration accuracy.}
\vspace{-0.1in}
\end{figure}
Most ANTS-based applications use symmetric diffeomorphic normalization.  However, 
ANTS also enables a simpler parameterization of a deformable mapping via a regularized 
vector space.  We term these types of transformations as ``elastic''.  The original Demons algorithm 
provides a classic example of using a regularized vector space for nonlinear registration.  
Caveats are that a regularized vector space may not preserve the underlying topology 
and may also prove too inflexible to capture the shape changes one is after.  Both of these 
shortcomings motivate the use of diffeomorphisms.

\noindent{\bf Elastic/Vector Space Transformations.} If we assume no affine transformation, then 
an elastic transformation involves computing a mapping from image $I(\x)$ to image $J(\x)$ 
through a deformation field $\disp( \x )$.  
\begin{wrapfigure}{r}{0.5\textwidth}
\label{fig:pull}
\includegraphics[width=0.45\textwidth]{./Figures/PullBack.pdf} 
\end{wrapfigure}  
The deformation is defined in the physical space of the image 
and dictates the positional difference between corresponding features in the two images.  
Thus, if a feature defined at $I(\x)$ matches a feature in $J$ at position $\y$ then the deformation 
field at $\x$ should give $\disp ( \x )=\y-\x$.   
Such a deformation field may be applied to deform 
image $J$ into image $I$ by composing the mapping $J_{deformed}(\x)=J(\x+\disp(\x))$.  
In a perfect world, then $I(\x)=J_{deformed}(\x)$, though this is rarely the case.  Figure~\ref{fig:pull} 
visualizes the deformation of an image under this standard model. 
Gradient descent optimization of an elastic mapping may be summarized (crudely) as: 
\begin{align}
\text{Compute the similarity gradient:   }  \nabla E = \partial_{\disp} \Pi(I,J(\x+\disp(\x))). \nonumber \\
\text{Update the deformation field:  }  \disp(\x) \leftarrow \disp(\x)+\delta \nabla E. \nonumber \\ 
\text{Regularize the deformation field:  }  \disp(\x) \leftarrow G_\sigma \star \disp(\x),
\end{align}
where $\Pi$ is the similarity, $\delta$ is a gradient step length and $G_\sigma$ is a gaussian smoother.
This optimization is captured in the following ANTS command:
\begin{verbatim}
ANTS 3 -m PR[AtlasHead.nii,crAnatomical.nii,1,4] -r Gauss[0,3]  
-o ElasticTest.nii.gz -i 30x20x10  -t Elast[1.5] 
\end{verbatim}
Here, we use the PR metric (a cross-correlation implementation) with
window radius 4, weight 1 and gradient step length 1.5.  The
optimization will be performed over three resolutions with a maximum
of 30 iterations at the coarsest level, 20 at the next coarsest and 10
at the full resolution.  We use a Gaussian regularizer with a sigma of
3 that operates only on the deformation field and not on the similarity gradient, as 
0 is passed as the first parameter to \verb Gauss .  One may see the
correspondence, yet again, between the ANTS call and the optimization
scheme.  The optimization will stop when either the energy cannot be 
smoothly minimized or the maximum number of iterations is reached. 
BSpline regularization is also available in ANTS -- see the DMFFD 
section below.

\noindent{\bf Warping and Invertibility.}ANTS uses physical space to define mappings. The origin etc is in the
coordinates of the world in which the picture (e.g. MRI) was
taken. Then, warping between physical coordinates is relatively
easy. Differences in bounding boxes etc present no problem -- assuming
you avoid inconsistent headers i.e. origins, directions, data
orientation. One may use PrintHeader to check the data and run simple,
fast tests (few iterations) to perform sanity checks before running
through loads of data.
An ANTS deformation consists of
a standard naming prefix plus a standard naming extension. We usually 
assume nii but other file types may be used. The standard naming
extension is Warpxvec.nii for the x component of the deformation with
similar naming for y and z. Each component is stored in a separate
scalar image. The value of a voxel of a deformation/warp component
image is the physical space displacement from that voxel. Note that
the inverse mapping is stored in InverseWarpxvec.nii and provides the
mapping in the opposite direction.  Note that an inverse -- in the 
digital domain -- is only approximate as shown in figure~\ref{fig:inv}.
\begin{figure}
\label{fig:inv}
\center \includegraphics[width=0.75\textwidth]{Figures/DeformationsInDigitalDomain.pdf} 
\itkcaption[Digital Invertibility]{Digital invertibility presents some limitations. 
Here, we see that invertibility is not exact but is gained only by interpolation.  
Thus, in three-dimensional scenarios, there are fundamental limits to the 
degree of invertibility that may be achieved.  The second and third voxels -- from left -- 
in image A undergo an expansion by a factor of 2.  That is, under the mapping, 
2 voxels are mapped to 4.  This gives the definition of the Jacobian -- computed 
by  ANTSJacobian -- which is a unitless measure defined by the 
ratio of volumes.  Thus, $\mathcal{J}(\x)=V_1(\phi(\x))/V_2(\x)$ where $V_i$ represents 
the volume in one domain versus the other.  Thus, if $\phi$ -- the mapping -- 
causes expansion, then $\mathcal{J}(\x) > 1$. }
\vspace{-0.1in}
\end{figure}  
A few important notes follow. 
(1) Deformation directionality: Warps/deformations applied to images occur in the opposite direction from warps/deformations applied to points. That is, if we map image B to A using 
\begin{verbatim}
ANTS -m PR[A,B,1,2] -o OUTPUT 
\end{verbatim}
 then we get a deformation called \verb OUTPUTWarp+extensions ~and an affine transform called \verb OUTPUTAffine.txt . 
This composite mapping - when applied to B - will transform B into the space of A. 
However, if we have points defined in B that we want to map to A, 
we have to use \verb OUTPUTInverseWarp ~and the inverse of \verb OUTPUTAffine.txt .  
This is because the domain and range of the map's definition need to be dense in a way that is appropriate for the data type. 
Figure~\ref{fig:inv} illustrates the concept. 
(2) Older Image Formats: older image formats (e.g. Analyze) may not
have proper origin/offset/direction support! In these cases, we
recommend converting to nii and verifying that data overlays properly.
(3) Transorm Composition: Composition of transforms may be achieved with
ComposeMultiTransform.  (4) Warping with inverses and concatenations --
viable when using diffeomorphisms -- are described in
figure~\ref{fig:warp}.
 \begin{figure}
\label{fig:warp}
\includegraphics[width=0.9\textwidth]{Figures/ANTSSyntax2.pdf} 
\itkcaption[ANTS Warping]{Here, we detail how one applies 
a deformation and associated inverses.  We also see how 
WarpImageMultiTransform can be used to concatenate a series 
of transformations.}
\vspace{-0.1in}
\end{figure}

\noindent{\bf Diffeomorphic Transformations.}  The elastic 
mapping space -- as indicated above -- may prove inadequate 
for some large deformation mapping scenarios.  Figures~\ref{fig:antshier} 
illustrate the changing performance one may get in switching from affine 
to elastic to diffeomorphic normalization.  
\begin{figure}
\label{fig:antssyn}
\includegraphics[width=0.9\textwidth]{./Figures/ANTSsyn.pdf} 
\itkcaption[ANTS Symmetric Normalization]{This example shows the benefit of the symmetric normalization model -- invertibility, symmetry, highly deformable and accurate registration.  The transformation model used here is highlighted in red.}
\vspace{-0.1in}
\end{figure}
Figure~\ref{fig:antssyn} shows 
how one may use ANTS to achieve a state-of-the-art diffeomorphic mapping. 
The ANTS diffeomorphic model chosen for this example -- symmetric normalization -- 
is invariant to the order of the input images (although the affine mapping is not).  
An additional advantage of the diffeomorphic model over the elastic model is 
that both forward and inverse transformations are computed thus allowing one 
to deform fixed to moving and also moving to fixed.   The transformation model 
chosen here -- \verb SyN[0.25] -- may be replaced with other 
diffeomorphic transformation models.  The most general is global-in-time SyN via \verb SyN[0.25,2,0.01]   , 
where the time step for integration is 0.01 (lower is more accurate).  A fast approximate 
solution may be gained through 
exponential mapping via \verb Exp[0.25,10] , where 10 integrations points are used. 
Update schemes for diffeomorphic optimization are similar to the elastic case 
and are described elsewhere \cite{Avants2008b,ANTStheory}.

\subsection{ANTS Similarity Terms}
From Gang.

{\em Note: ANTS does not like spaces between brackets in the call to the metrics.}

\subsection{Diffusion Tensor Normalization}
The ImageMath program -- via TensorFA and TensorMeanDiffusion -- 
may derive scalar images from DTI.  Such images may be used to 
distortion correct DTI to T1 or to map DTI together.   See the 
ANTS/Scripts program called \verb antswithdt.sh ~for a 
ready to go example.   Figure~\ref{fig:dterr} shows what might 
happen if your tensor entries are stored in the wrong order. 
ANTS expects the nifti standard ordering of the DT six vector.
\begin{wrapfigure}{r}{0.5\textwidth}
\label{fig:dterr}
\includegraphics[width=0.5\textwidth]{Figures/dtiordererr.pdf} 
\itkcaption[DT Entry Error]{The FA and mean diffusion of the tensor may be corrupted, 
as at right, if the order of the DT entries is wrong.}
\vspace{-0.1in}
\end{wrapfigure}  
\subsection{Multivariate Normalization with ANTS}
\begin{figure}
\label{fig:mvnorm}
\includegraphics[width=0.9\textwidth]{Figures/multivariatenorm.pdf} 
\itkcaption[Multivariate Normalization]{The current version of ANTS 
may perform normalization of multiple modalities by combining {\em intrasubject, intermodality} mappings with {\em intersubject, intramodality} maps to a 
group template.  In brain imaging, the intersubject maps are usually guided by the T1 component.}\vspace{-0.1in}
\end{figure}
Multivariate normalization may be performed in two ways with ANTS.  
First, as shown in figure~\ref{fig:mvnorm}, intra-subject mappings may 
be used to conform all modalities into a common subject space defined by 
the ``anchor'' modality.  In brain imaging, this is usually the T1 component.

A second type of multivariate normalization is able to use the T1 and DTI components 
directly in the optimization of the mapping.  An example of this type of call is below, 
where we assume that one has resampled the DT data to the space of T1: 
\begin{verbatim}
ANTS 3 -m PR[T1template.nii,T1subject.nii,1.25,4]
       -m PR[FAtemplate.nii,FAsubject.nii,1,4] 
-o MultiVar -i 30x30x20  -r Gauss[3,0] -t SyN[0.25]
\end{verbatim}
In this case, dense features within white matter are gained by using the DT component  
in addition to the T1 component.   The T1 PR metric, here, is weighted slightly more 
than the FA PR metric, 1.25 vs. 1.0.

\subsection{Optimal Template Construction with ANTS}
A useful script for setting up optimal template construction
 is available in ANTS/Scripts.  Two versions of available : serial and parallel.  
Data and a script for building an average face template is available 
the ANTS/Examples/Data directory.  Two key points about optimal templates:
1. The outcome stabilizes at around ten images, for most populations.
That is, if we randomly select images from individuals within the same 
demographic distribution, we will end up with very similar average images.  
2.  Optimality, for the ANTS-SyN approach, is defined by the minimum shape 
and appearance distance image. 
See the template page for some examples of this and users may 
contact ANTS developers for previously derived template 
images that may be useful.

\subsubsection{A Concrete Template Construction Example, Step-by-Step}
For this example, we will use the data in "ANTS/Examples/Data/B*.tiff."
The files are B1.tiff, B2.tiff, B3.tiff, B4.tiff and B5.tiff.   They are single slices 
from a dataset of real MRI brain images -- individuals around 15 years of age. 
Our goal is to derive a "most representative" single image from this population. 
The steps are: \begin{figure}
\label{fig:template}
\center \includegraphics[width=0.7\textwidth]{Figures/templateex.jpg} 
\itkcaption[ANTS Optimal Template]{This example may be recreated by the user.  The shared anatomy -- 
across this dataset -- is recovered in the derived optimal template.}
\end{figure}
\begin{enumerate}
\item   Make a directory for this example.   Copy or link all the B*tiff images into it.
\item Copy the file ANTS/Scripts/buildtemplate.sh into that directory.  
\item  On the command line, within that directory, run the following command: 
\verb  sh ~ \verb buildtemplate.sh ~ to get usage. 
\item If you follow the usage, you will then call the script something like this: 
\verb sh ~ \verb buildtemplate.sh ~ \verb 2 ~ \verb B ~ \verb tiff ~ \verb OUT ~ \verb 4 ~ . 
\item The script should exit and ask if the parameters are ok for your problem.  
Check the path to the programs, the fact that all the images you want to use are being 
listed and that the template output name is ok with you.  
Once everything is prepared, then edit the script to remove the exit call at the end of the parameter check. 
\item  Call the script as in step 4 and then wait for the output.  The template will 
be called something like (in this example)  \verb OUTtemplate.nii .
\end{enumerate}
The script is fairly easy to alter so that you can send the whole thing to distributed computing. 
Typically, one would use voxbo or the qsub program available in the Sun Grid Engine.  This 
is what is done in \verb buildtemplateparallel.sh .  
Figure~\ref{fig:template} shows the data and a result that may be recreated by the user.
\subsection{Notes on Large Deformation Mapping} 
Figure~\ref{fig:large} defines what one might expect from a 
high-resolution, large deformation, successful normalization.  
Major and many minor features are well aligned between 
these images to an extent that is approaching the limits 
allowable by biological variation.  
\begin{figure}
\label{fig:large}
\includegraphics[width=0.9\textwidth]{Figures/ANTSLargeDef.jpg} 
\itkcaption[ANTS LargeDeformation]{ANTS succeeds in a challenging 
normalization scenario.  Comparing the template mapping to the
individual shows that much of the cortex is well-aligned, as is the
hippocampus, despite the relatively large difference between the
initial template and the target.  Here, one might note a limitation of
whole brain mapping: occipital lobe sulcal variation is highly
idiosyncratic and extremely difficult to capture, in particular when
there is such severe atrophy nearby.
}
\end{figure}
\noindent{\bf Turning "Failures" into Successes:} 
Below are some pointers to follow if you are unable to recreate such normalization quality.  
Usually, reasons for registration failure are one of a very few things: 
\begin{enumerate}
\item The initialization is so off mark that affine registration fails -- thus meaning all subsequent deformable registration will not be meaningful.
\item The information within headers is inconsistent e.g. origins are defined in different ways, the "directions" are not correct or some combination of these.    The PrintHeader executable can help one in debugging this type of problem.  Also, the ImageMath function CompareHeadersAndImages can fix some problems of this type. 
\item  The similarity or transformation model is inappropriate or has too small a capture region for the problem. 
\end{enumerate}
An example of this final point is given here.   
A recent study used the following call to ANTS across a dataset of elderly subjects:
\begin{verbatim}
ANTS 3 -m PR[template.nii.gz,subject.nii.gz,1,2] -i 10x50x50x20
 -o subjectmap.nii -t SyN[0.25] -r Gauss[3,0]
\end{verbatim}
This succeeded for all subjects but one.  This subject had severe neurodegeneration which caused "under-normalization" to result.  A larger deformation mapping was required and so we increased the maximum number of iterations (the "-i" parameter).  One might also increase the gradient-descent step-size (SyN[0.25] => SyN[0.5]).   Large deformation mapping is challenging not only because of the amount of deformation, but also because the "true" solution is difficult to find.  Keeping this in mind, we also increased the "span" (radius) of the correlation window in the similarity metric (which also increases computation time). These modified parameters succeeded: 
\begin{verbatim}
ANTS 3 -m PR[template.nii.gz,subject.nii.gz,1,4] -i 100x100x100x20
 -o subjectmap.nii -t SyN[0.25] -r Gauss[3,0] 
\end{verbatim}
Key changes were to increase the radius of the correlation and to
allow more iterations at the coarsest resolution.  
This two coarsets levels of the computation take only about
10 percent of the time (about 5-10 minutes depending on the machine)
but accounts for the large majority of the shape variation.  Note that
this is a 3D example -- that is why the eyes appear only in the frame
at left.

\section{Image Segmentation and Labeling}
ANTS has tools for both tissue based segmentation and prior-based segmentation that 
leverages spatial priors, usually based on a template mapping.  
\subsection{Basic Segmentation}
A basic MRF segmentation algorithm is available in ANTS ImageMath.  We apply 
the segmentation to the example data in: ANTS/Examples/Data/r16slice.nii.  
\begin{verbatim}
ImageMath 2 r16slice.nii Segment r16slice.nii 4 0 
% output = r16slice_seg.nii, r16slice_prob_0.nii, r16slice_prob_1.nii, r16slice_prob_2.nii .  
\end{verbatim}
The parameter \verb 4 ~indicates 3 tissues plus background are sought.  The \verb 0 ~indicates 
that there are no prior images available to guide the segmentation.  Output is shown 
in figure~\ref{fig:seg} and -- with priors -- in figure~\ref{fig:seg2}.
\begin{figure}
\label{fig:seg}
\includegraphics[width=1\textwidth]{Figures/segmentation.pdf} 
\itkcaption[ANTS Segmentation]{ANTS provides basic 
Markov Random Field regularized Gaussian model tissue segmentation. 
The ANTS program N3BiasFieldCorrection is very valuable as preprocessing 
for this naive approach to segmentation.   The last panel -- far right -- 
shows the thickness derived from the white matter and gray matter 
probabilities where a prior on thickness was used to prevent thickness
overestimation.   Thickness was derived with DiReCT \cite{Das2009}, a software 
tool available in binary form only -- a future release may provide code.  
}
\end{figure}
\subsection{Prior and Template-based Image Segmentation}
The same algorithm may be augmented 
to perform Prior-based segmentation.  
\begin{verbatim}
ImageMath 2 r16sliceprior.nii Segment r16slice.nii 4 0.5 
   csfprior.nii wmprior.nii  gmprior.nii 
% output = r16sliceprior_seg.nii  r16sliceprior_prob_0.nii  
% r16sliceprior_prob_1.nii  r16sliceprior_prob_2.nii .  
\end{verbatim}
The parameter \verb 4 indicates 3 tissues plus background are sought.  The \verb 0.5 indicates 
we weight the priors equally as the data term and use a spatially varying set of Gaussians 
to estimate the segmentation.  In this way, data from the priors may be used to modify 
and guide the segmentation in a locally varying way, accomodating for both inhomogeneity 
and the different imaging signature that different tissues provide.   Prior images should be 
of the same size and dimension as the input data and should have intensities in the range $[0,1]$.
\begin{figure}
\label{fig:seg2}
\includegraphics[width=1\textwidth]{Figures/segmentation2.pdf}
\itkcaption[ANTS Segmentation]{ANTS extends basic Markov Random Field
  regularized Gaussian model tissue segmentation to include priors,
  which allow spatially varying tissue models to determine
  segmentation.  The ANTS program N3BiasFieldCorrection is less
  critical as preprocessing when using this approach to segmentation.
  Here, we see the improvement in segmentation as the locally varying 
  prior models are weighted more heavily.  
  The last panel -- far right -- shows the segmentation derived from
  data driven with an initialization founded purely on prior models 
  based on template mapping.  The prior model brings out the caudate 
and some CSF -- highlighted by arrows -- as their weight increases. 
}
\end{figure}
\subsection{Cortical Thickness}
Two forms of cortical thickness estimation from probability maps are available 
in ANTS: first, the traditional Laplacian cortical thickness estimation and, second, 
the more recently developed Diffeomorphic Registration-based Cortical Thickness 
(DiReCT) \cite{Das2009}.  Both methods estimate thickness of an object 
from probability or binary images that define object boundaries.  This tool 
is mainly of interest in brain mapping and cardiac imaging for morphometric 
studies.  Cortical thickness, for instance, is known to correlate with language 
development and IQ in adolescents.

\subsection{User/Label-Guided Normalization for Hippocampus Mapping}
See the link for details on hippocampus labeling:
\href{http://picsl.upenn.edu/ANTS/hipptutorial.php}{http://picsl.upenn.edu/ANTS/hipptutorial.php}.
\noindent{\bf Expectation-Based Point-Set Registration.}
Here, we apply the expectation-based point set registration method 
for mapping labeled points sets.   ITK-SNAP may be used to label 
images and exported segmentation images may be input to the 
PSE metric below, as labeled data.  The Frown and Smile data is used 
as example.  This data is available in the 
ANTS/Examples/Data/ directory.
\begin{verbatim}
ANTS 2 -o PSEtest  -i 91x70x55x40x30  -r Gauss[3,0.] -t SyN[0.3]  
   -m  PSE[Frown.nii,Smile.nii,Frown.nii,Smile.nii,1,0.2,11,0,25] 
    -m  MSQ[Frown.nii,Smile.nii,1,0] 
     --continue-affine 0  --number-of-affine-iterations 0

WarpImageMultiTransform  2 Frown.nii temp.nii -R Smile.nii
   -i PSEtestAffine.txt PSEtestInverseWarp.nii

WarpImageMultiTransform  2 Smile.nii temp2.nii -R Frown.nii 
    PSEtestWarp.nii PSEtestAffine.txt

CreateWarpedGridImage 2 PSEtestInverseWarp.nii grid1.nii
CreateWarpedGridImage 2 PSEtestWarp.nii grid2.nii
\end{verbatim}
This example should run on the downloaded ANTS data so you may see the results.

%\input{ANTSLabeledDataExample}
%\input{ANTSMaskedImageRegistrationExample}

\section{Application to Studies}
\subsection{Brain Mapping in the Presence of Lesions}
Many cases violate the basic assumption of a one-to-one mapping from a
template image to a target image. Brain lesions caused by stroke or
traumatic brain injury are a common instance of this.

The ANTS toolkit handles this situation by a constrained cost-function
masking approach. In short, the mapping within an excluded region (as
determined by an inclusive mask) is determined by the solution outside
of the region. To achieve this with ANTS, one would use a call similar
to this:
\begin{verbatim}
ANTS 3 -m PR[tp22_s1.nii,template.nii.gz,1,4] -i 50x20x0
   -o tp22map -t SyN[0.5] -x mask.nii.gz -r Gauss[3,0] 
\end{verbatim}
The additional parameter needed for this approach 
is the \verb -x ~ \verb mask.nii.gz ~ which 
is the mask defined in the "fixed" (or individual) image space.  
Mask values above 0.5 are considered non-outlier or non-lesioned regions.
The labeling may be performed with ITK-SNAP. 
Figure~\ref{fig:lesion} illustrates the approach.
\begin{figure}
\label{fig:lesion}
\includegraphics[width=1\textwidth]{Figures/lesionstudy.pdf}
\itkcaption[ANTS Lesions]{Here, we see a semi-automated approach for lesion analysis
based on diffeomorphic normalization.  The user outlines the lesion
site in the subject space, as shown in (a).  Diffeomorphic mapping is
then used to deform the subject image in (a) into the space of the
reference template.  The deformed subject is shown in panel (b).
Panel (c) shows the reference template space with the estimated
position of the subject's lesion overlaid on the healthy tissue of the
template.  This approach enables one to compare the subject image
against prior information stored in the template space such as
anatomical labels or statistics on the appearance of lesioned and/or
normal tissue.}
\end{figure}
Once the ANTS mapping is solved, 
one is able to estimate the location of the lesion in the template space by mapping the inverse lesion mask to the template.
To take the negative image of the inclusive mask 
and warp the mask to the template :
\begin{verbatim}
ImageMath 3 lesionmask.nii.gz Neg inclusivemask.nii.gz
WarpImageMultiTransform 3 lesionmask.nii.gz lesiontotemplate.nii.gz 
   -R template.nii.gz -i tp22mapAffine.txt tp22mapInverseWarp.nii
\end{verbatim}
More details on this approach are being prepared for publication.
See also: \verb http://www.ncrrn.org/papers/methodology_papers/sp_norm_kim.pdf .

\subsection{Statistical Mapping with ANTS: Morphometry, Function, Jacobian, Thickness}
ANTS has been applied in a wide array of imaging studies
\cite{Avants2009a,Avants2009b,Avants2009,Das2009,Klein2009,Massimo2009,Pluta2008,Tustison2009,Yushkevich2009,Avants2008c,Avants2008b,M.Grossman10282008,Kim2008,Simon2008,Sun2008,Yushkevich2008,Aguirre2007,Avants2007,Avants2007i}.
All of these studies benefit in some way from normalization whether the topic is cortical
thickness, Jacobian-based morphometry, volumetric morphometry or functional studies. 
From a broad perspective, each of these applications requires the same steps:
\begin{enumerate}
\item Preprocess images -- bias correction, segmentation, etc.  Potentially construct an optimal template. 
\item Normalize images -- run \verb ants.sh ~ to map a population of images to a template and store the deformations. 
\item Derive any data from the deformation that may be necessary, e.g. the log-Jacobian.
\item Apply the warp to any images one may want to analyze in template space.  
\item Perform statistics in template space over the region of interest, e.g. all cortex. 
\end{enumerate}
We now illustrate this procedure using ANTS tools and example images from ANTS/Examples/Data. 
The expected output is in figure~\ref{fig:morph}.
\begin{verbatim}
1.  ImageMath 2 r64slice.nii Segment r64slice.nii 4 0
1.  ImageMath 2 r16slice.nii Segment r16slice.nii 4 0
2.  sh ants.sh 2 r16slice.nii r64slice.nii SPM 100x100x100
3.  SmoothImage 2 r64slice_prob_1.nii  1. r64slice_prob_1.nii  
3.  WarpImageMultiTransform 2 r64slice_prob_1.nii  r64slice_prob_1_norm.nii
     -R r16slice.nii SPMWarp.nii SPMAffine.txt 
3.  ANTSJacobian 2 SPMWarp.nii SPM 1  # take the log
4.  ImageMath 2 SPMlogjacobianmask.nii m SPMlogjacobian.nii r64slice_prob_1_norm.nii
5.  Repeat for a population and run statistics on log jacobians.
\end{verbatim}
\begin{figure}
\label{fig:morph}
\includegraphics[width=1\textwidth]{Figures/morphometry.pdf}
\itkcaption[ANTS Morphometry]{The output from the ANTS morphometry example.}
\end{figure}
This example uses a standard Jacobian-based approach to morphometry.
Here, we use the log-Jacobian because it is symmetric about zero and
mask with the gray matter segmentation to restrict the analysis to the
cortex.  The Jacobian is discussed in figure~\ref{fig:inv} and visualization of
the Jacobian is shown in figure~\ref{fig:jac}.  This Jacobian comes
from the mapping shown in figure~\ref{fig:large}.
\begin{figure}
\label{fig:jac}
\includegraphics[width=1\textwidth]{Figures/Jacobian.pdf}
\itkcaption[ANTS Jacobian]{The Jacobian: Note that the bright values
  in the template image ventricles (left) indicate that the ventricles
  are relatively larger in the subject image. Similarly, the dark
  jacobian values in the individual image show that the template has
  smaller ventricles. This example is described on the large
  deformation page Large Def. Once one gains the Jacobian (or more
  appropriately, the log-Jacobian), then one may compute statistics
  across the population.}
\end{figure}
Similar analyses may be performed on thickness images, functional 
images or fractional anisotropy images derived from the diffusion 
tensor modality.  In all these cases, steps similar to those above 
would be performed.

\subsection{Statistics with ANTS and ``R''}
{\bf R} is an open-source, cutting-edge statistical package 
used by statisticians world-wide.   ANTS is designed to 
interface with {\bf R} via an ImageMath tool that converts 
image sets into an {\bf R} compatible format.  
Three steps are required.  First, convert image data into 
vector and matrix formats through a mask of your statistical 
region of interest.  Second, read that data into {\bf R} and 
run your statistics.  Third, write out the data from {\bf R} and 
convert back to an image.  Here are the steps, in code examples,
\begin{enumerate}
\item Create a vector for your output: 
\begin{verbatim} 
ImageMath 3 testresultvec.nii ConvertImageSetToMatrix 1 mask.nii image1.nii
\end{verbatim}
Here, we assume all images are called image1.nii, image2.nii, ... , imageN.nii so we 
can use a wildcard to pass to ImageMath and 
create a matrix for all your input data: 
\begin{verbatim} 
 ImageMath 3 datamatrix.nii ConvertImageSetToMatrix 1 mask.nii image*.nii 
\end{verbatim}
\item   Run {\bf R} by reading your data above, performing whatever statistics you want and then 
writing out.  
\item   Convert the test result back to an image.  
\begin{verbatim} 
ImageMath 3 testresult.nii ConvertVectorToImage mask.nii testresultvec.nii
\end{verbatim}
\end{enumerate}
Most statistical requirements may be met with this setup. 
\section{Dependencies and Related Software}
\subsection{Dependencies and Compilation}
ANTS depends on the most recent version of the Insight ToolKit (ITK)
-- www.itk.org.   File types that may be read and written are 
all of those available in itk and a few more \href{http://picsl.upenn.edu/ANTS/ioants.php}{http://picsl.upenn.edu/ANTS/ioants.php}.

ANTS and ITK are both compiled through pointing CMake -- www.cmake.org
-- to a CMakeLists.txt file.  The ANTS CMakeLists.txt is in the
Examples directory.

See below for more details on compilation and binaries: \newline
\href{http://picsl.upenn.edu/ANTS/download.php}{Compile and download: http://picsl.upenn.edu/ANTS/download.php}.
We maintain a dashboard to give users an idea of what to expect 
from compilation:\newline
\href{http://picsl.upenn.edu/cdash/index.php?project=ANTS}{http://picsl.upenn.edu/cdash/index.php?project=ANTS}

\subsection{Visualization and Quantification of ANTS Results}
Use \href{www.itksnap.org}{itk SNAP : www.itksnap.org } or mricro / mricron for viewing
images.  The ANTS program StackSlices can be used to conveniently scan
through a dataset in any of its axes.  It is useful for checking both
input data quality and output normalization quality.  MeasureImageSimilarity 
will supply numerical measures of image similarity via three different metrics.

\subsection{Statistics Beyond ANTS}
Use SPM, {\bf R} or npm (from Chris Rorden's mricron) for computing population statistics.

\subsection{Pipelining with ANTS}
Use PipeDream for automating large-scale processing:\newline
\href{https://sourceforge.net/projects/neuropipedream/}{PipeDream
  Homepage :
  https://sourceforge.net/projects/neuropipedream/} \newline PipeDream
is integrated with ANTS and automates more complex studies such as
multivariate diffusion tensor and T1 cortical thickness studies,
longitudinal mapping and reconstruction of nifti images from Dicom.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  Insert the bibliography using BibTeX
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliographystyle{plain}
\bibliography{ants}

\end{document}

### 2009-06-04 Documentation/Tex/antstheory.tex
tex and documentation
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@5 fd2da0a4-d968-48df-9cb5-d8c496287cbf

% Template article for preprint document class `elsart'
% with harvard style bibliographic references
% SP 2006/04/26

\documentclass[preprint,twocolumn,authoryear,5p]{elsarticle}

\newcommand{\fv}{\mbox{\boldmath $f$}}
\newcommand{\Iv}{\mbox{\boldmath $I$}}
\newcommand{\xv}{\mbox{\boldmath $x$}}
\newcommand{\velocity}{\mbox{\boldmath $v$}}
\newcommand{\velocityinv}{\mbox{\boldmath $v$}^{-1}}
\newcommand{\Velocityinv}{\mbox{\boldmath $V$}^{-1}}
\newcommand{\Velocity}{\mbox{\boldmath $V$}}
\newcommand{\welocity}{\mbox{\boldmath $w$}}
\newcommand{\spatialimagegradient}{\mbox{\boldmath $I$}\mbox{$_{x}$}}
\newcommand{\perpspatialimagegradient}{\mbox{\boldmath $I$}\mbox{$^{\perp}_{x}$}}
\newcommand{\temporalimagederivative}{\mbox{$I_t$}}
\newcommand{\finitestrain}{\mbox{\boldmath $\varepsilon^{\ast}$}}
\newcommand{\smallstrain}{\mbox{\boldmath $\varepsilon$}}
\newcommand{\dispgradient}{{\bf \nabla \displace}}
\newcommand{\displace}{{\bf u}} 
\newcommand{\Id}{\text{\bf Id}}
\newcommand{\ode}{{\em O.D.E.}}
\newcommand{\odes}{{\em O.D.E.}s}
\newcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\phiinv}{\phi^{-1}}
\newcommand{\psiinv}{\psi^{-1}}
\newcommand{\dM}{DM}
\newcommand{\DM}{Diffeomorphometry}
\newcommand{\diff}{diffeomorphism}
\newcommand{\Diff}{Diffeomorphism}
\newcommand{\orbit}{\mathcal{O}}
\newcommand{\avg}{\mathcal{A}}
\newcommand{\avgn}{\mathcal{A}_n}
\newcommand{\avgna}{\mathcal{A}^a_{n}}
\newcommand{\nset}{\{ J_i \}_n}
\newcommand{\bari}{\bar{I}}
\newcommand{\bart}{\bar{t}}
\newcommand{\jac}{\mathcal{J}}
\newcommand{\pert}{\mbox{\boldmath $w$}}
\newcommand{\barpert}{\bar{\mbox{\boldmath $w$}}}
\newcommand{\half}{0.5}

\newcommand{\X}{{\bf X}}
\newcommand{\x}{{\bf x}}
\newcommand{\Z}{{\bf Z}}
\newcommand{\z}{{\bf z}}
\newcommand{\p}{{\bf p}}
\newcommand{\Y}{{\bf Y}}
\newcommand{\y}{{\bf y}}
\newcommand{\ytild}{\tilde{{\bf y}}}
\newcommand{\q}{{\bf q}}
\newcommand{\surf}{\mathcal{S}}
\newcommand{\phij}{\phi_{ij}}
\newcommand{\aphij}{\bar{\phi}_{j}}
\newcommand{\apsij}{\bar{\psi}_{j}}
\newcommand{\aphi}{\bar{\phi}}
\newcommand{\aphiinv}{\bar{\phi}^{-1}}

\newcommand{\domain}{\Omega}
\newcommand{\meanshape}{\bf \bar{x}}
\newcommand{\g}{\mbox{\boldmath $g$}}
\newcommand{\h}{\mbox{\boldmath $h$}}
\newcommand{\yv}{\mbox{\boldmath $y$}}

\newcommand{\group}{Diff}

\newcommand{\evol}{{E_{\text{Vol}}}}
\newcommand{\epair}{{E_{\text{Pair}}}}
\newcommand{\esec}{{E_{S}}}
\newcommand{\inv}{^{-1}}
\newcommand{\vinit}{ \velocity^0_{ij}(\x,\tau=0) }
\newcommand{\avinit}{ \bar{\velocity}^0_{j}(\x,t) }
\newcommand{\avinitz}{ \bar{\velocity}^0_{j}(\x,t_a) }
\newcommand{\gz}{\nabla_{\z}}

\newcommand{\Section}[1]{\vspace{-8pt}\section{\hskip-1em.~~#1}\vspace{-3pt}}
\newcommand{\SubSection}[1]{\vspace{-3pt}\subsection{\hskip -1em.~~#1}\vspace{-3pt}}

\newcommand{\BASE}{.}%/Users/stnava/Work/}
\newcommand{\TEXDIR}{.}%{\BASE Words/Writing/}
\newcommand{\ROOT}{.}%{\BASE Words/}
\newcommand{\FIGDIR}{.}%{\BASE Words/Writing/MEDIAITK/wbir/}
\newcommand{\FIGDIRAIBS}{.}%{\BASE Words/Writing/IPMI2003/jpegs/}
\newcommand{\FIGDIRP}{proposal/}%{\BASE Words/Writing/proposal/}
\newcommand{\FIGDIRPJ}{proposal/jpegs/}%{\BASE Words/Writing/proposal/jpegs/}
\newcommand{\FIGDIRADAD}{.}%{\BASE Words/Writing/ADAD/}
\newcommand{\FIGS}{.}%{\BASE Words/Writing/neuroimageipam-final/FIGS/}

% Use the option doublespacing or reviewcopy to obtain double line spacing
 %\documentclass[doublespacing]{elsart}

% the natbib package allows both number and author-year (Harvard)
% style referencing;
\usepackage{natbib}

% if you use PostScript figures in your article
% use the graphics package for simple commands
% \usepackage{graphics}
% or use the graphicx package for more complicated commands
 \usepackage{graphicx}
% or use the epsfig package if you prefer to use the old commands
% \usepackage{epsfig}

% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb,amsmath}
\usepackage{color}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{rotating}
\usepackage{multirow,booktabs,ctable,array}
\usepackage{listing}						
\usepackage{listings}

\graphicspath{{/Users/nick/Documents/Academic/Gee/SubmittedPapers/JournalPapers/MedIA2009_BrianANTS/Figures/}
%{/Users/nick/Documents/Academic/Gee/SubmittedPapers/Conferences/ISBI2007/Images/}
%                          {/Users/nick/Documents/Academic/Gee/Projects/Helium-3/GrantSubmissions/R21/resplan/Images/}
%                          {/Users/nick/Documents/Academic/Gee/SubmittedPapers/JournalPapers/MRM2008/Figures/}
%                          {/Users/nick/Documents/Academic/Gee/SubmittedPapers/Conferences/ISBI2008/Figures/}
                          }

% The lineno packages adds line numbers. Start line numbering with
% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
% for the whole article with \linenumbers.
 \usepackage{lineno}

%\linenumbers
\begin{document}

%Setup listing environment
\definecolor{listcomment}{rgb}{0.0,0.5,0.0}
\definecolor{listkeyword}{rgb}{0.0,0.0,0.5}
\definecolor{listbackground}{gray}{0.965}

\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listbackground}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = none,
        tabsize = 2,
        language=[ISO]C++,
        floatplacement=!h
        }

\begin{frontmatter}

% Title, authors and addresses

% use the thanksref command within \title, \author or \address for footnotes;
% use the corauthref command within \author for corresponding author footnotes;
% use the ead command for the email address,
% and the form \ead[url] for the home page:
% \title{}
% \thanks[label1]{}
% \author{Name\corauthref{cor1}\thanksref{label2}}
% \ead{tustison@grasp.cis.upenn.edu}
% \ead[url]{home page}
% \thanks[label2]{}
% \corauth[cor1]{}
% \address{Address\thanksref{label3}}
% \thanks[label3]{}

\title{ANTS:  Advanced Open-Source Tools for Normalization And Neuroanatomy}
\author[label1]{PICSL}

% use optional labels to link authors explicitly to addresses:
% \author[label1,label2]{}
\address[label1]{Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, PA, USA}
%\address[label2]{Department of Mathematics, University of Pennsylvania, Philadelphia, PA, USA}

\begin{abstract}
Computational anatomy (CA) seeks to quantify natural variation in
biological shape and function with roots that reach back to the
seminal works of Charles Darwin and D'arcy Thompson.  CA is currently
applied to study health, disease and epidemiology and uses deformable
mappings between images as a basic technique.  However, there is a
lack of standards and reproducibility in the field that is due, in
part, to the use of proprietary software and private data.  To
facilitate reproducibility in CA measurements and advancement of
imaging sciences, NIH has recently committed significant support to
open-source data and software resources.  Here, we report a recent
product of this commitment: Advanced Normalization Tools (ANTS), an
ITK-based toolkit for CA and related areas.  The ANTS open-source
library consists of a well-evaluated suite of state-of-the-art
normalization and template-building tools for quantitative
morphometric analysis.  We highlight the prominent features of ANTS
and demonstrate its utility by performing a detailed analysis on
openly-available anatomically labeled brain data from the non-rigid
image registration evaluation project (NIREP).  The results from this
analysis evidences the high level of accuracy achievable with ANTS
using intensity-based registration alone.  In addition, we show the
significant performance gains may be achieved by coupling
intensity-based image metrics and point-set-based metrics from
specific, sensibly selected cortical structures.
%Computational anatomy is the quantitative study of biological shape and its natural variations which traces its roots to the seminal works of Charles Darwin and D'arcy Thompson.  Integral to this research field is the calculation of deformable mappings between images that serve as invaluable measurement tools.  Technological evolution has spawned much innovation in the computational development of such deformable mapping algorithms.  However, not only is algorithmic development essential for increasing scientific knowledge but the availability of these technologies is equally crucial for facilitating the progress of the scientific community at large.  This is the motivation behind such projects as the Insight Toolkit (ITK) of the National Institutes of Health which serves as a repository for many popular image segmentation and registration routines.  In addition, the ITK library provides a well-structured framework for several stand-alone projects such as Elastix, ITK-SNAP, ManagedITK, the Medical Imaging Toolkit (MITK), WrapITK, and many others.  Our own  contribution to the field of computational anatomy, also built on an ITK foundation, is called Advanced Normalization Tools (ANTS).  The ANTS open-source library consists of an entire suite of state-of-the-art normalization and template-building tools for quantitative morphometric analysis. After highlighting the prominent features of ANTS, we demonstrate its utility by performing a detailed analysis on anatomically labeled brain data from the non-rigid image registration evaluation project (NIREP).  The results from this analysis evidences the high level of accuracy achievable with ANTS using intensity-based registration alone.  In addition, we show the performance gains achieved by coupling intensity-based image metrics and point-set-based metrics from specific, sensibly selected cortical structures.

%In a recent evaluation of 14 non-linear deformation algorithms \cite{Klein2009}, the SyN algorithm of ANTS was consistently one of the top performers.

\end{abstract}

\begin{keyword}
ANTS \sep image registration \sep nueroanatomy \sep neuroimaging \sep normalization \sep  open-source \sep point-set registration
% keywords here, in the form: keyword \sep keyword
% PACS codes here, in the form: \PACS code \sep code
\end{keyword}

\end{frontmatter}

\section{Introduction}

The rapid advancement of biological and medical imaging technologies has caused a proliferation in the development of quantitative tools for computational anatomy.  The principal tools of this emerging field are meaningful deformable mappings between images whether they be driven by similarity metrics which are intensity-based,  point-set based, or both.  Several categories of mappings have been proposed in the literature.  Of particular recent interest are diffeomorphic transformations which, by definition, preserve topology.  Topology preservation is fundamental to making comparisons between objects in the natural world as such transformations permit comparisons to be made across time points in an individual's disease process or to study development patterns across a large population.

Despite the number of proposed algorithms, our limited assessment of published research mirrors the experience of many others who prefer a working paradigm of what has been referred to as {\em reproducible research}.  As described by Dr. Kovacevic ``[reproducible research] refers to the idea that, in `computational' sciences, the ultimate product is not a published paper but, rather, the entire environment used to produce the results in the paper (data, software,etc.).''  After an informal survey of 15 published papers, she found ``none had code available'' and ``in only about half the cases were the parameters [of the algorithm] specified'' (\cite{Kovacevic2006}).  Recent discussions within the computational sciences research community, particularly among advocates of ``open science,'' have also voiced similar concerns (\cite{Yoo2005,Ibanez2006}).   In this paper, we discuss our contribution to the open-source medical image analysis research community which we call ANTS (\underline{A}dvanced \underline{N}ormalization \underline{T}ool\underline{s}).  Built on an ITK framework, this software package comprises a suite of tools for image normalization and template building based on previously published research.

Perhaps the most persuasive evidence motivating the use of our contributions discussed in this paper is the recent outcome of a large-scale comparative image registration algorithm assessment (\cite{Klein2009}). In the largest evaluation study to date involving 14 popular non-linear registration algorithms, our Symmetric Normalization (SyN) transformation model (\cite{Avants2008}) discussed below, was consistently one of the top two performers across all tests.  Overlap and distance measures used for assessment employed three completely independent analysis methods (permutation tests, one-way ANOVA tests, and indifference zone ranking).  Unlike some of the other algorithms used in this brain registration evaluation study, all of our methods (not just SyN) are available as open-source.

We first provide an overview of the different components included in ANTS, such as the available transformation models and similarity metrics offered.  We follow this up with an extensive validation \ldots

\section{Theoretical Overview of ANTS}
A useful classification schema of normalization techniques is based upon the following three principal components (\cite{Brown1992,Ibanez2002}):
\begin{itemize}
  \item the {\em transformation model},
  \item the {\em similarity (or correspondence) measures},  and
  \item the {\em optimization strategy}.
\end{itemize}
In general, image normalization is the process of finding the optimal transformation, $\phi$, within a specified transformation space which maps each $\mathbf{x}$ of image $\mathcal{I}(\mathbf{x})$ to a location in image $\mathcal{J}(\mathbf{z})$ such that a specified cost function, $\mathcal{C}$, describing the similarity between $\mathcal{I}$ and $\mathcal{J}$, is minimized.   A summary of available transformation models and similarity measures are provided in Table \ref{table:chart}.  Details are given in subsequent sections.

\subsection{ANTS Transformation Models}

%\begin{figure*}
%  \centering
%    \begin{tabular}{c}
%      \includegraphics[width=170mm]{TransformationTable}
%    \end{tabular}
%  \caption{Available transformations and similarity metrics available in ANTS.  Similarity metric acronyms:  CC = cross correlation, MSQ = mean squares, MI = mutual information, JHCT = Jensen-Havrda-Charvat-Tsallis divergence, PSE = point-set expectation.  }
%  \label{fig:chart}
%\end{figure*}

\begin{table*}
  \centering
    \begin{tabular}{c c c c}
    {\bf Category} & {\bf Transformation, $\phi$} & {\bf Similarity Measures} & {\bf Brief Description} \\
    \toprule     
    \multirow{2}*{\bf Linear}
           & Rigid$^\dagger$ & MI, MSQ & Rigid registration. \\
       {} & Affine$^\dagger$ & MI, MSQ & Affine registration. \\
       \cmidrule(l){2-4}
    \multirow{2}*{\bf Elastic}
           & Deformable & CC, PR, MI, MSQ, PSE & Demons-like algorithm. \\
       {} & DMFFD & CC, PR, MI, MSQ, PSE  & FFD variant. \\
       \cmidrule(l){2-4}
    \multirow{2}*{\bf Diffeo.}
           & Exponential$^\dagger$ &  CC,PR, MI, MSQ, PSE  & $~\min~\velocity(\x)$ \\
       {} & Greedy SyN$^\dagger$ &  CC, PR, MI, MSQ, PSE  & ~~locally in time~$\min~\velocity(\x,t)$\\
       {} & Geodesic SyN$^\dagger$ &  CC, PR, MI, MSQ, PSE  &  $~\min~\velocity(\x,t)$ over all time \\
    \bottomrule
    \end{tabular}
  \caption{Transformations and a subset of the similarity metrics available in ANTS.  Similarity metric acronyms:  CC = cross correlation, PR = probabilistic matching, MSQ = mean squared difference, MI = mutual information, PSE = point-set expectation.  ANTS also provides the inverse of those transformations denoted by the `$\dagger$' symbol.  The brief descriptions of the diffeomorphic algorithms contrast the way in which the velocity field is optimized and used to parameterize $\phi$, the mapping.  All 
ANTS Diff algorithms generate $\phi(\x,t)$ over $t \in [ 0, 1]$ through gradient descent.}
  \label{table:chart}
\end{table*}    
\vspace{0in}

A variety of transformation models have been proposed in the literature with varying degrees of freedom (illustrated in Figure \ref{fig:hierarchy}).  For deformable transformations, one approach is to optimize within the space of non-topology preserving, yet physically motivated transformations---an approach pioneered by Bajcsy (\cite{Bajcsy1989}).  Elastic and related models, such as HAMMER (\cite{Shen2002}), statistical parametric mapping (SPM) (\cite{Ashburner2000}), free-form deformations (FFD) (\cite{Rueckert1999}), and
Thirion's demons (\cite{Thirion1998}) operate in the space of vector fields, which does not preserve topology.  In other words, barring ad hoc constraints to prevent otherwise, these algorithms allow the topology to change in an uncontrolled way which makes the deformable mappings difficult to interpret in functional or anatomical studies.

These topology concerns have led to current advocation within the computational anatomy research community of diffeomorphic mappings for transformation modeling.   Optimizing directly within this space has shown remarkable success in various computational anatomy studies involving longitudinal (\cite{Avants2007b,Fox2001}), functional (\cite{Miller2005}), and population data (\cite{Avants2007b}).  We include three such diffeomorphic algorithms in the ANTS toolbox based on previously reported research.

Regardless of current research trends, however, we recognize that selection of the transformation model is ultimately application-specific, that no single choice is optimal for all scenarios (\cite{Wolpert1997}), and, therefore, the transformation model must be chosen in a principled fashion.  In fact, several non-diffeomorphic algorithms performed quite well in Klein's comparative study of nonrigid registration algorithms (\cite{Klein2009}).  For this reason, we also include elastic-related methods for a panoply of transformation model options.

\begin{figure}
  \centering
    \begin{tabular}{c}
      % \includegraphics[width=80mm]{hierarchy}
    \end{tabular}
  \caption{Diagrammatic illustration of the transformation model hierarchy where the encompassing transformation spaces are characterized by increasing degrees of freedom.  }
  \label{fig:hierarchy}
\end{figure}

\subsubsection{Linear Transformations--Rigid and Affine:}
Image registration strategies often begin with a linear transformation
for initial global alignment which follows a deformable
transformation with increased degrees of freedom.  
Affine transformations within ANTS optimize either a
mean-squared difference metric or a mutual information (MI) image
similarity metric.  ANTS affine transformations are optimized
w.r.t. translation, rotation, scaling and shearing in succession.
This strategy allows careful control of each component of the affine
transformation.  Internally, ANTS also composes the affine
transformation with the deformation field before performing any
interpolation or downsampling.  In this way, ANTS normalization never
requires more than a single image interpolation step and is able to
always refer back to the original full-resolution image sources.

\subsubsection{Diffeomorphic Transformations}
In contrast to many transformation models which reside in the domain of vector spaces, a diffeomorphism is a differentiable mapping with a differentiable inverse (\cite{Ebin1970,Mumford1998}).  Modeling transformations with diffeomorphisms ensures certain desirable topological properties that cannot be guaranteed with other methods.

ANTS assumes the diffeomorphism (differentiable mapping with a differentiable
inverse), $\phi$, is defined on the image domain, $\Omega$, and
maintains an affine transform at the boundary such that
$\phi(\partial \Omega) = A(\mathbf{Id})$ where $A(\mathbf{Id})$ is an
affine mapping applied to the identity transformation.  The $\phi$,
over time, parameterizes a family of diffeomorphisms,
$\phi(\mathbf{x}, t) : \Omega \times t
\rightarrow \Omega$, which can be generated by integrating a
time-dependent, regularized velocity field, $\boldsymbol{v} : \Omega \times
t \rightarrow \mathbb{R}^d$, through an ordinary differential equation
(o.d.e.).  
\begin{align} \label{eq:ode}
\frac{d \phi(\mathbf{x}, t)}{dt} = \boldsymbol{v}(\phi(\mathbf{x}, t), t), \,\,\, \phi(\mathbf{x}, 0) = \mathbf{x}.
\end{align}
The existence and uniqueness theorem for o.d.e.'s implies that integrating Equation (\ref{eq:ode}) generates a diffeomorphism.  
The deformation field provided by $\phi$ is
$\mathbf{u}(\mathbf{x})=\phi(\mathbf{x}, 1) - \mathbf{x}$.

Dupuis et al. (\cite{Dupuis1998}) motivated the usage of diffeomorphisms for CA  by showing that the variational form
\begin{align}\label{eq:D}
  D(\mathcal{I}, \mathcal{J}) =  \int_0^1 ||L\boldsymbol{v}(x,t)||^2 dt , \,\, \mathcal{I}(\phi(\mathbf{x}, 1)) = \mathcal{J}(\mathbf{z})
\end{align}
represents a true mathematical metric between anatomical instances $\mathcal{I}$ and $\mathcal{J}$ given an appropriate norm, $L$, on the velocity field, $\boldsymbol{v}$.  An optimal solution, $\boldsymbol{v}^*$, minimizes the metric $D(\mathcal{I}, \mathcal{J})$ with respect to $L$.   Dupuis (\cite{Dupuis1998}) also showed that such a solution is guaranteed to be well-posed.  Intuitively, Equation (\ref{eq:D}) provides a sense of distance between two anatomical shapes.  It also illustrates that the optimal diffeomorphic solution is analogous to finding the geodesic curve between two points in a curved space.
\footnote{
It is important to note the similarity between the definition of curve length, $\int ||\mathcal{C}'(t)||dt$, for the parametric curve $\mathcal{C}(t)$ and Equation (\ref{eq:D}).  In this sense, the solution for Equation (\ref{eq:D}) is the geodesic diffeomorphism, where $\boldsymbol{v}$ is the tangent vector of the diffeomorphism, such that the shape distance, $D$, between $\mathcal{I}$ and $\mathcal{J}$ is minimized.
}

In most real-world applications, however, a diffeomorphic path connecting the anatomical instance $\mathcal{J}$ with $\mathcal{I}$ is non-existent (due, for example, to the photometric variation or the presence/absence of a tumor in neuroanatomical images).  Therefore, the following minimizing variational form is used for optimization in diffeomorphic normalization to accommodate inexact matching  (\cite{Dupuis1998,Miller2002})
\begin{align} \label{eq:lddmm}
  \boldsymbol{v}^* = \underset{\boldsymbol{v}}{\operatorname{argmin}}  \left\{ \int_0^1  ||L\boldsymbol{v}||^2  dt + \lambda || \mathcal{I} \circ \phi(\mathbf{x},1) - \mathcal{J}  ||  \right\}.
\end{align}
The Euler-Lagrange equations characterizing the optimizing time-varying velocity field, $\boldsymbol{v}^*$, were derived in \cite{Miller2002} and later used in formulating the gradient-descent optimization scheme known as {\em large deformation diffeomorphic metric-matching} (LDDMM) (\cite{Beg2005}).   The similarity metric, or data term, for LDDMM is the {\em squared intensity difference} with weight $\lambda$.

In medical image registration, one typically encounters more complex intensity transfers 
between one anatomical instance $\mathcal{J}$ and another instance $\mathcal{I}$.  Thus, 
ANTS enables a variety of similarity terms beyond the squared difference:
\begin{align} \label{eq:diff}
  \boldsymbol{v}^* = \underset{\boldsymbol{v}}{\operatorname{argmin}}  \left\{ \int_0^1  ||L\boldsymbol{v}||^2 dt  +~\lambda~\Pi( \mathcal{I}, \phi(\mathbf{x},1) , \mathcal{J} )  \right\}.
\end{align}
Here, $\Pi$ is a similarity metric depending on the images and the
mapping and $\lambda$ controls the degree of exactness in the
matching.  If $\Pi$ is selected as cross-correlation, then one is
estimating the diffeomorphism under more robust illumination
constraints, as described in \cite{Avants2008}.  A similar equation
for elastic/FFD/DMFFD matching may be minimized in ANTS.

Exploiting the fact that the diffeomorphism, $\phi$, can be decomposed into two components 
$\phi_1$ and $\phi_2$, Avants et al. construct a {\em symmetric} alternative to Equation (\ref{eq:diff}) (\cite{Avants2007b}).  This leads to the symmetric variant of Equation (\ref{eq:diff})
\begin{align}\label{eq:diffs}
  \{\boldsymbol{v}_1^*,\boldsymbol{v}_2^*\}  =
     \underset{\boldsymbol{v_1},\boldsymbol{v_2}}{\operatorname{argmin}}  
      \Bigg\{ & \int_0^{0.5} ( ||L\boldsymbol{v}_1(x,t)||^2 + ||L\boldsymbol{v}_2(x,t)||^2   )dt  \nonumber \\
     &+ || \mathcal{I} \circ \phi_1(\mathbf{x},0.5) - \mathcal{J}\circ \phi_2(\mathbf{x},0.5)   ||\Bigg\}.
\end{align}
The corresponding symmetric Euler-Lagrange equations are similar to (\cite{Miller2002}).  Finding $\boldsymbol{v_1}^*$ minimizes the variational energy from $t = 0$ whereas $\boldsymbol{v_2}^*$ minimizes from $t = 1$.  Thus, gradient-based iterative convergence deforms $\mathcal{I}$ and $\mathcal{J}$ along the geodesic diffeomorphism, $\phi$, to a fixed  point midway (intuited by the notion of shape distance) between $\mathcal{I}$ and $\mathcal{J}$.

ANTS allows a variety and combination of data terms to be optimized
with diffeomorphisms and, at the same time,
enables a variety of models for parameterizing the $\phi(\cdot)$.  
Briefly, ANTS Diff models are: 
\begin{enumerate}
\item {\bf Exponential Mapping:}  This constant velocity field parameterization does not produce a distance metric; maps are not dense in the space of diffeomorphisms and are not inverse consistent.
\item {\bf Greedy SyN:} Approximates the distance metric with a fast approach.  Evaluated in \cite{Klein2009}.  Enforces $\phiinv(\phi(\x,1))=\x$ in the discrete domain.  Theoretically and computationally inverse consistent / symmetric. 
\item {\bf Geodesic SyN:}  Symmetrically minimizes the velocity field / distance metric explicitly over all time.  Computationally more expensive than, but produces similar solutions to, Greedy SyN for most medical imaging problems.
\end{enumerate}
All of these transformation models may be used with any of the ANTS similarity metrics.  In particular, ANTS allows $\Pi$ 
to be composed of a summation/weighted combination of available similarity metrics/data terms.

\paragraph{Exponential Mapping}
The key difference between a time-varying diffeomorphism and a diffeomorphism 
generated by an exponential mapping (Ashburner200x) is that the exponential 
mapping maintains only a single vector field that is constant in time.   Theoretically, 
the restricting the velocity field to be constant in time reduces the size of the 
space that may be generated \cite{ArnoldODE} in a way that is similar to the difference 
between real and rational numbers, the latter of which are sparsely distributed through 
the reals.  
In contrast, 
the full diffeomorphic space may be explored by using time-varying velocity fields 
as does SyN and LDDMM.  The exponentiation of a constant velocity field generates a 
diffeomorphism through, 
\begin{align} \label{eq:odec}
\frac{d \phi(\mathbf{x}, t)}{dt} = \boldsymbol{v}(\phi(\mathbf{x},t) ), \,\,\, \phi(\mathbf{x}, 0) = \mathbf{x}.
\end{align}
Note that there is no time parameter in the velocity field, here.   The Dartel algorithm uses 
the well-known ``scaling and squaring'' approach to efficiently generate $\phi$ from a velocity field $\boldsymbol{v}(x)$.

\paragraph{Greedy SyN}
Greedy optimization of equation~\ref{eq:diffs}, involves gradient and update steps 
as follows, 
\begin{align}
\text{gradient:   }  \nabla E = \partial_{\phi_i} \Pi ( \mathcal{I}(\phiinv_1(\x,0.5)) ,  \mathcal{J}(\phiinv_2(\x,0.5)) ) \nonumber \\ 
\text{regularize:  } \nabla E \leftarrow G_\sigma \star \nabla E \nonumber \\ 
\text{update:  }  \phi_i(\x,0.5) \leftarrow \phi_i(\x,0.5)+\delta \nabla E ( \phi_i(\x,0.5) ). \nonumber \\
\text{update the inverse s.t.:  } \phi_i(\phiinv_i) = \Id.  
\end{align}
Though the gradients differ for each metric, the update scheme is the
same.  In words, the gradient is taken at the mid-point of the full
diffeomorphism, $\phi$, by decomposing with respect to its components,
$\phi_1$ and $\phi_2$.  The gradient with respect to each half is
computed, regularized with a Gaussian of variance $\sigma$ and then
mapped back to the origin of each diffeomorphism to update the full
mapping.  Hermosillo provides a good discussion of the gradients of
the similarity metrics available in ANTS
\cite{hermosillo}.  Note that, here, we supply a specific 
regularization scheme but others are available in ANTS.  
That is, one may directly regularize the $\phi_i$ or use 
no regularization at all.

\paragraph{Time-Varying SyN}
Global in time optimization of equation~\ref{eq:diffs}, involves a global gradient 
where, at each time point $t \in [0, 1]$, we have:
\begin{align}
\text{gradient:   }  \nabla E(\x,t) = \partial_{\phi_i} \Pi ( \mathcal{I}(\phiinv_1(\x,t)) ,  \mathcal{J}(\phiinv_2(\x,1-t)) ), \nonumber \\ 
\text{update:  }  \velocity(\x,t)=\velocity(\x,t)+G_\sigma \star \nabla E(\x,t). \nonumber \\
\text{reintegrate the o.d.e.s, via Runge-Kutta},  \nonumber \\
\text{that generate the $\phi_i$ in $[0,1]$.}  
\end{align}
The time-varying SyN algorithm is theoretically cleaner than greedy
SyN.  However, the computational and memory cost is greater due to the
dense-in-time gradient and requisite reintegration of the
diffeomorphisms after each update.  Note that, here, we supply a
specific regularization scheme but others are available in ANTS.

%\input{vectorspace.tex}

\subsubsection{Vector Field Operators:} Deformable normalization
strategies typically regularize a deformation, $\displace$, or
velocity field, $\velocity$, with a linear operator such as the
Laplacian or Navier-Stokes operator. One may write this regularization
as a variational minimization in terms of its linear operator or in
terms of a kernel function operating of the field itself, e.g., $
\displace_{smooth}= K \star \displace_{not-smooth}$, where $K$ is 
the Green's kernel for the linear operator, $L$ and $\star$ is
convolution.  ANTS enables a variety of choices for the kernel, $K$,
including the Gaussian with varying $\sigma$ and a variety of BSpline
functions, both of which induce adequate regularity for normalization
models used in ANTS.  Regularization models operate on either/both the
whole mapping, $\phi$, or/and the gradient of the similarity term.
The same schema is applied for diffeomorphic and the recently proposed directly-manipulated 
free-form-deformation
(DMFFD)\cite{Tustison2009}, allowing regularization of both total 
deformation and deformation update.  While additional physical operators will be
included in future releases, current BSpline options provide many orders 
of flexibility \cite{Tustison2005}.

\subsubsection{Vector Space Transformations}
Potential mapping solutions to the image matching problem operating in vector spaces are 
constructed in a similar variational form as that for the diffeomorphic formulation.
We write this general variational energy, $\Pi$, as 
\begin{align} \label{eq:energy}
  \Pi(\mathcal{I}, \mathcal{J}, \phi) = \int_{\Omega} \left(\Pi_{\sim}(\mathcal{I}, \mathcal{J}, \phi)(\mathbf{x}) + 
                  \Pi_{R}(\phi)(\mathbf{x})\right) d\Omega,
\end{align}
where $\mathcal{I}$ and $\mathcal{J}$ are, again, the moving and fixed images, respectively, and $\phi$ is the transformation which maps between $\mathcal{I}$ and $\mathcal{J}$.  $\Pi_{\sim}$ is the similarity metric and $\Pi_{R}$ is the explicit regularization term.

\paragraph{Gaussian-Regularized Elastic Deformation}

A simple and efficient, yet powerful image normalization algorithm is Thirion's demons (\cite{Thirion1998}).  Using an optical flow based similarity, the solution is obtained by iterating between the calculation of image forces and subsequent Gaussian regularization.  In ANTS, we extend this basic approach to include both cross-correlation and mean-squared difference.

\paragraph{Directly Manipulated Free-Form Deformation}
Another top performer in Klein's study (\cite{Klein2009}) was the Image Registration Toolkit (IRTK) based on the research originally reported in \cite{Rueckert1999} in which mutual information and a free-form deformation (FFD) transformation model were used to analyze breast deformation.
In ANTS we provide an implementation of a variant of the well-known FFD transformation model for image registration known as {\em directly manipulated free-form deformation} (\cite{Tustison2009}).   The DMFFD model replaces the standard FFD gradient used in \cite{Rueckert1999} with an intuitive preconditioned gradient to overcome problematic energy topographies intrinsic with the traditional approach.

For $n$-D images, the FFD transformational model, $\phi_{FFD}$, is defined as 
\begin{align}\label{eq:TFFD}
  \phi_{FFD} =  \sum_{i_1=1}^{M_1}\ldots\sum_{i_n=1}^{M_n} \mathbf{P}_{i_1, \ldots, i_n} \prod_{j=1}^nB_{i_j,d_j}(u_{j})     
\end{align}
where $\mathbf{P}_{i_1, \ldots, i_n}$ is an $n$-D grid of control points and $B_{i_j,d_j}(u_{j})$ is the B-spline in the $i_j^{th}$ direction of order $d_j$.

The gradient of the image normalization energy, $\Pi$, with respect to the control points used during gradient-based optimization is easily calculated to be
\begin{align} \label{eq:FFDgrad}
  \frac{\partial \Pi}{\partial \mathbf{P}_{i_1, \ldots, i_n}} =  \sum_{c=1}^{N_{\Omega}} \left(
  \frac{\partial \Pi_{\sim}}{\partial \phi} + \frac{\partial \Pi_{R}}{\partial \phi} 
  \right)_c \prod_{j=1}^nB_{i_j,d_j}(u^c_{j})
\end{align}
which is the gradient used in (\cite{Rueckert1999}).  In contrast, the DMFFD approach uses a preconditioned gradient given by
\begin{align} \label{eq:DMFFDgrad}
 \frac{\partial \Pi}{\partial \mathbf{P}_{i_1, \ldots, i_n}} =& 
        \Bigg( \sum_{c=1}^{N_{\Omega}}   \left( \frac{\partial \Pi_{\sim}}{\partial \phi} + \frac{\partial \Pi_{R}}{\partial \phi}  \right)_c          \prod_{j=1}^n B_{i_j, d_j}(u^c_j)   \nonumber \\
  &  \cdot \frac{\prod_{j=1}^n B^2_{i_j, d_j}(u^c_j)}{\sum_{k_1=1}^{d_1+1}\ldots\sum_{k_n=1}^{d_n+1} \prod_{j=1}^n B^2_{k_j, d_j}(u^c_j)} \Bigg)  \nonumber \\
          &\cdot \left( \frac{1}{ \sum_{c=1}^{N_\Omega} \prod_{j=1}^n B^2_{i_j, d_j}(u^c_j)  } \right).
\end{align}
The difference between the two gradients is seen to reside strictly in terms of the B-spline shape functions which serve to normalize the DMFFD gradient in a unique fashion so as to minimize its susceptibility to hemstitching during the course of optimization.
%A multi-resolution approach (\cite{Lee1997}) is included such that the B-spline mesh resolution is doubled at each stage of the optimization from a specified initial mesh resolution.  Also, all feasible B-spline orders are available (not just cubic).

\subsection{ANTS Intensity Based Similarity Metrics}
Several intensity-based image metrics have been proposed in the literature with varying levels of performance dependent upon specific applications.  We have included several of the most popular similarity metrics within ANTS.  In addition, our software framework facilitates the development of other image metrics if that is desired by the user.

Both mutual information (\cite{Viola1997}) and mean-squared difference similarity metrics are available for the linear transformations.  Also included are the cross-correlation (\cite{Avants2007b}), probabilistic matching (more aggressive version of cross-correlation), local mutual information (\cite{Rueckert1999,Pluim2003}), and mean squared difference similarity metrics for the non-linear transformation models.  For symmetric transformations, the inverse metrics and gradients are also calculated.  Parameters for the different metrics are also available (see the ANTS website (\cite{ANTS2009}) for further details).

\subsection{ANTS Point-Set Based Similarity Metrics}
In addition to intensity-based metrics, ANTS also contains similarity metrics for 
registering point-sets.  These include a landmark matching metric and 
two point-set metrics which can accommodate point-sets of different cardinality.
These point-set metrics can be used alone for strict point-set registration or
in conjunction with intensity-based metrics for dual intensity/point-set registration.

\subsubsection{Exact Landmark Matching}
Generalizing the B-spline fitting algorithm of Lee et al. (\cite{Lee1997}), 
we developed a scattered data approximation algorithm (\cite{Tustison2006b}) 
and contributed the code to the ITK library (\cite{Tustison2005}).  
This code is included in ANTS and forms the basis of our exact landmark 
matching where this metric seeks to minimize the weighted sum of distances between 
corresponding landmarks using a hierarchical approach.  One can also associate relative confidence values with each landmark for fine-tuning exact landmark matching results.

\subsubsection{Point-Set Expectation}
In (\cite{Pluta2008}), the point-set matching problem was formulated in the context 
of incomplete label matching but is equally applicable to the general scenario of 
registering point-sets not necessarily of equivalent cardinality.  Given two point-sets, 
$X$ and $Y$, the essential idea underlying the point-set expectation matching 
algorithm is that the optimal solution minimizes the distance between each point $y \in Y$
with its corresponding {\em expected} point in $X$.

We calculate the expected point using a Bayesian formulation and a non-parametric
Parzen windowing scheme.  This allows one to define the probability of the point $x \in X$ 
given a point $y\in Y$ as
\begin{align}
  \mathbf{P}(X = x | Y = y ) = G(y; x, \sigma_X )
\end{align}
where $G(y; x, \sigma_X)$ is a normalized Gaussian with mean $x$ and
standard deviation $\sigma_X$.  
The expected point $\mathrm{E}(X|y)$ is then calculated to be
\begin{align} \label{eq:expected}
  \mathrm{E}(X | Y = y) &= \sum_{j = 1}^{|X|} \mathbf{P}(X = x_j | Y = y ) x_j\nonumber \\
                                        &= \frac{1}{|X|}\sum_{j = 1}^{|X|} G(y; x_j, \sigma_X) x_j  
\end{align}
where $|\cdot|$ denotes cardinality.
The weighted sum of distances 
between the points in $Y$ and their corresponding expected points in $X$ is calculated from 
Equation (\ref{eq:expected}), i.e.
\begin{align}
\mathrm{PSE}(X,Y) = \frac{1 }{|Y|}\sum_{i=1}^{|Y|} \left\|y_i - \frac{1}{|X|} \sum_{j = 1}^{|X|} G(y_i; x_j, \sigma_X) x_j\right\|^2.
\end{align}

\subsubsection{Jensen-Havrda-Charvat-Tsallis Divergence}
Recent information theoretic approaches have been proposed for point-set registration.  A previous open-source contribution (\cite{Tustison2008b}) generalizes the Jensen-Shannon divergence to the Jensen-Havrda-Charvat-Tsallis (JCHT) divergence which
permits a fine-tuning  of the divergence measure such that emphasis can vary between robustness and sensitivity for 
application-speciÞc tailoring.

Each point-set is represented as a PDF via a Gaussian mixture model (GMM).  Assuming $K$ point-sets denoted by $\{X_k, k \in \{1,\ldots,K\}\}$, the $k^{th}$ point-set is denoted by $\{x^k_1, \ldots, x^k_{|X_k|}\}$.  The corresponding $k^{th}$ PDF is calculated from the $k^{th}$ point-set as
\begin{align}
  \mathbf{P}_k(s) = \frac{1}{|X_k|} \sum_{i=1}^{|X_k|} G( s; x^k_i, C^k_i )
\end{align}
where $G(s; x^k_i, C^k_i)$ is a normalized Gaussian with mean $x^k_i$ and covariance $C^k_i$ evaluated at $s$.  For each point, $x_i$, the associated weighted covariance matrix, $C_{\mathcal{K}_i}$, is given by 
\begin{align} \label{eq:C}
  C_{\mathcal{K}_i} = \frac{ \sum_{x_j \in \mathcal{N}_i, x_j \neq x_i} \mathcal{K}(x_i; x_j) (x_i - x_j)^\mathrm{T} (x_i - x_j)}{\sum_{x_j \in \mathcal{N}_i, x_j \neq x_i} \mathcal{K}(x_i; x_j) }
\end{align}   
where $\mathcal{N}_i$ is the local neighborhood of the point $x_i$ and $\mathcal{K}$ is a user-selected neighborhood  weighting kernel.   We use an isotropic Gaussian for $\mathcal{K}$ with variance $\sigma_{\mathcal{K}_i}^2$ as well as a k-d tree structure for efficient determination of $\mathcal{N}_i$ (\cite{Berg2000}).  Calculation of the gradient requires the inverse of each covariance matrix.  To avoid ill-conditioned covariance matrices, we use the modified covariance  
$C_i = C_{\mathcal{K}_i} + \sigma_n^2 I$
where $I$ is the identity matrix and $\sigma_n$ is a parameter denoting added isotropic Gaussian noise.

We designate the number of sample points generated for each of the $K$ probability density functions as $\{ M_1, \ldots, M_K \}$ and the $k^{th}$ set of points as $\{ s_1^k, \ldots, s_{M_k}^k \}$.  The JHCT divergence is then calculated using the $K$ sets of points and the formula
\begin{align} \label{eq:JHCT2}
  \mathrm{JHCT}_\alpha(&\mathbf{P}_1, \ldots, \mathbf{P}_K) =  \frac{1}{ 1-\alpha}  
  \nonumber \\
   &\left[\frac{1}{M} \left( \sum_{k=1}^K \sum_{j=1}^{M_k} \left[ \mathbf{P}^*(s_j^k) \right]^{\alpha-1} -1 \right)\right. \nonumber \\
   +&  \left. \frac{1}{N}\sum_{k=1}^K \frac{|X_k|}{M_k}   \left( \sum_{j=1}^{M_k} 
       \left[ \mathbf{P}_k(s_j^k) \right]^{\alpha-1} - 1 \right) \right]
\end{align}
where 
\begin{align}
\mathbf{P}^*(X) = \frac{1}{N}\sum_{k=1}^K \sum_{i=1}^{|X_k|} G(x;x_i^k, C_i^k), 
\end{align}
$N = \sum_{k = 1}^K |X_k|$, and $M = \sum_{k = 1}^K M_k$.
The prior weighting values are calculated from  $\gamma_k = |X_k|/N$ such that the larger point-sets are weighted more heavily.

\section{ANTS Implementation and Usage}
ANTS, built upon an ITK foundation, maintains the same coding style as its base.  For much of its functionality, ANTS requires ITK, necessitating the installation of ITK prior to installing ANTS.  All ANTS source code is available via the online source code repository SourceForge.%
\footnote{
http://sourceforge.net/projects/advants/
}
Binaries are also available from the same online location.

Based on our experience with standard command line argument parsing packages (e.g. \verb#getopt#), we developed our own set of classes for a more intuitive command line interface (at least from our perspective).  A summary of command line arguments are given in Table \ref{table:args}.  We invite the interested reader to visit our evolving ANTS web site (\cite{ANTS2009})for further details.
These ANTS argument parsing classes provide an intuitive compromise between parsers where every variable requires a unique flag and strict ordering requirements on the command line.  A typical command line call to ANTS  is  given by
\begin{lstlisting}
>ANTS 3 
  --metric MSQ[fixedImage.nii,movingImage.nii,0.75] 
  --metric PSE[fixedPoints.vtk,movingPoints.vtk,0.25] 
  --transformation SyN[0.25] 
  --regularization Gauss[3.0,0.5]
  --iterations 50x20x10x5 
  --output-naming results.nii
\end{lstlisting}
The correspondence between the ANTS command line specification and the diffeomorphic image normalization formulation with labeled point-sets $X$ and $Y$ illustrates the motivation for our command line interface,
\begin{align}
  \Pi(&\mathcal{I},\mathcal{J},X,Y,\phi) \nonumber \\ 
     =& \underbrace{\int_0^1 ||L\boldsymbol{v}||dt}_{\texttt{-t SyN[}\cdot \texttt{]},\texttt{ -r Gauss[}\cdot\texttt{,}\cdot\texttt{]}}  + \underbrace{\lambda_1\int_0^1 || \mathcal{I} \circ \phi(\mathbf{x},t) - \mathcal{J}  || dt}_{\texttt{-m MSQ[}\mathcal{J}\texttt{,}\mathcal{I}\texttt{,}\lambda_1\texttt{]}} \nonumber \\
&+\underbrace{\frac{\lambda_2 }{|Y|}\sum_{i=1}^{|Y|} \left\|y_i - \frac{1}{|X|} \sum_{j = 1}^{|X|} G(y_i; x_j, \sigma_X) x_j\right\|^2}_{\texttt{-m PSE[}Y\texttt{,}X\texttt{,}\lambda_2\texttt{]}}.
\end{align}

\begin{table*}
  \centering
    \begin{tabular}{c c c c c}
   {\bf } & {\bf Argument} & {\bf Flag} & {\bf Variables}  & {\bf Sample Parameters}\\
    \toprule
    \multirow{3}*{\bf Linear}
    & Iterations & \verb#--linear-iterations# & {} & $N_1$\verb#x#$N_2$\verb#x#$N_3$\verb#x#\ldots \\ 
    & Similarity & \verb#--linear-metric# & \verb#MI,MSQ# & \verb#[#$N_{bins}$,$N_{samples}$\verb#]# \\ 
    {} & Affine or Rigid &  \verb#--do-rigid# & {} & \verb#0# \\
    \cmidrule(l){2-5}
    \multirow{5}*{\bf Deform.}
    & Image Similarity & \verb#--metric,-m# & \verb#MI,CC,PR,MSQ# & \verb#[#$\mathcal{I},\mathcal{J}$,\verb#radius]# \\
    & Point-Set Similarity & \verb#--metric,-m# & \verb#PSE,JHCT# & \verb#[#$\mathcal{I},\mathcal{J},X,Y$\verb#]# \\
    {} & Iterations/Level & \verb#--iterations,-i# & {} & $N_1$\verb#x#$N_2$\verb#x#$N_3$\verb#x#\ldots \\ 
    {} & Regularization &  \verb#--regularization,-r# & \verb#Gauss,DMFFD# & \verb#[#$\sigma_{total}$,$\sigma_{gradient}$\verb#]# \\
    {} & Transformation & \verb#--transformation,-t# & \verb#Elast,SyN,Exp# & \verb#[#$\Delta_{gradient}$\verb#]# \\
    \cmidrule(l){2-5}
    \multirow{2}*{\bf Misc.}
    & Histogram Match $\mathcal{I},\mathcal{J}$ & \verb#--use-histogram-matching# & {}  & \verb#1# \\
    {} & NN Interpolation & \verb#--use-NN# & {} & \verb#0#\\
    {} & Mask Image &  \verb#--mask,-x# & {} & \verb#mask.nii# \\
    {} & Output Naming     & \verb#--output-naming,-o# & {} & \verb#filename.nii# \\
    \bottomrule
    \end{tabular}
  \caption{  Further information can be found on the ANTS website (\cite{ANTS2009}). }
  \label{table:args}
\end{table*}

% use section* for acknowledgement
\section*{Acknowledgment}
ANTS is supported by Grant 1R01EB006266-01 from the National Institute Of Biomedical Imaging and Bioengineering and administered through the UCLA Center for Computational Biology.
% optional entry into table of contents (if used)
%\addcontentsline{toc}{section}{Acknowledgment}
%The authors would like to thank...

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section
% NOTE: BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/

% can use a bibliography generated by BibTeX as a .bbl file
% standard IEEE bibliography style from:
% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/IEEEtran/bibtex
%\bibliographystyle{IEEEtran.bst}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)

\bibliographystyle{plain}
%elsarticle-num-names.bst}
\bibliography{references,ants}

\end{document}

### 2009-06-04 Documentation/Tex/runtex.sh

@@ -1,0 +1,11 @@
+
+for FL in ants antstheory ; do 
+for i in 1 2 3 
+do
+pdflatex $FL 
+bibtex   $FL
+done
+cp ${FL}.pdf ../
+done 
+rm -f *.bbl *.log *.aux *.toc *.blg *.brf *.out *.spl *~ 
+

### 2009-06-05 Documentation/Tex/ants.tex
figs and minor updates
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@10 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 ANTS/Examples/Data/ directory.
+\begin{figure}
+\label{fig:frown}
+\includegraphics[width=0.45\textwidth]{Figures/frowntosmile.png}
+\includegraphics[width=0.45\textwidth]{Figures/smiletofrown.png}
+\itkcaption[Frown To Smile]{Expected output for the frown to smile 
+shows a smooth, though large deformation.  The grids are overlaid on 
+the deformed images. }
+\end{figure}
 \begin{verbatim}
-ANTS 2 -o PSEtest  -i 91x70x55x40x30  -r Gauss[3,0.] -t SyN[0.3]  
-   -m  PSE[Frown.nii,Smile.nii,Frown.nii,Smile.nii,1,0.2,11,0,25] 
-    -m  MSQ[Frown.nii,Smile.nii,1,0] 
-     --continue-affine 0  --number-of-affine-iterations 0
+ANTS 2 -o PSEtest  -i 91x70x55x40x30  -r Gauss[3,0.] -t SyN[0.2]
+   -m  PSE[Frown.nii,Smile.nii,Frown.nii,Smile.nii,0.75,0.1,11,0,10] 
+   -m  MSQ[Frown.nii,Smile.nii,1,0]    --number-of-affine-iterations 0
 
-WarpImageMultiTransform  2 Frown.nii temp.nii -R Smile.nii
+WarpImageMultiTransform  2 Frown.nii FrownToSmile.nii -R Smile.nii
    -i PSEtestAffine.txt PSEtestInverseWarp.nii
 
-WarpImageMultiTransform  2 Smile.nii temp2.nii -R Frown.nii 
+WarpImageMultiTransform  2 Smile.nii SmileToFrown.nii -R Frown.nii 
     PSEtestWarp.nii PSEtestAffine.txt

### 2009-06-05 Documentation/Tex/ants.tex
updates to pdfs
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@11 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 
-\noindent{\bf Warping and Invertibility.}ANTS uses physical space to define mappings. The origin etc is in the
-coordinates of the world in which the picture (e.g. MRI) was
-taken. Then, warping between physical coordinates is relatively
-easy. Differences in bounding boxes etc present no problem -- assuming
-you avoid inconsistent headers i.e. origins, directions, data
-orientation. One may use PrintHeader to check the data and run simple,
-fast tests (few iterations) to perform sanity checks before running
-through loads of data.
+\noindent{\bf Warping and Invertibility.}  ANTS uses physical space to
+define mappings. The origin etc is in the coordinates of the world in
+which the picture (e.g. MRI) was taken. Then, warping between physical
+coordinates is relatively easy. Differences in bounding boxes etc
+present no problem -- assuming you avoid inconsistent headers
+i.e. origins, directions, data orientation. One may use PrintHeader to
+check the data and run simple, fast tests (few iterations) to perform
+sanity checks before running through loads of data.
 An ANTS deformation consists of
@@
 Here, we see that invertibility is not exact but is gained only by interpolation.  
-Thus, in three-dimensional scenarios, there are fundamental limits to the 
+Thus, in three-dimensional scenarios in particular, there are fundamental limits to the 
 degree of invertibility that may be achieved.  The second and third voxels -- from left -- 
@@
 by  ANTSJacobian -- which is a unitless measure defined by the 
-ratio of volumes.  Thus, $\mathcal{J}(\x)=V_1(\phi(\x))/V_2(\x)$ where $V_i$ represents 
-the volume in one domain versus the other.  Thus, if $\phi$ -- the mapping -- 
+ratio of volumes.  Thus, $\mathcal{J}(\x)=V(\phi(\x))/V(\x)$ where $V$ represents 
+the volume operation and $\x$, here, may be a small object.  Thus, if $\phi$ -- the mapping -- 
 causes expansion, then $\mathcal{J}(\x) > 1$. }
@@
 \begin{verbatim}
-ANTS -m PR[A,B,1,2] -o OUTPUT 
+ANTS 2 -m PR[A,B,1,2] -o OUTPUT 
 \end{verbatim}
@@
 \subsection{ANTS Similarity Terms}
-From Gang.
-
+Here is a script that will let you experiment with different similarity term 
+and transformation model combinations.  A few are listed, with parameters 
+that are a reasonable starting point.
+\begin{verbatim}
+II=r16slice.nii # change to your own images 
+JJ=r64slice.nii 
+DIM=2 # two-dimensional images 
+ITS=" -i 100x100x30 " # 3 optimization levels 
+TSYNWITHTIME=" -t SyN[1,2,0.05]  -r Gauss[3,0.5] "
+TGREEDYSYN=" -t SyN[0.25]  -r Gauss[3,0] "
+TELAST=" -t Elast[1] -r Gauss[0,3] " 
+TEXP="  -t Exp[0.5,10] -r Gauss[0,3] " 
+LABELGUIDED=" -m PSE[${II},${JJ},${II},${JJ},0.75,0.1,25,0,10] "
+INTMSQ=" -m MSQ[${II},${JJ},1,0.1] "
+INTMI=" -m MI[${II},${JJ},1,32] "
+INTPR=" -m PR[${II},${JJ},1,4] "
+NAME=TEST
+INT=$INTPR
+TRAN=$TSYN
+ANTS $DIM -o $NAME $ITS  $TRAN  $INT
+INVW=" -i ${NAME}Affine.txt ${NAME}InverseWarp.nii "
+FWDW=" ${NAME}Warp.nii ${NAME}Affine.txt "
+WarpImageMultiTransform  2 ${II} a.nii -R ${JJ} $INVW
+WarpImageMultiTransform  2 ${JJ} b.nii -R ${II} $FWDW
+\end{verbatim}
+Convergence occurs under two conditions: (1) either the maximum number of 
+iterations are reached at a given optimization level or (2) the slope 
+of the change in the optimization objective is negative or very small. 
 {\em Note: ANTS does not like spaces between brackets in the call to the metrics.}
@@
 in addition to the T1 component.   The T1 PR metric, here, is weighted slightly more 
-than the FA PR metric, 1.25 vs. 1.0.
+than the FA PR metric, 1.25 vs. 1.0.  {\em The convergence criterion for 
+the multivariate scenario is a slave to the {\bf last} metric you 
+pass on the ANTS command line}.
 

### 2009-06-05 ImageRegistration/itkAvantsMutualInformationRegistrationFunction.h
bug in MI - wrong Spline Order
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@12 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -632,12 +632,7 @@ public:
 	mag1=sqrt(mag1);
 	if (mag1 > 1.e-5 ) fixedGradientNorm/=mag1;  
 	
-	//	for (int imd=0; imd<ImageDimension; imd++)
-	  {
-	    //  fdvec1[imd]=fixedGradientNorm[imd]*spacing[imd]*0.5;
-	    // fdvec2[imd]=fixedGradientNorm[imd]*(-1.)*spacing[imd]*0.5;
-	  }
-	  double nccm1=0;
+	double nccm1=0;
 	loce=this->GetValueAndDerivative(oindex,nccm1,fdvec1,fdvec2); 
 	float sign=(1.) * loce;//nccp1-nccm1; 
 	

### 2009-06-09 Utilities/itkN3MRIBiasFieldCorrectionImageFilter.h
STYLE:  Changed variable BiasFieldControlPointLattice to LogBiasFieldControlPointLattice which is more accurate.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@18 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -3,8 +3,8 @@
   Program:   Insight Segmentation & Registration Toolkit
   Module:    $RCSfile: itkN3MRIBiasFieldCorrectionImageFilter.h,v $
   Language:  C++
-  Date:      $Date: 2009/04/30 20:40:31 $
-  Version:   $Revision: 1.3 $
+  Date:      $Date: 2009/04/29 19:05:28 $
+  Version:   $Revision: 1.1 $
 
   Copyright (c) Insight Software Consortium. All rights reserved.
   See ITKCopyright.txt or http://www.itk.org/HTML/Copyright.htm for details.
@@ -172,7 +172,7 @@ public:
   itkSetMacro( NumberOfControlPoints, ArrayType );
   itkGetConstMacro( NumberOfControlPoints, ArrayType );
   
-  itkGetConstMacro( BiasFieldControlPointLattice, 
+  itkGetConstMacro( LogBiasFieldControlPointLattice, 
     typename BiasFieldControlPointLatticeType::Pointer );
 
 protected:
@@ -197,7 +197,7 @@ private:
   MaskPixelType                               m_MaskLabel;
 
   /**
-   * Parameters associated 
+   * Parameters for deconvolution with Weiner filter
    */
   unsigned int                                m_NumberOfHistogramBins;
   RealType                                    m_WeinerFilterNoise;
@@ -213,10 +213,10 @@ private:
    * B-spline fitting parameters
    */
   typename 
-    BiasFieldControlPointLatticeType::Pointer    m_BiasFieldControlPointLattice;
-  unsigned int                                   m_SplineOrder;
-  ArrayType                                      m_NumberOfControlPoints;
-  ArrayType                                      m_NumberOfFittingLevels;
+    BiasFieldControlPointLatticeType::Pointer m_LogBiasFieldControlPointLattice;
+  unsigned int                                m_SplineOrder;
+  ArrayType                                   m_NumberOfControlPoints;
+  ArrayType                                   m_NumberOfFittingLevels;
   
 
 

### 2009-06-09 Documentation/Tex/metric.tex
add metric.tex as a separate tex file, may be removed later
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@22 fd2da0a4-d968-48df-9cb5-d8c496287cbf

ANTS supports both volumetric registration and point set registration. The image / point set similarity metrics in ANTS are unified in the form of a function on the images or the point sets:
$$\textbf{Simlarity}[\textbf{fixedImage}, \textbf{movingImage}, \textbf{weight}, \textbf{parameters}]$$

The similarity type for the deformation transformation is specified by -m option. This option contains two parts: simarity type and the parameters inside the brackets. (Note: no white spaces exist between parameters.) The possible similarity metrics are: 
\begin{itemize}
 \item Intensity based:
    \begin{itemize}
        \item Cross correlation

        -m CC/cross-correlation/CrossCorrelation[fixedImage,movingImage,weight,radius]

        This metric works well for intra-modality image registration.

        Example: \textbf{-m CC[fixed.nii,moving.nii,1,5]} specifies:
            \begin{itemize}
             \item the fixed image: fixed.nii
             \item the moving image: moving.nii             
             \item weight for this metric is 1 (i.e. only this metric drives the registration).      \item the region radius for computing cross correlation is 5
            \end{itemize}

        \item Mutual information

        -m MI/mutual-information/MutualInformation[fixedImage,movingImage,weight,number of histogram bins]

        This metric works both well for intra-modality and inter-modality image registration.

        Example: \textbf{-m MI[fixed.nii,moving.nii,1,32]} specifies:
            \begin{itemize}
             \item the fixed image: fixed.nii
             \item the moving image: moving.nii             
             \item weight for this metric is 1 (i.e. only this metric drives the registration).      \item the number of bins in computing mutual information is 32
            \end{itemize}

        \item PR

        -m PR/probabilistic/Probabilistic[fixedImage,movingImage,weight,radius]

        TODO: This metric works for intra-modality image registration.

        Example: \textbf{-m PR[fixed.nii,moving.nii,1,5]} specifies:
            \begin{itemize}
             \item the fixed image: fixed.nii
             \item the moving image: moving.nii             
             \item weight for this metric is 1 (i.e. only this metric drives the registration).      \item the region radius for computing cross correlation is 5
            \end{itemize}

        \item Mean square difference

        -m MSQ/mean-squares/MeanSquares[fixedImage,movingImage,weight, 0]

        This metric works for intra-modality image registration.

        Example: \textbf{-m MSQ[fixed.nii,moving.nii,1,0]} specifies:
            \begin{itemize}
             \item the fixed image: fixed.nii
             \item the moving image: moving.nii             
             \item weight for this metric is 1 (i.e. only this metric drives the registration).      \item 0 is a padding value of no real meaning.
            \end{itemize}

    \end{itemize}

 \item Point set based:
    \begin{itemize}
        \item Point set expectation

        -m PSE/point-set-expectation/PointSetExpectation [fixedImage,movingImage,fixedPoints,movingPoints,weight,pointSetPercentage,pointSetSigma, boundaryPointsOnly,kNeighborhood,PartialMatchingIterations=100000]

        the partial matching option assumes the complete labeling is in the first set of label parameters ... more iterations leads to more symmetry in the matching  - 0 iterations means full asymmetry

        TODO: Example: \textbf{-m JTB[fixed.nii,moving.nii,fixed.nii,moving.nii,?????]} specifies:
            \begin{itemize}
             \item the fixed image: fixed.nii
             \item the moving image: moving.nii             
             \item the fixed point set: fixed.nii
             \item the moving point set: moving.nii
             \item weight for this metric is 1 (i.e. only this metric drives the registration).
             \item TODO: What are the parameters?
            \end{itemize}

        \item Jensen-Tsallis BSpline

        -m JTB/jensen-tsallis-bspline/JensenTsallisBSpline [fixedImage,movingImage,fixedPoints,movingPoints,weight,pointSetPercentage,pointSetSigma, boundaryPointsOnly,kNeighborhood,alpha,meshResolution,splineOrder,numberOfLevels, useAnisotropicCovariances]

        the partial matching option assumes the complete labeling is in the first set of label parameters ... more iterations leads to more symmetry in the matching  - 0 iterations means full asymmetry

            TODO: Example: \textbf{-m JTB[fixed.nii,moving.nii,fixed.nii,moving.nii,?????]} specifies:
            \begin{itemize}
             \item the fixed image: fixed.nii
             \item the moving image: moving.nii             
             \item the fixed point set: fixed.nii
             \item the moving point set: moving.nii
             \item weight for this metric is 1 (i.e. only this metric drives the registration).
             \item TODO: What are the parameters?
            \end{itemize}

    \end{itemize}

 \item For the affine registration, there are two types of similarity metrics, which are specified using \textbf{$--$affine-metric-type}:

    \begin{itemize}

    \item Mutual information, specified using \textbf{MI}.
         Example: \textbf{$--$affine-metric-type MI}. Options can be specified as:

         \begin{itemize}
            \item \textbf{$--$MI-option} specifies the number of bins and samples: MI\_bins x MI\_samples, for example: \textbf{$--$MI-option 32x8000}
         \end{itemize}

    \item Mean square difference, specified using \textbf{MSE}.
         Example: \textbf{$--$affine-metric-type MSE}
         \begin{itemize}
            \item \textbf{$--$MI-option} specifies the number of bins and samples: MI\_bins x MI\_samples, for example: \textbf{$--$MI-option 32x8000}
        \end{itemize}

    \end{itemize}

\end{itemize}

### 2009-06-10 Documentation/Tex/metric.tex
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@23 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
-ANTS supports both volumetric registration and point set registration. The image / point set similarity metrics in ANTS are unified in the form of a function on the images or the point sets:
-$$\textbf{Simlarity}[\textbf{fixedImage}, \textbf{movingImage}, \textbf{weight}, \textbf{parameters}]$$
+ANTS supports both volumetric registration and point set registration.
+The image / point set similarity metrics in ANTS are unified in the form of a function on the images or the point sets:       $$\textbf{Simlarity}[\textbf{fixedImage}, \textbf{movingImage}, \textbf{weight}, \textbf{parameters}].$$
 
-The similarity type for the deformation transformation is specified by -m option.
-This option contains two parts: simarity type and the parameters inside the brackets.
-(Note: no white spaces exist between parameters.) The possible similarity metrics are:
+The similarity type for the deformation transformation is specified by \verb -m  option, which contains two parts: simarity type and the parameters inside the brackets.
+(Note: no white spaces exist between parameters.) The possible similarity metrics for volumetric images are:
 \begin{itemize}
- \item Intensity based:
-    \begin{itemize}
-        \item Cross correlation
-
-        -m CC/cross-correlation/CrossCorrelation[fixedImage,movingImage,weight,radius]
-
+    \item Cross correlation:  \verb"-m CC[fixedImage,movingImage,weight,radius]".
 This metric works well for intra-modality image registration.
-
-        Example: \textbf{-m CC[fixed.nii,moving.nii,1,5]} specifies:
+For exxample, \verb"-m CC[fixed.nii,moving.nii,1,5]" specifies:
     \begin{itemize}
@@
         \item the moving image: moving.nii             
-             \item weight for this metric is 1 (i.e. only this metric drives the registration).      \item the region radius for computing cross correlation is 5
+        \item weight for this metric is 1 (i.e. only this metric drives the registration).      
+        \item the region radius for computing cross correlation is 5
     \end{itemize}
 
-        \item Mutual information
+    \item Mutual information: \verb"-m MI[fixedImage,movingImage,weight,number-of-histogram-bins]".
+This metric works both well for intra-modality and inter-modality image registration.
+For example, the first three parameters in \verb"-m MI[fixed.nii,moving.nii,1,32]" is similar to the example above in cross correlation, except that the last parameter means that the number of bins in computing mutual information is 32.
 
-        -m MI/mutual-information/MutualInformation[fixedImage,movingImage,weight,number of histogram bins]
+    \item PR: \verb"-m PR[fixedImage,movingImage,weight,radius]".
+This metric works for intra-modality image registration.
+[TODO: what is different with CC???].
+The meaning of parameters are similar to cross correlation.
 
-        This metric works both well for intra-modality and inter-modality image registration.
+    \item Mean square difference: \verb"-m MSQ[fixedImage,movingImage,weight,0]".
+This metric works for intra-modality image registration.
+The last parameter \verb"0" is a padding value of no real meaning.
+For example, \verb"-m MSQ[fixed.nii,moving.nii,1,0]".
 
-        Example: \textbf{-m MI[fixed.nii,moving.nii,1,32]} specifies:
-            \begin{itemize}
-             \item the fixed image: fixed.nii
-             \item the moving image: moving.nii             
-             \item weight for this metric is 1 (i.e. only this metric drives the registration).      \item the number of bins in computing mutual information is 32
 \end{itemize}
 
-        \item PR
-
-        -m PR/probabilistic/Probabilistic[fixedImage,movingImage,weight,radius]
-
-        TODO: This metric works for intra-modality image registration.
-
-        Example: \textbf{-m PR[fixed.nii,moving.nii,1,5]} specifies:
+ANTS also support registration of point sets. The supported formats for point sets can be found in I/O section. The similarity metrics for point sets are:
 \begin{itemize}
-             \item the fixed image: fixed.nii
-             \item the moving image: moving.nii             
-             \item weight for this metric is 1 (i.e. only this metric drives the registration).      \item the region radius for computing cross correlation is 5
+    \item Point set expectation: 
+     \begin{verbatim}
+-m PSE [fixedImage,movingImage,fixedPoints,movingPoints,weight,
+pointSetPercentage,pointSetSigma,boundaryPointsOnly,
+kNeighborhood,PartialMatchingIterations=100000]
+    \end{verbatim}
+    \begin{itemize}
+        \item \verb"fixedImage": defines the space domain of the fixed point set.
+        \item \verb"movingImage": defines the space domain of the moving point set.
+        \item \verb"fixedPoints": defines the coordinates of the fixed point set.
+It can be an image with discrete positive labels, a VTK format point set file, or a text file.
+Details can be found in I/O section (TODO).
+        \item \verb"movingPoints": defines the coordinates of the moving point set.
+        \item \verb"weight": weight for this metric. \verb"1" means that only this metric drives the registration.
+        \item \verb"pointSetPercentage": the percentage of points to be randomly sampled used in the registration.
+        \item \verb"pointSetSigma": the ??? 
+        \item \verb"boundaryPointsOnly": [TODO] \verb"1" (or ``true'') means only the boundary points in the point sets are used to drive registration.
+        \item \verb"kNeighborhood" is a positive discrete number. The first $k$ neighbors are used to compute the deformation during the registration. 
+        \item \verb"PartialMatchingIterations" controls the symmtry in the matching.
+This option assumes the complete labeling is in the first set of label parameters ...
+more iterations leads to more symmetry in the matching  - 0 iterations means full asymmetry
     \end{itemize}
 
-        \item Mean square difference
-
-        -m MSQ/mean-squares/MeanSquares[fixedImage,movingImage,weight, 0]
-
-        This metric works for intra-modality image registration.
-
-        Example: \textbf{-m MSQ[fixed.nii,moving.nii,1,0]} specifies:
+    \item Jensen-Tsallis BSpline
+    \begin{verbatim}
+-m JTB[fixedImage,movingImage,fixedPoints,movingPoints,weight,
+pointSetPercentage,pointSetSigma,boundaryPointsOnly,kNeighborhood,alpha,
+meshResolution,splineOrder,numberOfLevels,useAnisotropicCovariances] 
+    \end{verbatim}
     \begin{itemize}
-             \item the fixed image: fixed.nii
-             \item the moving image: moving.nii             
-             \item weight for this metric is 1 (i.e. only this metric drives the registration).      \item 0 is a padding value of no real meaning.
+        \item \verb"fixedImage": defines the space domain of the fixed point set.
+        \item \verb"movingImage": defines the space domain of the moving point set.
+        \item \verb"fixedPoints": defines the coordinates of the fixed point set.
+It can be an image with discrete positive labels, a VTK format point set file, or a text file.
+Details can be found in I/O section (TODO).
+        \item \verb"movingPoints": defines the coordinates of the moving point set.
+        \item \verb"weight": weight for this metric. \verb"1" means that only this metric drives the registration.
+        \item \verb"pointSetPercentage": the percentage of points to be randomly sampled used in the registration.
+        \item \verb"pointSetSigma": the ??? 
+        \item \verb"boundaryPointsOnly": [TODO] \verb"1" (or ``true'') means only the boundary points in the point sets are used to drive registration.
+        \item \verb"kNeighborhood" is a positive discrete number. The first $k$ neighbors are used to compute the deformation during the registration. 
+        \item \verb"alpha"
+        \item \verb"meshResolution"
+        \item \verb"splineOrder"
+        \item \verb"numberOfLevels"
+        \item \verb"useAnisotropicCovariances"
     \end{itemize}
@@
 
- \item Point set based:
+In current implementation, the affine registration only supports two types of similarity metrics on volumetric images, which are specified using \verb"--affine-metric-type":
 \begin{itemize}
-        \item Point set expectation
+    \item Mutual information, specified as \verb"--affine-metric-type MI".
+This usually works for both inter and intra-modalities in 3D.
+Also, the options used in computing mututal information in affine registration can be specified as \verb"--MI-option N1xN2".
+The first parameter \verb"N1" specifies the number of bins.
+The second parameter \verb"N2" specifies the number of samples.
+For example: \verb"--MI-option 32x8000".
 
-        -m PSE/point-set-expectation/PointSetExpectation [fixedImage,movingImage,fixedPoints,movingPoints,weight,pointSetPercentage,pointSetSigma, boundaryPointsOnly,kNeighborhood,PartialMatchingIterations=100000]
-
-        the partial matching option assumes the complete labeling is in the first set of label parameters ...
-more iterations leads to more symmetry in the matching  - 0 iterations means full asymmetry
-
-        TODO: Example: \textbf{-m JTB[fixed.nii,moving.nii,fixed.nii,moving.nii,?????]} specifies:
-            \begin{itemize}
-             \item the fixed image: fixed.nii
-             \item the moving image: moving.nii             
-             \item the fixed point set: fixed.nii
-             \item the moving point set: moving.nii
-             \item weight for this metric is 1 (i.e. only this metric drives the registration).
-             \item TODO: What are the parameters?
+    \item Mean square difference, specified using \verb"--affine-metric-type MSE". There is no options for this metric.
 \end{itemize}
 
-        \item Jensen-Tsallis BSpline
+Fig. \ref{fig:metric_example} shows the registration result using intensity difference with the following command:
+\begin{verbatim}
+ANTS 2 -m MSQ[r16slice.nii,r64slice.nii,1,0] -r Gauss[3,0] -t SyN[0.5] -i 50x50x30
+\end{verbatim}
 
-        -m JTB/jensen-tsallis-bspline/JensenTsallisBSpline [fixedImage,movingImage,fixedPoints,movingPoints,weight,pointSetPercentage,pointSetSigma, boundaryPointsOnly,kNeighborhood,alpha,meshResolution,splineOrder,numberOfLevels, useAnisotropicCovariances]
+%        \begin{wrapfigure}{r}{0.5\textwidth}
+\begin{figure}
+    \label{fig:metric_example}
+    \centering
+    \begin{tabular}[h]{c|c|c}
+        \includegraphics[width=0.16\textwidth]{Figures/r16slice.jpg} &
+        \includegraphics[width=0.16\textwidth]{Figures/r64slice.jpg} &
+        \includegraphics[width=0.16\textwidth]{Figures/resMSQ.jpg} \\
+        fixed image &
+        moving image & 
+        MSE  \\
+    \end{tabular} 
+    \itkcaption{registration using mean square intensity difference}
+\end{figure}
+    %    \end{wrapfigure}
+Here is an example script to register a pair of images using different image metrics:
+\begin{verbatim}
+#use cross-correlation with cc region radius 2
+ANTS 2 -m CC[r16slice.nii,r64slice.nii,1,2] -r Gauss[3,0] -t SyN[1] -i 50x50x30
+WarpImageMultiTransform 2 r64slice.nii resCC2.nii Warp.nii Affine.txt
+MeasureImageSimilarity 2 0 r16slice.nii resCC2.nii metricexamplelog.txt
+MeasureImageSimilarity 2 1 r16slice.nii resCC2.nii metricexamplelog.txt
+MeasureImageSimilarity 2 2 r16slice.nii resCC2.nii metricexamplelog.txt
+ConvertToJpg resCC2.nii resCC2.jpg
 
-        the partial matching option assumes the complete labeling is in the first set of label parameters ...
-more iterations leads to more symmetry in the matching  - 0 iterations means full asymmetry
+#use cross-correlation with cc region radius 4
+ANTS 2 -m CC[r16slice.nii,r64slice.nii,1,4] -r Gauss[3,0] -t SyN[1] -i 50x50x30
+WarpImageMultiTransform 2 r64slice.nii resCC4.nii Warp.nii Affine.txt
+MeasureImageSimilarity 2 0 r16slice.nii resCC4.nii metricexamplelog.txt
+MeasureImageSimilarity 2 1 r16slice.nii resCC4.nii metricexamplelog.txt
+MeasureImageSimilarity 2 2 r16slice.nii resCC4.nii metricexamplelog.txt
+ConvertToJpg resCC4.nii resCC4.jpg
 
-            TODO: Example: \textbf{-m JTB[fixed.nii,moving.nii,fixed.nii,moving.nii,?????]} specifies:
-            \begin{itemize}
-             \item the fixed image: fixed.nii
-             \item the moving image: moving.nii             
-             \item the fixed point set: fixed.nii
-             \item the moving point set: moving.nii
-             \item weight for this metric is 1 (i.e. only this metric drives the registration).
-             \item TODO: What are the parameters?
-            \end{itemize}
+#use cross-correlation with cc region radius 2
+ANTS 2 -m PR[r16slice.nii,r64slice.nii,1,2] -r Gauss[3,0] -t SyN[1] -i 50x50x30
+WarpImageMultiTransform 2 r64slice.nii resPR2.nii Warp.nii Affine.txt
+MeasureImageSimilarity 2 0 r16slice.nii resPR2.nii metricexamplelog.txt
+MeasureImageSimilarity 2 1 r16slice.nii resPR2.nii metricexamplelog.txt
+MeasureImageSimilarity 2 2 r16slice.nii resPR2.nii metricexamplelog.txt
+ConvertToJpg resPR2.nii resPR2.jpg
 
-    \end{itemize}
+#use cross-correlation with cc region radius 4
+ANTS 2 -m PR[r16slice.nii,r64slice.nii,1,4] -r Gauss[3,0] -t SyN[1] -i 50x50x30
+WarpImageMultiTransform 2 r64slice.nii resPR4.nii Warp.nii Affine.txt
+MeasureImageSimilarity 2 0 r16slice.nii resPR4.nii metricexamplelog.txt
+MeasureImageSimilarity 2 1 r16slice.nii resPR4.nii metricexamplelog.txt
+MeasureImageSimilarity 2 2 r16slice.nii resPR4.nii metricexamplelog.txt
+ConvertToJpg resPR4.nii resPR4.jpg
 
- \item For the affine registration, there are two types of similarity metrics, which are specified using \textbf{$--$affine-metric-type}:
+#use intensity difference with radius 0 -- radius no effect on intensity difference
+ANTS 2 -m MSQ[r16slice.nii,r64slice.nii,1,0] -r Gauss[3,0] -t SyN[0.5] -i 50x50x30
+WarpImageMultiTransform 2 r64slice.nii resMSQ.nii Warp.nii Affine.txt
+MeasureImageSimilarity 2 0 r16slice.nii resMSQ.nii metricexamplelog.txt
+MeasureImageSimilarity 2 1 r16slice.nii resMSQ.nii metricexamplelog.txt
+MeasureImageSimilarity 2 2 r16slice.nii resMSQ.nii metricexamplelog.txt
+ConvertToJpg resMSQ.nii resMSQ.jpg
 
-    \begin{itemize}
+# use mutual information with radius 0 and step-size 1.0
+ANTS 2 -m MI[r16slice.nii,r64slice.nii,1,32] -r Gauss[3,0] -t SyN[1] -i 50x50x30
+WarpImageMultiTransform 2 r64slice.nii resMI2.nii Warp.nii Affine.txt
+MeasureImageSimilarity 2 0 r16slice.nii resMI2.nii metricexamplelog.txt
+MeasureImageSimilarity 2 1 r16slice.nii resMI2.nii metricexamplelog.txt
+MeasureImageSimilarity 2 2 r16slice.nii resMI2.nii metricexamplelog.txt
+ConvertToJpg resMI2.nii resMI2.jpg
 
-    \item Mutual information, specified using \textbf{MI}.
-         Example: \textbf{$--$affine-metric-type MI}. Options can be specified as:
-
-         \begin{itemize}
-            \item \textbf{$--$MI-option} specifies the number of bins and samples: MI\_bins x MI\_samples, for example: \textbf{$--$MI-option 32x8000}
-         \end{itemize}
-
-    \item Mean square difference, specified using \textbf{MSE}.
-         Example: \textbf{$--$affine-metric-type MSE}
-         \begin{itemize}
-            \item \textbf{$--$MI-option} specifies the number of bins and samples: MI\_bins x MI\_samples, for example: \textbf{$--$MI-option 32x8000}
-        \end{itemize}
-
-    \end{itemize}
-
-\end{itemize}
+\end{verbatim}

### 2009-06-11 Documentation/Tex/ants.tex
add metric.tex as a separate tex file, may be removed later
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@22 fd2da0a4-d968-48df-9cb5-d8c496287cbf
need Figures/r16...
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@24 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 similarity and optimization criteria as options.  The program is
-wrapped in \verb ants.sh for normalization with ``out of the box''
-parameters and in \verb buildtemplateparallel.sh for 
+wrapped in \verb ants.sh ~for normalization with ``out of the box''
+parameters and in \verb buildtemplateparallel.sh ~for 
 computationally distributed optimal template construction.
@@
 Some operations output text files, some output images and some output only to the terminal.
-ImageMath allows one to multiply images together (m), to negate images (Neg), to take an image to a power (pow), to test the invertibility of transformations (InvId), to compute the fractional anisotropy of an image (TensorFA) and to perform Markov Random Field segmentation with our without priors (Segment).
+ImageMath allows one to multiply images together (m), to negate images (Neg), to take an image to a power (pow), to test the invertibility of transformations (InvId), to compute the fractional anisotropy of an image (TensorFA) and to perform Markov Random Field segmentation with or without priors (Segment).
 Many other operations are available.
@@
 through a deformation field $\disp( \x )$.  
-\begin{wrapfigure}{r}{0.5\textwidth}
+\begin{figure}
 \label{fig:pull}
-\includegraphics[width=0.45\textwidth]{./Figures/PullBack.pdf} 
-\end{wrapfigure}  
+\center \includegraphics[width=0.75\textwidth]{./Figures/PullBack.pdf} 
+\itkcaption[ANTS Deformation]{A deformation in a digital domain.}
+\end{figure}  
 The deformation is defined in the physical space of the image 
@@
 {\em Note: ANTS does not like spaces between brackets in the call to the metrics.}
+
+\subsection{Choosing a Metric}
+\input{metric.tex}
 

### 2009-06-11 Documentation/Tex/antstheory.tex
Moved things around.  Added text regarding experiments.  Added a couple figures.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@25 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 \title{ANTS:  Advanced Open-Source Tools for Normalization And Neuroanatomy}
-\author[label1]{PICSL}
-
-% use optional labels to link authors explicitly to addresses:
+%\author[label1]{Brian B. Avants}
+%\author[label1]{Nicholas J. Tustison}
+%\author[label1]{Gang Song}
+%\author[label1]{James C. Gee}
+\author{Brian B. Avants}
+\author{Nicholas J. Tustison}
+\author{Gang Song}
+\author{James C. Gee}% use optional labels to link authors explicitly to addresses:
 % \author[label1,label2]{}
-\address[label1]{Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, PA, USA}
+%\address[label1]{Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, PA, USA}
+\address{Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, PA, USA}
 %\address[label2]{Department of Mathematics, University of Pennsylvania, Philadelphia, PA, USA}
@@
 We first provide an overview of the different components included in ANTS, such as the available transformation models and similarity metrics offered.
-We follow this up with an extensive validation \ldots
+This is followed by an extensive experimental analysis that builds upon the results from the recent image normalization evaluation of \cite{Klein2009} which was limited to a single configuration of ANTS.
+This will provide  a useful benchmark for evaluating other possible ANTS configurations.
 
@@
     \multirow{2}*{\bf Linear}
-           & Rigid$^\dagger$ & MI, MSQ & Rigid registration. \\
-       {} & Affine$^\dagger$ & MI, MSQ & Affine registration. \\
+           & Rigid$^\dagger$ & MI, MSQ & translation and rotation \\
+       {} & Affine$^\dagger$ & MI, MSQ & rigid, scaling, and shear \\
        \cmidrule(l){2-4}
     \multirow{2}*{\bf Elastic}
-           & Deformable & CC, PR, MI, MSQ, PSE & Demons-like algorithm. \\
-       {} & DMFFD & CC, PR, MI, MSQ, PSE  & FFD variant. \\
+           & Deformable & CC, PR, MI, MSQ, JHCT, PSE & Demons-like algorithm \\
+       {} & DMFFD & CC, PR, MI, MSQ, JHCT, PSE  & FFD variant \\
        \cmidrule(l){2-4}
-    \multirow{2}*{\bf Diffeo.}
-           & Exponential$^\dagger$ &  CC,PR, MI, MSQ, PSE  & $~\min~\velocity(\x)$ \\
-       {} & Greedy SyN$^\dagger$ &  CC, PR, MI, MSQ, PSE  & ~~locally in time~$\min~\velocity(\x,t)$\\
-       {} & Geodesic SyN$^\dagger$ &  CC, PR, MI, MSQ, PSE  &  $~\min~\velocity(\x,t)$ over all time \\
+    \multirow{2}*{\bf Diffeomorphic}
+           & Exponential &  CC,PR, MI, MSQ, JHCT, PSE  & minimizes $\boldsymbol{v}(\mathbf{x})$ \\
+       {} & Greedy SyN$^\dagger$ &  CC, PR, MI, MSQ, JHCT, PSE  & minimizes $\boldsymbol{v}(\mathbf{x,t})$ locally in time \\
+       {} & Geodesic SyN$^\dagger$ &  CC, PR, MI, MSQ, JHCT, PSE  & minimizes $\boldsymbol{v}(\mathbf{x,t})$ over all time \\
     \bottomrule
     \end{tabular}
-  \caption{Transformations and a subset of the similarity metrics available in ANTS.
-Similarity metric acronyms:  CC = cross correlation, PR = probabilistic matching, MSQ = mean squared difference, MI = mutual information, PSE = point-set expectation.
-ANTS also provides the inverse of those transformations denoted by the `$\dagger$' symbol.
-The brief descriptions of the diffeomorphic algorithms contrast the way in which the velocity field is optimized and used to parameterize $\phi$, the mapping.
-All
-ANTS Diff algorithms generate $\phi(\x,t)$ over $t \in [ 0, 1]$ through gradient descent.}
+  \caption{Available transformations and similarity metrics available in ANTS.
+Similarity metric acronyms:  CC = cross correlation, PR = probabilistic matching, MSQ = mean squared difference, MI = mutual information, JHCT = Jensen-Havrda-Charvat-Tsallis divergence, PSE = point-set expectation.
+ANTS also provides the inverse of those transformations denoted by the `$\dagger$' symbol.}
   \label{table:chart}
 \end{table*}
-\vspace{0in}
 
@@
     \begin{tabular}{c}
-      % \includegraphics[width=80mm]{hierarchy}
+      \includegraphics[width=80mm]{hierarchy}
     \end{tabular}
@@
 
-\subsubsection{Linear Transformations--Rigid and Affine:}
-Image registration strategies often begin with a linear transformation
-for initial global alignment which follows a deformable
-transformation with increased degrees of freedom.  
-Affine transformations within ANTS optimize either a
-mean-squared difference metric or a mutual information (MI) image
-similarity metric.  ANTS affine transformations are optimized
-w.r.t. translation, rotation, scaling and shearing in succession.
-This strategy allows careful control of each component of the affine
-transformation.  Internally, ANTS also composes the affine
-transformation with the deformation field before performing any
-interpolation or downsampling.  In this way, ANTS normalization never
-requires more than a single image interpolation step and is able to
-always refer back to the original full-resolution image sources.
+\subsubsection{Rigid and Affine Linear Transformations }
+
+Image registration strategies often begin with a linear transformation for initial global alignment which precedes a deformable transformation with increased degrees of freedom.
+The linear transformations available within ANTS optimize either a mean-squared difference (MSQ) or mutual information (MI) similarity metric which are optimized with respect to translation, rotation, and, in the case of affine transformations, scaling and shearing.
+The successive optimization of each component of the linear transformation allows for careful control over increasing degrees of freedom.
+ANTS also composes the affine transformation with the deformable transformation field before performing any interpolation or downsampling.
+In this way, ANTS normalization never requires more than a single image interpolation step and is able to always refer back to the original full-resolution image sources.
+%We have implemented quaternion-based rigid and affine transformations within ANTS.
+Not only does this facilitate accessibility of a sound global alignment strategy, but the parameters of the calculated optimal linear transformation can be maintained internally to ANTS to avoid unnecessary image interpolations.
+Either a mean-squared difference metric or a mutual information metric can be used for the linear alignment.
+
+\subsubsection{Vector Field Operators for Regularization} Antecedent to discussion of available deformable transformation models, we point out that  deformable normalization strategies typically invoke a deformation regularization step which smooths the displacement field, $\displace$, or velocity field, $\mathbf{v}$,  or both by a linear operator such as the Laplacian or Navier-Stokes operator.
+One may write this regularization as a variational minimization in terms of its linear operator or in
+terms of a kernel function operating of the field itself, e.g., $
+\displace_{smooth}= K \star \displace_{not \,\, smooth}$, where $K \star$ denotes 
+convolution with the Green's kernel, $K$, for the linear operator, $L$.
+
+Regularization models operate on either the
+whole mapping $\phi$ or the gradient of the similarity term or both.
+The same regularization schema is available for both diffeomorphic and the recently 
+proposed directly manipulated free-form deformation
+(DMFFD) \cite{Tustison2009}, allowing regularization of both total 
+deformation and deformation update.  Viewed from this perspective, hybrid configurations incorporating discretized FFD strategies and diffeomorphisms can be combined for novel 
+image normalization approaches.
+
+ANTS enables a variety of choices for $K$
+including the Gaussian with varying $\sigma$ and a variety of B-spline
+functions, both of which induce adequate regularity for normalization
+models used in ANTS.  While additional physical operators will be
+included in future releases, current B-spline options provide many orders 
+of flexibility \cite{Tustison2005}.
 
 \subsubsection{Diffeomorphic Transformations}
+
 In contrast to many transformation models which reside in the domain of vector spaces, a diffeomorphism is a differentiable mapping with a differentiable inverse (\cite{Ebin1970,Mumford1998}).
@@
 
-ANTS assumes the diffeomorphism (differentiable mapping with a differentiable
-inverse), $\phi$, is defined on the image domain, $\Omega$, and
-maintains an affine transform at the boundary such that
-$\phi(\partial \Omega) = A(\mathbf{Id})$ where $A(\mathbf{Id})$ is an
-affine mapping applied to the identity transformation.  The $\phi$,
-over time, parameterizes a family of diffeomorphisms,
-$\phi(\mathbf{x}, t) : \Omega \times t
-\rightarrow \Omega$, which can be generated by integrating a
-time-dependent, regularized velocity field, $\boldsymbol{v} : \Omega \times
-t \rightarrow \mathbb{R}^d$, through an ordinary differential equation
-(o.d.e.).  
+ANTS assumes the diffeomorphism, $\phi$, is defined on the image domain, $\Omega$, an maintains an affine transform at the boundary such that $\phi(\partial \Omega) = A(\mathbf{Id})$ where $A(\mathbf{Id})$ is an affine mapping applied to the identity transformation.
+$\phi$, over time, parameterizes a family of diffeomorphisms, $\phi(\mathbf{x}, t) : \Omega \times t \rightarrow  \Omega$, which can be generated by integrating a time-dependent, smooth velocity field, $\boldsymbol{v} : \Omega \times t \rightarrow \mathbb{R}^d$,  through the ordinary differential equation (o.d.e.)
 \begin{align} \label{eq:ode}
@@
 The existence and uniqueness theorem for o.d.e.'s implies that integrating Equation (\ref{eq:ode}) generates a diffeomorphism.
-The deformation field provided by $\phi$ is
-$\mathbf{u}(\mathbf{x})=\phi(\mathbf{x}, 1) - \mathbf{x}$.
+The deformation field yielded by $\phi$ is $\mathbf{u}(\mathbf{x}) = \phi(\mathbf{x},1) - \mathbf{x}$.
 
@@
 \begin{align}\label{eq:D}
-  D(\mathcal{I}, \mathcal{J}) =  \int_0^1 ||L\boldsymbol{v}(x,t)||^2 dt , \,\, \mathcal{I}(\phi(\mathbf{x}, 1)) = \mathcal{J}(\mathbf{z})
+  D(\mathcal{I}, \mathcal{J}) =  \int_0^1 ||L\boldsymbol{v}||^2 dt , \,\, \mathcal{I}(\phi(\mathbf{x}, 1)) = \mathcal{J}(\mathbf{z})
 \end{align}
@@
 Intuitively, Equation (\ref{eq:D}) provides a sense of distance between two anatomical shapes.
-It also illustrates that the optimal diffeomorphic solution is analogous to finding the geodesic curve between two points in a curved space.
+It also illustrates that the optimal diffeomorphic solution is analogous to finding the geodesic curve between two points in a curved space.%
 \footnote{
@@
 \begin{align} \label{eq:lddmm}
-  \boldsymbol{v}^* = \underset{\boldsymbol{v}}{\operatorname{argmin}}  \left\{ \int_0^1  ||L\boldsymbol{v}||^2  dt + \lambda || \mathcal{I} \circ \phi(\mathbf{x},1) - \mathcal{J}  ||  \right\}.
+  \boldsymbol{v}^* = \underset{\boldsymbol{v}}{\operatorname{argmin}}  \left\{ \int_0^1  ||L\boldsymbol{v}||^2dt  + \lambda\int_{\Omega} || \mathcal{I} \circ \phi(\mathbf{x},1) - \mathcal{J}  || d\Omega \right\}.
 \end{align}
-The Euler-Lagrange equations characterizing the optimizing time-varying velocity field, $\boldsymbol{v}^*$, were derived in \cite{Miller2002} and later used in formulating the gradient-descent optimization scheme known as {\em large deformation diffeomorphic metric-matching} (LDDMM) (\cite{Beg2005}).
-The similarity metric, or data term, for LDDMM is the {\em squared intensity difference} with weight $\lambda$.
+The Euler-Lagrange equations characterizing the optimizing time-varying velocity field, $\boldsymbol{v}^*$, were derived in \cite{Miller2002} and later used in formulating the gradient-descent optimization scheme known as {\em large deformation diffeomorphic metric-matching} (LDDMM) (\cite{Beg2005}) with the similarity metric, or data term, for LDDMM being the squared intensity difference with weight $\lambda$.
 
-In medical image registration, one typically encounters more complex intensity transfers 
-between one anatomical instance $\mathcal{J}$ and another instance $\mathcal{I}$.  Thus, 
-ANTS enables a variety of similarity terms beyond the squared difference:
+To accommodate a variety of medical image normalization tasks, one typically encounters more complex intensity transfers between one anatomical instance $\mathcal{J}$ and another instance $\mathcal{I}$.
+Thus,
+ANTS enables not only a variety of similarity metric possibilities beyond the conventional squared difference
+metric but it also permits any number of different similarity metrics for a particular image normalization task.  
+This leads to the following generalization of  Equation (\ref{eq:lddmm}):
 \begin{align} \label{eq:diff}
-  \boldsymbol{v}^* = \underset{\boldsymbol{v}}{\operatorname{argmin}}  \left\{ \int_0^1  ||L\boldsymbol{v}||^2 dt  +~\lambda~\Pi( \mathcal{I}, \phi(\mathbf{x},1) , \mathcal{J} )  \right\}.
+  \boldsymbol{v}^* = \underset{\boldsymbol{v}}{\operatorname{argmin}}  \left\{ \int_0^1  ||L\boldsymbol{v}||^2 dt  +\lambda\int_{\Omega}\Pi_{\sim}( \mathcal{I}, \phi(\mathbf{x},1) , \mathcal{J} )d\Omega  \right\}
 \end{align}
-Here, $\Pi$ is a similarity metric depending on the images and the
+where $\Pi_{\sim}$ is a similarity metric depending on the images and the
 mapping and $\lambda$ controls the degree of exactness in the
-matching.  If $\Pi$ is selected as cross-correlation, then one is
+matching.  If $\Pi_{\sim}$ is selected as cross-correlation, then one is
 estimating the diffeomorphism under more robust illumination
-constraints, as described in \cite{Avants2008}.  A similar equation
-for elastic/FFD/DMFFD matching may be minimized in ANTS.
+constraints, as described in \cite{Avants2008}.  
+%A similar equation for elastic/FFD/DMFFD matching may be minimized in ANTS.
 
 Exploiting the fact that the diffeomorphism, $\phi$, can be decomposed into two components 
-$\phi_1$ and $\phi_2$, Avants et al. construct a {\em symmetric} alternative to Equation (\ref{eq:diff}) (\cite{Avants2007b}).  This leads to the symmetric variant of Equation (\ref{eq:diff})
+$\phi_1$ and $\phi_2$, \cite{Avants2007b} construct a {\em symmetric} alternative to Equation (\ref{eq:diff}).  This leads to the symmetric variant of Equation (\ref{eq:diff})
 \begin{align}\label{eq:diffs}
-  \{\boldsymbol{v}_1^*,\boldsymbol{v}_2^*\}  =
+  \{\boldsymbol{v}_1^*,\boldsymbol{v}_2^*\}  = &
      \underset{\boldsymbol{v_1},\boldsymbol{v_2}}{\operatorname{argmin}}  
-      \Bigg\{ & \int_0^{0.5} ( ||L\boldsymbol{v}_1(x,t)||^2 + ||L\boldsymbol{v}_2(x,t)||^2   )dt  \nonumber \\
-     &+ || \mathcal{I} \circ \phi_1(\mathbf{x},0.5) - \mathcal{J}\circ \phi_2(\mathbf{x},0.5)   ||\Bigg\}.
+      \Bigg\{ \nonumber \\
+      & \int_0^{0.5} \left( ||L\boldsymbol{v}_1(x,t)||^2 + ||L\boldsymbol{v}_2(x,t)||^2   \right)dt  \nonumber \\
+     &+ \lambda\int_{\Omega}\Pi_{\sim}\left( \mathcal{I} \circ \phi_1(\mathbf{x},0.5), \mathcal{J}\circ \phi_2(\mathbf{x},0.5)\right)d\Omega\Bigg\}.
 \end{align}
@@
 Finding $\boldsymbol{v_1}^*$ minimizes the variational energy from $t = 0$ whereas $\boldsymbol{v_2}^*$ minimizes from $t = 1$.
-Thus, gradient-based iterative convergence deforms $\mathcal{I}$ and $\mathcal{J}$ along the geodesic diffeomorphism, $\phi$, to a fixed  point midway (intuited by the notion of shape distance) between $\mathcal{I}$ and $\mathcal{J}$.
+Thus, gradient-based iterative convergence deforms $\mathcal{I}$ and $\mathcal{J}$ along the geodesic diffeomorphism, $\phi$, to a fixed  point midway (intuited by the notion of shape distance) between $\mathcal{I}$ and $\mathcal{J}$ thus motivating the denotation of the solution strategy as Symmetric Normalization (SyN).
 
-ANTS allows a variety and combination of data terms to be optimized
-with diffeomorphisms and, at the same time,
-enables a variety of models for parameterizing the $\phi(\cdot)$.  
-Briefly, ANTS Diff models are: 
-\begin{enumerate}
-\item {\bf Exponential Mapping:}  This constant velocity field parameterization does not produce a distance metric; maps are not dense in the space of diffeomorphisms and are not inverse consistent.
-\item {\bf Greedy SyN:} Approximates the distance metric with a fast approach.
+Other diffeomorphic flavors have since been reported in the research literature (e.g.
+DARTEL (\cite{Ashburner2007}) and Diffeomorphic Demons (\cite{Vercauteren2007,Vercauteren2009}).
+We include three diffeomorphic transformation models for parameterizing $\phi(\cdot)$.
+These include Geodesic SyN, Greedy SyN, and exponential mapping.
+As summarized in Table \ref{table:chart}, each of these transformation models can utilize a host of similarity measures both individually and in mutual combination.
+
+%Briefly, ANTS diffeomorphic transformation models are: 
+%\begin{enumerate}
+%\item {\bf Exponential Mapping:}  This constant velocity field parameterization does not produce a distance metric; maps are not dense in the space of diffeomorphisms and are not inverse consistent.
+%\item {\bf Greedy SyN:} Approximates the distance metric with a fast approach.
 Evaluated in \cite{Klein2009}.
@@
 Theoretically and computationally inverse consistent / symmetric.
-\item {\bf Geodesic SyN:}  Symmetrically minimizes the velocity field / distance metric explicitly over all time.
+%\item {\bf Geodesic SyN:}  Symmetrically minimizes the velocity field / distance metric explicitly over all time.
 Computationally more expensive than, but produces similar solutions to, Greedy SyN for most medical imaging problems.
-\end{enumerate}
-All of these transformation models may be used with any of the ANTS similarity metrics.  In particular, ANTS allows $\Pi$ 
-to be composed of a summation/weighted combination of available similarity metrics/data terms.
+%\end{enumerate}
+%All of these transformation models may be used with any of the ANTS similarity metrics.  In particular, ANTS allows $\Pi$ 
+%to be composed of a summation/weighted combination of available similarity metrics/data terms.
+
+\paragraph{Geodesic SyN}
+Using a gradient-based optimization strategy, minimization implied by Equation (\ref{eq:diffs}), which we denote as $\Pi$, first requires a specified discretization of $t \in [0,1]$.
+Calculation of the gradient of $\Pi$ with respect to the diffeomorphisms $\phi_1$ and $\phi_2$ is performed at each of the $t$ values which is given by
+\begin{align}
+  \nabla \Pi(\boldsymbol{x},t) = \frac{\partial}{\partial \phi_i} \Pi_{\sim}( \mathcal{I}(\phi_{1}^{-1}(\boldsymbol{x},t)), \mathcal{J}(\phi_{2}^{-1}(\boldsymbol{x},1-t)) 
+\end{align}
+for $i \in \{1,2\}$.   The corresponding velocity fields at each $t$ are then updated from the previous iteration according to 
+\begin{align}
+  \boldsymbol{v}(\mathbf{x},t) = \boldsymbol{v}(\mathbf{x},t) + K \star \nabla \Pi(\mathbf{x},t).
+\end{align}
+We then integrate Equation (\ref{eq:ode}) using Runge-Kutta methods to yield $\phi_i(\mathbf{x},t)$ for each $t \in [0,1]$ and $i \in \{1,2\}$.
+We cycle through these steps until convergence or iterative exhaustion.
+
+\paragraph{Greedy SyN}
+Although the Geodesic SyN algorithm conforms most closely to the theoretical diffeomorphic foundations culminating with Equation (\ref{eq:diffs}), the computational and memory cost is significant due to the dense-in-time gradient calculations and requisite reintegration of the diffeomorphisms after each iterative update.
+As a lower-cost alternative, we offer a greedy variant which performs quite well for most medical image normalization problems we have encountered.
+Additionally, this was the strategy used in the large-scale comparative image registration algorithm assessment of \cite{Klein2009}.
+
+Greedy optimization of Equation (\ref{eq:diffs}) requires calculation of the gradient of the gradient taken at the mid-point of the full diffeomorphism, i.e. at $t = 0.5$
+\begin{align}
+\nabla \Pi =\frac{\partial}{\partial \phi_i} \Pi_{\sim}( \mathcal{I}(\phiinv_1(\x,0.5)) ,  \mathcal{J}(\phiinv_2(\x,0.5)) )
+\end{align}
+for $i \in \{1,2\}$.  $\phi_1(\mathbf{x},0.5)$ and $\phi_2(\mathbf{x},0.5)$ are then updated from the previous iteration according to 
+\begin{align}
+  \phi_i(\mathbf{x},0.5) = \phi_i(\mathbf{x},0.5)+\delta  (K \star \nabla \Pi ( \phi_i(\x,0.5) )).
+\end{align}
+where $\delta$ is a user-specifed step parameter.  The gradient is then mapped back to the origin of each diffeomorphism to update the full mapping by enforcing $\phiinv(\phi(\x,1))=\x$.
 
 \paragraph{Exponential Mapping}
-The key difference between a time-varying diffeomorphism and a diffeomorphism 
-generated by an exponential mapping (Ashburner200x) is that the exponential 
-mapping maintains only a single vector field that is constant in time.   Theoretically, 
-the restricting the velocity field to be constant in time reduces the size of the 
-space that may be generated \cite{ArnoldODE} in a way that is similar to the difference 
-between real and rational numbers, the latter of which are sparsely distributed through 
-the reals.  
-In contrast, 
-the full diffeomorphic space may be explored by using time-varying velocity fields 
-as does SyN and LDDMM.  The exponentiation of a constant velocity field generates a 
-diffeomorphism through, 
+Ashburner introduced DARTEL (Diffeomorphic Anatomical Registration using Exponentiated Lie algebra) as a rapidly computed alternative to time parameterized diffeomorphic schemes (\cite{Ashburner2007}).
+The key difference between a time-varying diffeomorphism and a diffeomorphism generated by an exponential mapping (\cite{Ashburner2007}) is that the exponential mapping maintains only a single vector field that is constant in time.
+
+By exponentiation of a constant velocity field, a diffeomorphism can be generated from the following o.d.e (cf Equation (\ref{eq:ode}))
 \begin{align} \label{eq:odec}
@@
 \end{align}
-Note that there is no time parameter in the velocity field, here.   The Dartel algorithm uses 
-the well-known ``scaling and squaring'' approach to efficiently generate $\phi$ from a velocity field $\boldsymbol{v}(x)$.
-
-\paragraph{Greedy SyN}
-Greedy optimization of equation~\ref{eq:diffs}, involves gradient and update steps 
-as follows, 
-\begin{align}
-\text{gradient:   }  \nabla E = \partial_{\phi_i} \Pi ( \mathcal{I}(\phiinv_1(\x,0.5)) ,  \mathcal{J}(\phiinv_2(\x,0.5)) ) \nonumber \\ 
-\text{regularize:  } \nabla E \leftarrow G_\sigma \star \nabla E \nonumber \\ 
-\text{update:  }  \phi_i(\x,0.5) \leftarrow \phi_i(\x,0.5)+\delta \nabla E ( \phi_i(\x,0.5) ). \nonumber \\
-\text{update the inverse s.t.:  } \phi_i(\phiinv_i) = \Id.  
-\end{align}
-Though the gradients differ for each metric, the update scheme is the
-same.  In words, the gradient is taken at the mid-point of the full
-diffeomorphism, $\phi$, by decomposing with respect to its components,
-$\phi_1$ and $\phi_2$.  The gradient with respect to each half is
-computed, regularized with a Gaussian of variance $\sigma$ and then
-mapped back to the origin of each diffeomorphism to update the full
-mapping.  Hermosillo provides a good discussion of the gradients of
-the similarity metrics available in ANTS
-\cite{hermosillo}.  Note that, here, we supply a specific 
-regularization scheme but others are available in ANTS.  
-That is, one may directly regularize the $\phi_i$ or use 
-no regularization at all.
-
-\paragraph{Time-Varying SyN}
-Global in time optimization of equation~\ref{eq:diffs}, involves a global gradient 
-where, at each time point $t \in [0, 1]$, we have:
-\begin{align}
-\text{gradient:   }  \nabla E(\x,t) = \partial_{\phi_i} \Pi ( \mathcal{I}(\phiinv_1(\x,t)) ,  \mathcal{J}(\phiinv_2(\x,1-t)) ), \nonumber \\ 
-\text{update:  }  \velocity(\x,t)=\velocity(\x,t)+G_\sigma \star \nabla E(\x,t). \nonumber \\
-\text{reintegrate the o.d.e.s, via Runge-Kutta},  \nonumber \\
-\text{that generate the $\phi_i$ in $[0,1]$.}  
-\end{align}
-The time-varying SyN algorithm is theoretically cleaner than greedy
-SyN.  However, the computational and memory cost is greater due to the
-dense-in-time gradient and requisite reintegration of the
-diffeomorphisms after each update.  Note that, here, we supply a
-specific regularization scheme but others are available in ANTS.
-
-%\input{vectorspace.tex}
-
-\subsubsection{Vector Field Operators:} Deformable normalization
-strategies typically regularize a deformation, $\displace$, or
-velocity field, $\velocity$, with a linear operator such as the
-Laplacian or Navier-Stokes operator. One may write this regularization
-as a variational minimization in terms of its linear operator or in
-terms of a kernel function operating of the field itself, e.g., $
-\displace_{smooth}= K \star \displace_{not-smooth}$, where $K$ is 
-the Green's kernel for the linear operator, $L$ and $\star$ is
-convolution.  ANTS enables a variety of choices for the kernel, $K$,
-including the Gaussian with varying $\sigma$ and a variety of BSpline
-functions, both of which induce adequate regularity for normalization
-models used in ANTS.  Regularization models operate on either/both the
-whole mapping, $\phi$, or/and the gradient of the similarity term.
-The same schema is applied for diffeomorphic and the recently proposed directly-manipulated 
-free-form-deformation
-(DMFFD)\cite{Tustison2009}, allowing regularization of both total 
-deformation and deformation update.  While additional physical operators will be
-included in future releases, current BSpline options provide many orders 
-of flexibility \cite{Tustison2005}.
+Note that there is no explicit time parameter in the velocity field.
+Theoretically, restricting the velocity field to be constant in time reduces the size of the space that may be generated (\cite{Arnold1991}) in a way that is similar to the difference between real and rational numbers, the latter of which are sparsely distributed through the reals.
 
@@
 
-A simple and efficient, yet powerful image normalization algorithm is Thirion's demons (\cite{Thirion1998}).
+A simple and efficient, yet powerful image normalization algorithm is the approach known as Thirion's demons (\cite{Thirion1998}).
 Using an optical flow based similarity, the solution is obtained by iterating between the calculation of image forces and subsequent Gaussian regularization.
-In ANTS, we extend this basic approach to include both cross-correlation and mean-squared difference.
+In ANTS, we extend this basic approach to include the similarity metrics available for deformable registration (see Table \ref{table:chart}).
 
@@
 
-For $n$-D images, the FFD transformational model, $\phi_{FFD}$, is defined as 
+For $n$-D images, the FFD (and DMFFD) transformational model, $\phi_{FFD}$, is defined as 
 \begin{align}\label{eq:TFFD}
@@
 
-\subsection{ANTS Intensity Based Similarity Metrics}
+\subsection{ANTS Intensity-Based Similarity Metrics}
 Several intensity-based image metrics have been proposed in the literature with varying levels of performance dependent upon specific applications.
 We have included several of the most popular similarity metrics within ANTS.
-In addition, our software framework facilitates the development of other image metrics if that is desired by the user.
+In addition, our software framework facilitates the development of other image metrics.
 
 Both mutual information (\cite{Viola1997}) and mean-squared difference similarity metrics are available for the linear transformations.
-Also included are the cross-correlation (\cite{Avants2007b}), probabilistic matching (more aggressive version of cross-correlation), local mutual information (\cite{Rueckert1999,Pluim2003}), and mean squared difference similarity metrics for the non-linear transformation models.
-For symmetric transformations, the inverse metrics and gradients are also calculated.
-Parameters for the different metrics are also available (see the ANTS website (\cite{ANTS2009}) for further details).
+Also included are the cross-correlation (\cite{Avants2007b}), probabilistic matching (a more aggressive version of cross-correlation), local mutual information (\cite{Rueckert1999,Pluim2003}), and mean squared difference similarity metrics for the non-linear transformation models.
+The parameters for the different metrics are discussed in the ANTS documentation (\cite{ANTS2009}).
 
-\subsection{ANTS Point-Set Based Similarity Metrics}
+\subsection{ANTS Point-Set-Based Similarity Metrics}
 In addition to intensity-based metrics, ANTS also contains similarity metrics for 
@@
 \subsubsection{Exact Landmark Matching}
-Generalizing the B-spline fitting algorithm of Lee et al. (\cite{Lee1997}), 
+Generalizing the B-spline fitting algorithm of \cite{Lee1997}, 
 we developed a scattered data approximation algorithm (\cite{Tustison2006b}) 
@@
 matching where this metric seeks to minimize the weighted sum of distances between 
-corresponding landmarks using a hierarchical approach.  One can also associate relative confidence values with each landmark for fine-tuning exact landmark matching results.
+corresponding landmarks using a hierarchical approach. 
+% One can also associate relative confidence values with each landmark for fine-tuning exact landmark matching results.
 
@@
 Recent information theoretic approaches have been proposed for point-set registration.
-A previous open-source contribution (\cite{Tustison2008b}) generalizes the Jensen-Shannon divergence to the Jensen-Havrda-Charvat-Tsallis (JCHT) divergence which
+A previous open-source contribution (\cite{Tustison2009}) generalizes the Jensen-Shannon divergence to the Jensen-Havrda-Charvat-Tsallis (JCHT) divergence which
 permits a fine-tuning  of the divergence measure such that emphasis can vary between robustness and sensitivity for 
-application-speciÞc tailoring.
+application-speciÞc tailoring \cite{Tustison2009a}.
 
@@
 }
-Binaries are also available from the same online location.
+Binaries are also available from the same online location.  For quality assurance and maintenance purposes we have established an ANTS test reporting open-source ``dashboard''%
+\footnote{
+http://www.cdash.org
+}
+ on our lab website%
+\footnote{
+http://www.picsl.upenn.edu/cdash/index.php?project=ANTS
+}
+to monitor compilation and testing of the ANTS program.
+A screenshot from a daily testing period is given in Figure \ref{fig:dashboard}.
+Such a configuration facilitates reporting of problems encountered by users on a multitude of computing platforms.
+
+\begin{figure*}
+\centering
+  \begin{tabular}{c}
+      \includegraphics[width=135mm]{ANTSdashboard} 
+  \end{tabular}
+\caption{The ANTS dashboard, which is hosted on the PICSL website, reports daily building and testing of the ANTS software.
+It also allows any user to submit their own building and testing configurations to help with debugging issues and maintenance for a variety of computing platforms.
+}
+\label{fig:dashboard}
+\end{figure*}
 
 Based on our experience with standard command line argument parsing packages (e.g.
-\verb#getopt#), we developed our own set of classes for a more intuitive command line interface (at least from our perspective).
+\verb#getopt#), we developed our own set of classes for an intuitive command line interface.
 A summary of command line arguments are given in Table \ref{table:args}.
-We invite the interested reader to visit our evolving ANTS web site (\cite{ANTS2009})for further details.
 These ANTS argument parsing classes provide an intuitive compromise between parsers where every variable requires a unique flag and strict ordering requirements on the command line.
@@
   --metric PSE[fixedPoints.vtk,movingPoints.vtk,0.25] 
-  --transformation SyN[0.25] 
+  --transformation Exp[0.75] 
   --regularization Gauss[3.0,0.5]
@@
 \end{lstlisting}
-The correspondence between the ANTS command line specification and the diffeomorphic image normalization formulation with labeled point-sets $X$ and $Y$ illustrates the motivation for our command line interface,
+The correspondence between the ANTS command line specification and the image normalization formulation with labeled point-sets $X$ and $Y$ illustrates the motivation for our command line interface,
 \begin{align}
   \Pi(&\mathcal{I},\mathcal{J},X,Y,\phi) \nonumber \\ 
-     =& \underbrace{\int_0^1 ||L\boldsymbol{v}||dt}_{\texttt{-t SyN[}\cdot \texttt{]},\texttt{ -r Gauss[}\cdot\texttt{,}\cdot\texttt{]}}  + \underbrace{\lambda_1\int_0^1 || \mathcal{I} \circ \phi(\mathbf{x},t) - \mathcal{J}  || dt}_{\texttt{-m MSQ[}\mathcal{J}\texttt{,}\mathcal{I}\texttt{,}\lambda_1\texttt{]}} \nonumber \\
+     =& \underbrace{\int_0^1 ||L\boldsymbol{v}||dt}_{\texttt{-t Exp[}\cdot \texttt{]},\texttt{ -r Gauss[}\cdot\texttt{,}\cdot\texttt{]}}  + \underbrace{\lambda_1\int_{\Omega} || \mathcal{I} \circ \phi(\mathbf{x},t) - \mathcal{J}  || d\Omega}_{\texttt{-t Exp[}\cdot \texttt{]},\texttt{ -m MSQ[}\mathcal{J}\texttt{,}\mathcal{I}\texttt{,}\lambda_1\texttt{]}} \nonumber \\
 &+\underbrace{\frac{\lambda_2 }{|Y|}\sum_{i=1}^{|Y|} \left\|y_i - \frac{1}{|X|} \sum_{j = 1}^{|X|} G(y_i; x_j, \sigma_X) x_j\right\|^2}_{\texttt{-m PSE[}Y\texttt{,}X\texttt{,}\lambda_2\texttt{]}}.
 \end{align}
+
+In Table \ref{table:args} we give a brief summary of the arguments available for the normalizations offered by the ANTS package.
+This includes the corresponding variable specification.
+More information can be found on the ANTS website.
 
@@
     {} & Iterations/Level & \verb#--iterations,-i# & {} & $N_1$\verb#x#$N_2$\verb#x#$N_3$\verb#x#\ldots \\ 
-    {} & Regularization &  \verb#--regularization,-r# & \verb#Gauss,DMFFD# & \verb#[#$\sigma_{total}$,$\sigma_{gradient}$\verb#]# \\
+    {} & Regularization &  \verb#--regularization,-r# & \verb#Gauss,DMFFD# & \verb#[#$\sigma^2_{gradient}$,$\sigma^2_{total}$\verb#]#, \verb#[1x1,3x3]#\\
     {} & Transformation & \verb#--transformation,-t# & \verb#Elast,SyN,Exp# & \verb#[#$\Delta_{gradient}$\verb#]# \\
@@
     \end{tabular}
-  \caption{  Further information can be found on the ANTS website (\cite{ANTS2009}). }
+  \caption{The various flags and variables for a variety of image registration possibilities.   Additional information can be found on the ANTS website (\cite{ANTS2009}). }
   \label{table:args}
 \end{table*}
+
+\section{Experimental Evaluation}
+
+There are many avenues for exploration of the various components of ANTS.
+However, due to space constraints, we limit experimental analysis within this paper to an extension of the normalization assessment carried out by \cite{Klein2009}.
+As previously mentioned, this large-scale assessment encompassed evaluation of 14 popular registration algorithms which were optimized, in terms of their parameters, by their respective authors before a thorough brain image normalization study.
+Although the Greedy SyN algorithm, outlined in an earlier section, was consistently one of the top two performers in Klein's study, for the benefit of the users of ANTS, we explore the other transformation model possibilities within ANTS and compare them with Greedy SyN.
+In terms of data, we utilize the NA0 evaluation database of the Non-Rigid Image Registration Evaluation Project (NIREP) for future comparison with evaluation studies that have been proposed by the NIREP initiative.%
+\footnote{http://www.nirep.org/
+}
+
+As outlined in the introduction, in addition to the transformation, the optimization strategy and similarity metric form the image normalization schematic troika.
+Since our optimization strategy is limited to gradient descent, experimental analysis includes an exploration of optimal gradient steps within a sensible window where the steps are scaled according to the voxel spacing.
+In terms of similarity metric, we limit exploration to cross-correlation while varying the radius within reasonable values.
+Other metrics were not explored since the labeled brain images entail a simple intensity relationship between image pairs obviating the need for similarity metrics for more complex intensity relationships (e.g.
+MI) in addition to the fact that the consistently top two performers in \cite{Klein2009} used cross-correlation.
+
+Briefly, two experiments were performed.
+The first experiment consisted of a more extensive parameter search over both the transformation model space and the cross-correlation metric radius using the exhaustive pairwise combination of the eight 2-D images illustrated in Figure \ref{fig:phantoms}.
+Based on the results of the first experiment, the parameter space was pruned and subsequently used for the registrations performed during the second experiment involving 10 randomly selected image pairs from the 16 labeled NA0 NIREP brain images.
+
+\subsection{2-D Simulated Image Normalization Evaluation}
+
+\begin{figure}
+\centering
+  \begin{tabular}{c}
+      \includegraphics[width=85mm]{phantomwmgm} 
+  \end{tabular}
+\caption{The eight 2-D simulated images used for the initial parameter search.  Each image is of size $[102,95]$ and are available with the ANTS source distribution.}  
+\label{fig:phantoms}
+\end{figure}
+
+\subsection{3-D NIREP Brain Image Normalization Evaluation}
+
+\section{Discussion}
 
@@
 
-\bibliographystyle{plain}
-%elsarticle-num-names.bst}
-\bibliography{references,ants}
+\bibliographystyle{elsarticle-num-names.bst}
+\bibliography{/Users/nick/Pending/references.bib}
 

### 2009-06-11 Documentation/Tex/metric.tex
need Figures/r16...
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@24 fd2da0a4-d968-48df-9cb5-d8c496287cbf
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@32 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
     \begin{tabular}[h]{c|c|c}
-        \includegraphics[width=0.16\textwidth]{Figures/r16slice.jpg} &
-        \includegraphics[width=0.16\textwidth]{Figures/r64slice.jpg} &
-        \includegraphics[width=0.16\textwidth]{Figures/resMSQ.jpg} \\
+%        \includegraphics[width=0.16\textwidth]{Figures/r16slice.jpg} &
+%        \includegraphics[width=0.16\textwidth]{Figures/r64slice.jpg} &
+ %       \includegraphics[width=0.16\textwidth]{Figures/resMSQ.jpg} \\
         fixed image &
@@
     %    \end{wrapfigure}
-Here is an example script to register a pair of images using different image metrics:
+Here is also an example script to register a pair of images using mean square intensity difference and computing the metrics of the registration image.
 \begin{verbatim}
-#use cross-correlation with cc region radius 2
-ANTS 2 -m CC[r16slice.nii,r64slice.nii,1,2] -r Gauss[3,0] -t SyN[1] -i 50x50x30
-WarpImageMultiTransform 2 r64slice.nii resCC2.nii Warp.nii Affine.txt
-MeasureImageSimilarity 2 0 r16slice.nii resCC2.nii metricexamplelog.txt
-MeasureImageSimilarity 2 1 r16slice.nii resCC2.nii metricexamplelog.txt
-MeasureImageSimilarity 2 2 r16slice.nii resCC2.nii metricexamplelog.txt
-ConvertToJpg resCC2.nii resCC2.jpg
-
-#use cross-correlation with cc region radius 4
-ANTS 2 -m CC[r16slice.nii,r64slice.nii,1,4] -r Gauss[3,0] -t SyN[1] -i 50x50x30
-WarpImageMultiTransform 2 r64slice.nii resCC4.nii Warp.nii Affine.txt
-MeasureImageSimilarity 2 0 r16slice.nii resCC4.nii metricexamplelog.txt
-MeasureImageSimilarity 2 1 r16slice.nii resCC4.nii metricexamplelog.txt
-MeasureImageSimilarity 2 2 r16slice.nii resCC4.nii metricexamplelog.txt
-ConvertToJpg resCC4.nii resCC4.jpg
-
-#use cross-correlation with cc region radius 2
-ANTS 2 -m PR[r16slice.nii,r64slice.nii,1,2] -r Gauss[3,0] -t SyN[1] -i 50x50x30
-WarpImageMultiTransform 2 r64slice.nii resPR2.nii Warp.nii Affine.txt
-MeasureImageSimilarity 2 0 r16slice.nii resPR2.nii metricexamplelog.txt
-MeasureImageSimilarity 2 1 r16slice.nii resPR2.nii metricexamplelog.txt
-MeasureImageSimilarity 2 2 r16slice.nii resPR2.nii metricexamplelog.txt
-ConvertToJpg resPR2.nii resPR2.jpg
-
-#use cross-correlation with cc region radius 4
-ANTS 2 -m PR[r16slice.nii,r64slice.nii,1,4] -r Gauss[3,0] -t SyN[1] -i 50x50x30
-WarpImageMultiTransform 2 r64slice.nii resPR4.nii Warp.nii Affine.txt
-MeasureImageSimilarity 2 0 r16slice.nii resPR4.nii metricexamplelog.txt
-MeasureImageSimilarity 2 1 r16slice.nii resPR4.nii metricexamplelog.txt
-MeasureImageSimilarity 2 2 r16slice.nii resPR4.nii metricexamplelog.txt
-ConvertToJpg resPR4.nii resPR4.jpg
-
 #use intensity difference with radius 0 -- radius no effect on intensity difference
@@
 ConvertToJpg resMSQ.nii resMSQ.jpg
-
-# use mutual information with radius 0 and step-size 1.0
-ANTS 2 -m MI[r16slice.nii,r64slice.nii,1,32] -r Gauss[3,0] -t SyN[1] -i 50x50x30
-WarpImageMultiTransform 2 r64slice.nii resMI2.nii Warp.nii Affine.txt
-MeasureImageSimilarity 2 0 r16slice.nii resMI2.nii metricexamplelog.txt
-MeasureImageSimilarity 2 1 r16slice.nii resMI2.nii metricexamplelog.txt
-MeasureImageSimilarity 2 2 r16slice.nii resMI2.nii metricexamplelog.txt
-ConvertToJpg resMI2.nii resMI2.jpg
-
 \end{verbatim}

### 2009-06-11 Scripts/buildtemplateparallel.sh
Changed the script I/O, increased gradient step size to -0.25, and added a script to monitor qstat periodically to check on your jobs.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@37 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -5,29 +5,50 @@ NUMPARAMS=$#
 if [ $NUMPARAMS -lt 4  ]
 then 
 echo " USAGE ::  "
-echo "  sh   buildtemplate.sh  ImageDimension WildcardingSide1 WildcardingSide2 Naming  OptionalIterationLimit-Default=4  OptionalDoQSub "
-echo " " 
-echo " the files used for template construction will  be  of the type     WildcardingSide1*WildcardingSide2  " 
-echo " " 
-echo " We assume all files to be added to the template are in the same directory -- you can modify the script if you prefer something different "
-echo " things you should change for your needs are highlighted by   EDIT THIS "
-echo " if the template file does not exist yet, we create an unbiased starting point by averaging the input dataset "
+echo "   sh buildtemplateparallel.sh ImageDimension OutputRoot IterationLimit <images>"
+echo  
+echo " ImageDimension  -  Dimension of your image, eg 3 for 3D."
+echo " OutputRoot      -  Root file name of the output. Should be a file root only, no path information. "
+echo " IterationLimit  -  Number of times to build the template. The previous template is used to initialize "
+echo "                    each iteration. "
+echo " <images>        -  List of images in the current directory, eg *_t1.nii.gz. "
+echo  
+echo " This script builds a template iteratively from the input images and uses SGE to parallelize the "
+echo " registration of every subject to the template. "
+echo
+echo " We assume all files to be added to the template are in the current directory. You can modify the "
+echo " script if you want to relax this assumption, but be sure that the qsubbed jobs get the correct "
+echo " absolute path to the images."
+echo 
[…]

### 2009-06-11 Scripts/waitForSGEQJobs.pl

@@ -1,0 +1,142 @@
+#!/usr/bin/perl -w
+
+use strict;
+
+# Usage: waitForSGEQJobs.pl <verbose [1 or 0]> <delay in seconds in range 10-600> [job IDs]
+# 
+#
+# Takes as args a string of qsub job IDs and periodically monitors them. Once they all finish, it returns 0
+#
+# If any of the jobs go into error state, an error is printed to stderr and the program waits for the non-error
+# jobs to finish, then returns 1
+#
+
+# Usual qstat format - check this at run time
+# job-ID  prior   name       user         state submit/start at     queue                          slots ja-task-ID
+
+
+# First thing to do is parse our input
+
+my ($verbose, $delay, @jobIDs) = @ARGV;
+
+# Check for user stupidity
+if ($delay < 10) {
+    print STDERR "Sleep period is too short, will poll queue once every 10 seconds\n";
+    $delay = 10;
+}
+elsif ($delay > 600) {
+    print STDERR "Sleep period is too long, will poll queue once every 10 minutes\n";
+    $delay = 600;
+}
+
+print "Waiting for " . scalar(@jobIDs) . " jobs: @jobIDs\n";
+
+my $user=`whoami`;
+
+my $qstatOutput = `qstat -u $user`;
+
+if (!scalar(@jobIDs) || !$qstatOutput) {
+    # Nothing to do
+    exit 0;
+}
+
+my @qstatLines = split("\n", $qstatOutput);
+
+my @header = split('\s+', trim($qstatLines[0]));
+
+# Position in qstat output of tokens we want
+my $jobID_Pos = -1;
+my $statePos = -1; 
+ 
+foreach my $i (0..$#header) {
+    if ($header[$i] eq "job-ID") {
+	$jobID_Pos = $i;
+    }
+    elsif ($header[$i] eq "state") {
+	$statePos = $i;
+    }
+}
+
+
+# If we can't parse the job IDs, something is very wrong
+if ($jobID_Pos < 0 || $statePos < 0) {
+    die "Cannot find job-ID and state field in qstat output, cannot monitor jobs\n";
+}
+
+
+
+# Now check on all of our jobs
+my $jobsIncomplete = 1;
+
+# Set to 1 for any job in an error state
+my $haveErrors = 0;
+
+while ($jobsIncomplete) {
+    
+    # Jobs that are still showing up in qstat
+    $jobsIncomplete = 0;
+
[…]

### 2009-06-24 Documentation/Tex/ants.tex
updates to documentation and output of programs ....
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@44 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 \itkcaption[ANTS Command Line]{The relationship between the variational 
-form that defines the optimization goals and the ANTS command line.}
+form that defines the optimization goals and the ANTS command line.  The 
+SyN option implements the method in \cite{Avants2008b} and evaluated in 
+\cite{Klein2009}.}
 \vspace{-0.1in}
@@
   \caption{Transformations and a subset of the similarity metrics available in ANTS.
-Similarity metric acronyms:  CC = fast cross correlation, PR = pure cross correlation (the preferred metric), MSQ = mean squared difference, MI = mutual information, PSE = point-set expectation.
+Similarity metric acronyms:  CC = fast cross correlation, PR = pure cross correlation (the preferred metric), MSQ = mean squared difference, MI = mutual information, PSE = point-set expectation \cite{Pluta2008}.
 ANTS also provides the inverse of those transformations denoted by the `$\dagger$' symbol.
@@
 how one may use ANTS to achieve a state-of-the-art diffeomorphic mapping. 
-The ANTS diffeomorphic model chosen for this example -- symmetric normalization -- 
+The ANTS diffeomorphic model chosen for this example -- symmetric normalization \cite{Avants2008b,Kim2008} -- 
 is invariant to the order of the input images (although the affine mapping is not).  
@@
 A useful script for setting up optimal template construction
- is available in ANTS/Scripts.  Two versions of available : serial and parallel.  
+ is available in ANTS/Scripts.  This method is described in \cite{Avants2004,Avants2006d,Kim2008,Yushkevich2009}, 
+where more recent publications are more descriptive and incorporate 
+more recent features and optimization strategies. 
+Two versions of this algorithm are available : serial and parallel.  
 Data and a script for building an average face template is available 
@@
 studies.  Cortical thickness, for instance, is known to correlate with language 
-development and IQ in adolescents.
+development and IQ in adolescents.  The laplacian approach to 
+cortical thickness may be used via the program LaplacianThickness, 
+which implements the method described in ``Three-dimensional mapping of cortical thickness using Laplace's Equation'' by 
+Jones, et al \cite{}.  Super-sampling 
+and controlling segmentation accuracy for input to this program is up 
+to the user.  Otherwise, the Laplacian method may grossly overestimate 
+thickness in closed sulci.
 
@@
 Here, we apply the expectation-based point set registration method 
-for mapping labeled points sets.   ITK-SNAP may be used to label 
+for mapping labeled points sets, as described in \cite{Pluta2008}.   ITK-SNAP may be used to label 
 images and exported segmentation images may be input to the 
@@
 
-\subsection{Statistics with ANTS and ``R''}
+\subsection{Statistics with ANTS and {\bf R}}
 {\bf R} is an open-source, cutting-edge statistical package 

### 2009-06-25 ImageRegistration/ANTS_affine_registration2.h
fix a bug when the initial gradient in affine registration is all zero.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@45 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -831,10 +831,16 @@ bool RegisterImageAffineMutualInformationMultiResolution(RunningAffineCacheType 
             double inner_product_last_current_gradient = 0.0;
             for(int j=0; j<kParaDim; j++) inner_product_last_current_gradient += current_gradient[j]*last_gradient[j];
             if (inner_product_last_current_gradient < 0) current_step_length *= relaxation_factor;
-            if (current_step_length < minimum_step_length) {
+            if (current_step_length < minimum_step_length || gradient_magnitude == 0.0 ) {
             	is_converged = true;
                 break;
             }
+
+//            for(int j = 0; j < kParaDim; j++)  std::cerr<< current_gradient[i] << ", ";
+//            std::cerr<< std::endl;
+
+
+
 
             for(int j = 0; j < kParaDim; j++)  current_para[j] += (-1.0) * current_gradient[j] * current_step_length / gradient_magnitude;
 
@@ -950,6 +956,9 @@ void ComputeSingleAffineTransform2D3D(ImagePointerType fixed_image, ImagePointer
     transform->SetCenter(opt.transform_initial->GetCenter());
 
     double rval_init = TestCostValueMMI(fixed_image, moving_image, opt.transform_initial->GetParameters(), opt.transform_initial->GetCenter(), transform);
+
+    // std::cerr << "ABCDABCD: " << transform << std::endl;
+
     double rval_final = TestCostValueMMI(fixed_image, moving_image, para_final, opt.transform_initial->GetCenter(), transform);
 
     std::cout << "outputput affine center: " << transform->GetCenter() << std::endl;

### 2009-06-26 Documentation/Tex/antstheory.tex
Changed bibliography path.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@26 fd2da0a4-d968-48df-9cb5-d8c496287cbf
ok - updated to compile
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@31 fd2da0a4-d968-48df-9cb5-d8c496287cbf
Changed the format to IEEE and updated the citations to reflect IEEE usage.  Also, changed the graphics path.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@46 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
-% Template article for preprint document class `elsart'
-% with harvard style bibliographic references
-% SP 2006/04/26
-
-\documentclass[preprint,twocolumn,authoryear,5p]{elsarticle}
+%\documentclass[11pt,draftcls,onecolumn]{IEEEtran}
+\documentclass[10pt,twocolumn]{IEEEtran}
 
@@
 
-% Use the option doublespacing or reviewcopy to obtain double line spacing
- %\documentclass[doublespacing]{elsart}
+%% bare_jrnl.tex
+%% V1.2
+%% 2002/11/18
+%% by Michael Shell
+%% mshell@ece.gatech.edu
+%%
+%% NOTE: This text file uses MS Windows line feed conventions. When (human)
+%% reading this file on other platforms, you may have to use a text
+%% editor that can handle lines terminated by the MS Windows line feed
+%% characters (0x0D 0x0A).
+%%
+%% This is a skeleton file demonstrating the use of IEEEtran.cls
+%% (requires IEEEtran.cls version 1.6b or later) with an IEEE journal paper.
+%%
+%% Support sites:
+%% http://www.ieee.org
+%% and/or
+%% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/IEEEtran/
+%%
+%% This code is offered as-is - no warranty - user assumes all risk.
+%% Free to use, distribute and modify.
 
-% the natbib package allows both number and author-year (Harvard)
-% style referencing;
-\usepackage{natbib}
+% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
+% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
+% *** with production work. IEEE's font choices can trigger bugs that do  ***
+% *** not appear when using other class files.                            ***
+% Testflow can be obtained at:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/testflow
 
-% if you use PostScript figures in your article
-% use the graphics package for simple commands
-% \usepackage{graphics}
-% or use the graphicx package for more complicated commands
- \usepackage{graphicx}
-% or use the epsfig package if you prefer to use the old commands
-% \usepackage{epsfig}
-
-% The amssymb package provides various useful mathematical symbols
+% Note that the a4paper option is mainly intended so that authors in
+% countries using A4 can easily print to A4 and see how their papers will
+% look in print. Authors are encouraged to use U.S. letter paper when 
+% submitting to IEEE. Use the testflow package mentioned above to verify
+% correct handling of both paper sizes by the author's LaTeX system.
+%
+% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
+% should be used if it is desired that the figures are to be displayed in
+% draft mode.
+%
+% This example can be formatted using the peerreview
+% (instead of journal) mode.
+%\documentclass[11pt,draftcls,onecolumn]{IEEEtran}
 \usepackage{amssymb,amsmath}
@@
 \usepackage{listings}					
+\graphicspath{./Figures/}
 
-\graphicspath{{/Users/nick/Documents/Academic/Gee/SubmittedPapers/JournalPapers/MedIA2009_BrianANTS/Figures/}
-%{/Users/nick/Documents/Academic/Gee/SubmittedPapers/Conferences/ISBI2007/Images/}
-%                          {/Users/nick/Documents/Academic/Gee/Projects/Helium-3/GrantSubmissions/R21/resplan/Images/}
-%                          {/Users/nick/Documents/Academic/Gee/SubmittedPapers/JournalPapers/MRM2008/Figures/}
-%                          {/Users/nick/Documents/Academic/Gee/SubmittedPapers/Conferences/ISBI2008/Figures/}
-                          }
+% If the IEEEtran.cls has not been installed into the LaTeX system files,
+% manually specify the path to it:
+% \documentclass[journal]{../sty/IEEEtran}
+
+% some very useful LaTeX packages include:
+
+%\usepackage{cite}      % Written by Donald Arseneau
+                        % V1.6 and later of IEEEtran pre-defines the format
+                        % of the cite.sty package \cite{} output to follow
+                        % that of IEEE. Loading the cite package will
+                        % result in citation numbers being automatically
+                        % sorted and properly "ranged". i.e.,
+                        % [1], [9], [2], [7], [5], [6]
+                        % (without using cite.sty)
+                        % will become:
+                        % [1], [2], [5]--[7], [9] (using cite.sty)
+                        % cite.sty's \cite will automatically add leading
+                        % space, if needed. Use cite.sty's noadjust option
+                        % (cite.sty V3.8 and later) if you want to turn this
+                        % off. cite.sty is already installed on most LaTeX
+                        % systems. The latest version can be obtained at:
+                        % http://www.ctan.org/tex-archive/macros/latex/contrib/supported/cite/
+
+%\usepackage{graphicx}  % Written by David Carlisle and Sebastian Rahtz
+                        % Required if you want graphics, photos, etc.
+                        % graphicx.sty is already installed on most LaTeX
+                        % systems. The latest version and documentation can
+                        % be obtained at:
+                        % http://www.ctan.org/tex-archive/macros/latex/required/graphics/
+                        % Another good source of documentation is "Using
+                        % Imported Graphics in LaTeX2e" by Keith Reckdahl
+                        % which can be found as esplatex.ps and epslatex.pdf
+                        % at: http://www.ctan.org/tex-archive/info/
+% NOTE: for dual use with latex and pdflatex, instead load graphicx like:
+%\ifx\pdfoutput\undefined
+%\usepackage{graphicx}
+%\else
+%\usepackage[pdftex]{graphicx}
+%\fi
+
+% However, be warned that pdflatex will require graphics to be in PDF
+% (not EPS) format and will preclude the use of PostScript based LaTeX
+% packages such as psfrag.sty and pstricks.sty. IEEE conferences typically
+% allow PDF graphics (and hence pdfLaTeX). However, IEEE journals do not
+% (yet) allow image formats other than EPS or TIFF. Therefore, authors of
+% journal papers should use traditional LaTeX with EPS graphics.
+%
+% The path(s) to the graphics files can also be declared: e.g.,
+% \graphicspath{{../eps/}{../ps/}}
+% if the graphics files are not located in the same directory as the
+% .tex file. This can be done in each branch of the conditional above
+% (after graphicx is loaded) to handle the EPS and PDF cases separately.
+% In this way, full path information will not have to be specified in
+% each \includegraphics command.
+%
+% Note that, when switching from latex to pdflatex and vice-versa, the new
+% compiler will have to be run twice to clear some warnings.
+
+%\usepackage{psfrag}    % Written by Craig Barratt, Michael C. Grant,
+                        % and David Carlisle
+                        % This package allows you to substitute LaTeX
+                        % commands for text in imported EPS graphic files.
+                        % In this way, LaTeX symbols can be placed into
+                        % graphics that have been generated by other
+                        % applications. You must use latex->dvips->ps2pdf
+                        % workflow (not direct pdf output from pdflatex) if
+                        % you wish to use this capability because it works
+                        % via some PostScript tricks. Alternatively, the
+                        % graphics could be processed as separate files via
+                        % psfrag and dvips, then converted to PDF for
+                        % inclusion in the main file which uses pdflatex.
+                        % Docs are in "The PSfrag System" by Michael C. Grant
+                        % and David Carlisle. There is also some information 
+                        % about using psfrag in "Using Imported Graphics in
+                        % LaTeX2e" by Keith Reckdahl which documents the
+                        % graphicx package (see above). The psfrag package
+                        % and documentation can be obtained at:
+                        % http://www.ctan.org/tex-archive/macros/latex/contrib/supported/psfrag/
+
+%\usepackage{subfigure} % Written by Steven Douglas Cochran
+                        % This package makes it easy to put subfigures
+                        % in your figures. i.e., "figure 1a and 1b"
+                        % Docs are in "Using Imported Graphics in LaTeX2e"
+                        % by Keith Reckdahl which also documents the graphicx
+                        % package (see above). subfigure.sty is already
+                        % installed on most LaTeX systems. The latest version
+                        % and documentation can be obtained at:
+                        % http://www.ctan.org/tex-archive/macros/latex/contrib/supported/subfigure/
+
+%\usepackage{url}       % Written by Donald Arseneau
+                        % Provides better support for handling and breaking
+                        % URLs. url.sty is already installed on most LaTeX
+                        % systems. The latest version can be obtained at:
+                        % http://www.ctan.org/tex-archive/macros/latex/contrib/other/misc/
+                        % Read the url.sty source comments for usage information.
+
+%\usepackage{stfloats}  % Written by Sigitas Tolusis
+                        % Gives LaTeX2e the ability to do double column
+                        % floats at the bottom of the page as well as the top.
+                        % (e.g., "\begin{figure*}[!b]" is not normally
+                        % possible in LaTeX2e). This is an invasive package
+                        % which rewrites many portions of the LaTeX2e output
+                        % routines. It may not work with other packages that
+                        % modify the LaTeX2e output routine and/or with other
+                        % versions of LaTeX. The latest version and
+                        % documentation can be obtained at:
+                        % http://www.ctan.org/tex-archive/macros/latex/contrib/supported/sttools/
+                        % Documentation is contained in the stfloats.sty
+                        % comments as well as in the presfull.pdf file.
+                        % Do not use the stfloats baselinefloat ability as
+                        % IEEE does not allow \baselineskip to stretch.
+                        % Authors submitting work to the IEEE should note
+                        % that IEEE rarely uses double column equations and
+                        % that authors should try to avoid such use.
+                        % Do not be tempted to use the cuted.sty or
+                        % midfloat.sty package (by the same author) as IEEE
+                        % does not format its papers in such ways.
+
+%\usepackage{amsmath}   % From the American Mathematical Society
+                        % A popular package that provides many helpful commands
+                        % for dealing with mathematics. Note that the AMSmath
+                        % package sets \interdisplaylinepenalty to 10000 thus
+                        % preventing page breaks from occurring within multiline
+                        % equations. Use:
+%\interdisplaylinepenalty=2500
+                        % after loading amsmath to restore such page breaks
+                        % as IEEEtran.cls normally does. amsmath.sty is already
+                        % installed on most LaTeX systems. The latest version
+                        % and documentation can be obtained at:
+                        % http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
+
+% Other popular packages for formatting tables and equations include:
+
+%\usepackage{array}
+% Frank Mittelbach's and David Carlisle's array.sty which improves the
+% LaTeX2e array and tabular environments to provide better appearances and
+% additional user controls. array.sty is already installed on most systems.
+% The latest version and documentation can be obtained at:
+% http://www.ctan.org/tex-archive/macros/latex/required/tools/
+
+% Mark Wooding's extremely powerful MDW tools, especially mdwmath.sty and
+% mdwtab.sty which are used to format equations and tables, respectively.
+% The MDWtools set is already installed on most LaTeX systems. The lastest
+% version and documentation is available at:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/mdwtools/
+
+% V1.6 of IEEEtran contains the IEEEeqnarray family of commands that can
+% be used to generate multiline equations as well as matrices, tables, etc.
+
+% Also of notable interest:
+
+% Scott Pakin's eqparbox package for creating (automatically sized) equal
+% width boxes. Available:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/eqparbox/
+
+% Notes on hyperref:
+% IEEEtran.cls attempts to be compliant with the hyperref package, written
+% by Heiko Oberdiek and Sebastian Rahtz, which provides hyperlinks within
+% a document as well as an index for PDF files (produced via pdflatex).
+% However, it is a tad difficult to properly interface LaTeX classes and
+% packages with this (necessarily) complex and invasive package. It is
+% recommended that hyperref not be used for work that is to be submitted
+% to the IEEE. Users who wish to use hyperref *must* ensure that their
+% hyperref version is 6.72u or later *and* IEEEtran.cls is version 1.6b
+% or later. The latest version of hyperref can be obtained at:
+%
+% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/hyperref/
+%
+% Also, be aware that cite.sty (as of version 3.9, 11/2001) and hyperref.sty
+% (as of version 6.72t, 2002/07/25) do not work optimally together.
+% To mediate the differences between these two packages, IEEEtran.cls, as
+% of v1.6b, predefines a command that fools hyperref into thinking that
+% the natbib package is being used - causing it not to modify the existing
+% citation commands, and allowing cite.sty to operate as normal. However,
+% as a result, citation numbers will not be hyperlinked. Another side effect
+% of this approach is that the natbib.sty package will not properly load
+% under IEEEtran.cls. However, current versions of natbib are not capable
+% of compressing and sorting citation numbers in IEEE's style - so this
+% should not be an issue. If, for some strange reason, the user wants to
+% load natbib.sty under IEEEtran.cls, the following code must be placed
+% before natbib.sty can be loaded:
+%
+% \makeatletter
+% \let\NAT@parse\undefined
+% \makeatother
+%
+% Hyperref should be loaded differently depending on whether pdflatex
+% or traditional latex is being used:
+%
+%\ifx\pdfoutput\undefined
+%\usepackage[hypertex]{hyperref}
+%\else
+%\usepackage[pdftex,hypertexnames=false]{hyperref}
+%\fi
+%
+% Pdflatex produces superior hyperref results and is the recommended
+% compiler for such use.
+
+% *** Do not adjust lengths that control margins, column widths, etc. ***
+% *** Do not use packages that alter fonts (such as pslatex).         ***
+% There should be no need to do such things with IEEEtran.cls V1.6 and later.
+
+% correct bad hyphenation here
+\hyphenation{op-tical net-works semi-conduc-tor}
 
@@
 
-\begin{frontmatter}
+%
+% paper title
+\title{ANTS:  Advanced Open-Source Tools for Normalization And Neuroanatomy}
+%
+%
+% author names and IEEE memberships
+% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
+% a structure at a ~ so this keeps an author's name from being broken across
+% two lines.
+% use \thanks{} to gain access to the first footnote area
+% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
+% was not built to handle multiple paragraphs
+\author{Brian B. Avants, Nicholas J. Tustison, Gang Song, and James C. Gee
+%               Penn Image Computing and Science Laboratory \\
+%               University of Pennsylvania \\
+%               3600 Market Street, Suite 370 \\
+%               Philadelphia,  PA  19104 \\
+%               contact email:  tustison@grasp.cis.upenn.edu
+%\thanks{Manuscript received XXX; revised XXX.
+ %       }% <-this % stops a space
+}
 
-% Title, authors and addresses
+% note the % following the last \IEEEmembership and also the first \thanks - 
+% these prevent an unwanted space from occurring between the last author name
+% and the end of the author line. i.e., if you had this:
+% 
+% \author{....lastname \thanks{...} \thanks{...} }
+%                     ^------------^------------^----Do not want these spaces!
+%
+% a space would be appended to the last name and could cause every name on that
+% line to be shifted left slightly. This is one of those "LaTeX things". For
+% instance, "A\textbf{} \textbf{}B" will typeset as "A B" not "AB". If you want
+% "AB" then you have to do: "A\textbf{}\textbf{}B"
+% \thanks is no different in this regard, so shield the last } of each \thanks
+% that ends a line with a % and do not let a space in before the next \thanks.
+% Spaces after \IEEEmembership other than the last one are OK (and needed) as
+% you are supposed to have spaces between the names. For what it is worth,
+% this is a minor point as most people would not even notice if the said evil
+% space somehow managed to creep in.
+%
+% The paper headers
+\markboth{IEEE Transactions on Medical Imaging,~Vol.~X, No.~XX,~November~200X}{Tustison\MakeLowercase{\textit{et al.}}: Advanced Normalization Tools}
+% The only time the second header will appear is for the odd numbered pages
+% after the title page when using the twoside option.
+% 
+% *** Note that you probably will NOT want to include the author's name in ***
+% *** the headers of peer review papers.                                   ***
 
-% use the thanksref command within \title, \author or \address for footnotes;
-% use the corauthref command within \author for corresponding author footnotes;
-% use the ead command for the email address,
-% and the form \ead[url] for the home page:
-% \title{}
-% \thanks[label1]{}
-% \author{Name\corauthref{cor1}\thanksref{label2}}
-% \ead{tustison@grasp.cis.upenn.edu}
-% \ead[url]{home page}
-% \thanks[label2]{}
-% \corauth[cor1]{}
-% \address{Address\thanksref{label3}}
-% \thanks[label3]{}
+% If you want to put a publisher's ID mark on the page
+% (can leave text blank if you just want to see how the
+% text height on the first page will be reduced by IEEE)
+%\pubid{0000--0000/00\$00.00~\copyright~2002 IEEE}
 
-\title{ANTS:  Advanced Open-Source Tools for Normalization And Neuroanatomy}
-%\author[label1]{Brian B. Avants}
-%\author[label1]{Nicholas J. Tustison}
-%\author[label1]{Gang Song}
-%\author[label1]{James C. Gee}
-\author{Brian B. Avants}
-\author{Nicholas J. Tustison}
-\author{Gang Song}
-\author{James C. Gee}% use optional labels to link authors explicitly to addresses:
-% \author[label1,label2]{}
-%\address[label1]{Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, PA, USA}
-\address{Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, PA, USA}
-%\address[label2]{Department of Mathematics, University of Pennsylvania, Philadelphia, PA, USA}
+% use only for invited papers
+%\specialpapernotice{(Invited Paper)}
+
+% make the title area
+
+\maketitle
 
@@
 
-\begin{keyword}
-ANTS \sep image registration \sep nueroanatomy \sep neuroimaging \sep normalization \sep  open-source \sep point-set registration
-% keywords here, in the form: keyword \sep keyword
-% PACS codes here, in the form: \PACS code \sep code
-\end{keyword}
+\IEEEkeywords{
+ANTS, image registration, nueroanatomy, neuroimaging, normalization, open-source, point-set registration
+}
 
-\end{frontmatter}
+\IEEEpeerreviewmaketitle
 
@@
 As described by Dr.
-Kovacevic ``[reproducible research] refers to the idea that, in `computational' sciences, the ultimate product is not a published paper but, rather, the entire environment used to produce the results in the paper (data, software,etc.).''  After an informal survey of 15 published papers, she found ``none had code available'' and ``in only about half the cases were the parameters [of the algorithm] specified'' (\cite{Kovacevic2006}).
-Recent discussions within the computational sciences research community, particularly among advocates of ``open science,'' have also voiced similar concerns (\cite{Yoo2005,Ibanez2006}).
+Kovacevic ``[reproducible research] refers to the idea that, in `computational' sciences, the ultimate product is not a published paper but, rather, the entire environment used to produce the results in the paper (data, software,etc.).''  After an informal survey of 15 published papers, she found ``none had code available'' and ``in only about half the cases were the parameters [of the algorithm] specified'' \cite{Kovacevic2006}.
+Recent discussions within the computational sciences research community, particularly among advocates of ``open science,'' have also voiced similar concerns \cite{Yoo2005,Ibanez2006}.
 In this paper, we discuss our contribution to the open-source medical image analysis research community which we call ANTS (\underline{A}dvanced \underline{N}ormalization \underline{T}ool\underline{s}).
@@
 
-Perhaps the most persuasive evidence motivating the use of our contributions discussed in this paper is the recent outcome of a large-scale comparative image registration algorithm assessment (\cite{Klein2009}).
-In the largest evaluation study to date involving 14 popular non-linear registration algorithms, our Symmetric Normalization (SyN) transformation model (\cite{Avants2008}) discussed below, was consistently one of the top two performers across all tests.
+Perhaps the most persuasive evidence motivating the use of our contributions discussed in this paper is the recent outcome of a large-scale comparative image registration algorithm assessment \cite{Klein2009}.
+In the largest evaluation study to date involving 14 popular non-linear registration algorithms, our Symmetric Normalization (SyN) transformation model \cite{Avants2008} discussed below, was consistently one of the top two performers across all tests.
 Overlap and distance measures used for assessment employed three completely independent analysis methods (permutation tests, one-way ANOVA tests, and indifference zone ranking).
@@
 \section{Theoretical Overview of ANTS}
-A useful classification schema of normalization techniques is based upon the following three principal components (\cite{Brown1992,Ibanez2002}):
+A useful classification schema of normalization techniques is based upon the following three principal components \cite{Brown1992,Ibanez2002}:
 \begin{itemize}
@@
 A variety of transformation models have been proposed in the literature with varying degrees of freedom (illustrated in Figure \ref{fig:hierarchy}).
-For deformable transformations, one approach is to optimize within the space of non-topology preserving, yet physically motivated transformations---an approach pioneered by Bajcsy (\cite{Bajcsy1989}).
-Elastic and related models, such as HAMMER (\cite{Shen2002}), statistical parametric mapping (SPM) (\cite{Ashburner2000}), free-form deformations (FFD) (\cite{Rueckert1999}), and
-Thirion's demons (\cite{Thirion1998}) operate in the space of vector fields, which does not preserve topology.
+For deformable transformations, one approach is to optimize within the space of non-topology preserving, yet physically motivated transformations---an approach pioneered by Bajcsy \cite{Bajcsy1989}.
+Elastic and related models, such as HAMMER \cite{Shen2002}, statistical parametric mapping (SPM) \cite{Ashburner2000}, free-form deformations (FFD) \cite{Rueckert1999}, and
+Thirion's demons \cite{Thirion1998} operate in the space of vector fields, which does not preserve topology.
 In other words, barring ad hoc constraints to prevent otherwise, these algorithms allow the topology to change in an uncontrolled way which makes the deformable mappings difficult to interpret in functional or anatomical studies.
@@
 These topology concerns have led to current advocation within the computational anatomy research community of diffeomorphic mappings for transformation modeling.
-Optimizing directly within this space has shown remarkable success in various computational anatomy studies involving longitudinal (\cite{Avants2007b,Fox2001}), functional (\cite{Miller2005}), and population data (\cite{Avants2007b}).
+Optimizing directly within this space has shown remarkable success in various computational anatomy studies involving longitudinal \cite{Avants2007b,Fox2001}, functional \cite{Miller2005}, and population data \cite{Avants2007b}.
 We include three such diffeomorphic algorithms in the ANTS toolbox based on previously reported research.
 
-Regardless of current research trends, however, we recognize that selection of the transformation model is ultimately application-specific, that no single choice is optimal for all scenarios (\cite{Wolpert1997}), and, therefore, the transformation model must be chosen in a principled fashion.
-In fact, several non-diffeomorphic algorithms performed quite well in Klein's comparative study of nonrigid registration algorithms (\cite{Klein2009}).
+Regardless of current research trends, however, we recognize that selection of the transformation model is ultimately application-specific, that no single choice is optimal for all scenarios \cite{Wolpert1997}, and, therefore, the transformation model must be chosen in a principled fashion.
+In fact, several non-diffeomorphic algorithms performed quite well in Klein's comparative study of nonrigid registration algorithms \cite{Klein2009}.
 For this reason, we also include elastic-related methods for a panoply of transformation model options.
@@
     \begin{tabular}{c}
-      \includegraphics[width=80mm]{hierarchy}
+      \includegraphics[width=80mm]{Figures/hierarchy}
     \end{tabular}
@@
 
-In contrast to many transformation models which reside in the domain of vector spaces, a diffeomorphism is a differentiable mapping with a differentiable inverse (\cite{Ebin1970,Mumford1998}).
+In contrast to many transformation models which reside in the domain of vector spaces, a diffeomorphism is a differentiable mapping with a differentiable inverse \cite{Ebin1970,Mumford1998}.
 Modeling transformations with diffeomorphisms ensures certain desirable topological properties that cannot be guaranteed with other methods.
@@
 
-Dupuis et al. (\cite{Dupuis1998}) motivated the usage of diffeomorphisms for CA  by showing that the variational form
+Dupuis et al. \cite{Dupuis1998} motivated the usage of diffeomorphisms for CA  by showing that the variational form
 \begin{align}\label{eq:D}
@@
 An optimal solution, $\boldsymbol{v}^*$, minimizes the metric $D(\mathcal{I}, \mathcal{J})$ with respect to $L$.
-Dupuis (\cite{Dupuis1998}) also showed that such a solution is guaranteed to be well-posed.
+Dupuis \cite{Dupuis1998} also showed that such a solution is guaranteed to be well-posed.
 Intuitively, Equation (\ref{eq:D}) provides a sense of distance between two anatomical shapes.
@@
 In most real-world applications, however, a diffeomorphic path connecting the anatomical instance $\mathcal{J}$ with $\mathcal{I}$ is non-existent (due, for example, to the photometric variation or the presence/absence of a tumor in neuroanatomical images).
-Therefore, the following minimizing variational form is used for optimization in diffeomorphic normalization to accommodate inexact matching  (\cite{Dupuis1998,Miller2002})
+Therefore, the following minimizing variational form is used for optimization in diffeomorphic normalization to accommodate inexact matching  \cite{Dupuis1998,Miller2002}
 \begin{align} \label{eq:lddmm}
@@
 \end{align}
-The Euler-Lagrange equations characterizing the optimizing time-varying velocity field, $\boldsymbol{v}^*$, were derived in \cite{Miller2002} and later used in formulating the gradient-descent optimization scheme known as {\em large deformation diffeomorphic metric-matching} (LDDMM) (\cite{Beg2005}) with the similarity metric, or data term, for LDDMM being the squared intensity difference with weight $\lambda$.
+The Euler-Lagrange equations characterizing the optimizing time-varying velocity field, $\boldsymbol{v}^*$, were derived in \cite{Miller2002} and later used in formulating the gradient-descent optimization scheme known as {\em large deformation diffeomorphic metric-matching} (LDDMM) \cite{Beg2005} with the similarity metric, or data term, for LDDMM being the squared intensity difference with weight $\lambda$.
 
@@
 \end{align}
-The corresponding symmetric Euler-Lagrange equations are similar to (\cite{Miller2002}).
+The corresponding symmetric Euler-Lagrange equations are similar to \cite{Miller2002}.
 Finding $\boldsymbol{v_1}^*$ minimizes the variational energy from $t = 0$ whereas $\boldsymbol{v_2}^*$ minimizes from $t = 1$.
@@
 Other diffeomorphic flavors have since been reported in the research literature (e.g.
-DARTEL (\cite{Ashburner2007}) and Diffeomorphic Demons (\cite{Vercauteren2007,Vercauteren2009}).
+DARTEL \cite{Ashburner2007} and Diffeomorphic Demons \cite{Vercauteren2007,Vercauteren2009}.
 We include three diffeomorphic transformation models for parameterizing $\phi(\cdot)$.
@@
 \paragraph{Exponential Mapping}
-Ashburner introduced DARTEL (Diffeomorphic Anatomical Registration using Exponentiated Lie algebra) as a rapidly computed alternative to time parameterized diffeomorphic schemes (\cite{Ashburner2007}).
-The key difference between a time-varying diffeomorphism and a diffeomorphism generated by an exponential mapping (\cite{Ashburner2007}) is that the exponential mapping maintains only a single vector field that is constant in time.
+Ashburner introduced DARTEL (Diffeomorphic Anatomical Registration using Exponentiated Lie algebra) as a rapidly computed alternative to time parameterized diffeomorphic schemes \cite{Ashburner2007}.
+The key difference between a time-varying diffeomorphism and a diffeomorphism generated by an exponential mapping \cite{Ashburner2007} is that the exponential mapping maintains only a single vector field that is constant in time.
 
@@
 Note that there is no explicit time parameter in the velocity field.
-Theoretically, restricting the velocity field to be constant in time reduces the size of the space that may be generated (\cite{Arnold1991}) in a way that is similar to the difference between real and rational numbers, the latter of which are sparsely distributed through the reals.
+Theoretically, restricting the velocity field to be constant in time reduces the size of the space that may be generated \cite{Arnold1991} in a way that is similar to the difference between real and rational numbers, the latter of which are sparsely distributed through the reals.
 
@@
 
-A simple and efficient, yet powerful image normalization algorithm is the approach known as Thirion's demons (\cite{Thirion1998}).
+A simple and efficient, yet powerful image normalization algorithm is the approach known as Thirion's demons \cite{Thirion1998}.
 Using an optical flow based similarity, the solution is obtained by iterating between the calculation of image forces and subsequent Gaussian regularization.
@@
 \paragraph{Directly Manipulated Free-Form Deformation}
-Another top performer in Klein's study (\cite{Klein2009}) was the Image Registration Toolkit (IRTK) based on the research originally reported in \cite{Rueckert1999} in which mutual information and a free-form deformation (FFD) transformation model were used to analyze breast deformation.
-In ANTS we provide an implementation of a variant of the well-known FFD transformation model for image registration known as {\em directly manipulated free-form deformation} (\cite{Tustison2009}).
+Another top performer in Klein's study \cite{Klein2009} was the Image Registration Toolkit (IRTK) based on the research originally reported in \cite{Rueckert1999} in which mutual information and a free-form deformation (FFD) transformation model were used to analyze breast deformation.
+In ANTS we provide an implementation of a variant of the well-known FFD transformation model for image registration known as {\em directly manipulated free-form deformation} \cite{Tustison2009}.
 The DMFFD model replaces the standard FFD gradient used in \cite{Rueckert1999} with an intuitive preconditioned gradient to overcome problematic energy topographies intrinsic with the traditional approach.
@@
 \end{align}
-which is the gradient used in (\cite{Rueckert1999}).  In contrast, the DMFFD approach uses a preconditioned gradient given by
+which is the gradient used in \cite{Rueckert1999}.  In contrast, the DMFFD approach uses a preconditioned gradient given by
 \begin{align} \label{eq:DMFFDgrad}
@@
 
-Both mutual information (\cite{Viola1997}) and mean-squared difference similarity metrics are available for the linear transformations.
-Also included are the cross-correlation (\cite{Avants2007b}), probabilistic matching (a more aggressive version of cross-correlation), local mutual information (\cite{Rueckert1999,Pluim2003}), and mean squared difference similarity metrics for the non-linear transformation models.
-The parameters for the different metrics are discussed in the ANTS documentation (\cite{ANTS2009}).
+Both mutual information \cite{Viola1997} and mean-squared difference similarity metrics are available for the linear transformations.
+Also included are the cross-correlation \cite{Avants2007b}, probabilistic matching (a more aggressive version of cross-correlation), local mutual information \cite{Rueckert1999,Pluim2003}, and mean squared difference similarity metrics for the non-linear transformation models.
+The parameters for the different metrics are discussed in the ANTS documentation \cite{ANTS2009}.
 
@@
 Generalizing the B-spline fitting algorithm of \cite{Lee1997}, 
-we developed a scattered data approximation algorithm (\cite{Tustison2006b}) 
-and contributed the code to the ITK library (\cite{Tustison2005}).  
+we developed a scattered data approximation algorithm \cite{Tustison2006b} 
+and contributed the code to the ITK library \cite{Tustison2005}.  
 This code is included in ANTS and forms the basis of our exact landmark 
@@
 \subsubsection{Point-Set Expectation}
-In (\cite{Pluta2008}), the point-set matching problem was formulated in the context 
+In \cite{Pluta2008}, the point-set matching problem was formulated in the context 
 of incomplete label matching but is equally applicable to the general scenario of 
@@
 Recent information theoretic approaches have been proposed for point-set registration.
-A previous open-source contribution (\cite{Tustison2009}) generalizes the Jensen-Shannon divergence to the Jensen-Havrda-Charvat-Tsallis (JCHT) divergence which
+A previous open-source contribution \cite{Tustison2009} generalizes the Jensen-Shannon divergence to the Jensen-Havrda-Charvat-Tsallis (JCHT) divergence which
 permits a fine-tuning  of the divergence measure such that emphasis can vary between robustness and sensitivity for 
@@
 where $\mathcal{N}_i$ is the local neighborhood of the point $x_i$ and $\mathcal{K}$ is a user-selected neighborhood  weighting kernel.
-We use an isotropic Gaussian for $\mathcal{K}$ with variance $\sigma_{\mathcal{K}_i}^2$ as well as a k-d tree structure for efficient determination of $\mathcal{N}_i$ (\cite{Berg2000}).
+We use an isotropic Gaussian for $\mathcal{K}$ with variance $\sigma_{\mathcal{K}_i}^2$ as well as a k-d tree structure for efficient determination of $\mathcal{N}_i$ \cite{Berg2000}.
 Calculation of the gradient requires the inverse of each covariance matrix.
@@
   \begin{tabular}{c}
-      \includegraphics[width=135mm]{ANTSdashboard} 
+      \includegraphics[width=135mm]{Figures/ANTSdashboard} 
   \end{tabular}
@@
     \end{tabular}
-  \caption{The various flags and variables for a variety of image registration possibilities.   Additional information can be found on the ANTS website (\cite{ANTS2009}). }
+  \caption{The various flags and variables for a variety of image registration possibilities.   Additional information can be found on the ANTS website \cite{ANTS2009}. }
   \label{table:args}
@@
   \begin{tabular}{c}
-      \includegraphics[width=85mm]{phantomwmgm} 
+      \includegraphics[width=85mm]{Figures/phantomwmgm} 
   \end{tabular}
@@
 
-\bibliographystyle{elsarticle-num-names.bst}
-\bibliography{/Users/nick/Pending/references.bib}
+\bibliographystyle{IEEEtran.bst}
+\bibliography{./references.bib}
 

### 2009-06-28 Documentation/Tex/antstheory.tex
Updated the description of the 2-D and 3-D data used in the experiments.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@49 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 
-As outlined in the introduction, in addition to the transformation, the optimization strategy and similarity metric form the image normalization schematic troika.
+As outlined in the introduction, in addition to the transformation, the optimization strategy and similarity metric form the image normalization scheme.
 Since our optimization strategy is limited to gradient descent, experimental analysis includes an exploration of optimal gradient steps within a sensible window where the steps are scaled according to the voxel spacing.
@@
 \subsection{2-D Simulated Image Normalization Evaluation}
+
+Eight simulated 2-D images were created to model the types of deformations one would encounter in brain image normalization.
+Each image was created with isotropic spacing and of size $102\times95$.
+The foreground of each image was comprised of two labels representing the white matter and grey matter.
+Since the 2-D simulated image experiments were used to prune the transformation model space (and not the image similarity metric space) the foreground intensities were not created to model the intensity variation normally seen in the grey/white matter.
+Note that these images are distributed with the ANTS open source package.
 
@@
   \end{tabular}
-\caption{The eight 2-D simulated images used for the initial parameter search.  Each image is of size $[102,95]$ and are available with the ANTS source distribution.}  
+\caption{The eight 2-D simulated images used for the initial parameter search.  These images  are available with the ANTS source distribution.}  
 \label{fig:phantoms}
@@
 \subsection{3-D NIREP Brain Image Normalization Evaluation}
+
+The Non-Rigid Image Registration Project is a large-scale evaluation resource for deformable registration algorithms headed by Gary Christensen at the University of Iowa.
+In addition to the development and distribution of the necessary software tools for algorithmic validation, this project includes the distribution of appropriate image data and corresponding segmentations.
+One such database that has been made available is referred to as the ``NA0'' database consisting of 16 MR image volumes of normal adult human volunteers.
+A brief demographic sketch of the NA0 database is as follows:  8 males with mean age of 32.5 $\pm$ 8.4 years (range of 25 to 48 years) and  8 females with mean age of 29.8 $\pm$ 5.8 years (range of 24 to 41).
+
+Following acquisition, each image was resampled and padded to isotropic voxel spacing of $0.7\times0.7\times0.7$ mm$^3$ and total size of $256 \times 300 \times 256$ voxels.
+The cortex of each of the 16 MR image volumes was segmented into 32 regions \cite{Allen2002,Allen2003} using Brainvox which were later refined using manual editing.
+
+\begin{figure}
+\centering
+  \begin{tabular}{cc}
+      \includegraphics[width=40mm]{Figures/na01x_axial.png} &
+      \includegraphics[width=40mm]{Figures/na01x_axial_labels.png} \\
+      \includegraphics[width=40mm]{Figures/na01x_coronal.png} &
+      \includegraphics[width=40mm]{Figures/na01x_coronal_labels.png} \\
+      \reflectbox{\includegraphics[width=40mm]{Figures/na01x_sagittal.png}} &
+      \reflectbox{\includegraphics[width=40mm]{Figures/na01x_sagittal_labels.png}} \\
+  \end{tabular}
+\caption{Canonical image views of NA01 image from the NIREP data base (left column) and the corresponding segmentations (right column).  }  
+\label{fig:na01x}
+\end{figure}
 

### 2009-07-08 ImageRegistration/itkAvantsMutualInformationRegistrationFunction.h
mi typedef bug
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@52 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -254,7 +254,7 @@ public:
   typedef typename MarginalPDFType::IndexType MarginalPDFIndexType;
   /** Typedef for the joint PDF and PDF derivatives are stored as ITK Images. */
   typedef Image<PDFValueType,2> JointPDFType;
-  typedef Image<PDFValueType,2> JointPDFDerivativesType;
+  typedef Image<PDFValueType,3> JointPDFDerivativesType;
   typedef JointPDFType::IndexType                JointPDFIndexType;
   typedef JointPDFType::PixelType                JointPDFValueType;
   typedef JointPDFType::RegionType              JointPDFRegionType;

### 2009-07-08 Documentation/Tex/ants.tex
Changed the figure labels to appear after the captions which seems to have fixed the reference problem.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@47 fd2da0a4-d968-48df-9cb5-d8c496287cbf
ok
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@53 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 \begin{figure}
-\label{fig:cmd}
 \includegraphics[width=0.9\textwidth]{Figures/ANTSSyntax.pdf} 
@@
 \vspace{-0.1in}
+\label{fig:cmd}
 \end{figure}
@@
 
-\subsubsection{Linear Registration}
-\begin{figure}
-\label{fig:aff}
+\subsubsection{Affine and Rigid Registration}
+\begin{figure}
 \includegraphics[width=0.9\textwidth]{./Figures/ANTSanatomy.pdf} 
@@
 \vspace{-0.1in}
+\label{fig:aff}
 \end{figure}
@@
 
+\noindent{\bf Rigid Registration: }  ANTS will also perform rigid registration.  
+To force the linear component of registration to be purely rigid (no scaling or shearing), 
+then pass the parameter  \verb"--rigid-affine true " to 
+the ANTS command line.   For example,  
+\begin{verbatim}
+ANTS 3 -m MI[AtlasHead.nii,crAnatomical.nii,1,32] 
+-o TEST -i 0 --rigid-affine true 
+\end{verbatim}
+The jacobian of the resulting affine mapping should be unity.
+
 \subsubsection{Deformable Registration}
@@
 \begin{figure}
-\label{fig:antshier}
 \includegraphics[width=0.9\textwidth]{./Figures/ANTSElastToDiff.pdf} 
@@
 \vspace{-0.1in}
+\label{fig:antshier}
 \end{figure}
@@
 \begin{figure}
-\label{fig:pull}
 \center \includegraphics[width=0.75\textwidth]{./Figures/PullBack.pdf} 
 \itkcaption[ANTS Deformation]{A deformation in a digital domain.}
+\label{fig:pull}
 \end{figure}  
@@
 \begin{figure}
-\label{fig:inv}
 \center \includegraphics[width=0.75\textwidth]{Figures/DeformationsInDigitalDomain.pdf} 
@@
 \vspace{-0.1in}
+\label{fig:inv}
 \end{figure}  
@@
  \begin{figure}
-\label{fig:warp}
 \includegraphics[width=0.9\textwidth]{Figures/ANTSSyntax2.pdf} 
@@
 \vspace{-0.1in}
+\label{fig:warp}
 \end{figure}
@@
 \begin{figure}
-\label{fig:antssyn}
 \includegraphics[width=0.9\textwidth]{./Figures/ANTSsyn.pdf} 
@@
 \vspace{-0.1in}
+\label{fig:antssyn}
 \end{figure}
@@
 \begin{wrapfigure}{r}{0.5\textwidth}
-\label{fig:dterr}
 \includegraphics[width=0.5\textwidth]{Figures/dtiordererr.pdf} 
@@
 \vspace{-0.1in}
+\label{fig:dterr}
 \end{wrapfigure}  
@@
 \begin{figure}
-\label{fig:mvnorm}
 \includegraphics[width=0.9\textwidth]{Figures/multivariatenorm.pdf} 
@@
 group template.  In brain imaging, the intersubject maps are usually guided by the T1 component.}\vspace{-0.1in}
+\label{fig:mvnorm}
 \end{figure}
@@
 Our goal is to derive a "most representative" single image from this population. 
-The steps are: \begin{figure}
+The steps are: 
+\begin{figure}
+\center \includegraphics[width=0.7\textwidth]{Figures/templateex.jpg} 
+\itkcaption[ANTS Optimal Template]{This example may be recreated by the user.  
+The shared anatomy -- 
+across this dataset -- is recovered in the derived optimal template.}
 \label{fig:template}
-\center \includegraphics[width=0.7\textwidth]{Figures/templateex.jpg} 
-\itkcaption[ANTS Optimal Template]{This example may be recreated by the user.  The shared anatomy -- 
-across this dataset -- is recovered in the derived optimal template.}
 \end{figure}
@@
 \begin{figure}
-\label{fig:large}
 \includegraphics[width=0.9\textwidth]{Figures/ANTSLargeDef.jpg} 
@@
 }
+\label{fig:large}
 \end{figure}
@@
 \begin{figure}
-\label{fig:seg}
 \includegraphics[width=1\textwidth]{Figures/segmentation.pdf} 
@@
 }
+\label{fig:seg}
 \end{figure}
@@
 \begin{figure}
-\label{fig:seg2}
 \includegraphics[width=1\textwidth]{Figures/segmentation2.pdf}
@@
 }
+\label{fig:seg2}
 \end{figure}
@@
 \begin{figure}
-\label{fig:frown}
 \includegraphics[width=0.45\textwidth]{Figures/frowntosmile.png}
@@
 the deformed images. }
+\label{fig:frown}
 \end{figure}
@@
 \begin{figure}
-\label{fig:lesion}
 \includegraphics[width=1\textwidth]{Figures/lesionstudy.pdf}
@@
 normal tissue.}
+\label{fig:lesion}
 \end{figure}
@@
 \begin{figure}
-\label{fig:morph}
 \includegraphics[width=1\textwidth]{Figures/morphometry.pdf}
 \itkcaption[ANTS Morphometry]{The output from the ANTS morphometry example.}
+\label{fig:morph}
 \end{figure}
@@
 \begin{figure}
-\label{fig:jac}
 \includegraphics[width=1\textwidth]{Figures/Jacobian.pdf}
@@
   across the population.}
+\label{fig:jac}
 \end{figure}

### 2009-07-09 Temporary/itkFastMarchingImageFilter.h
Initial commit.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@54 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -1,0 +1,413 @@
+/*=========================================================================
+
+  Program:   Insight Segmentation & Registration Toolkit
+  Module:    $RCSfile: itkFastMarchingImageFilter.h,v $
+  Language:  C++
+  Date:      $Date: 2009-04-23 03:53:35 $
+  Version:   $Revision: 1.40 $
+
+  Copyright (c) Insight Software Consortium. All rights reserved.
+  See ITKCopyright.txt or http://www.itk.org/HTML/Copyright.htm for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkFastMarchingImageFilter_h
+#define __itkFastMarchingImageFilter_h
+
+#include "itkArray.h"
+#include "itkImageToImageFilter.h"
+#include "itkIndex.h"
+#include "itkLevelSet.h"
+#include "itkNeighborhoodIterator.h"
+
+#include "vnl/vnl_math.h"
+
+#include <functional>
+#include <queue>
+
+namespace itk
+{
+
+/** \class FastMarchingImageFilter
+ * \brief Solve an Eikonal equation using Fast Marching
+ *
+ * Fast marching solves an Eikonal equation where the speed is always
+ * non-negative and depends on the position only. Starting from an 
+ * initial position on the front, fast marching systematically moves the
+ * front forward one grid point at a time.
+ *
+ * Updates are preformed using an entropy satisfy scheme where only
+ * "upwind" neighborhoods are used. This implementation of Fast Marching
+ * uses a std::priority_queue to locate the next proper grid position to
+ * update. 
+ * 
+ * Fast Marching sweeps through N grid points in (N log N) steps to obtain
+ * the arrival time value as the front propagates through the grid.
+ *
+ * Implementation of this class is based on Chapter 8 of
+ * "Level Set Methods and Fast Marching Methods", J.A. Sethian,
+ * Cambridge Press, Second edition, 1999.
+ *
+ * This class is templated over the level set image type and the speed
+ * image type. The initial front is specified by two containers: one
+ * containing the known points and one containing the trial
+ * points.  Alive points are those that are already part of the
+ * object, and trial points are considered for inclusion.
+ * In order for the filter to evolve, at least some trial
[…]

### 2009-07-10 Tensor/itkPreservationOfPrincipalDirectionTensorReorientationImageFilter.h
reorientation of vectors
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@55 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -127,6 +127,23 @@ protected:
    *     ImageToImageFilter::GenerateData() */
   void GenerateData( void );
 
+
+  typename DeformationFieldType::PixelType 
+    TransformVector(DeformationFieldType* field, typename FloatImageType::IndexType index )
+    {
+      typename DeformationFieldType::PixelType vec=field->GetPixel(index);
+      typename DeformationFieldType::PixelType newvec; 
+      newvec.Fill(0);
+      
+      for (unsigned int row=0; row<ImageDimension; row++)
+	for (unsigned int col=0; col<ImageDimension; col++)
+	  newvec[row]+=vec[col]*field->GetDirection()[row][col];
+      
+      return newvec;
+    }
+
+
+
 private:
   PreservationOfPrincipalDirectionTensorReorientationImageFilter(const Self&); //purposely not implemented
   void operator=(const Self&); //purposely not implemented

### 2009-07-14 Documentation/Tex/ants.tex
update refs
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@58 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 
+\section{Annotated Bibliography}
+The original statement of the symmetric normalization and template
+construction methodology was given in \cite{Avants2004}.  A follow up
+study that used landmark guidance to compare the chimpanzee cortex to
+the human cortex was published here \cite{Avants2006} -- this study
+used {\em in vivo} MRI and template-based normalization to confirm
+volumetric numbers derived from an early 20th century post-mortem
+study comparing one human and one chimp.  This conference article has
+some additional detail and alternative updates to the methodology, in
+particular application to shape-based interpolation
+\cite{Avants2005b}.  Network based studies were performed here
+\cite{duda08miccai,duda08cvpr}.  The main SyN paper is here
+\cite{Avants2008}.  Applications to neurodegeration are here
+\cite{Avants2005,Avants2008a,Grossman2008,Avants2009,Das2009,Yushkevich2009,Massimo2009}.
+Hippocampus focused work is here \cite{Pluta2009,Yushkevich2009}.  The
+main evaluation papers include \cite{Avants2008} and \cite{Klein2009}
+for the cortex and deep brain structures whereas \cite{Pluta2009}
+evaluates the use of automated and semi-automated normalization for
+high-throughput hippocampus morphometry.  An additional evaluation
+paper is being developed.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2009-08-19 Documentation/Tex/ants.tex
tcsh issues
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@99 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 \subsection{ANTS Scripts} 
-The ANTS/Scripts directory contains (hopefully) user-friendly wrappings of ANTS tools 
-that enable higher-level error checking and combinations of basic ANTS functions. 
-These scripts are called as \verb sh  ~ \verb antsscriptname.sh ~ and provide usage when 
-called from the command line.  For instance, try \verb sh ~ \verb ants.sh .
+The ANTS/Scripts directory contains (hopefully) user-friendly
+wrappings of ANTS tools that enable higher-level error checking and
+combinations of basic ANTS functions.  These scripts are called as
+\verb sh ~ \verb antsscriptname.sh ~ and provide usage when called
+from the command line.  For instance, try \verb sh ~ \verb ants.sh .
+Additionally, if you run ants in tcsh shells, you will need to call it
+as follows:
+\begin{verbatim}
+~/Code/ANTSSourceForgeSVN/bin/ANTS 2 -m PR\[r16slice.nii,r64slice.nii,1,2\]
+\end{verbatim}
+instead of
+\begin{verbatim}
+~/Code/ANTSSourceForgeSVN/bin/ANTS 2 -m PR[r16slice.nii,r64slice.nii,1,2]
+\end{verbatim}
+because tcsh does no interpret the brackets as intended.  Using the
+bash shell script interface also avoids this.  The key change involves using the slash to force tcsh to 
+interpet the bracket literally.
 

### 2009-08-19 ImageRegistration/itkANTSImageRegistrationOptimizer.h
adding the        -F, --fixed-image-initial-affine  option
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@100 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -133,6 +133,9 @@ public:
     void SetInverseDeformationField(DeformationFieldPointer A) {this->m_InverseDeformationField=A;}
     void SetMaskImage( ImagePointer m) { this->m_MaskImage=m; }
 
+    void SetFixedImageAffineTransform(AffineTransformPointer A) {this->m_FixedImageAffineTransform=A;}
+    AffineTransformPointer GetFixedImageAffineTransform() {return this->m_FixedImageAffineTransform;}
+
     /** Get functions */
     AffineTransformPointer GetAffineTransform() {return this->m_AffineTransform;}
     DeformationFieldPointer GetDeformationField( ) {return this->m_DeformationField;}
@@ -359,7 +362,7 @@ public:
 
     }
 
-  PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer movingImage, PointSetPointer movingpoints,  AffineTransformPointer aff , DeformationFieldPointer totalField, bool doinverse )
+PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer movingImage, PointSetPointer movingpoints,  AffineTransformPointer aff , DeformationFieldPointer totalField, bool doinverse , AffineTransformPointer  fixedaff  )
   {
     if (!movingpoints) { std::cout << " NULL POINTS " << std::endl;  return NULL; }
 
@@ -368,6 +371,12 @@ public:
       {
       affinverse=AffineTransformType::New();
       aff->GetInverse(affinverse);
+      }
+    AffineTransformPointer fixedaffinverse=NULL;
+    if (fixedaff)
+      {
+      fixedaffinverse=AffineTransformType::New();
+      fixedaff->GetInverse(fixedaffinverse);
       }
        
       typedef itk::WarpImageMultiTransformFilter<ImageType,ImageType, DeformationFieldType, TransformType> WarperType;
@@ -378,11 +387,13 @@ public:
      if (!doinverse)
 	{
 	if (totalField) warper->PushBackDeformationFieldTransform(totalField);
-	if (aff) warper->PushBackAffineTransform(aff);
+	if (fixedaff) warper->PushBackAffineTransform(fixedaff);
+	else if (aff) warper->PushBackAffineTransform(aff);
 	}
       else
 	{	
 	if (aff) warper->PushBackAffineTransform( affinverse );
+	else if (fixedaff) warper->PushBackAffineTransform(fixedaffinverse);
[…]

### 2009-08-19 ImageRegistration/itkANTSImageTransformation.h

@@ -90,6 +90,8 @@ class ITK_EXPORT ANTSImageTransformation
   AffineTransformPointer GetAffineTransform() { return this->m_AffineTransform;  }
   DeformationFieldPointer GetDeformationField( ) { return this->m_DeformationField; }
   DeformationFieldPointer GetInverseDeformationField() { return this->m_InverseDeformationField; }
+  void SetFixedImageAffineTransform(AffineTransformPointer A) {this->m_FixedImageAffineTransform=A;}   
+  AffineTransformPointer GetFixedImageAffineTransform() {return this->m_FixedImageAffineTransform;}
 
   /** Initialize the mapping */
   void InitializeTransform()
@@ -144,6 +146,7 @@ private:
   void operator=( const Self& ); //purposely not implemented
 
   AffineTransformPointer    m_AffineTransform;
+  AffineTransformPointer    m_FixedImageAffineTransform;
   DeformationFieldPointer m_DeformationField;
   DeformationRegionOfInterestType  m_DeformationRegionOfInterest;
   DeformationRegionOfInterestCenterType  m_DeformationRegionOfInterestCenter;

### 2009-08-20 ImageRegistration/itkANTSImageRegistrationOptimizer.h
a first pass at DiReCT thickness in ANTS :
open issues include -- we need to introduce priors , a special
thickness-oriented similarity function, and a different optimization
of the diff mapping.



git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@103 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -632,7 +632,17 @@ PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer mo
       this->m_RoiNumbers = this->m_Parser->template ConvertVector<float>(temp);
       if ( temp.length() > 3 ) this->m_UseROI=true;
       }
-   
+
+    typename ParserType::OptionType::Pointer oOption 
+      = this->m_Parser->GetOption( "output-naming" );  
+    this->m_OutputNamingConvention=oOption->GetValue();   
+
+
+    typename ParserType::OptionType::Pointer thicknessOption 
+      = this->m_Parser->GetOption( "compute-thickness" );  
+    if( thicknessOption->GetValue() == "true" ) this->m_ComputeThickness=true;
+    else this->m_ComputeThickness=false;
+    std::cout <<" compute thickness? " << this->m_ComputeThickness << std::endl;
     /**
      * Get transformation model and associated parameters
      */
@@ -1539,7 +1549,8 @@ private:
   bool m_UseNN;
   unsigned int m_CurrentIteration;
   unsigned int m_CurrentLevel;
-   std::string m_TransformationModel;
+  std::string m_TransformationModel;
+  std::string m_OutputNamingConvention;
   PointSetPointer  m_FixedPointSet;
   PointSetPointer  m_MovingPointSet;
   std::vector<unsigned int> m_Iterations;
@@ -1569,6 +1580,11 @@ private:
   unsigned int m_BSplineFieldOrder;  
   ArrayType m_GradSmoothingMeshSize;
   ArrayType m_TotalSmoothingMeshSize;
+
+/** For thickness calculation */
+  ImagePointer m_HitImage; 
+  ImagePointer m_ThickImage; 
+  bool m_ComputeThickness; 
 
 };
 

### 2009-08-21 ImageRegistration/itkANTSImageRegistrationOptimizer.h
adding some new stuff -- masking etc for cortical thickness estimation + a special DiReCT optimizer  for the transformation
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@104 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -982,7 +982,9 @@ PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer mo
             }
           if(this->m_Iterations[currentLevel] > 0)
             {
-            if (this->m_SyNType) 
+            if (this->m_SyNType && this->m_ComputeThickness ) 
+              this->DiReCTUpdate(fixedImage, movingImage, this->m_SimilarityMetrics[0]->GetFixedPointSet(),  this->m_SimilarityMetrics[0]->GetMovingPointSet() );
+            else if (this->m_SyNType) 
               this->SyNTVRegistrationUpdate(fixedImage, movingImage, this->m_SimilarityMetrics[0]->GetFixedPointSet(),  this->m_SimilarityMetrics[0]->GetMovingPointSet() );
             else 
               this->SyNRegistrationUpdate(fixedImage, movingImage, this->m_SimilarityMetrics[0]->GetFixedPointSet(),  this->m_SimilarityMetrics[0]->GetMovingPointSet() );
@@ -1195,6 +1197,7 @@ PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer mo
   void SyNExpRegistrationUpdate(ImagePointer fixedImage, ImagePointer movingImage, PointSetPointer fpoints=NULL, PointSetPointer mpoints=NULL);
 
   void SyNTVRegistrationUpdate(ImagePointer fixedImage, ImagePointer movingImage, PointSetPointer fpoints=NULL, PointSetPointer mpoints=NULL);
+  void DiReCTUpdate(ImagePointer fixedImage, ImagePointer movingImage, PointSetPointer fpoints=NULL, PointSetPointer mpoints=NULL);
 
   /** allows one to copy or add a field to a time index within the velocity
 * field 

### 2009-09-30 Documentation/Tex/ants.tex
large def
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@158 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 \section{Introduction}
-The ANTS framework provides open-source, encapsulated,
-software-engineered functionality for deformable normalization and
-segmentation.  Independent evaluation of ANTS normalization tools, 
+The ANTS framework provides open-source functionality for deformable
+normalization with large deformations, as shown in
+figure~\ref{fig:chalf}.  Small deformation mappings and segmentation 
+tools are also available. 
+Independent evaluation of ANTS normalization tools, 
 applied to ``control'' data, 
@@
 \footnote{This document is a work in progress. Please check for updates with each release.}
+\begin{figure}
+\includegraphics[width=0.33\textwidth]{Figures/grid1100.pdf} 
+\includegraphics[width=0.33\textwidth]{Figures/grid1110.pdf} 
+\includegraphics[width=0.33\textwidth]{Figures/grid1119.pdf} 
+\itkcaption[ANTS Large Deformation]{The original goal of ANTS was to
+  develop public, open source large deformation image registration.
+  This is a classic example showing the progress of deforming a half
+  C to a full C along a geodesic diffeomorphism.  The deforming 
+  grid accompanies each deformed image.}
+\vspace{-0.1in}
+\label{fig:chalf}
+\end{figure}
 
@@
 interpet the bracket literally. 
-
+\newpage
 \section{ImageRegistration with ANTS}
@@
 A useful script for setting up optimal template construction
- is available in ANTS/Scripts.  This method is described in \cite{Avants2004,Avants2006d,Kim2008,Yushkevich2009}, 
+ is available in ANTS/Scripts.  This method is described in \cite{Avants2004,Avants2006d,Kim2008,Yushkevich2009,Avants2009c}, 
 where more recent publications are more descriptive and incorporate 
@@
 at left. 
-
+\newpage
 \section{Image Segmentation and Labeling}
@@
 This example should run on the downloaded ANTS data so you may see the results.
-
 %\input{ANTSLabeledDataExample}
 %\input{ANTSMaskedImageRegistrationExample}
-
+\newpage
 \section{Application to Studies}
@@
 Most statistical requirements may be met with this setup. 
+\newpage
 \section{Dependencies and Related Software}
@@
 longitudinal mapping and reconstruction of nifti images from Dicom.
-
+\newpage
 \section{Annotated Bibliography}
 The original statement of the symmetric normalization and template
-construction methodology was given in \cite{Avants2004}.  A follow up
-study that used landmark guidance to compare the chimpanzee cortex to
+construction methodology was given in \cite{Avants2004} with a more 
+recent template paper here \cite{Avants2009c}. 
+A follow up study that used landmark guidance to compare the chimpanzee cortex to
 the human cortex was published here \cite{Avants2006} -- this study

### 2009-09-30 Documentation/Tex/antstheory.tex
update refs
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@58 fd2da0a4-d968-48df-9cb5-d8c496287cbf
ok - minor improvements
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@160 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 lack of standards and reproducibility in the field that is due, in
-part, to the use of proprietary software and private data.  To
+part, to the use of proprietary software and private data that is difficult 
+to openly evaluate.  To
 facilitate reproducibility in CA measurements and advancement of
@@
 %In a recent evaluation of 14 non-linear deformation algorithms \cite{Klein2009}, the SyN algorithm of ANTS was consistently one of the top performers.
-
+{\bf FIXME -- make clear what transformations we are comparing with constant similarity ... }
 \end{abstract}
 
-\IEEEkeywords{
-ANTS, image registration, nueroanatomy, neuroimaging, normalization, open-source, point-set registration
-}
-
-\IEEEpeerreviewmaketitle
+%\IEEEkeywords{
+%ANTS, image registration, nueroanatomy, neuroimaging, normalization, open-source, point-set registration
+%}
+
+%\IEEEpeerreviewmaketitle
+\maketitle
 
@@
 
-The rapid advancement of biological and medical imaging technologies has caused a proliferation in the development of quantitative tools for computational anatomy.
-The principal tools of this emerging field are meaningful deformable mappings between images whether they be driven by similarity metrics which are intensity-based,  point-set based, or both.
-Several categories of mappings have been proposed in the literature.
-Of particular recent interest are diffeomorphic transformations which, by definition, preserve topology.
-Topology preservation is fundamental to making comparisons between objects in the natural world as such transformations permit comparisons to be made across time points in an individual's disease process or to study development patterns across a large population.
+The rapid advancement of biological and medical imaging technologies
+has caused a proliferation in the development of quantitative tools
+for computational anatomy.  The principal tools of this emerging field
+are deformable mappings between images whether they be driven by
+similarity metrics which are intensity-based, point-set based, or
+both.  Several categories of mappings have been proposed in the
+literature.  Of particular recent interest are diffeomorphic
+transformations which, by definition, preserve topology.  Topology
+preservation is fundamental to making comparisons between objects in
+the natural world that are thought to change in such a way that local
+neighborhoods are preserved.  Cytoarchitectonic brain mapping studies
+also suggest that the layout of cell types throughout the brain is
+generally preserved \cite{Schleicher2009}, further motivating diffeomorphic mapping
+in the context of the brain.
 
@@
 
-We first provide an overview of the different components included in ANTS, such as the available transformation models and similarity metrics offered.
-This is followed by an extensive experimental analysis that builds upon the results from the recent image normalization evaluation of \cite{Klein2009} which was limited to a single configuration of ANTS.
-This will provide  a useful benchmark for evaluating other possible ANTS configurations.
+We first provide an overview of the different components included in
+ANTS, such as the available transformation models and similarity
+metrics offered.  This is followed by an extensive experimental
+analysis that builds upon the results from the recent image
+normalization evaluation of \cite{Klein2009} which was limited to a
+single configuration of ANTS.  This will provide a useful benchmark
+for evaluating other possible ANTS configurations.  Importantly, this
+work also suggests---to our knowledge, for the first time---the clear
+practical benefit of time-parameterized diffeomorphic mapping over the
+greedy and exponential mapping approaches.
 
@@
 Thirion's demons \cite{Thirion1998} operate in the space of vector fields, which does not preserve topology.
-In other words, barring ad hoc constraints to prevent otherwise, these algorithms allow the topology to change in an uncontrolled way which makes the deformable mappings difficult to interpret in functional or anatomical studies.
-
-These topology concerns have led to current advocation within the computational anatomy research community of diffeomorphic mappings for transformation modeling.
-Optimizing directly within this space has shown remarkable success in various computational anatomy studies involving longitudinal \cite{Avants2007b,Fox2001}, functional \cite{Miller2005}, and population data \cite{Avants2007b}.
-We include three such diffeomorphic algorithms in the ANTS toolbox based on previously reported research.
-
-Regardless of current research trends, however, we recognize that selection of the transformation model is ultimately application-specific, that no single choice is optimal for all scenarios \cite{Wolpert1997}, and, therefore, the transformation model must be chosen in a principled fashion.
-In fact, several non-diffeomorphic algorithms performed quite well in Klein's comparative study of nonrigid registration algorithms \cite{Klein2009}.
-For this reason, we also include elastic-related methods for a panoply of transformation model options.
+In other words, barring ad hoc constraints to prevent otherwise, these algorithms may allow the topology to change in an uncontrolled way which makes the deformable mappings difficult to interpret in functional or anatomical studies.
+
+Thus, in addition to shape-based, biological motivation, diffeomorphic
+mapping is motivated by the desire for deformable transformations that
+provide well-behaved solutions with mathematical guarantees about
+distances and regularity.  Furthemore, the diffeomorphic space has
+group structure \cite{Arnold1991}.  Optimizing directly within this
+space has shown remarkable success in various computational anatomy
+studies involving longitudinal \cite{Avants2007b,Fox2001}, functional
+\cite{Miller2005}, and population data \cite{Avants2007b}.  We include
+three such diffeomorphic algorithms in the ANTS toolbox based on
+previously reported research and a new time-parameterized extension to
+the standard symmetric normalization algorithm \cite{Avants2008}.
+
+Regardless of current research trends, however, we recognize that
+selection of the transformation model is ultimately
+application-specific, that no single choice is optimal for all
+scenarios \cite{Wolpert1997}, and, therefore, the transformation model
+must be chosen in a principled fashion.  In fact, several
+non-diffeomorphic algorithms performed quite well in Klein's
+comparative study of nonrigid registration algorithms
+\cite{Klein2009}.  For this reason, we also include elastic-related
+methods for a panoply of transformation model options.
 
@@
 One may write this regularization as a variational minimization in terms of its linear operator or in
-terms of a kernel function operating of the field itself, e.g., $
+terms of a kernel function operating on the field itself, e.g., $
 \displace_{smooth}= K \star \displace_{not \,\, smooth}$, where $K \star$ denotes 
@@
 \subsubsection{Diffeomorphic Transformations}
-
-In contrast to many transformation models which reside in the domain of vector spaces, a diffeomorphism is a differentiable mapping with a differentiable inverse \cite{Ebin1970,Mumford1998}.
-Modeling transformations with diffeomorphisms ensures certain desirable topological properties that cannot be guaranteed with other methods.
-
-ANTS assumes the diffeomorphism, $\phi$, is defined on the image domain, $\Omega$, an maintains an affine transform at the boundary such that $\phi(\partial \Omega) = A(\mathbf{Id})$ where $A(\mathbf{Id})$ is an affine mapping applied to the identity transformation.
+Diffeomorphisms form a group of differentiable maps with
+differentiable inverse \cite{Ebin1970,Mumford1998} that is closed
+under composition.  In contrast, the vector space that most deformable
+image registration methods use is closed under {\em addition}, an
+operation that cannot guarantee topology preservation.  That is, two
+topology-preserving vector fields added together are still a vector
+field but the result may no longer preserve topology.  Furthermore,
+most regularization models for vector spaces use a quadratic penalty,
+thus making large deformations difficult to realize.  Modeling
+transformations with diffeomorphisms, on the other hand, ensures a
+flexible, linear penalty on deformation while guaranteeing topology 
+preservation.  Additionally, distance metrics in the space of diffeomorphisms 
+allow geodesic properties to be explored \cite{Miller2001,Beg2005}.
+
+ANTS assumes the diffeomorphism, $\phi$, is defined on the image domain, $\Omega$, and maintains an affine transform at the boundary such that $\phi(\partial \Omega) = A(\mathbf{Id})$ where $A(\mathbf{Id})$ is an affine mapping applied to the identity transformation.
 $\phi$, over time, parameterizes a family of diffeomorphisms, $\phi(\mathbf{x}, t) : \Omega \times t \rightarrow  \Omega$, which can be generated by integrating a time-dependent, smooth velocity field, $\boldsymbol{v} : \Omega \times t \rightarrow \mathbb{R}^d$,  through the ordinary differential equation (o.d.e.)
@@
 \begin{align}\label{eq:D}
-  D(\mathcal{I}, \mathcal{J}) =  \int_0^1 ||L\boldsymbol{v}||^2 dt , \,\, \mathcal{I}(\phi(\mathbf{x}, 1)) = \mathcal{J}(\mathbf{z})
+  D(\mathcal{I}, \mathcal{J}) =  \int_0^1 ||L\boldsymbol{v}|| dt , \,\, \mathcal{I}(\phi(\mathbf{x}, 1)) = \mathcal{J}(\mathbf{z})
 \end{align}
@@
 Exploiting the fact that the diffeomorphism, $\phi$, can be decomposed into two components 
-$\phi_1$ and $\phi_2$, \cite{Avants2007b} construct a {\em symmetric} alternative to Equation (\ref{eq:diff}).  This leads to the symmetric variant of Equation (\ref{eq:diff})
+$\phi_1$ and $\phi_2$, one may construct a {\em symmetric} alternative to Equation (\ref{eq:diff}).  
+Now define, in $t \in [0,0.5]$, $\boldsymbol{v}(x,t) = \boldsymbol{v}_1(x,t) $ and 
+$\boldsymbol{v}(x,t) = \boldsymbol{v}_2(x,1-t)$ when $t \in [0.5,1]$. 
+This leads to the symmetric variant of Equation (\ref{eq:diff}), 
 \begin{align}\label{eq:diffs}
-  \{\boldsymbol{v}_1^*,\boldsymbol{v}_2^*\}  = &
-     \underset{\boldsymbol{v_1},\boldsymbol{v_2}}{\operatorname{argmin}}  
+  \{ \boldsymbol{v}^{*}_1 , \boldsymbol{v}^{*}_2 \}= &
+     \underset{\boldsymbol{v}_{1,2}}{\operatorname{argmin}}  
       \Bigg\{ \nonumber \\
-      & \int_0^{0.5} \left( ||L\boldsymbol{v}_1(x,t)||^2 + ||L\boldsymbol{v}_2(x,t)||^2   \right)dt  \nonumber \\
+      & \int_0^{0.5} ||L\boldsymbol{v}_1(x,t)||^2 ~dt +  \int_0^{0.5} ||L\boldsymbol{v}_2(x,t)||^2 ~dt  \nonumber \\
      &+ \lambda\int_{\Omega}\Pi_{\sim}\left( \mathcal{I} \circ \phi_1(\mathbf{x},0.5), \mathcal{J}\circ \phi_2(\mathbf{x},0.5)\right)d\Omega\Bigg\}.
 \end{align}
+Note that the regularization term, here, is equivalent to that in equation~\ref{eq:lddmm}.  The only change is the splitting of the integral into 
+two time intervals reflecting the underlying optimized components of the velocity field. 
 The corresponding symmetric Euler-Lagrange equations are similar to \cite{Miller2002}.
-Finding $\boldsymbol{v_1}^*$ minimizes the variational energy from $t = 0$ whereas $\boldsymbol{v_2}^*$ minimizes from $t = 1$.
+The difference, here, is that in finding $\boldsymbol{v}^*$, we minimize the variational energy from either end-point towards the mid-point of the transformation, as indicated by the data term.
+This strategy ``splits'' the optimization dependence equally between both images.
 Thus, gradient-based iterative convergence deforms $\mathcal{I}$ and $\mathcal{J}$ along the geodesic diffeomorphism, $\phi$, to a fixed  point midway (intuited by the notion of shape distance) between $\mathcal{I}$ and $\mathcal{J}$ thus motivating the denotation of the solution strategy as Symmetric Normalization (SyN).
@@
 \paragraph{Geodesic SyN}
-Using a gradient-based optimization strategy, minimization implied by Equation (\ref{eq:diffs}), which we denote as $\Pi$, first requires a specified discretization of $t \in [0,1]$.
-Calculation of the gradient of $\Pi$ with respect to the diffeomorphisms $\phi_1$ and $\phi_2$ is performed at each of the $t$ values which is given by
+Using a gradient-based optimization strategy for minimizing equation
+(\ref{eq:diffs}) first requires a specified discretization of $t$,
+s.t. $t \in [0,1] \approx \{ 0 , 1/k , 2/k , \cdots , 1 \} $ where integer $k > 1$
+is the desired number of discretized intervals.  Calculation of the
+gradient of $\Pi$ with respect to the diffeomorphisms $\phi_1$ and
+$\phi_2$ is performed at each of these $t$ values, $t_k$, 
 \begin{align}
-  \nabla \Pi(\boldsymbol{x},t) = \frac{\partial}{\partial \phi_i} \Pi_{\sim}( \mathcal{I}(\phi_{1}^{-1}(\boldsymbol{x},t)), \mathcal{J}(\phi_{2}^{-1}(\boldsymbol{x},1-t)) 
-\end{align}
-for $i \in \{1,2\}$.   The corresponding velocity fields at each $t$ are then updated from the previous iteration according to 
+\label{eq:grad}
+  \nabla \Pi_i(\boldsymbol{x},t_k) = \frac{\partial}{\partial \phi_i} \Pi_{\sim}( \mathcal{I}(\phi_{1}^{-1}(\boldsymbol{x},t_k)), \mathcal{J}(\phi_{2}^{-1}(\boldsymbol{x},1-t_k)) 
+\end{align}
+for $i \in \{1,2\}$.   The total velocity field is then updated from the previous iteration according to, 
+for each $i$, 
 \begin{align}
-  \boldsymbol{v}(\mathbf{x},t) = \boldsymbol{v}(\mathbf{x},t) + K \star \nabla \Pi(\mathbf{x},t).
-\end{align}
-We then integrate Equation (\ref{eq:ode}) using Runge-Kutta methods to yield $\phi_i(\mathbf{x},t)$ for each $t \in [0,1]$ and $i \in \{1,2\}$.
-We cycle through these steps until convergence or iterative exhaustion.
+  \boldsymbol{v}(\mathbf{x},t) = \boldsymbol{v}(\mathbf{x},t) + \delta K \star \nabla \Pi_i(\mathbf{x},t),
+\end{align}
+where $\delta$ is the user-specified gradient descent parameter.  We
+then integrate Equation (\ref{eq:ode}) using Runge-Kutta methods to
+yield $\phi_i(\mathbf{x},t)$ for each $t \in [0,1]$ and $i \in
+\{1,2\}$.  We cycle through these steps until convergence or iterative
+exhaustion.
 
 \paragraph{Greedy SyN}
-Although the Geodesic SyN algorithm conforms most closely to the theoretical diffeomorphic foundations culminating with Equation (\ref{eq:diffs}), the computational and memory cost is significant due to the dense-in-time gradient calculations and requisite reintegration of the diffeomorphisms after each iterative update.
-As a lower-cost alternative, we offer a greedy variant which performs quite well for most medical image normalization problems we have encountered.
-Additionally, this was the strategy used in the large-scale comparative image registration algorithm assessment of \cite{Klein2009}.
-
-Greedy optimization of Equation (\ref{eq:diffs}) requires calculation of the gradient of the gradient taken at the mid-point of the full diffeomorphism, i.e. at $t = 0.5$
-\begin{align}
-\nabla \Pi =\frac{\partial}{\partial \phi_i} \Pi_{\sim}( \mathcal{I}(\phiinv_1(\x,0.5)) ,  \mathcal{J}(\phiinv_2(\x,0.5)) )
-\end{align}
+Although the Geodesic SyN algorithm conforms most closely to the
+theoretical diffeomorphic foundations culminating with Equation
+(\ref{eq:diffs}), the computational and memory cost is significant due
+to the dense-in-time gradient calculations and requisite reintegration
+of the diffeomorphisms after each iterative update.  As a lower-cost
+alternative, we offer a greedy variant which performs quite well for
+most medical image normalization problems we have encountered.
+Additionally, this was the strategy used in the large-scale
+comparative image registration algorithm assessment of
+\cite{Klein2009}.
+
+Greedy optimization of Equation (\ref{eq:diffs}) calculates the gradient only at the mid-point of the full diffeomorphism, i.e. at $t = 0.5$ 
+in equation~\ref{eq:grad}
+% \begin{align}
+% \nabla \Pi =\frac{\partial}{\partial \phi_i} \Pi_{\sim}( \mathcal{I}(\phiinv_1(\x,0.5)) ,  \mathcal{J}(\phiinv_2(\x,0.5)) )
+% \end{align}
 for $i \in \{1,2\}$.  $\phi_1(\mathbf{x},0.5)$ and $\phi_2(\mathbf{x},0.5)$ are then updated from the previous iteration according to 
 \begin{align}
-  \phi_i(\mathbf{x},0.5) = \phi_i(\mathbf{x},0.5)+\delta  (K \star \nabla \Pi ( \phi_i(\x,0.5) )).
-\end{align}
-where $\delta$ is a user-specifed step parameter.  The gradient is then mapped back to the origin of each diffeomorphism to update the full mapping by enforcing $\phiinv(\phi(\x,1))=\x$.
+  \phi_i(\mathbf{x},0.5) = \phi_i(\mathbf{x},0.5)+ (\delta K \star \nabla \Pi_i(\mathbf{x},0.5) ) \circ  \phi_i(\x,0.5).
+\end{align}
+That is, the gradient at the mid-point is mapped back to the origin of each diffeomorphism.  
+We then update the full mapping by explicitly enforcing $\phiinv(\phi(\x,1))=\x$ in the discrete 
+domain, as described in \cite{Avants2008}.
 
@@
   \Pi(&\mathcal{I},\mathcal{J},X,Y,\phi) \nonumber \\ 
-     =& \underbrace{\int_0^1 ||L\boldsymbol{v}||dt}_{\texttt{-t Exp[}\cdot \texttt{]},\texttt{ -r Gauss[}\cdot\texttt{,}\cdot\texttt{]}}  + \underbrace{\lambda_1\int_{\Omega} || \mathcal{I} \circ \phi(\mathbf{x},t) - \mathcal{J}  || d\Omega}_{\texttt{-t Exp[}\cdot \texttt{]},\texttt{ -m MSQ[}\mathcal{J}\texttt{,}\mathcal{I}\texttt{,}\lambda_1\texttt{]}} \nonumber \\
+     =& \underbrace{\int_0^1 ||L\boldsymbol{v}||^2 dt}_{\texttt{-t Exp[}\cdot \texttt{]},\texttt{ -r Gauss[}\cdot\texttt{,}\cdot\texttt{]}}  + \underbrace{\lambda_1\int_{\Omega} || \mathcal{I} \circ \phi(\mathbf{x},t) - \mathcal{J}  || d\Omega}_{\texttt{-t Exp[}\cdot \texttt{]},\texttt{ -m MSQ[}\mathcal{J}\texttt{,}\mathcal{I}\texttt{,}\lambda_1\texttt{]}} \nonumber \\
 &+\underbrace{\frac{\lambda_2 }{|Y|}\sum_{i=1}^{|Y|} \left\|y_i - \frac{1}{|X|} \sum_{j = 1}^{|X|} G(y_i; x_j, \sigma_X) x_j\right\|^2}_{\texttt{-m PSE[}Y\texttt{,}X\texttt{,}\lambda_2\texttt{]}}.
@@
 ANTS is supported by Grant 1R01EB006266-01 from the National Institute Of Biomedical Imaging and Bioengineering and administered through the UCLA Center for Computational Biology.
-% optional entry into table of contents (if used)
-%\addcontentsline{toc}{section}{Acknowledgment}
-%The authors would like to thank...
-
-% trigger a \newpage just before the given reference
-% number - used to balance the columns on the last page
-% adjust value as needed - may need to be readjusted if
-% the document is modified later
-%\IEEEtriggeratref{8}
-% The "triggered" command can be changed if desired:
-%\IEEEtriggercmd{\enlargethispage{-5in}}
-
-% references section
-% NOTE: BibTeX documentation can be easily obtained at:
-% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
-
-% can use a bibliography generated by BibTeX as a .bbl file
-% standard IEEE bibliography style from:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/IEEEtran/bibtex
-%\bibliographystyle{IEEEtran.bst}
-% argument is your BibTeX string definitions and bibliography database(s)
-%\bibliography{IEEEabrv,../bib/paper}
-%
-% <OR> manually copy in the resultant .bbl file
-% set second argument of \begin to the number of references
-% (used to reserve space for the reference number labels box)
 
 \bibliographystyle{IEEEtran.bst}
-\bibliography{./references.bib}
+\bibliography{./references.bib,ants,avantsAll}
 

### 2009-09-30 Documentation/Tex/runtex.sh

@@ -1,5 +1,6 @@
 
-for FL in ants antstheory ; do 
+for FL in ants # antstheory 
+  do 
 for i in 1 2 3 
 do
 pdflatex $FL 

### 2009-09-30 ImageRegistration/itkANTSImageRegistrationOptimizer.h
changing DT to input value in final def integration
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@162 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -1179,12 +1179,8 @@ PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer mo
         //	 this->m_SyNMInv = this->IntegrateVelocity(1,0.5);
         //	 this->m_SyNM=this->CopyDeformationField(mdiffmap);
         //	 this->m_SyNF=this->CopyDeformationField(fdiffmap);
-      	 this->m_DeltaTime *= 0.1;
-        if (this->m_DeltaTime < 0.01) 
-          this->m_DeltaTime = 0.01;
       	 this->m_DeformationField = this->IntegrateVelocity(0,1);	 
         //	 ImagePointer wmimage= this->WarpMultiTransform(  this->m_SmoothFixedImages[0],this->m_SmoothMovingImages[0], this->m_AffineTransform, this->m_DeformationField, false , this->m_ScaleFactor );
-
       	 this->m_InverseDeformationField=this->IntegrateVelocity(1,0);
         }
       else

### 2009-10-01 Utilities/itkBSplineScatteredDataPointSetToImageFilter.h
ENH: Reconfigured the code to be partly multi-threaded.  Specifically, I threaded the function GenerateControlPointLattice() which originally iterated through all the input points in series at each of the multi-resolution levels.  However, as this can be parallelized, the majority of this function is now in ThreadedGenerateData() which gives a significant speed-up for large B-spline fitting problems.  I also changed some of the documentation for readability and completeness.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@164 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -3,8 +3,8 @@
   Program:   Insight Segmentation & Registration Toolkit
   Module:    $RCSfile: itkBSplineScatteredDataPointSetToImageFilter.h,v $
   Language:  C++
-  Date:      $Date: 2009/04/16 18:22:08 $
-  Version:   $Revision: 1.17 $
+  Date:      $Date: 2009-04-20 14:53:42 $
+  Version:   $Revision: 1.8 $
 
   Copyright (c) Insight Software Consortium. All rights reserved.
   See ITKCopyright.txt or http://www.itk.org/HTML/Copyright.htm for details.
@@ -37,22 +37,20 @@ namespace itk
  * a fast approximation to that irregularly spaced data using uniform
  * B-splines.  The traditional method of inverting the observation
  * matrix to find a least-squares fit is made obsolete.  Therefore,
- * memory issues are not a concern and inverting large matrices are
- * unnecessary.  The reference below gives the algorithm for 2-D images
- * and cubic splines.  This class generalizes that work to encompass n-D
- * images and any *feasible* B-spline order.
+ * memory issues are not a concern and inverting large matrices is
+ * not applicable.  In addition, this allows fitting to be multi-threaded.
+ * This class generalizes from Lee's original paper to encompass
+ * n-D data in m-D parametric space and any *feasible* B-spline order as well
+ * as the option of specifying a confidence value for each point.
  *
  * In addition to specifying the input point set, one must specify the number
- * of control points.  If one wishes to use the multilevel component of
+ * of control points.  The specified number of control points must be
+ * > m_SplineOrder.  If one wishes to use the multilevel component of
  * this algorithm, one must also specify the number of levels in the
  * hierarchy.  If this is desired, the number of control points becomes
  * the number of control points for the coarsest level.  The algorithm
  * then increases the number of control points at each level so that
- * the B-spline n-D grid is refined to twice the previous level.  The
- * scattered data is specified by the pixel values.
- *
- * Note that the specified number of control points must be > m_SplineOrder.
- *
+ * the B-spline n-D grid is refined to twice the previous level.
  *
  * \author Nicholas J. Tustison
  *
@@ -65,6 +63,7 @@ namespace itk
  * with Multilevel B-Splines", IEEE Transactions on Visualization and
[…]

### 2009-10-02 Scripts/buildtemplateparallel.sh
sge comments
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@165 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -20,6 +20,8 @@ echo " We assume all files to be added to the template are in the current direct
 echo " script if you want to relax this assumption, but be sure that the qsubbed jobs get the correct "
 echo " absolute path to the images."
 echo 
+echo " We use the sun grid engine (SGE) to distribute computing via its qsub command. "
+echo 
 echo " Things within the script that you may need to change for your needs are highlighted by EDIT THIS "
 echo
 echo " The template will be written to [OutputRoot]template.nii. If the template file exists, it is used as the starting point for "

### 2009-11-24 Scripts/buildtemplateparallel.sh
minor bug in AverageImages initialization
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@210 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -65,7 +65,7 @@ IMAGESETVARIABLE=$*
 
 if [ ! -s $TEMPLATE ] ; then 
 echo " No initial template exists. Creating population average image"
- ${ANTSPATH}AverageImages $DIM $TEMPLATE  $IMAGESETVARIABLE 
+ ${ANTSPATH}AverageImages $DIM $TEMPLATE 1  $IMAGESETVARIABLE 
 fi
 
 echo  " ANTSPATH  $ANTSPATH "

### 2009-11-25 ImageRegistration/itkANTSImageRegistrationOptimizer.h
bug fix in evaluation of gradient in bspliner convergence
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@212 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -1056,7 +1056,7 @@ PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer mo
          * the desired window will start at the user-specified origin and 
          * end at the current iteration).
          */  
-	unsigned int domtar=10;
+	unsigned int domtar=6;
 	if( this->m_CurrentIteration > domtar )
           {            
           typedef BSplineScatteredDataPointSetToImageFilter
@@ -1112,11 +1112,11 @@ PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer mo
           bspliner->Update();
 	  
 	  ProfilePointType endPoint;
-	  //  endPoint[0] = static_cast<float>( this->m_CurrentIteration-1 );
-          endPoint[0] = 0.9 * static_cast<float>( this->m_CurrentIteration-1 );
+          endPoint[0] = static_cast<float>( this->m_CurrentIteration-domainsize*0.5 );
 	  typename BSplinerType::GradientType gradient;
+	  gradient.Fill(0);
 	  bspliner->EvaluateGradientAtPoint( endPoint, gradient ); 
-	  if (  gradient[0][0]  < 0.0001 && this->m_CurrentIteration > 9) converged=true;
+	  if (  gradient[0][0]  < 0.0001 && this->m_CurrentIteration > domtar) converged=true;
 	  std::cout << " E-Slope " << gradient[0][0] ;//<< std::endl;
 	  }
         for ( unsigned int qq=0; qq < this->m_Energy.size(); qq++ )

### 2009-11-25 ImageRegistration/itkANTSImageRegistrationOptimizer.h
bug fix in evaluation of gradient in bspliner convergence
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@213 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -1056,7 +1056,7 @@ PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer mo
          * the desired window will start at the user-specified origin and 
          * end at the current iteration).
          */  
-	unsigned int domtar=6;
+	unsigned int domtar=12;
 	if( this->m_CurrentIteration > domtar )
           {            
           typedef BSplineScatteredDataPointSetToImageFilter

### 2009-11-25 ImageRegistration/itkANTSImageRegistrationOptimizer.h
large-deformation option
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@214 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -664,7 +664,7 @@ PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer mo
     this->m_OutputNamingConvention=oOption->GetValue();   
 
     typename ParserType::OptionType::Pointer thicknessOption 
-      = this->m_Parser->GetOption( "compute-thickness" );  
+      = this->m_Parser->GetOption( "large-deformation" );  
     if( thicknessOption->GetValue() == "true" ||  thicknessOption->GetValue() == "1" ) { this->m_ComputeThickness=1; this->m_SyNFullTime=2; }// asymm forces
     else if(  thicknessOption->GetValue() == "2" )  { this->m_ComputeThickness=1; this->m_SyNFullTime=1; } // symmetric forces 
     else this->m_ComputeThickness=0; // not full time varying stuff 

### 2009-11-25 Documentation/Tex/antstheory.tex
final design for theory paper
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@215 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 \documentclass[10pt,twocolumn]{IEEEtran}
-
 \newcommand{\fv}{\mbox{\boldmath $f$}}
@@
 \newcommand{\half}{0.5}
-
 \newcommand{\X}{{\bf X}}
@@
 \newcommand{\aphiinv}{\bar{\phi}^{-1}}
-
 \newcommand{\domain}{\Omega}
@@
 \newcommand{\yv}{\mbox{\boldmath $y$}}
-
 \newcommand{\group}{Diff}
-
 \newcommand{\evol}{{E_{\text{Vol}}}}
@@
 \newcommand{\gz}{\nabla_{\z}}
-
 \newcommand{\Section}[1]{\vspace{-8pt}\section{\hskip-1em.~~#1}\vspace{-3pt}}
 \newcommand{\SubSection}[1]{\vspace{-3pt}\subsection{\hskip -1em.~~#1}\vspace{-3pt}}
-
 \newcommand{\BASE}{.}%/Users/stnava/Work/}
@@
 \newcommand{\FIGS}{.}%{\BASE Words/Writing/neuroimageipam-final/FIGS/}
-
-%% bare_jrnl.tex
-%% V1.2
-%% 2002/11/18
-%% by Michael Shell
-%% mshell@ece.gatech.edu
-%%
-%% NOTE: This text file uses MS Windows line feed conventions. When (human)
-%% reading this file on other platforms, you may have to use a text
-%% editor that can handle lines terminated by the MS Windows line feed
-%% characters (0x0D 0x0A).
-%%
-%% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.6b or later) with an IEEE journal paper.
-%%
-%% Support sites:
-%% http://www.ieee.org
-%% and/or
-%% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/IEEEtran/
-%%
-%% This code is offered as-is - no warranty - user assumes all risk.
-%% Free to use, distribute and modify.
-
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. IEEE's font choices can trigger bugs that do  ***
-% *** not appear when using other class files.                            ***
-% Testflow can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/testflow
-
-% Note that the a4paper option is mainly intended so that authors in
-% countries using A4 can easily print to A4 and see how their papers will
-% look in print. Authors are encouraged to use U.S. letter paper when 
-% submitting to IEEE. Use the testflow package mentioned above to verify
-% correct handling of both paper sizes by the author's LaTeX system.
-%
-% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
-% should be used if it is desired that the figures are to be displayed in
-% draft mode.
 %
@@
 \graphicspath{./Figures/}
-
-% If the IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it:
-% \documentclass[journal]{../sty/IEEEtran}
-
-% some very useful LaTeX packages include:
-
-%\usepackage{cite}      % Written by Donald Arseneau
-                        % V1.6 and later of IEEEtran pre-defines the format
-                        % of the cite.sty package \cite{} output to follow
-                        % that of IEEE. Loading the cite package will
-                        % result in citation numbers being automatically
-                        % sorted and properly "ranged". i.e.,
-                        % [1], [9], [2], [7], [5], [6]
-                        % (without using cite.sty)
-                        % will become:
-                        % [1], [2], [5]--[7], [9] (using cite.sty)
-                        % cite.sty's \cite will automatically add leading
-                        % space, if needed. Use cite.sty's noadjust option
-                        % (cite.sty V3.8 and later) if you want to turn this
-                        % off. cite.sty is already installed on most LaTeX
-                        % systems. The latest version can be obtained at:
-                        % http://www.ctan.org/tex-archive/macros/latex/contrib/supported/cite/
-
-%\usepackage{graphicx}  % Written by David Carlisle and Sebastian Rahtz
-                        % Required if you want graphics, photos, etc.
-                        % graphicx.sty is already installed on most LaTeX
-                        % systems. The latest version and documentation can
-                        % be obtained at:
-                        % http://www.ctan.org/tex-archive/macros/latex/required/graphics/
-                        % Another good source of documentation is "Using
-                        % Imported Graphics in LaTeX2e" by Keith Reckdahl
-                        % which can be found as esplatex.ps and epslatex.pdf
-                        % at: http://www.ctan.org/tex-archive/info/
-% NOTE: for dual use with latex and pdflatex, instead load graphicx like:
-%\ifx\pdfoutput\undefined
-%\usepackage{graphicx}
-%\else
-%\usepackage[pdftex]{graphicx}
-%\fi
-
-% However, be warned that pdflatex will require graphics to be in PDF
-% (not EPS) format and will preclude the use of PostScript based LaTeX
-% packages such as psfrag.sty and pstricks.sty. IEEE conferences typically
-% allow PDF graphics (and hence pdfLaTeX). However, IEEE journals do not
-% (yet) allow image formats other than EPS or TIFF. Therefore, authors of
-% journal papers should use traditional LaTeX with EPS graphics.
-%
-% The path(s) to the graphics files can also be declared: e.g.,
-% \graphicspath{{../eps/}{../ps/}}
-% if the graphics files are not located in the same directory as the
-% .tex file. This can be done in each branch of the conditional above
-% (after graphicx is loaded) to handle the EPS and PDF cases separately.
-% In this way, full path information will not have to be specified in
-% each \includegraphics command.
-%
-% Note that, when switching from latex to pdflatex and vice-versa, the new
-% compiler will have to be run twice to clear some warnings.
-
-%\usepackage{psfrag}    % Written by Craig Barratt, Michael C. Grant,
-                        % and David Carlisle
-                        % This package allows you to substitute LaTeX
-                        % commands for text in imported EPS graphic files.
-                        % In this way, LaTeX symbols can be placed into
-                        % graphics that have been generated by other
-                        % applications. You must use latex->dvips->ps2pdf
-                        % workflow (not direct pdf output from pdflatex) if
-                        % you wish to use this capability because it works
-                        % via some PostScript tricks. Alternatively, the
-                        % graphics could be processed as separate files via
-                        % psfrag and dvips, then converted to PDF for
-                        % inclusion in the main file which uses pdflatex.
-                        % Docs are in "The PSfrag System" by Michael C. Grant
-                        % and David Carlisle. There is also some information 
-                        % about using psfrag in "Using Imported Graphics in
-                        % LaTeX2e" by Keith Reckdahl which documents the
-                        % graphicx package (see above). The psfrag package
-                        % and documentation can be obtained at:
-                        % http://www.ctan.org/tex-archive/macros/latex/contrib/supported/psfrag/
-
-%\usepackage{subfigure} % Written by Steven Douglas Cochran
-                        % This package makes it easy to put subfigures
-                        % in your figures. i.e., "figure 1a and 1b"
-                        % Docs are in "Using Imported Graphics in LaTeX2e"
-                        % by Keith Reckdahl which also documents the graphicx
-                        % package (see above). subfigure.sty is already
-                        % installed on most LaTeX systems. The latest version
-                        % and documentation can be obtained at:
-                        % http://www.ctan.org/tex-archive/macros/latex/contrib/supported/subfigure/
-
-%\usepackage{url}       % Written by Donald Arseneau
-                        % Provides better support for handling and breaking
-                        % URLs. url.sty is already installed on most LaTeX
-                        % systems. The latest version can be obtained at:
-                        % http://www.ctan.org/tex-archive/macros/latex/contrib/other/misc/
-                        % Read the url.sty source comments for usage information.
-
-%\usepackage{stfloats}  % Written by Sigitas Tolusis
-                        % Gives LaTeX2e the ability to do double column
-                        % floats at the bottom of the page as well as the top.
-                        % (e.g., "\begin{figure*}[!b]" is not normally
-                        % possible in LaTeX2e). This is an invasive package
-                        % which rewrites many portions of the LaTeX2e output
-                        % routines. It may not work with other packages that
-                        % modify the LaTeX2e output routine and/or with other
-                        % versions of LaTeX. The latest version and
-                        % documentation can be obtained at:
-                        % http://www.ctan.org/tex-archive/macros/latex/contrib/supported/sttools/
-                        % Documentation is contained in the stfloats.sty
-                        % comments as well as in the presfull.pdf file.
-                        % Do not use the stfloats baselinefloat ability as
-                        % IEEE does not allow \baselineskip to stretch.
-                        % Authors submitting work to the IEEE should note
-                        % that IEEE rarely uses double column equations and
-                        % that authors should try to avoid such use.
-                        % Do not be tempted to use the cuted.sty or
-                        % midfloat.sty package (by the same author) as IEEE
-                        % does not format its papers in such ways.
-
-%\usepackage{amsmath}   % From the American Mathematical Society
-                        % A popular package that provides many helpful commands
-                        % for dealing with mathematics. Note that the AMSmath
-                        % package sets \interdisplaylinepenalty to 10000 thus
-                        % preventing page breaks from occurring within multiline
-                        % equations. Use:
-%\interdisplaylinepenalty=2500
-                        % after loading amsmath to restore such page breaks
-                        % as IEEEtran.cls normally does. amsmath.sty is already
-                        % installed on most LaTeX systems. The latest version
-                        % and documentation can be obtained at:
-                        % http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
-
-% Other popular packages for formatting tables and equations include:
-
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty which improves the
-% LaTeX2e array and tabular environments to provide better appearances and
-% additional user controls. array.sty is already installed on most systems.
-% The latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/tools/
-
-% Mark Wooding's extremely powerful MDW tools, especially mdwmath.sty and
-% mdwtab.sty which are used to format equations and tables, respectively.
-% The MDWtools set is already installed on most LaTeX systems. The lastest
-% version and documentation is available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/mdwtools/
-
-% V1.6 of IEEEtran contains the IEEEeqnarray family of commands that can
-% be used to generate multiline equations as well as matrices, tables, etc.
-
-% Also of notable interest:
-
-% Scott Pakin's eqparbox package for creating (automatically sized) equal
-% width boxes. Available:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/eqparbox/
-
-% Notes on hyperref:
-% IEEEtran.cls attempts to be compliant with the hyperref package, written
-% by Heiko Oberdiek and Sebastian Rahtz, which provides hyperlinks within
-% a document as well as an index for PDF files (produced via pdflatex).
-% However, it is a tad difficult to properly interface LaTeX classes and
-% packages with this (necessarily) complex and invasive package. It is
-% recommended that hyperref not be used for work that is to be submitted
-% to the IEEE. Users who wish to use hyperref *must* ensure that their
-% hyperref version is 6.72u or later *and* IEEEtran.cls is version 1.6b
-% or later. The latest version of hyperref can be obtained at:
-%
-% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/hyperref/
-%
-% Also, be aware that cite.sty (as of version 3.9, 11/2001) and hyperref.sty
-% (as of version 6.72t, 2002/07/25) do not work optimally together.
-% To mediate the differences between these two packages, IEEEtran.cls, as
-% of v1.6b, predefines a command that fools hyperref into thinking that
-% the natbib package is being used - causing it not to modify the existing
-% citation commands, and allowing cite.sty to operate as normal. However,
-% as a result, citation numbers will not be hyperlinked. Another side effect
-% of this approach is that the natbib.sty package will not properly load
-% under IEEEtran.cls. However, current versions of natbib are not capable
-% of compressing and sorting citation numbers in IEEE's style - so this
-% should not be an issue. If, for some strange reason, the user wants to
-% load natbib.sty under IEEEtran.cls, the following code must be placed
-% before natbib.sty can be loaded:
-%
-% \makeatletter
-% \let\NAT@parse\undefined
-% \makeatother
-%
-% Hyperref should be loaded differently depending on whether pdflatex
-% or traditional latex is being used:
-%
-%\ifx\pdfoutput\undefined
-%\usepackage[hypertex]{hyperref}
-%\else
-%\usepackage[pdftex,hypertexnames=false]{hyperref}
-%\fi
-%
-% Pdflatex produces superior hyperref results and is the recommended
-% compiler for such use.
-
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-
-% correct bad hyphenation here
 \hyphenation{op-tical net-works semi-conduc-tor}
-
-% The lineno packages adds line numbers. Start line numbering with
-% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
-% for the whole article with \linenumbers.
 \usepackage{lineno}
-
-%\linenumbers
 \begin{document}
-
-%Setup listing environment
 \definecolor{listcomment}{rgb}{0.0,0.5,0.0}
@@
 \definecolor{listbackground}{gray}{0.965}
-
 \lstset{frame = tb,
@@
 % paper title
-\title{ANTS:  Advanced Open-Source Tools for Normalization And Neuroanatomy}
+\title{ANTS: Open-Source Tools for Normalization And Neuroanatomy}
 %
@@
 lack of standards and reproducibility in the field that is due, in
-part, to the use of proprietary software and private data that is difficult 
-to openly evaluate.  To
-facilitate reproducibility in CA measurements and advancement of
-imaging sciences, NIH has recently committed significant support to
-open-source data and software resources.  Here, we report a recent
-product of this commitment: Advanced Normalization Tools (ANTS), an
-ITK-based toolkit for CA and related areas.  The ANTS open-source
-library consists of a well-evaluated suite of state-of-the-art
-normalization and template-building tools for quantitative
-morphometric analysis.  We highlight the prominent features of ANTS
-and demonstrate its utility by performing a detailed analysis on
-openly-available anatomically labeled brain data from the non-rigid
-image registration evaluation project (NIREP).  The results from this
-analysis evidences the high level of accuracy achievable with ANTS
-using intensity-based registration alone.  In addition, we show the
-significant performance gains may be achieved by coupling
-intensity-based image metrics and point-set-based metrics from
-specific, sensibly selected cortical structures.
-%Computational anatomy is the quantitative study of biological shape and its natural variations which traces its roots to the seminal works of Charles Darwin and D'arcy Thompson.
-Integral to this research field is the calculation of deformable mappings between images that serve as invaluable measurement tools.
-Technological evolution has spawned much innovation in the computational development of such deformable mapping algorithms.
-However, not only is algorithmic development essential for increasing scientific knowledge but the availability of these technologies is equally crucial for facilitating the progress of the scientific community at large.
-This is the motivation behind such projects as the Insight Toolkit (ITK) of the National Institutes of Health which serves as a repository for many popular image segmentation and registration routines.
-In addition, the ITK library provides a well-structured framework for several stand-alone projects such as Elastix, ITK-SNAP, ManagedITK, the Medical Imaging Toolkit (MITK), WrapITK, and many others.
-Our own  contribution to the field of computational anatomy, also built on an ITK foundation, is called Advanced Normalization Tools (ANTS).
-The ANTS open-source library consists of an entire suite of state-of-the-art normalization and template-building tools for quantitative morphometric analysis.
-After highlighting the prominent features of ANTS, we demonstrate its utility by performing a detailed analysis on anatomically labeled brain data from the non-rigid image registration evaluation project (NIREP).
-The results from this analysis evidences the high level of accuracy achievable with ANTS using intensity-based registration alone.
-In addition, we show the performance gains achieved by coupling intensity-based image metrics and point-set-based metrics from specific, sensibly selected cortical structures.
-
+part, to the use of proprietary software and private data that is
+difficult to openly evaluate.  To facilitate reproducibility in CA
+measurements and advancement of imaging sciences, NIH has recently
+committed significant support to open-source data and software
+resources.  Here, we report a recent product of this commitment:
+Advanced Normalization Tools (ANTS), an ITK-based toolkit for CA and
+related areas.  The ANTS open-source library consists of a
+well-evaluated suite of state-of-the-art normalization, segmentation
+and template-building tools for quantitative morphometric analysis.
+We highlight the prominent features of ANTS, including diffeomorphic
+normalization methods, and demonstrate its utility by performing a
+detailed analysis on openly-available anatomically labeled brain data
+from the non-rigid image registration evaluation project (NIREP).  The
+results from this analysis evidences the high level of accuracy
+achievable with ANTS using intensity-based registration and
+segmentation.  In addition, we show the significant performance gains
+may be achieved by coupling intensity-based image metrics and
+label-based metrics from specific, sensibly selected cortical
+structures.  Additional features are highlighted in the appendix. 
 %In a recent evaluation of 14 non-linear deformation algorithms \cite{Klein2009}, the SyN algorithm of ANTS was consistently one of the top performers.
-{\bf FIXME -- make clear what transformations we are comparing with constant similarity ... }
+%{\bf FIXME -- make clear what transformations we are comparing with constant similarity ... }
 \end{abstract}
-
 %\IEEEkeywords{
@@
 %}
-
 %\IEEEpeerreviewmaketitle
@@
 \section{Introduction}
-
 The rapid advancement of biological and medical imaging technologies
@@
 also suggest that the layout of cell types throughout the brain is
-generally preserved \cite{Schleicher2009}, further motivating diffeomorphic mapping
-in the context of the brain.
-
-Despite the number of proposed algorithms, our limited assessment of published research mirrors the experience of many others who prefer a working paradigm of what has been referred to as {\em reproducible research}.
-As described by Dr.
-Kovacevic ``[reproducible research] refers to the idea that, in `computational' sciences, the ultimate product is not a published paper but, rather, the entire environment used to produce the results in the paper (data, software,etc.).''  After an informal survey of 15 published papers, she found ``none had code available'' and ``in only about half the cases were the parameters [of the algorithm] specified'' \cite{Kovacevic2006}.
-Recent discussions within the computational sciences research community, particularly among advocates of ``open science,'' have also voiced similar concerns \cite{Yoo2005,Ibanez2006}.
-In this paper, we discuss our contribution to the open-source medical image analysis research community which we call ANTS (\underline{A}dvanced \underline{N}ormalization \underline{T}ool\underline{s}).
-Built on an ITK framework, this software package comprises a suite of tools for image normalization and template building based on previously published research.
-
-Perhaps the most persuasive evidence motivating the use of our contributions discussed in this paper is the recent outcome of a large-scale comparative image registration algorithm assessment \cite{Klein2009}.
-In the largest evaluation study to date involving 14 popular non-linear registration algorithms, our Symmetric Normalization (SyN) transformation model \cite{Avants2008} discussed below, was consistently one of the top two performers across all tests.
-Overlap and distance measures used for assessment employed three completely independent analysis methods (permutation tests, one-way ANOVA tests, and indifference zone ranking).
-Unlike some of the other algorithms used in this brain registration evaluation study, all of our methods (not just SyN) are available as open-source.
+generally preserved \cite{Schleicher2009}, further motivating
+diffeomorphic mapping in the context of the brain.
+
+Despite the number of proposed algorithms, our limited assessment of
+published research mirrors the experience of many others who prefer a
+working paradigm of what has been referred to as {\em reproducible
+  research}.  As described by Dr. Kovacevic ``[reproducible research]
+refers to the idea that, in `computational' sciences, the ultimate
+product is not a published paper but, rather, the entire environment
+used to produce the results in the paper (data, software,etc.).''
+After an informal survey of 15 published papers, she found ``none had
+code available'' and ``in only about half the cases were the
+parameters [of the algorithm] specified'' \cite{Kovacevic2006}.
+Recent discussions within the computational sciences research
+community, particularly among advocates of ``open science,'' have also
+voiced similar concerns \cite{Yoo2005,Ibanez2006}.  In this paper, we
+discuss our contribution to the open-source medical image analysis
+research community which we call ANTS (\underline{A}dvanced
+\underline{N}ormalization \underline{T}ool\underline{s}).  Built on an
+ITK framework, this software package comprises a suite of tools for
+image normalization and template building based on previously
+published research.
+
+Perhaps the most persuasive evidence motivating the use of our
+contributions discussed in this paper is the recent outcome of a
+large-scale comparative image registration algorithm assessment
+\cite{Klein2009}. In the largest evaluation study to date involving 14
+popular non-linear registration algorithms, our Symmetric
+Normalization (SyN) transformation model \cite{Avants2008} discussed
+below, was consistently one of the top two performers across all
+tests.  Overlap and distance measures used for assessment employed
+three completely independent analysis methods (permutation tests,
+one-way ANOVA tests, and indifference zone ranking).  Unlike some of
+the other algorithms used in this brain registration evaluation study,
+all of our methods (not just SyN) are available as open-source.
 
@@
 normalization evaluation of \cite{Klein2009} which was limited to a
-single configuration of ANTS.  This will provide a useful benchmark
-for evaluating other possible ANTS configurations.  Importantly, this
-work also suggests---to our knowledge, for the first time---the clear
+single configuration of ANTS.  A set of ANTS tools are evaluated on
+the sixteen subjects from the public, expert-labeled neuroanatomy in
+the NIREP dataset.  This will provide a useful benchmark for
+evaluating other possible ANTS configurations.  Importantly, this work
+also suggests---to our knowledge, for the first time---the clear
 practical benefit of time-parameterized diffeomorphic mapping over the
@@
 
+The paper is organized as follows.  Section~\ref{sec:theory} gives an
+overview of the transformation models and similarity metrics available
+in ANTS, along with optimization schemes.  Section~\ref{sec:imp}
+connects the theory with the user interface.  Section~\ref{sec:expt},
+first, illustrates the performance of different transformation models
+on a pair of ``classic'' image registration examples.  Second, we
+report results on a series of large-scale experiments that profile
+ANTS registration schema against a cortical labeling problem using the
+NIREP evaluation dataset.  Finally, we close with discussion of the
+toolkit and our findings.
+
 \section{Theoretical Overview of ANTS}
-A useful classification schema of normalization techniques is based upon the following three principal components \cite{Brown1992,Ibanez2002}:
+\label{sec:theory}
+A useful classification schema of normalization techniques is based
+upon the following three principal components
+\cite{Brown1992,Ibanez2002}:
 \begin{itemize}
@@
 \end{itemize}
-In general, image normalization is the process of finding the optimal transformation, $\phi$, within a specified transformation space which maps each $\mathbf{x}$ of image $\mathcal{I}(\mathbf{x})$ to a location in image $\mathcal{J}(\mathbf{z})$ such that a specified cost function, $\mathcal{C}$, describing the similarity between $\mathcal{I}$ and $\mathcal{J}$, is minimized.
-A summary of available transformation models and similarity measures are provided in Table \ref{table:chart}.
-Details are given in subsequent sections.
+In general, image normalization is the process of finding the optimal
+transformation, $\phi$, within a specified transformation space which
+maps each $\mathbf{x}$ of image $\mathcal{I}(\mathbf{x})$ to a
+location in image $\mathcal{J}(\mathbf{z})$ such that a specified cost
+function, $\mathcal{C}$, describing the similarity between
+$\mathcal{I}$ and $\mathcal{J}$, is minimized.  A summary of available
+transformation models and similarity measures are provided in Table
+\ref{table:chart}.  Details are given in subsequent sections.
 
 \subsection{ANTS Transformation Models}
-
 %\begin{figure*}
@@
 %\end{figure*}
-
 \begin{table*}
@@
 
-A variety of transformation models have been proposed in the literature with varying degrees of freedom (illustrated in Figure \ref{fig:hierarchy}).
-For deformable transformations, one approach is to optimize within the space of non-topology preserving, yet physically motivated transformations---an approach pioneered by Bajcsy \cite{Bajcsy1989}.
-Elastic and related models, such as HAMMER \cite{Shen2002}, statistical parametric mapping (SPM) \cite{Ashburner2000}, free-form deformations (FFD) \cite{Rueckert1999}, and
-Thirion's demons \cite{Thirion1998} operate in the space of vector fields, which does not preserve topology.
-In other words, barring ad hoc constraints to prevent otherwise, these algorithms may allow the topology to change in an uncontrolled way which makes the deformable mappings difficult to interpret in functional or anatomical studies.
+A variety of transformation models have been proposed in the
+literature with varying degrees of freedom (illustrated in Figure
+\ref{fig:hierarchy}).  For deformable transformations, one approach is
+to optimize within the space of non-topology preserving, yet
+physically motivated transformations---an approach pioneered by Bajcsy
+\cite{Bajcsy1989}.  Elastic and related models, such as HAMMER
+\cite{Shen2002}, statistical parametric mapping (SPM)
+\cite{Ashburner2000}, free-form deformations (FFD)
+\cite{Rueckert1999}, and Thirion's demons \cite{Thirion1998} operate
+in the space of vector fields, which does not preserve topology.  In
+other words, barring ad hoc constraints to prevent otherwise, these
+algorithms may allow the topology to change in an uncontrolled way
+which makes the deformable mappings difficult to interpret in
+functional or anatomical studies.
 
@@
 comparative study of nonrigid registration algorithms
-\cite{Klein2009}.  For this reason, we also include elastic-related
-methods for a panoply of transformation model options.
-
+\cite{Klein2009}.  For this reason, ANTS also includes elastic-related
+methods for a panoply of transformation model options.  As we show
+below, all of these methods may be implemented, when using gradient
+descent optimization, in a generic framework that generates
+transformations (when necessary) and updates transformations according
+to the needs of each model.  ANTS implementation follows this general form 
+for each model. 
 \begin{figure}
@@
 \end{figure}
-
 \subsubsection{Rigid and Affine Linear Transformations }
-
-Image registration strategies often begin with a linear transformation for initial global alignment which precedes a deformable transformation with increased degrees of freedom.
-The linear transformations available within ANTS optimize either a mean-squared difference (MSQ) or mutual information (MI) similarity metric which are optimized with respect to translation, rotation, and, in the case of affine transformations, scaling and shearing.
-The successive optimization of each component of the linear transformation allows for careful control over increasing degrees of freedom.
-ANTS also composes the affine transformation with the deformable transformation field before performing any interpolation or downsampling.
-In this way, ANTS normalization never requires more than a single image interpolation step and is able to always refer back to the original full-resolution image sources.
-%We have implemented quaternion-based rigid and affine transformations within ANTS.
-Not only does this facilitate accessibility of a sound global alignment strategy, but the parameters of the calculated optimal linear transformation can be maintained internally to ANTS to avoid unnecessary image interpolations.
-Either a mean-squared difference metric or a mutual information metric can be used for the linear alignment.
-
-\subsubsection{Vector Field Operators for Regularization} Antecedent to discussion of available deformable transformation models, we point out that  deformable normalization strategies typically invoke a deformation regularization step which smooths the displacement field, $\displace$, or velocity field, $\mathbf{v}$,  or both by a linear operator such as the Laplacian or Navier-Stokes operator.
-One may write this regularization as a variational minimization in terms of its linear operator or in
-terms of a kernel function operating on the field itself, e.g., $
-\displace_{smooth}= K \star \displace_{not \,\, smooth}$, where $K \star$ denotes 
-convolution with the Green's kernel, $K$, for the linear operator, $L$.
-
-Regularization models operate on either the
-whole mapping $\phi$ or the gradient of the similarity term or both.
-The same regularization schema is available for both diffeomorphic and the recently 
-proposed directly manipulated free-form deformation
-(DMFFD) \cite{Tustison2009}, allowing regularization of both total 
-deformation and deformation update.  Viewed from this perspective, hybrid configurations incorporating discretized FFD strategies and diffeomorphisms can be combined for novel 
-image normalization approaches.
-
-ANTS enables a variety of choices for $K$
-including the Gaussian with varying $\sigma$ and a variety of B-spline
-functions, both of which induce adequate regularity for normalization
-models used in ANTS.  While additional physical operators will be
-included in future releases, current B-spline options provide many orders 
-of flexibility \cite{Tustison2005}.
+Image registration strategies often begin with a linear transformation
+for initial global alignment which precedes a deformable
+transformation with increased degrees of freedom.  The linear
+transformations available within ANTS optimize either a mean-squared
+difference (MSQ) or mutual information (MI) similarity metric which
+are optimized with respect to translation, rotation, and, in the case
+of affine transformations, scaling and shearing.  The successive
+optimization of each component of the linear transformation allows for
+careful control over increasing degrees of freedom.  ANTS also
+composes the affine transformation with the deformable transformation
+field before performing any interpolation or downsampling.  In this
+way, ANTS normalization never requires more than a single image
+interpolation step and is able to always refer back to the original
+full-resolution image sources.  The ANTS implementation of rigid
+mapping is quaternion-based with additional scaling and shearing terms
+when affine mapping is desired (via the {\verb --rigid-affine } ~{\verb true } ~flag).
+
+\subsubsection{Vector Field Operators for Regularization} 
+Deformable normalization strategies typically invoke a deformation
+regularization step which smooths the displacement field, $\displace$,
+or velocity field, $\mathbf{v}$, or both by a linear operator such as
+the Laplacian or Navier-Stokes operator. One may write this
+regularization as a variational minimization in terms of its linear
+operator or in terms of a kernel function operating on the field
+itself, e.g., $ \displace_{smooth}= K \star \displace_{not \,\,
+  smooth}$, where $K \star$ denotes convolution with the Green's
+kernel, $K$, for the linear operator, $L$.  Regularization models
+operate on either the whole mapping $\phi$ or the gradient of the
+similarity term or both.  The same regularization schema is available
+for both diffeomorphic and the recently proposed directly manipulated
+free-form deformation (DMFFD) \cite{Tustison2009}, allowing
+regularization of both total deformation and deformation update.
+Viewed from this perspective, hybrid configurations incorporating
+discretized FFD strategies and diffeomorphisms can be combined for
+novel image normalization approaches.  ANTS enables a variety of
+choices for $K$ including the Gaussian with varying $\sigma$ and a
+variety of B-spline functions, both of which induce adequate
+regularity for normalization models used in ANTS.  While additional
+physical operators will be included in future releases, current
+B-spline options provide many orders of flexibility
+\cite{Tustison2005}.
 
@@
 flexible, linear penalty on deformation while guaranteeing topology
-preservation.  Additionally, distance metrics in the space of diffeomorphisms 
-allow geodesic properties to be explored \cite{Miller2001,Beg2005}.
-
-ANTS assumes the diffeomorphism, $\phi$, is defined on the image domain, $\Omega$, and maintains an affine transform at the boundary such that $\phi(\partial \Omega) = A(\mathbf{Id})$ where $A(\mathbf{Id})$ is an affine mapping applied to the identity transformation.
-$\phi$, over time, parameterizes a family of diffeomorphisms, $\phi(\mathbf{x}, t) : \Omega \times t \rightarrow  \Omega$, which can be generated by integrating a time-dependent, smooth velocity field, $\boldsymbol{v} : \Omega \times t \rightarrow \mathbb{R}^d$,  through the ordinary differential equation (o.d.e.)
+preservation.  Additionally, distance metrics in the space of
+diffeomorphisms allow geodesic properties to be explored
+\cite{Miller2001,Beg2005}.
+
+ANTS assumes the diffeomorphism, $\phi$, is defined on the image
+domain, $\Omega$, and maintains an affine transform at the boundary
+such that $\phi(\partial \Omega) = A(\mathbf{Id})$ where
+$A(\mathbf{Id})$ is an affine mapping applied to the identity
+transformation.  The map $\phi$, over time, parameterizes a family of
+diffeomorphisms, $\phi(\mathbf{x}, t) : \Omega \times t \rightarrow
+\Omega$, which can be generated by integrating a (potentially) time-dependent,
+smooth velocity field, $\boldsymbol{v} : \Omega \times t \rightarrow
+\mathbb{R}^d$, through the ordinary differential equation (o.d.e.)
 \begin{align} \label{eq:ode}
-\frac{d \phi(\mathbf{x}, t)}{dt} = \boldsymbol{v}(\phi(\mathbf{x}, t), t), \,\,\, \phi(\mathbf{x}, 0) = \mathbf{x}.
-\end{align}
-The existence and uniqueness theorem for o.d.e.'s implies that integrating Equation (\ref{eq:ode}) generates a diffeomorphism.
-The deformation field yielded by $\phi$ is $\mathbf{u}(\mathbf{x}) = \phi(\mathbf{x},1) - \mathbf{x}$.
-
-Dupuis et al. \cite{Dupuis1998} motivated the usage of diffeomorphisms for CA  by showing that the variational form
+\frac{d \phi(\mathbf{x}, t)}{dt} = \boldsymbol{v}(\phi(\mathbf{x}, t),
+t), \,\,\, \phi(\mathbf{x}, 0) = \mathbf{x}.
+\end{align}
+The existence and uniqueness theorem for o.d.e.'s implies that
+integrating Equation (\ref{eq:ode}) generates a diffeomorphism.  The
+deformation field yielded by $\phi$ is $\mathbf{u}(\mathbf{x}) =
+\phi(\mathbf{x},1) - \mathbf{x}$.
+
+Dupuis et al. \cite{Dupuis1998} motivated the usage of diffeomorphisms
+for CA by showing that the variational form
 \begin{align}\label{eq:D}
-  D(\mathcal{I}, \mathcal{J}) =  \int_0^1 ||L\boldsymbol{v}|| dt , \,\, \mathcal{I}(\phi(\mathbf{x}, 1)) = \mathcal{J}(\mathbf{z})
-\end{align}
-represents a true mathematical metric between anatomical instances $\mathcal{I}$ and $\mathcal{J}$ given an appropriate norm, $L$, on the velocity field, $\boldsymbol{v}$.
-An optimal solution, $\boldsymbol{v}^*$, minimizes the metric $D(\mathcal{I}, \mathcal{J})$ with respect to $L$.
-Dupuis \cite{Dupuis1998} also showed that such a solution is guaranteed to be well-posed.
-Intuitively, Equation (\ref{eq:D}) provides a sense of distance between two anatomical shapes.
-It also illustrates that the optimal diffeomorphic solution is analogous to finding the geodesic curve between two points in a curved space.%
-\footnote{
-It is important to note the similarity between the definition of curve length, $\int ||\mathcal{C}'(t)||dt$, for the parametric curve $\mathcal{C}(t)$ and Equation (\ref{eq:D}).
-In this sense, the solution for Equation (\ref{eq:D}) is the geodesic diffeomorphism, where $\boldsymbol{v}$ is the tangent vector of the diffeomorphism, such that the shape distance, $D$, between $\mathcal{I}$ and $\mathcal{J}$ is minimized.
-}
-
-In most real-world applications, however, a diffeomorphic path connecting the anatomical instance $\mathcal{J}$ with $\mathcal{I}$ is non-existent (due, for example, to the photometric variation or the presence/absence of a tumor in neuroanatomical images).
-Therefore, the following minimizing variational form is used for optimization in diffeomorphic normalization to accommodate inexact matching  \cite{Dupuis1998,Miller2002}
+  D(\mathcal{I}, \mathcal{J}) = \int_0^1 ||L\boldsymbol{v}|| dt , \,\,
+  \mathcal{I}(\phi(\mathbf{x}, 1)) = \mathcal{J}(\mathbf{z})
+\end{align}
+represents a true mathematical metric between anatomical instances
+$\mathcal{I}$ and $\mathcal{J}$ given an appropriate norm, $L$, on the
+velocity field, $\boldsymbol{v}$.  An optimal solution,
+$\boldsymbol{v}^*$, minimizes the metric $D(\mathcal{I}, \mathcal{J})$
+with respect to $L$.  Dupuis \cite{Dupuis1998} also showed that such a
+solution is guaranteed to be well-posed.  Intuitively, Equation
+(\ref{eq:D}) provides a sense of distance between two anatomical
+shapes.  It also illustrates that the optimal diffeomorphic solution
+is analogous to finding the geodesic curve between two points in a
+curved space.
+\footnote{ It is important to note the similarity between the
+  definition of curve length, $\int ||\mathcal{C}'(t)||dt$, for the
+  parametric curve $\mathcal{C}(t)$ and Equation (\ref{eq:D}).  In
+  this sense, the solution for Equation (\ref{eq:D}) is the geodesic
+  diffeomorphism, where $\boldsymbol{v}$ is the tangent vector of the
+  diffeomorphism, such that the shape distance, $D$, between
+  $\mathcal{I}$ and $\mathcal{J}$ is minimized.  }
+
+In most real-world applications, however, a diffeomorphic path
+connecting the anatomical instance $\mathcal{J}$ with $\mathcal{I}$ is
+non-existent (due, for example, to photometric variation, idiosyncratic cortical folding or the
+presence/absence of a tumor in neuroanatomical images).  Therefore,
+the following minimizing variational form is used for optimization in
+diffeomorphic normalization to accommodate inexact matching
+\cite{Dupuis1998,Miller2002}
 \begin{align} \label{eq:lddmm}
-  \boldsymbol{v}^* = \underset{\boldsymbol{v}}{\operatorname{argmin}}  \left\{ \int_0^1  ||L\boldsymbol{v}||^2dt  + \lambda\int_{\Omega} || \mathcal{I} \circ \phi(\mathbf{x},1) - \mathcal{J}  || d\Omega \right\}.
-\end{align}
-The Euler-Lagrange equations characterizing the optimizing time-varying velocity field, $\boldsymbol{v}^*$, were derived in \cite{Miller2002} and later used in formulating the gradient-descent optimization scheme known as {\em large deformation diffeomorphic metric-matching} (LDDMM) \cite{Beg2005} with the similarity metric, or data term, for LDDMM being the squared intensity difference with weight $\lambda$.
-
-To accommodate a variety of medical image normalization tasks, one typically encounters more complex intensity transfers between one anatomical instance $\mathcal{J}$ and another instance $\mathcal{I}$.
-Thus,
-ANTS enables not only a variety of similarity metric possibilities beyond the conventional squared difference
-metric but it also permits any number of different similarity metrics for a particular image normalization task.  
-This leads to the following generalization of  Equation (\ref{eq:lddmm}):
+  \boldsymbol{v}^* = \underset{\boldsymbol{v}}{\operatorname{argmin}}
+  \left\{ \int_0^1 ||L\boldsymbol{v}||^2dt + \lambda\int_{\Omega} ||
+  \mathcal{I} \circ \phi(\mathbf{x},1) - \mathcal{J} || d\Omega
+  \right\}.
+\end{align}
+The Euler-Lagrange equations characterizing the optimizing
+time-varying velocity field, $\boldsymbol{v}^*$, were derived in
+\cite{Miller2002} and later used in formulating the gradient-descent
+optimization scheme known as {\em large deformation diffeomorphic
+  metric-matching} (LDDMM) \cite{Beg2005} with the similarity metric,
+or data term, for LDDMM being the squared intensity difference with
+weight $\lambda$.
+
+To accommodate a variety of medical image normalization tasks, one
+typically encounters more complex intensity transfers between one
+anatomical instance $\mathcal{J}$ and another instance $\mathcal{I}$.
+Thus, ANTS enables not only a variety of similarity metric
+possibilities beyond the conventional squared difference metric but it
+also permits any number of different similarity metrics for a
+particular image normalization task.  This leads to the following
+generalization of Equation (\ref{eq:lddmm}):
 \begin{align} \label{eq:diff}
-  \boldsymbol{v}^* = \underset{\boldsymbol{v}}{\operatorname{argmin}}  \left\{ \int_0^1  ||L\boldsymbol{v}||^2 dt  +\lambda\int_{\Omega}\Pi_{\sim}( \mathcal{I}, \phi(\mathbf{x},1) , \mathcal{J} )d\Omega  \right\}
-\end{align}
-where $\Pi_{\sim}$ is a similarity metric depending on the images and the
-mapping and $\lambda$ controls the degree of exactness in the
-matching.  If $\Pi_{\sim}$ is selected as cross-correlation, then one is
-estimating the diffeomorphism under more robust illumination
+  \boldsymbol{v}^* = \underset{\boldsymbol{v}}{\operatorname{argmin}}
+  \left\{ \int_0^1 ||L\boldsymbol{v}||^2 dt
+  +\lambda\int_{\Omega}\Pi_{\sim}( \mathcal{I}, \phi(\mathbf{x},1) ,
+  \mathcal{J} )d\Omega \right\}
+\end{align}
+where $\Pi_{\sim}$ is a similarity metric depending on the images and
+the mapping and $\lambda$ controls the degree of exactness in the
+matching.  If $\Pi_{\sim}$ is selected as cross-correlation, then one
+is estimating the diffeomorphism under more robust illumination
 constraints, as described in \cite{Avants2008}.
-%A similar equation for elastic/FFD/DMFFD matching may be minimized in ANTS.
-
-Exploiting the fact that the diffeomorphism, $\phi$, can be decomposed into two components 
-$\phi_1$ and $\phi_2$, one may construct a {\em symmetric} alternative to Equation (\ref{eq:diff}).  
-Now define, in $t \in [0,0.5]$, $\boldsymbol{v}(x,t) = \boldsymbol{v}_1(x,t) $ and 
+
+Exploiting the fact that the diffeomorphism, $\phi$, can be decomposed
+into two components $\phi_1$ and $\phi_2$, one may construct a {\em
+  symmetric} alternative to Equation (\ref{eq:diff}).  Now define, in
+$t \in [0,0.5]$, $\boldsymbol{v}(x,t) = \boldsymbol{v}_1(x,t) $ and
 $\boldsymbol{v}(x,t) = \boldsymbol{v}_2(x,1-t)$ when $t \in [0.5,1]$.
@@
   \{ \boldsymbol{v}^{*}_1 , \boldsymbol{v}^{*}_2 \}= &
-     \underset{\boldsymbol{v}_{1,2}}{\operatorname{argmin}}  
-      \Bigg\{ \nonumber \\
-      & \int_0^{0.5} ||L\boldsymbol{v}_1(x,t)||^2 ~dt +  \int_0^{0.5} ||L\boldsymbol{v}_2(x,t)||^2 ~dt  \nonumber \\
-     &+ \lambda\int_{\Omega}\Pi_{\sim}\left( \mathcal{I} \circ \phi_1(\mathbf{x},0.5), \mathcal{J}\circ \phi_2(\mathbf{x},0.5)\right)d\Omega\Bigg\}.
-\end{align}
-Note that the regularization term, here, is equivalent to that in equation~\ref{eq:lddmm}.  The only change is the splitting of the integral into 
-two time intervals reflecting the underlying optimized components of the velocity field. 
-The corresponding symmetric Euler-Lagrange equations are similar to \cite{Miller2002}.
-The difference, here, is that in finding $\boldsymbol{v}^*$, we minimize the variational energy from either end-point towards the mid-point of the transformation, as indicated by the data term.
-This strategy ``splits'' the optimization dependence equally between both images.
-Thus, gradient-based iterative convergence deforms $\mathcal{I}$ and $\mathcal{J}$ along the geodesic diffeomorphism, $\phi$, to a fixed  point midway (intuited by the notion of shape distance) between $\mathcal{I}$ and $\mathcal{J}$ thus motivating the denotation of the solution strategy as Symmetric Normalization (SyN).
-
-Other diffeomorphic flavors have since been reported in the research literature (e.g.
-DARTEL \cite{Ashburner2007} and Diffeomorphic Demons \cite{Vercauteren2007,Vercauteren2009}.
-We include three diffeomorphic transformation models for parameterizing $\phi(\cdot)$.
-These include Geodesic SyN, Greedy SyN, and exponential mapping.
-As summarized in Table \ref{table:chart}, each of these transformation models can utilize a host of similarity measures both individually and in mutual combination.
+  \underset{\boldsymbol{v}_{1,2}}{\operatorname{argmin}} \Bigg\{
+  \nonumber \\ & \int_0^{0.5} ||L\boldsymbol{v}_1(x,t)||^2 ~dt +
+  \int_0^{0.5} ||L\boldsymbol{v}_2(x,t)||^2 ~dt \nonumber \\ &+
+  \lambda\int_{\Omega}\Pi_{\sim}\left( \mathcal{I} \circ
+  \phi_1(\mathbf{x},0.5), \mathcal{J}\circ
+  \phi_2(\mathbf{x},0.5)\right)d\Omega\Bigg\}.
+\end{align}
+Note that the regularization term, here, is equivalent to that in
+equation~\ref{eq:lddmm}.  The only change is the splitting of the
+integral into two time intervals reflecting the underlying optimized
+components of the velocity field.  The corresponding symmetric
+Euler-Lagrange equations are similar to \cite{Miller2002}.  The
+difference, here, is that in finding $\boldsymbol{v}^*$, we minimize
+the variational energy from either end-point towards the mid-point of
+the transformation, as indicated by the data term.  This strategy
+``splits'' the optimization dependence equally between both images.
+Thus, gradient-based iterative convergence deforms $\mathcal{I}$ and
+$\mathcal{J}$ along the geodesic diffeomorphism, $\phi$, to a fixed
+point midway (intuited by the notion of shape distance) between
+$\mathcal{I}$ and $\mathcal{J}$ thus motivating the denotation of the
+solution strategy as Symmetric Normalization (SyN).
+
+Other diffeomorphic algorithms have since been reported in the
+research literature e.g. DARTEL \cite{Ashburner2007} and Diffeomorphic
+Demons \cite{Vercauteren2007,Vercauteren2009}, both of which use the
+constant velocity, exponential model for generating diffeomorphisms.
+We thus include three diffeomorphic transformation models for
+parameterizing $\phi(\cdot)$.  These include Geodesic SyN, Greedy SyN,
+and exponential mapping.  As summarized in Table \ref{table:chart},
+each of these transformation models can utilize a host of similarity
+measures both individually and in mutual combination.
 
@@
 (\ref{eq:diffs}) first requires a specified discretization of $t$,
-s.t. $t \in [0,1] \approx \{ 0 , 1/k , 2/k , \cdots , 1 \} $ where integer $k > 1$
-is the desired number of discretized intervals.  Calculation of the
-gradient of $\Pi$ with respect to the diffeomorphisms $\phi_1$ and
-$\phi_2$ is performed at each of these $t$ values, $t_k$, 
+s.t. $t \in [0,1] \approx \{ 0 , 1/k , 2/k , \cdots , 1 \} $ where
+integer $k > 1$ is the desired number of discretized intervals.
+Calculation of the gradient of $\Pi$ with respect to the
+diffeomorphisms $\phi_1$ and $\phi_2$ is performed at each of these
+$t$ values, $t_k$,
 \begin{align}
 \label{eq:grad}
-  \nabla \Pi_i(\boldsymbol{x},t_k) = \frac{\partial}{\partial \phi_i} \Pi_{\sim}( \mathcal{I}(\phi_{1}^{-1}(\boldsymbol{x},t_k)), \mathcal{J}(\phi_{2}^{-1}(\boldsymbol{x},1-t_k)) 
-\end{align}
-for $i \in \{1,2\}$.   The total velocity field is then updated from the previous iteration according to, 
-for each $i$, 
+  \nabla \Pi_i(\boldsymbol{x},t_k) = \frac{\partial}{\partial \phi_i}
+  \Pi_{\sim}( \mathcal{I}(\phi_{1}^{-1}(\boldsymbol{x},t_k)),
+  \mathcal{J}(\phi_{2}^{-1}(\boldsymbol{x},1-t_k))
+\end{align}
+for $i \in \{1,2\}$.  The total velocity field is then updated from
+the previous iteration according to, for each $i$,
 \begin{align}
-  \boldsymbol{v}(\mathbf{x},t) = \boldsymbol{v}(\mathbf{x},t) + \delta K \star \nabla \Pi_i(\mathbf{x},t),
-\end{align}
-where $\delta$ is the user-specified gradient descent parameter.  We
-then integrate Equation (\ref{eq:ode}) using Runge-Kutta methods to
-yield $\phi_i(\mathbf{x},t)$ for each $t \in [0,1]$ and $i \in
-\{1,2\}$.  We cycle through these steps until convergence or iterative
-exhaustion.
+  \boldsymbol{v}(\mathbf{x},t) = \boldsymbol{v}(\mathbf{x},t) + \delta
+  K \star \nabla \Pi_i(\mathbf{x},t),
+\end{align}
+where $\delta$ is the user-specified gradient descent parameter.  Note
+that the update, here, is a $N+1$-dimensional vector field and $K$ is
+a $N+1$-dimensional operator, when the images are $N$ dimensions.  We
+generate $\phi_i(\mathbf{x},t)$ for each $t \in [0,1]$ and $i \in
+\{1,2\}$ by integrating Equation (\ref{eq:ode}) using Runge-Kutta methods.  
+We cycle through these steps until convergence or iterative exhaustion.
 
@@
 
-Greedy optimization of Equation (\ref{eq:diffs}) calculates the gradient only at the mid-point of the full diffeomorphism, i.e. at $t = 0.5$ 
-in equation~\ref{eq:grad}
-% \begin{align}
-% \nabla \Pi =\frac{\partial}{\partial \phi_i} \Pi_{\sim}( \mathcal{I}(\phiinv_1(\x,0.5)) ,  \mathcal{J}(\phiinv_2(\x,0.5)) )
-% \end{align}
-for $i \in \{1,2\}$.  $\phi_1(\mathbf{x},0.5)$ and $\phi_2(\mathbf{x},0.5)$ are then updated from the previous iteration according to 
+Greedy optimization of Equation (\ref{eq:diffs}) calculates the
+gradient only at the mid-point of the full diffeomorphism, i.e. at $t
+= 0.5$ in equation~\ref{eq:grad}
 \begin{align} 
-  \phi_i(\mathbf{x},0.5) = \phi_i(\mathbf{x},0.5)+ (\delta K \star \nabla \Pi_i(\mathbf{x},0.5) ) \circ  \phi_i(\x,0.5).
-\end{align}
-That is, the gradient at the mid-point is mapped back to the origin of each diffeomorphism.  
-We then update the full mapping by explicitly enforcing $\phiinv(\phi(\x,1))=\x$ in the discrete 
-domain, as described in \cite{Avants2008}.
+\nabla \Pi =\frac{\partial}{\partial \phi_i} \Pi_{\sim}(
+\mathcal{I}(\phiinv_1(\x,0.5)) , \mathcal{J}(\phiinv_2(\x,0.5)) )
+\end{align} 
+for $i \in \{1,2\}$.  $\phi_1(\mathbf{x},0.5)$ and
+$\phi_2(\mathbf{x},0.5)$ are then updated from the previous iteration
+according to
+\begin{align}
+  \phi_i(\mathbf{x},0.5) = \phi_i(\mathbf{x},0.5)+ (\delta K \star
+  \nabla \Pi_i(\mathbf{x},0.5) ) \circ \phi_i(\x,0.5).
+\end{align}
+That is, the gradient at the mid-point is mapped back to the origin of
+each diffeomorphism.  We then update the full mapping by explicitly
+enforcing $\phiinv(\phi(\x,1))=\x$ in the discrete domain, as
+described in \cite{Avants2008}.
 
 \paragraph{Exponential Mapping}
-Ashburner introduced DARTEL (Diffeomorphic Anatomical Registration using Exponentiated Lie algebra) as a rapidly computed alternative to time parameterized diffeomorphic schemes \cite{Ashburner2007}.
-The key difference between a time-varying diffeomorphism and a diffeomorphism generated by an exponential mapping \cite{Ashburner2007} is that the exponential mapping maintains only a single vector field that is constant in time.
-
-By exponentiation of a constant velocity field, a diffeomorphism can be generated from the following o.d.e (cf Equation (\ref{eq:ode}))
+Ashburner introduced DARTEL (Diffeomorphic Anatomical Registration
+using Exponentiated Lie algebra) as a rapidly computed alternative to
+time parameterized diffeomorphic schemes \cite{Ashburner2007}.  The
+key difference between a time-varying diffeomorphism and a
+diffeomorphism generated by an exponential mapping
+\cite{Ashburner2007} is that the exponential mapping maintains only a
+single vector field that is constant in time.
+
+By exponentiation of a constant velocity field, a diffeomorphism can
+be generated from the following o.d.e (cf Equation (\ref{eq:ode}))
 \begin{align} \label{eq:odec}
-\frac{d \phi(\mathbf{x}, t)}{dt} = \boldsymbol{v}(\phi(\mathbf{x},t) ), \,\,\, \phi(\mathbf{x}, 0) = \mathbf{x}.
+\frac{d \phi(\mathbf{x}, t)}{dt} = \boldsymbol{v}(\phi(\mathbf{x},t)
+), \,\,\, \phi(\mathbf{x}, 0) = \mathbf{x}.
 \end{align}
 Note that there is no explicit time parameter in the velocity field.
-Theoretically, restricting the velocity field to be constant in time reduces the size of the space that may be generated \cite{Arnold1991} in a way that is similar to the difference between real and rational numbers, the latter of which are sparsely distributed through the reals.
+Theoretically, restricting the velocity field to be constant in time
+reduces the size of the space that may be generated \cite{Arnold1991}
+in a way that is similar to the difference between real and rational
+numbers, the latter of which are sparsely distributed through the
+reals.
 
@@
   \Pi(\mathcal{I}, \mathcal{J}, \phi) = \int_{\Omega} \left(\Pi_{\sim}(\mathcal{I}, \mathcal{J}, \phi)(\mathbf{x}) + 
-                  \Pi_{R}(\phi)(\mathbf{x})\right) d\Omega,
+                \lambda~  \Pi_{R}(\phi)(\mathbf{x})\right) d\Omega,
 \end{align}
@@
 \paragraph{Gaussian-Regularized Elastic Deformation}
-
-A simple and efficient, yet powerful image normalization algorithm is the approach known as Thirion's demons \cite{Thirion1998}.
-Using an optical flow based similarity, the solution is obtained by iterating between the calculation of image forces and subsequent Gaussian regularization.
-In ANTS, we extend this basic approach to include the similarity metrics available for deformable registration (see Table \ref{table:chart}).
+A simple and efficient, yet powerful image normalization algorithm is
+the approach known as Thirion's demons \cite{Thirion1998}.  Using an
+optical flow based similarity, the solution is obtained by iterating
+between the calculation of image forces and subsequent Gaussian
+regularization.  In ANTS, we extend this basic approach to include the
+similarity metrics available for deformable registration (see Table
+\ref{table:chart}).  This leads to an update form that is performed 
+as follows: 
+\begin{eqnarray} 
+{\bf U}=\frac{\partial}{\partial \phi} \Pi_{\sim}(\mathcal{I}(\x), \mathcal{J}(\phiinv(\x))) \notag \\ 
+\mathbf{u}(\mathbf{x}) = \mathbf{u}(\mathbf{x}) + \delta  K \star {\bf U},
+\end{eqnarray}
+Thus, ANTS vector field transformation models default to a 
+Demons style optimization.  Like the Demons algorithm, this 
+method uses a greedy descent on the energy and does not 
+truly optimize the total variational form.  The update strategy 
+for the exponential mapping is very similar to the above, but 
+with an additional composition on the gradient that maps the gradient 
+back to time zero.  See \cite{Vercauteren2009} for a nice discussion 
+of the distinction between additive and compositive updates in these methods.
 
 \paragraph{Directly Manipulated Free-Form Deformation}
-Another top performer in Klein's study \cite{Klein2009} was the Image Registration Toolkit (IRTK) based on the research originally reported in \cite{Rueckert1999} in which mutual information and a free-form deformation (FFD) transformation model were used to analyze breast deformation.
-In ANTS we provide an implementation of a variant of the well-known FFD transformation model for image registration known as {\em directly manipulated free-form deformation} \cite{Tustison2009}.
-The DMFFD model replaces the standard FFD gradient used in \cite{Rueckert1999} with an intuitive preconditioned gradient to overcome problematic energy topographies intrinsic with the traditional approach.
+Another top performer in Klein's study \cite{Klein2009} was the Image
+Registration Toolkit (IRTK) based on the research originally reported
+in \cite{Rueckert1999} in which mutual information and a free-form
+deformation (FFD) transformation model were used to analyze breast
+deformation.  In ANTS we provide an implementation of a variant of the
+well-known FFD transformation model for image registration known as
+{\em directly manipulated free-form deformation} \cite{Tustison2009}.
+The DMFFD model replaces the standard FFD gradient used in
+\cite{Rueckert1999} with an intuitive preconditioned gradient to
+overcome problematic energy topographies intrinsic with the
+traditional approach.
 
@@
 \subsection{ANTS Intensity-Based Similarity Metrics}
-Several intensity-based image metrics have been proposed in the literature with varying levels of performance dependent upon specific applications.
-We have included several of the most popular similarity metrics within ANTS.
-In addition, our software framework facilitates the development of other image metrics.
-
-Both mutual information \cite{Viola1997} and mean-squared difference similarity metrics are available for the linear transformations.
-Also included are the cross-correlation \cite{Avants2007b}, probabilistic matching (a more aggressive version of cross-correlation), local mutual information \cite{Rueckert1999,Pluim2003}, and mean squared difference similarity metrics for the non-linear transformation models.
-The parameters for the different metrics are discussed in the ANTS documentation \cite{ANTS2009}.
+Several intensity-based image metrics have been proposed in the
+literature with varying levels of performance dependent upon specific
+applications.  We have included several of the most popular similarity
+metrics within ANTS.  In addition, our software framework facilitates
+the development of other image metrics.  Both mutual information
+\cite{Viola1997} and mean-squared difference similarity metrics are
+available for the linear transformations.  Also included are the
+cross-correlation (the PR metric) \cite{Avants2007b}, local mutual
+information \cite{Rueckert1999,Pluim2003}, and mean squared difference
+similarity metrics for the non-linear transformation models.  The
+parameters for the different metrics are discussed in the ANTS
+documentation \cite{ANTS2009}.
 
 \subsection{ANTS Point-Set-Based Similarity Metrics}
-In addition to intensity-based metrics, ANTS also contains similarity metrics for 
-registering point-sets.  These include a landmark matching metric and 
-two point-set metrics which can accommodate point-sets of different cardinality.
-These point-set metrics can be used alone for strict point-set registration or
-in conjunction with intensity-based metrics for dual intensity/point-set registration.
+In addition to intensity-based metrics, ANTS also contains similarity
+metrics for registering point-sets.  These include a landmark matching
+metric and two point-set metrics which can accommodate point-sets of
+different cardinality.  These point-set metrics can be used alone for
+strict point-set registration or in conjunction with intensity-based
+metrics for dual intensity/point-set registration.
 
 \subsubsection{Exact Landmark Matching}
-Generalizing the B-spline fitting algorithm of \cite{Lee1997}, 
-we developed a scattered data approximation algorithm \cite{Tustison2006b} 
-and contributed the code to the ITK library \cite{Tustison2005}.  
-This code is included in ANTS and forms the basis of our exact landmark 
-matching where this metric seeks to minimize the weighted sum of distances between 
-corresponding landmarks using a hierarchical approach. 
-% One can also associate relative confidence values with each landmark for fine-tuning exact landmark matching results.
+Generalizing the B-spline fitting algorithm of \cite{Lee1997}, we
+developed a scattered data approximation algorithm
+\cite{Tustison2006b} and contributed the code to the ITK library
+\cite{Tustison2005}.  This code is included in ANTS and forms the
+basis of our exact landmark matching where this metric seeks to
+minimize the weighted sum of distances between corresponding landmarks
+using a hierarchical approach.  One can also associate relative
+confidence values with each landmark for fine-tuning exact landmark
+matching results.
 
 \subsubsection{Point-Set Expectation}
-In \cite{Pluta2008}, the point-set matching problem was formulated in the context 
-of incomplete label matching but is equally applicable to the general scenario of 
-registering point-sets not necessarily of equivalent cardinality.  Given two point-sets, 
-$X$ and $Y$, the essential idea underlying the point-set expectation matching 
-algorithm is that the optimal solution minimizes the distance between each point $y \in Y$
-with its corresponding {\em expected} point in $X$.
-
-We calculate the expected point using a Bayesian formulation and a non-parametric
-Parzen windowing scheme.  This allows one to define the probability of the point $x \in X$ 
-given a point $y\in Y$ as
+In \cite{Pluta2009}, the point-set matching problem was formulated in
+the context of incomplete label matching but is equally applicable to
+the general scenario of registering point-sets not necessarily of
+equivalent cardinality.  Given two point-sets, $X$ and $Y$, the
+essential idea underlying the point-set expectation matching algorithm
+is that the optimal solution minimizes the distance between each point
+$y \in Y$ with its corresponding {\em expected} point in $X$.
+
+We calculate the expected point using a Bayesian formulation and a
+non-parametric Parzen windowing scheme.  This allows one to define the
+probability of the point $x \in X$ given a point $y\in Y$ as
 \begin{align}
@@
 where $G(y; x, \sigma_X)$ is a normalized Gaussian with mean $x$ and
-standard deviation $\sigma_X$.  
-The expected point $\mathrm{E}(X|y)$ is then calculated to be
+standard deviation $\sigma_X$.  The expected point $\mathrm{E}(X|y)$
+is then calculated to be
 \begin{align} \label{eq:expected}
-  \mathrm{E}(X | Y = y) &= \sum_{j = 1}^{|X|} \mathbf{P}(X = x_j | Y = y ) x_j\nonumber \\
-                                        &= \frac{1}{|X|}\sum_{j = 1}^{|X|} G(y; x_j, \sigma_X) x_j  
-\end{align}
-where $|\cdot|$ denotes cardinality.
-The weighted sum of distances 
-between the points in $Y$ and their corresponding expected points in $X$ is calculated from 
-Equation (\ref{eq:expected}), i.e.
+  \mathrm{E}(X | Y = y) &= \sum_{j = 1}^{|X|} \mathbf{P}(X = x_j | Y =
+  y ) x_j\nonumber \\ &= \frac{1}{|X|}\sum_{j = 1}^{|X|} G(y; x_j,
+  \sigma_X) x_j
+\end{align}
+where $|\cdot|$ denotes cardinality.  The weighted sum of distances
+between the points in $Y$ and their corresponding expected points in
+$X$ is calculated from Equation (\ref{eq:expected}), i.e.
 \begin{align}
-\mathrm{PSE}(X,Y) = \frac{1 }{|Y|}\sum_{i=1}^{|Y|} \left\|y_i - \frac{1}{|X|} \sum_{j = 1}^{|X|} G(y_i; x_j, \sigma_X) x_j\right\|^2.
-\end{align}
-
-\subsubsection{Jensen-Havrda-Charvat-Tsallis Divergence}
-Recent information theoretic approaches have been proposed for point-set registration.
-A previous open-source contribution \cite{Tustison2009} generalizes the Jensen-Shannon divergence to the Jensen-Havrda-Charvat-Tsallis (JCHT) divergence which
-permits a fine-tuning  of the divergence measure such that emphasis can vary between robustness and sensitivity for 
-application-speciÞc tailoring \cite{Tustison2009a}.
-
-Each point-set is represented as a PDF via a Gaussian mixture model (GMM).
-Assuming $K$ point-sets denoted by $\{X_k, k \in \{1,\ldots,K\}\}$, the $k^{th}$ point-set is denoted by $\{x^k_1, \ldots, x^k_{|X_k|}\}$.
-The corresponding $k^{th}$ PDF is calculated from the $k^{th}$ point-set as
-\begin{align}
-  \mathbf{P}_k(s) = \frac{1}{|X_k|} \sum_{i=1}^{|X_k|} G( s; x^k_i, C^k_i )
-\end{align}
-where $G(s; x^k_i, C^k_i)$ is a normalized Gaussian with mean $x^k_i$ and covariance $C^k_i$ evaluated at $s$.
-For each point, $x_i$, the associated weighted covariance matrix, $C_{\mathcal{K}_i}$, is given by
-\begin{align} \label{eq:C}
-  C_{\mathcal{K}_i} = \frac{ \sum_{x_j \in \mathcal{N}_i, x_j \neq x_i} \mathcal{K}(x_i; x_j) (x_i - x_j)^\mathrm{T} (x_i - x_j)}{\sum_{x_j \in \mathcal{N}_i, x_j \neq x_i} \mathcal{K}(x_i; x_j) }
-\end{align}   
-where $\mathcal{N}_i$ is the local neighborhood of the point $x_i$ and $\mathcal{K}$ is a user-selected neighborhood  weighting kernel.
-We use an isotropic Gaussian for $\mathcal{K}$ with variance $\sigma_{\mathcal{K}_i}^2$ as well as a k-d tree structure for efficient determination of $\mathcal{N}_i$ \cite{Berg2000}.
-Calculation of the gradient requires the inverse of each covariance matrix.
-To avoid ill-conditioned covariance matrices, we use the modified covariance
-$C_i = C_{\mathcal{K}_i} + \sigma_n^2 I$
-where $I$ is the identity matrix and $\sigma_n$ is a parameter denoting added isotropic Gaussian noise.
-
-We designate the number of sample points generated for each of the $K$ probability density functions as $\{ M_1, \ldots, M_K \}$ and the $k^{th}$ set of points as $\{ s_1^k, \ldots, s_{M_k}^k \}$.
-The JHCT divergence is then calculated using the $K$ sets of points and the formula
-\begin{align} \label{eq:JHCT2}
-  \mathrm{JHCT}_\alpha(&\mathbf{P}_1, \ldots, \mathbf{P}_K) =  \frac{1}{ 1-\alpha}  
-  \nonumber \\
-   &\left[\frac{1}{M} \left( \sum_{k=1}^K \sum_{j=1}^{M_k} \left[ \mathbf{P}^*(s_j^k) \right]^{\alpha-1} -1 \right)\right. \nonumber \\
-   +&  \left. \frac{1}{N}\sum_{k=1}^K \frac{|X_k|}{M_k}   \left( \sum_{j=1}^{M_k} 
-       \left[ \mathbf{P}_k(s_j^k) \right]^{\alpha-1} - 1 \right) \right]
-\end{align}
-where 
-\begin{align}
-\mathbf{P}^*(X) = \frac{1}{N}\sum_{k=1}^K \sum_{i=1}^{|X_k|} G(x;x_i^k, C_i^k), 
-\end{align}
-$N = \sum_{k = 1}^K |X_k|$, and $M = \sum_{k = 1}^K M_k$.
-The prior weighting values are calculated from  $\gamma_k = |X_k|/N$ such that the larger point-sets are weighted more heavily.
+\mathrm{PSE}(X,Y) = \frac{1 }{|Y|}\sum_{i=1}^{|Y|} \left\|y_i -
+\frac{1}{|X|} \sum_{j = 1}^{|X|} G(y_i; x_j, \sigma_X) x_j\right\|^2.
+\end{align}
+A related similarity metric, based on maximizing the Jensen-Havrda-Charvat-Tsallis Divergence 
+between the point sets, is described in the appendix.
 
 \section{ANTS Implementation and Usage}
+\label{sec:imp}
 ANTS, built upon an ITK foundation, maintains the same coding style as its base.
@@
 }
-Binaries are also available from the same online location.  For quality assurance and maintenance purposes we have established an ANTS test reporting open-source ``dashboard''%
+Binaries for Windows, OSX, 32 and 64-bit Linux are also available from the same online location.
+For quality assurance and maintenance purposes we have established an ANTS test reporting open-source ``dashboard''%
 \footnote{
@@
 \end{align}
-
+Here, we apply the expectation-based point set registration method 
+for mapping labeled points sets, as described in \cite{Pluta2008}.   ITK-SNAP may be used to label 
+images and exported segmentation images may be input to the 
+PSE metric below, as labeled data.  The Frown and Smile data is used 
+as example.  This data is available in the 
+ANTS/Examples/Data/ directory.
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{Figures/frowntosmile.png}
+\includegraphics[width=0.45\textwidth]{Figures/smiletofrown.png}
+\caption{Expected output for the frown to smile 
+shows a smooth, though large deformation.  The grids are overlaid on 
+the deformed images. }
+\label{fig:frown}
+\end{figure}
+This example should run on the downloaded ANTS data so you may see the results.
+ANTS documentation provides the specific command used for the data.  
 In Table \ref{table:args} we give a brief summary of the arguments available for the normalizations offered by the ANTS package.
@@
     & Similarity & \verb#--linear-metric# & \verb#MI,MSQ# & \verb#[#$N_{bins}$,$N_{samples}$\verb#]# \\ 
-    {} & Affine or Rigid &  \verb#--do-rigid# & {} & \verb#0# \\
+    {} & Affine or Rigid &  \verb#--do-rigid# & {} & \verb#true / false # \\
     \cmidrule(l){2-5}
@@
     {} & Transformation & \verb#--transformation,-t# & \verb#Elast,SyN,Exp# & \verb#[#$\Delta_{gradient}$\verb#]# \\   
+    {} & Transformation & \verb#--large-deformation# & \verb#SyN# & \verb#[#$\Delta_{gradient}$,\# time points,dT\verb#]# \\
     \cmidrule(l){2-5}
@@
 \section{Experimental Evaluation}
-
+\label{sec:expt}
+The first part of our experimental evaluation will assess ANTS
+performance on qualitative, ``classic'' examples of large deformation
+image registration.  Two examples will suffice, both of which are
+based on the letter ``{\bf C}'' examples pioneered by
+\cite{Christensen96} and also used in \cite{Vercauteren2009}.  The
+purpose of these examples is to show large deformation capabilities
+with different transformation options.  We follow this example with
+experiments that assess whether, in the cortical labeling problem with
+normal control data, the added flexibility of true large deformation
+implementations conveys a performance advantage.
+
+\subsection{Classic Registration Examples}
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.33\textwidth]{Figures/grid1100.pdf} 
+\includegraphics[width=0.33\textwidth]{Figures/grid1110.pdf} 
+\includegraphics[width=0.33\textwidth]{Figures/grid1119.pdf} 
+\end{center}
+\caption{ANTS Large Deformation: The original goal of ANTS was to
+  develop public, open source large deformation image registration.
+  This is a classic example showing the progress of deforming a half
+  C to a full C along a geodesic diffeomorphism.  The deforming 
+  grid accompanies each deformed image.}
+\vspace{-0.1in}
+\label{fig:chalf}
+\end{figure}
+(via the {\verb --large-deformation } ~{\verb 1 } ~flag).
+
+\subsection{Cortical Labeling}
 There are many avenues for exploration of the various components of ANTS.
@@
 
+Build template parallel with affine update .... scripts etc .
+
 \section{Discussion}
@@
 
+\section{Additional ANTS Features}
+
+\subsubsection{Jensen-Havrda-Charvat-Tsallis Divergence}
+Recent information theoretic approaches have been proposed for
+point-set registration.  A previous open-source contribution
+\cite{Tustison2009} generalizes the Jensen-Shannon divergence to the
+Jensen-Havrda-Charvat-Tsallis (JCHT) divergence which permits a
+fine-tuning of the divergence measure such that emphasis can vary
+between robustness and sensitivity for application-speciﬁc tailoring
+\cite{Tustison2009a}.
+
+Each point-set is represented as a PDF via a Gaussian mixture model
+(GMM).  Assuming $K$ point-sets denoted by $\{X_k, k \in
+\{1,\ldots,K\}\}$, the $k^{th}$ point-set is denoted by $\{x^k_1,
+\ldots, x^k_{|X_k|}\}$.  The corresponding $k^{th}$ PDF is calculated
+from the $k^{th}$ point-set as
+\begin{align}
+  \mathbf{P}_k(s) = \frac{1}{|X_k|} \sum_{i=1}^{|X_k|} G( s; x^k_i,
+  C^k_i )
+\end{align}
+where $G(s; x^k_i, C^k_i)$ is a normalized Gaussian with mean $x^k_i$
+and covariance $C^k_i$ evaluated at $s$.  For each point, $x_i$, the
+associated weighted covariance matrix, $C_{\mathcal{K}_i}$, is given
+by
+\begin{align} \label{eq:C}
+  C_{\mathcal{K}_i} = \frac{ \sum_{x_j \in \mathcal{N}_i, x_j \neq
+      x_i} \mathcal{K}(x_i; x_j) (x_i - x_j)^\mathrm{T} (x_i -
+    x_j)}{\sum_{x_j \in \mathcal{N}_i, x_j \neq x_i} \mathcal{K}(x_i;
+    x_j) }
+\end{align}   
+where $\mathcal{N}_i$ is the local neighborhood of the point $x_i$ and
+$\mathcal{K}$ is a user-selected neighborhood weighting kernel.  We
+use an isotropic Gaussian for $\mathcal{K}$ with variance
+$\sigma_{\mathcal{K}_i}^2$ as well as a k-d tree structure for
+efficient determination of $\mathcal{N}_i$ \cite{Berg2000}.
+Calculation of the gradient requires the inverse of each covariance
+matrix.  To avoid ill-conditioned covariance matrices, we use the
+modified covariance $C_i = C_{\mathcal{K}_i} + \sigma_n^2 I$ where $I$
+is the identity matrix and $\sigma_n$ is a parameter denoting added
+isotropic Gaussian noise.
+
+We designate the number of sample points generated for each of the $K$ probability density functions as $\{ M_1, \ldots, M_K \}$ and the $k^{th}$ set of points as $\{ s_1^k, \ldots, s_{M_k}^k \}$.
+The JHCT divergence is then calculated using the $K$ sets of points and the formula
+\begin{align} \label{eq:JHCT2}
+  \mathrm{JHCT}_\alpha(&\mathbf{P}_1, \ldots, \mathbf{P}_K) =  \frac{1}{ 1-\alpha}  
+  \nonumber \\
+   &\left[\frac{1}{M} \left( \sum_{k=1}^K \sum_{j=1}^{M_k} \left[ \mathbf{P}^*(s_j^k) \right]^{\alpha-1} -1 \right)\right. \nonumber \\
+   +&  \left. \frac{1}{N}\sum_{k=1}^K \frac{|X_k|}{M_k}   \left( \sum_{j=1}^{M_k} 
+       \left[ \mathbf{P}_k(s_j^k) \right]^{\alpha-1} - 1 \right) \right]
+\end{align}
+where 
+\begin{align}
+\mathbf{P}^*(X) = \frac{1}{N}\sum_{k=1}^K \sum_{i=1}^{|X_k|} G(x;x_i^k, C_i^k), 
+\end{align}
+$N = \sum_{k = 1}^K |X_k|$, and $M = \sum_{k = 1}^K M_k$.
+The prior weighting values are calculated from  $\gamma_k = |X_k|/N$ such that the larger point-sets are weighted more heavily.
+
 \bibliographystyle{IEEEtran.bst}

### 2009-11-25 Scripts/buildtemplateparallel.sh
fusing shapeupdate and buildtemplateparallel
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@216 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -51,14 +51,14 @@ QSUBOPTS="" # EDIT THIS
   TEMPLATE=${TEMPLATENAME}.nii
 
   # Gradient step size, smaller in magnitude means more smaller (more cautious) steps
-  GRADIENTSTEP="-0.25"
+  GRADIENTSTEP="0.25"
 
 ITERATIONLIMIT=$3
 
 shift 3
 
 # Optionally disable qsub for debugging purposes - runs jobs in series
-DOQSUB=0
+DOQSUB=1
 
 
 IMAGESETVARIABLE=$*
@@ -201,43 +201,9 @@ fi
 
 echo " finished $i " >> ${TEMPLATENAME}metriclog.txt 
 
-${ANTSPATH}AverageImages $DIM ${TEMPLATE} 1 ${OUTPUTNAME}*formed.nii     
-#sh sygnccavg.sh 0.1  $TEMPLATE 
+sh ${ANTSPATH}shapeupdatetotemplate.sh $DIM $OUTPUTNAME $GRADIENTSTEP $IMAGESETVARIABLE
+# NIREP2  0.5 na*x.nii
 
-
-# below, a cheap approach to integrating the negative velocity field 
-# in the absence of other code and saving the velocity fields. 
-# additionally, this works for all types of registration algorithms 
-LESSLIMIT=$((${ITERATIONLIMIT} - 1))
-if [  $i  -lt  ${LESSLIMIT} ]
- then
-   rm -f  ${OUTPUTNAME}*InverseWarp*vec.nii  
-     ${ANTSPATH}AverageImages $DIM ${TEMPLATENAME}warpxvec.nii 0 ${OUTPUTNAME}*Warpxvec.nii  
-     ${ANTSPATH}AverageImages $DIM ${TEMPLATENAME}warpyvec.nii 0 ${OUTPUTNAME}*Warpyvec.nii  
-if [ $DIM -gt 2  ]
-then 
-     ${ANTSPATH}AverageImages $DIM ${TEMPLATENAME}warpzvec.nii 0 ${OUTPUTNAME}*Warpzvec.nii  
-fi
-     ${ANTSPATH}MultiplyImages  $DIM ${TEMPLATENAME}warpxvec.nii $GRADIENTSTEP ${TEMPLATENAME}warpxvec.nii 
[…]

### 2009-11-25 Scripts/buildtemplateparallel.sh
minor bug fix
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@217 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -46,9 +46,9 @@ QSUBOPTS="" # EDIT THIS
 
 
 # Mapping Parameters 
-  MAXITERATIONS=30x100x20  # EDIT THIS 
+  MAXITERATIONS=30x90x20  # EDIT THIS 
   TEMPLATENAME=${OUTPUTNAME}template
-  TEMPLATE=${TEMPLATENAME}.nii
+  TEMPLATE=${TEMPLATENAME}.nii.gz 
 
   # Gradient step size, smaller in magnitude means more smaller (more cautious) steps
   GRADIENTSTEP="0.25"
@@ -143,15 +143,15 @@ echo $NUMLEVELS
 count=0
 ANTSSCRIPTNAME=${ANTSPATH}ants.sh
 ANTSAFFSCRIPTNAME=${ANTSPATH}antsaffine.sh
-if [ -f $ANTSSCRIPTNAME  ]
-    then
-    echo " FILE $ANTSSCRIPTNAME OK ! "
-       else 
-       echo " FILE $ANTSSCRIPTNAME DOES NOT EXIST !!! "
+SHAPESCRIPT=${ANTSPATH}shapeupdatetotemplate.sh 
+for FLE in $ANTSSCRIPTNAME  $SHAPESCRIPT ; do 
+if [ ! -f $FLE  ] ; then
+       echo " FILE $FLE DOES NOT EXIST !!! "
        echo " copy the file to this directory & repeat " 
        echo " Available in ANTS/Scripts directory "
        exit
 fi
+done
 
 # Job IDs of jobs submitted to queue in loop below
 jobIDs=""
@@ -201,21 +201,16 @@ fi
 
 echo " finished $i " >> ${TEMPLATENAME}metriclog.txt 
 
-sh ${ANTSPATH}shapeupdatetotemplate.sh $DIM $OUTPUTNAME $GRADIENTSTEP $IMAGESETVARIABLE
+sh $SHAPESCRIPT $DIM $OUTPUTNAME $GRADIENTSTEP $IMAGESETVARIABLE
 # NIREP2  0.5 na*x.nii
 
-
-if [ $DIM -lt 3  ]
-then 
-    ${ANTSPATH}ConvertToJpg ${TEMPLATE} ${TEMPLATENAME}$i.jpg
-fi
[…]

### 2009-11-28 ImageRegistration/itkANTSImageRegistrationOptimizer.h
adding diffeomorphic demons style transformation model -- GreedyExp
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@221 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -617,6 +617,11 @@ PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer mo
         {
             std::cout << "SyN diffeomorphic model for transformation. " << std::endl;
         }
+        else if ( this->m_TransformationModel  == std::string("GreedyExp") ) 
+        {
+            std::cout << "Greedy Exp Diff model for transformation.   Similar to Diffeomorphic Demons.  Params same as Exp model. " << std::endl;
+            this->m_TransformationModel=std::string("GreedyExp");
+        }
         else
         {
             std::cout << "Exp Diff model for transformation. " << std::endl;
@@ -1029,6 +1034,13 @@ PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer mo
             this->DiffeomorphicExpRegistrationUpdate(fixedImage, movingImage,this->m_SimilarityMetrics[0]->GetFixedPointSet(),  this->m_SimilarityMetrics[0]->GetMovingPointSet() );
             }
           }
+        else if (this->GetTransformationModel() == std::string("GreedyExp"))
+          {
+          if(this->m_Iterations[currentLevel] > 0)
+            {
+            this->GreedyExpRegistrationUpdate(fixedImage, movingImage,this->m_SimilarityMetrics[0]->GetFixedPointSet(),  this->m_SimilarityMetrics[0]->GetMovingPointSet() );
+            }
+          }
 
         this->m_CurrentIteration++;
 	
@@ -1191,9 +1203,23 @@ PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer mo
       }
     else if (this->GetTransformationModel() == std::string("Exp"))
       {
-      DeformationFieldPointer diffmap = this->IntegrateConstantVelocity( this->m_DeformationField, (unsigned int)this->m_NTimeSteps , 1.0);///(float)this->m_NTimeSteps
+	DeformationFieldPointer diffmap =  this->IntegrateConstantVelocity( this->m_DeformationField, (unsigned int)this->m_NTimeSteps , 1.0);///(float)this->m_NTimeSteps
       DeformationFieldPointer invdiffmap = this->IntegrateConstantVelocity(this->m_DeformationField,(unsigned int) this->m_NTimeSteps, -1.0);///(float)this->m_NTimeSteps*(-1.0)
[…]

### 2009-11-28 ImageRegistration/itkANTSImageRegistrationOptimizer.h
ok - updating Exp again  - tests probably dont work
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@224 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -1203,8 +1203,8 @@ PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer mo
       }
     else if (this->GetTransformationModel() == std::string("Exp"))
       {
-	DeformationFieldPointer diffmap =  this->IntegrateConstantVelocity( this->m_DeformationField, (unsigned int)this->m_NTimeSteps , 1.0);///(float)this->m_NTimeSteps
-      DeformationFieldPointer invdiffmap = this->IntegrateConstantVelocity(this->m_DeformationField,(unsigned int) this->m_NTimeSteps, -1.0);///(float)this->m_NTimeSteps*(-1.0)
+	DeformationFieldPointer diffmap =  this->IntegrateConstantVelocity( this->m_DeformationField, (unsigned int)this->m_NTimeSteps , 1.0/ (float)this->m_NTimeSteps);
+	DeformationFieldPointer invdiffmap = this->IntegrateConstantVelocity(this->m_DeformationField,(unsigned int) this->m_NTimeSteps, -1.0/(float)this->m_NTimeSteps);
       this->m_InverseDeformationField=invdiffmap;
       this->m_DeformationField=diffmap;
       AffineTransformPointer invaff =NULL;

### 2009-11-28 ImageRegistration/itkANTSImageRegistrationOptimizer.h
update just Exp map
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@228 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -1203,8 +1203,8 @@ PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer mo
       }
     else if (this->GetTransformationModel() == std::string("Exp"))
       {
-	DeformationFieldPointer diffmap =  this->IntegrateConstantVelocity( this->m_DeformationField, (unsigned int)this->m_NTimeSteps , 1.0/ (float)this->m_NTimeSteps);
-	DeformationFieldPointer invdiffmap = this->IntegrateConstantVelocity(this->m_DeformationField,(unsigned int) this->m_NTimeSteps, -1.0/(float)this->m_NTimeSteps);
+	DeformationFieldPointer diffmap =  this->IntegrateConstantVelocity( this->m_DeformationField, (unsigned int)this->m_NTimeSteps , 1 ); // 1.0/ (float)this->m_NTimeSteps);
+	DeformationFieldPointer invdiffmap = this->IntegrateConstantVelocity(this->m_DeformationField,(unsigned int) this->m_NTimeSteps, -1 ); // -1.0/(float)this->m_NTimeSteps);
       this->m_InverseDeformationField=invdiffmap;
       this->m_DeformationField=diffmap;
       AffineTransformPointer invaff =NULL;

### 2009-11-30 Documentation/Tex/runtex.sh
--geodesic option
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@229 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -1,5 +1,5 @@
 
-for FL in ants # antstheory 
+for FL in antstheory # ants
   do 
 for i in 1 2 3 
 do

### 2009-11-30 ImageRegistration/itkANTSImageRegistrationOptimizer.h

@@ -669,7 +669,7 @@ PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer mo
     this->m_OutputNamingConvention=oOption->GetValue();   
 
     typename ParserType::OptionType::Pointer thicknessOption 
-      = this->m_Parser->GetOption( "large-deformation" );  
+      = this->m_Parser->GetOption( "geodesic" );  
     if( thicknessOption->GetValue() == "true" ||  thicknessOption->GetValue() == "1" ) { this->m_ComputeThickness=1; this->m_SyNFullTime=2; }// asymm forces
     else if(  thicknessOption->GetValue() == "2" )  { this->m_ComputeThickness=1; this->m_SyNFullTime=1; } // symmetric forces 
     else this->m_ComputeThickness=0; // not full time varying stuff 
@@ -1128,8 +1128,9 @@ PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer mo
 	  typename BSplinerType::GradientType gradient;
 	  gradient.Fill(0);
 	  bspliner->EvaluateGradientAtPoint( endPoint, gradient ); 
-	  if (  gradient[0][0]  < 0.0001 && this->m_CurrentIteration > domtar) converged=true;
-	  std::cout << " E-Slope " << gradient[0][0] ;//<< std::endl;
+	  this->m_ESlope=gradient[0][0]  ;
+	  if (  this->m_ESlope < 0.0001 && this->m_CurrentIteration > domtar) converged=true;
+	  std::cout << " E-Slope " <<  this->m_ESlope;//<< std::endl;
 	  }
         for ( unsigned int qq=0; qq < this->m_Energy.size(); qq++ )
           {
@@ -1617,7 +1618,6 @@ private:
     AffineTransformPointer m_FixedImageAffineTransform;
     DeformationFieldPointer m_DeformationField;
     DeformationFieldPointer m_InverseDeformationField;
-    DeformationFieldPointer m_StaticVelocityField;
 
 
     std::vector<float> m_GradientDescentParameters;
@@ -1649,6 +1649,7 @@ private:
   float m_NTimeSteps;
   float m_GaussianTruncation;
   float m_DeltaTime;
+  float m_ESlope; 
 
 /** energy stuff */
   std::vector<float> m_Energy;
@@ -1661,6 +1662,8 @@ private:
[…]

### 2009-11-30 Scripts/antsbashstats.sh

@@ -55,6 +55,7 @@ arith_mean ()
 
   while read value   # Read one data point at a time.
   do
+      echo $value 
     rt=$(echo "scale=$SC; $rt + $value" | bc)
     (( ct++ ))
   done

### 2009-11-30 Utilities/ReadWriteImage.h
Fixed upper/lower triangular format issues for NIFTI DTI data
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@230 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -26,7 +26,7 @@
 // the function below should be modified/eliminated.
 
 template <class TImageType>
-void NiftiDTICheck(itk::SmartPointer<TImageType> &target, const char *file)
+void NiftiDTICheck(itk::SmartPointer<TImageType> &target, const char *file, bool makeLower)
 {
   typedef typename TImageType::PixelType PixType;
   
@@ -37,29 +37,43 @@ void NiftiDTICheck(itk::SmartPointer<TImageType> &target, const char *file)
   if ((pos1 == std::string::npos) && (pos2 == std::string::npos))
     return;
     
-  //typedef itk::ImageFileReader< TImageType >      FileSourceType;
-  //typename FileSourceType::Pointer reader = FileSourceType::New();
-  //reader->SetFileName( file );
-  //reader->Update();
-  //if (reader->GetImageIO()->GetNumberOfComponents() != 6)
-  //  return;
   if (PixType::Length != 6)
     return;
     
-  std::cout << "Performing lower/upper triangular format swap for Nifti DTI" << std::endl;
+  std::cout << "Performing lower/upper triangular format check for Nifti DTI" << std::endl;
     
   // swap elements 2 and 3 for lower<->upper conversion  
   itk::ImageRegionIteratorWithIndex<TImageType> 
   iter(target,target->GetLargestPossibleRegion());
+  
+  double yy=0.0;
+  double yz=0.0;
+  double count=0.0;
   while (!iter.IsAtEnd())
   {
-    PixType pix = iter.Get();
-    typename PixType::ValueType temp;
-    temp = pix[2];
-    pix[2] = pix[3];
-    pix[3] = temp;
-    iter.Set(pix);
+    yz += iter.Get()[2];
+    yy += iter.Get()[3];
+    count += 1;
     ++iter;
+  }
+  yy /= count;
+  yz /= count;
+  
+  if ( ((yy < yz) && makeLower) || ((yy > yz) && !makeLower) )
+  {
+    std::cout << "Performing lower/upper triangular format swap for Nifti DTI" << std::endl;
+  
+    iter.GoToBegin();
+    while (!iter.IsAtEnd())
+    {
+      PixType pix = iter.Get();
+      typename PixType::ValueType temp;
+      temp = pix[2];
+      pix[2] = pix[3];
+      pix[3] = temp;
+      iter.Set(pix);
+      ++iter;
+    }
   }
   
 }
[…]

### 2010-01-06 Documentation/Tex/ants.tex
--geodesic option
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@229 fd2da0a4-d968-48df-9cb5-d8c496287cbf
updating with I/O and other stuff ...
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@239 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 % The author and/or editor can define 'significant' however they like.
-\release{1.00}
+\release{1.5}
 
@@
 neighbor interpolation, otherwise linear interpolation is used.
+
+Also, please take a look at 
+\href{http://picsl.upenn.edu/ANTS/ioants.php}{http://picsl.upenn.edu/ANTS/ioants.php} 
+and the section below on I/O.
+
+\subsection{I/O Data Formats In ANTS}
+ANTS support both 2D and 3D images. 
+Since ANTS is implemented in ITK, 
+it is able to read and write the popular data formats supported through ITK. 
+There are four basic types of data for ANTS.
+\subsubsection{Image Volumes}
+ANTS support both 2D and 3D images, including
+\begin{itemize}
+\item Nifti (.nii, .nii.gz)
+\item Analyze (.hdr + .img / .img.gz)
+\item MetaImage (.mha)
+\item Other formats through itk::ImageFileWriter / itk::ImageFileWriter such as jpg, tiff, etc.  See ITK documentation. 
+\end{itemize}
+\subsubsection{Affine Transformation File}
+ANTS uses the itk::TransformFileReader / itk::TransformFileWriter to handle affine transformation files. 
+These are text files recording the matrix and offset for the affine transform {\bf FIXME NEED MORE DETAIL on definition and geometry of affine matrix}.
+\subsubsection{Deformation Field File}
+ANTS writes the deformation field using one file for the deformation in each dimension. 
+For example, if the deformation field is named ABCD.nii.gz and exists in three dimensions, the corresponding files are:
+ABCDxvec.nii.gz, 
+ABCDyvec.nii.gz, 
+ABCDzvec.nii.gz (for 3D).  Note that ANTS programs {\em assume} the $*$vec part of the name will exist.  Thus, 
+one refers to the warp file as  ABCD.nii.gz , dropping the $*$vec from the name. 
+\subsubsection{Labeled Point Sets}
+There are three formats supported
+\begin{itemize}
+\item  VTK format: point + scalar label in a .vtk file. {\bf FIXME NEED MORE DETAIL -- what type of vtk data ?  polydata?}.
+\item  Labeled point sets are saved as an image volume with the intensity = 1,2,3 for label = 1,2,3. For example, 
+figure~\ref{fig:frown} shows 2D images representing a 3-class labeled point set. 
+\item Avants Format. In this simple format, the point sets are saved as a text file. The first and last line is always 0 0 0 0. Each line in between for a point as X Y Z L (L is the label.)
+\end{itemize}
+The names and formats of the output files are both specified by -o option to the ANTS executable. 
+For example, for in 3D with -o RegA2B.nii.gz , the typical output files are:
+Affine transform file: 
+RegA2BAffine.txt 
+Warping deformation field: 
+RegA2BWarpxvec.nii.gz, RegA2BWarpyvec.nii.gz, RegA2BWarpzvec.nii.gz
+Inverse warping deformation field:
+RegA2BInverseWarpxvec.nii.gz, RegA2BInverseWarpyvec.nii.gz, RegA2BInverseWarpzvec.nii.gz.  
+To apply the affine transform and the deformation field to the moving image, use WarpImageMultiTransform. 
+For example: to apply the transform to the moving image:
+
+WarpImageMultiTransform 2 A.nii.gz ADeformed.nii.gz -R B.nii.gz RegA2BWarp.nii.gz RegA2BAffine.txt
+
+to apply the inverse transform to the fixed image:
+
+WarpImageMultiTransform 2 B.nii.gz BDeformed.nii.gz -R A.nii.gz -i RegA2BAffine.txt RegA2BInverseWarp.nii.gz
 

### 2010-01-06 Documentation/Tex/runtex.sh

@@ -1,5 +1,5 @@
 
-for FL in antstheory # ants
+for FL in ants # antstheory # ants
   do 
 for i in 1 2 3 
 do

### 2010-01-06 ImageRegistration/itkANTSImageRegistrationOptimizer.h
modification to parameter default in InvertField
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@240 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -1348,7 +1348,7 @@ PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer mo
 
   float InvertField(DeformationFieldPointer field, 
 		    DeformationFieldPointer inverseField, float weight=1.0, 
-		    float toler=0.5, int maxiter=20, bool print = false)
+		    float toler=0.1, int maxiter=20, bool print = false)
 {
   
   VectorType zero; zero.Fill(0);

### 2010-01-07 Scripts/antsTutorial.sh
Niels van Strien's ANTS tutorial
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@242 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -1,0 +1,474 @@
+# !/bin/sh 
+
+VERSION="0.0.2"
+
+function Usage {
+    cat <<USAGE
+
+ants.sh will perform an elastic or diffeomorphic transformation of the moving image 
+into the fixed image space. 
+
+this script encodes some reasonable defaults for the parameters and is useful for testing and comparing methods
+
+Usage: 
+
+sh ants.sh -d ImageDimension -r fixed.ext -i moving.ext 
+
+Compulsory arguments:
+
+-d:  ImageDimension: 2 or 3 (for 2 or 3 Dimensional registration)
+
+-r:  reference image 
+
+-i:  input image
+
+Optional arguments
+
+-o:  OUTPREFIX; A prefix that is prepended to all output files.
+
+-m:  max-iterations
+ 
+	Max-Iterations in form: JxKxL where
+	J = max iterations at coarsest resolution (here, reduce by power of 2^2)
+	K = middle resolution iterations (here,reduce by power of 2)
+	L = fine resolution iterations (here, full resolution) !!this level takes much
+            more time per iteration!!
+
+	Adding an extra value before JxKxL (i.e. resulting in IxJxKxL) would add another 
+	iteration level.
+
+-n3  N3BiasFieldCorrection of moving image ( 0 = off; 1 = on (default) )
+
+-l:  Labels-In-Fixed-Image-Space-To-Deform-To-Moving-Image (default is off)
+
+-tm: type of similarity metric used for registration. 
+
+	For intramodal image registration, use:
+	CC = cross-correlation
+	MI = mutual information
+	PR = probability mapping (default)
+	MSQ = mean square difference
+	
+	For intermodal image registration, use:
+	MI = mutual information
+	PR = probability mapping (default)
+
+	You can change the values for each of these methods in this script.
+
+-tt: type of transformation model used for registration. 
+
+	For elastic image registration, use:
+	EL = elastic transformation model (less deformation possible)
+	
+	For diffeomorphic image registration, use:
+	SY = SyN with time (default) with arbitrary number of time points in time discretization 
+	S2 = SyN with time optimized specifically for 2 time points in the time discretization
+	GR = Greedy SyN 
+	EX = Exponential
+        DD = Diffeomorphic Demons style exponential mapping 
+
[…]

### 2010-01-07 Scripts/antswithdt.sh
updating to use ants.sh for default warp of scalar image
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@243 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -3,11 +3,11 @@
 NUMPARAMS=$#
 
 MAXITERATIONS=30x90x20
-
+export ANTSPATH=${ANTSPATH:="$HOME/bin/ants/"}
 if [ $NUMPARAMS -lt 3  ]
 then 
 echo " USAGE ::  "
-echo "  sh   ants.sh  ImageDimension  fixed.ext  moving.ext  OPTIONAL-OUTPREFIX   OPTIONAL-max-iterations  OPTIONAL-Subject/Moving-DT-To-Deform-To-Fixed-Image  OPTIONAL-Subject/Moving-BZero-To-DistortionCorrect-To-Moving-T1-Image " 
+echo "  sh   ants.sh  ImageDimension  fixed.ext  moving.ext Subject/Moving-DT-To-Deform-To-Fixed-Image  OPTIONAL-Subject/Moving-BZero-To-DistortionCorrect-To-Moving-T1-Image " 
 echo " be sure to set ANTSPATH environment variable " 
 echo " Max-Iterations in form :    JxKxL where " 
 echo "  J = max iterations at coarsest resolution (here, reduce by power of 2^2) "  
@@ -56,20 +56,10 @@ then
 OUTPUTNAME=${MOVING%.*} 
 fi  
 
-if [ $NUMPARAMS -gt 3  ]
-then 
-  OUTPUTNAME=${4}
-fi
-
-if [ $NUMPARAMS -gt 4  ]
-then
- MAXITERATIONS=$5
-fi
-
 MOVINGDT=0
-if [ $NUMPARAMS -gt 5  ]
+if [ $NUMPARAMS -gt 3  ]
 then
- MOVINGDT=$6
+ MOVINGDT=$4 
 if [ ! -f $MOVINGDT ]
 then 
 echo " THIS DTI DOES NOT EXIST : $MOVINGDT " 
@@ -78,9 +68,9 @@ fi
 fi
 
 MOVINGBZ=0
-if [ $NUMPARAMS -gt 6  ]
+if [ $NUMPARAMS -gt 4  ]
 then
- MOVINGBZ=$7
+ MOVINGBZ=$5
 if [ ! -f $MOVINGBZ ]
 then 
 echo " THIS BZero DOES NOT EXIST : $MOVINGBZ " 
@@ -99,11 +89,11 @@ fi
 
 
 # Mapping Parameters 
[…]

### 2010-01-08 Scripts/antsTutorial.sh
comments and fixes
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@245 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -1,6 +1,8 @@
 # !/bin/sh 
 
-VERSION="0.0.2"
+VERSION="0.0.3"
+
+export ANTSPATH=${ANTSPATH:="$HOME/bin/ants/"} # EDIT THIS 
 
 function Usage {
     cat <<USAGE
@@ -10,6 +12,9 @@ into the fixed image space.
 
 this script encodes some reasonable defaults for the parameters and is useful for testing and comparing methods
 
+ANTS will test for convergence after 10 iterations to prevent the program from running forever. 
+Convergence is determined by occurrence of a flat or inflection point in the similarity term time series. 
+
 Usage: 
 
 sh ants.sh -d ImageDimension -r fixed.ext -i moving.ext 
@@ -18,15 +23,15 @@ Compulsory arguments:
 
 -d:  ImageDimension: 2 or 3 (for 2 or 3 Dimensional registration)
 
--r:  reference image 
+-i:  Input image
 
--i:  input image
+-r:  Reference image 
 
 Optional arguments
 
--o:  OUTPREFIX; A prefix that is prepended to all output files.
+-l:  Labels-In-Fixed-Image-Space-To-Deform-To-Moving-Image (default is off)
 
--m:  max-iterations
+-m:  Max-iterations
  
 	Max-Iterations in form: JxKxL where
 	J = max iterations at coarsest resolution (here, reduce by power of 2^2)
@@ -37,11 +42,13 @@ Optional arguments
 	Adding an extra value before JxKxL (i.e. resulting in IxJxKxL) would add another 
 	iteration level.
 
--n3  N3BiasFieldCorrection of moving image ( 0 = off; 1 = on (default) )
+-n:  N3BiasFieldCorrection of moving image ( 0 = off; 1 = on (default) )
 
--l:  Labels-In-Fixed-Image-Space-To-Deform-To-Moving-Image (default is off)
+-o:  OUTPREFIX; A prefix that is prepended to all output files.
 
--tm: type of similarity metric used for registration. 
+-q:  Perform a Quality Check (QC) of the result ( 0 = off (default); 1 = on ).
+
+-s:  Type of similarity metric used for registration. 
 
 	For intramodal image registration, use:
 	CC = cross-correlation
@@ -55,7 +62,7 @@ Optional arguments
 
 	You can change the values for each of these methods in this script.
 
--tt: type of transformation model used for registration. 
[…]

### 2010-01-08 Scripts/ants.sh
default output is now nii.gz
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@246 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -108,18 +108,18 @@ echo " "
 
 
 # finally check the image headers
-compareheaders=`${ANTSPATH}ImageMath $DIM ${OUTPUTNAME}repaired.nii  CompareHeadersAndImages $FIXED $MOVING  | grep FailureState | cut -d ' ' -f 4  `
+compareheaders=`${ANTSPATH}ImageMath $DIM ${OUTPUTNAME}repaired.nii.gz   CompareHeadersAndImages $FIXED $MOVING  | grep FailureState | cut -d ' ' -f 4  `
 if [ $compareheaders -ne 0 ]
 then
 echo " You may have a problem with your header definition "
-echo " The repaired image is in : ${OUTPUTNAME}repaired.nii " 
+echo " The repaired image is in : ${OUTPUTNAME}repaired.nii.gz  " 
 echo " Call ImageMath's   CompareHeadersAndImages  on your Fixed and Moving image " 
 exit
 fi
 
-${ANTSPATH}N3BiasFieldCorrection $DIM ${OUTPUTNAME}repaired.nii  ${OUTPUTNAME}repaired.nii 4
+${ANTSPATH}N3BiasFieldCorrection $DIM ${OUTPUTNAME}repaired.nii.gz   ${OUTPUTNAME}repaired.nii.gz  4
 
-exe=" ${ANTSPATH}ANTS $DIM -m  ${METRIC}${FIXED},${OUTPUTNAME}repaired.nii,${METRICPARAMS}  -t $TRANSFORMATION  -r $REGULARIZATION -o ${OUTPUTNAME}   -i $MAXITERATIONS   --use-Histogram-Matching "
+exe=" ${ANTSPATH}ANTS $DIM -m  ${METRIC}${FIXED},${OUTPUTNAME}repaired.nii.gz,${METRICPARAMS}  -t $TRANSFORMATION  -r $REGULARIZATION -o ${OUTPUTNAME}   -i $MAXITERATIONS   --use-Histogram-Matching "
 
  echo " $exe "
 
@@ -128,11 +128,11 @@ exe=" ${ANTSPATH}ANTS $DIM -m  ${METRIC}${FIXED},${OUTPUTNAME}repaired.nii,${MET
   #below, some affine options 
[…]

### 2010-01-08 Scripts/antsTutorial.sh

@@ -163,7 +163,7 @@ Data integrity check failed
 There seems to be a problem with the header definitions of the input files. This
 script has tried to repair this issue automatically by invoking:
 
-${ANTSPATH}ImageMath $DIM ${OUTPUTNAME}repaired.nii CompareHeadersAndImages $FIXED $MOVING
+${ANTSPATH}ImageMath $DIM ${OUTPUTNAME}repaired.nii.gz  CompareHeadersAndImages $FIXED $MOVING
 
 The repaired image is: 
 
@@ -273,7 +273,7 @@ fi
 
 # check the image headers
 echo "input files: checking"
-compareheaders=`${ANTSPATH}ImageMath $DIM ${OUTPUTNAME}repaired.nii CompareHeadersAndImages $FIXED $MOVING  | grep FailureState | cut -d ' ' -f 4 `
+compareheaders=`${ANTSPATH}ImageMath $DIM ${OUTPUTNAME}repaired.nii.gz  CompareHeadersAndImages $FIXED $MOVING  | grep FailureState | cut -d ' ' -f 4 `
 
 if [ $compareheaders -ne 0 ]
 then
@@ -386,7 +386,7 @@ echo
 echo "--------------------------------------------------------------------------------------"
 echo "Applying forward transformation to ${MOVING}"
 echo "--------------------------------------------------------------------------------------"
-${ANTSPATH}WarpImageMultiTransform $DIM ${MOVING} ${OUTPUTNAME}deformed.nii ${OUTPUTNAME}Warp.nii ${OUTPUTNAME}Affine.txt -R ${FIXED} 
+${ANTSPATH}WarpImageMultiTransform $DIM ${MOVING} ${OUTPUTNAME}deformed.nii.gz  ${OUTPUTNAME}Warp.nii.gz  ${OUTPUTNAME}Affine.txt -R ${FIXED} 
 
 # Apply inverse transformation to FIXED
 if [ "${TRANSFORMATIONTYPE}" == "EL" ] 
@@ -401,13 +401,13 @@ echo "--------------------------------------------------------------------------
 echo "Applying inverse transformation to ${FIXED}"
[…]

### 2010-01-08 Scripts/shapeupdatetotemplate.sh
adding gz warp compatibility
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@247 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -72,7 +72,7 @@ echo " if the files and parameters are all ok then uncomment the exit call below
 echo " " 
 #exit
 
-   rm -f  ${OUTPUTNAME}*InverseWarp*vec.nii  
+   rm -f  ${OUTPUTNAME}*InverseWarp*vec.nii* 
 
 deformedimages=(` ls ${OUTPUTNAME}*formed.nii*     `)
 NUM=${#deformedimages[@]} 
@@ -92,20 +92,20 @@ fi
 fi
 
 
-deformx=(` ls ${OUTPUTNAME}*Warpxvec.nii     `)
+deformx=(` ls ${OUTPUTNAME}*Warpxvec.nii*     `)
 if [ ${#deformx[@]} -le 1 ] ; then 
 echo " you do not have images of the type :  ${OUTPUTNAME}*Warpxvec.nii  "
 echo " if they dont exist, you need to run ants.sh to map your input images $IMAGESETVARIABLE to the template : $TEMPLATE " 
 exit
 fi
-deformy=(` ls ${OUTPUTNAME}*Warpyvec.nii     `)
+deformy=(` ls ${OUTPUTNAME}*Warpyvec.nii*     `)
 if [ ${#deformy[@]} -le 1 ] ; then 
 echo " you do not have images of the type :  ${OUTPUTNAME}*Warpyvec.nii  "
 echo " if they dont exist, you need to run ants.sh to map your input images $IMAGESETVARIABLE to the template : $TEMPLATE " 
 exit
 fi
 if [ $DIM -eq 3 ] ; then 
-deformz=(` ls ${OUTPUTNAME}*Warpzvec.nii     `)
+deformz=(` ls ${OUTPUTNAME}*Warpzvec.nii*     `)
 if [ ${#deformz[@]} -le 1 ] ; then 
 echo " you do not have images of the type :  ${OUTPUTNAME}*Warpzvec.nii  "
 echo " if they dont exist, you need to run ants.sh to map your input images $IMAGESETVARIABLE to the template : $TEMPLATE " 
@@ -121,17 +121,17 @@ ${ANTSPATH}AverageImages $DIM ${TEMPLATE} 1 ${OUTPUTNAME}*formed.nii*
[…]

### 2010-02-03 Utilities/itkVectorGaussianInterpolateImageFunction.h
had to add a bunch of checks for nans ....
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@301 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -86,9 +86,13 @@ public:
       nt[d] = (int)(bb_end[d] - bb_start[d] + 0.5);
       dx[d].set_size(nt[d]);
       gx[d].set_size(nt[d]);
-      sf[d] = 1.0 / (sqrt(2.0) * sigma[d] / img->GetSpacing()[d]);
-      cut[d] = sigma[d] * alpha / img->GetSpacing()[d];
+      this->sigma[d]=1;
+      sf[d] = 1.0 / (sqrt(2.0) * this->sigma[d] / img->GetSpacing()[d]);
+      //      std::cout << " sigma " << this->sigma[d] << " spc " << img->GetSpacing()[d] << " sf " << sf[d] << std::endl;
+      cut[d] = this->sigma[d] * alpha / img->GetSpacing()[d];
       }
+
+    this->m_ImageSize=this->GetInputImage()->GetLargestPossibleRegion().GetSize();
     }
 
   /** Set input */
@@ -103,7 +107,7 @@ public:
     {
     // Set the parameters
     for(size_t d = 0; d < VDim; d++)
-      this->sigma[d] = sigma[d];
+      this->sigma[d] = 1.0; //sigma[d];
     this->alpha = alpha;
 
     // If the image already set, recompute
@@ -128,14 +132,17 @@ public:
     const ContinuousIndexType &index,
     OutputType *grad) const
     {
-      PixelType Vout;
+      OutputType Vout;
       Vout.Fill(0);
 
       // The bound variables for x, y, z
       int i0[VDim], i1[VDim];
       // Compute the ERF difference arrays
+      //      std::cout << " index " << index << " VD " << VDim << std::endl;
+
       for(size_t d = 0; d < VDim; d++)
         {
+	  if ( index[d] <= 0 || index[d] >= this->m_ImageSize[d]-1  || vnl_math_isnan(index[d]) || vnl_math_isinf(index[d]) ) return Vout; 
         double *pdx = const_cast<double *>(dx[d].data_block());
         double *pgx = grad ?  const_cast<double *>(gx[d].data_block()) : NULL;
         compute_erf_array(pdx, i0[d], i1[d], bb_start[d], nt[d], cut[d], index[d], sf[d], pgx);
@@ -198,13 +205,14 @@ public:
         for(size_t q = 0; q < VDim; q++)
           {
           grad[q] = (dsum_me[q] - rc * dsum_m[q]) / sum_m;
-          grad[q] /= -1.4142135623730951 * sigma[q];
[…]

### 2010-02-03 Utilities/itkApocritaSegmentationImageFilter.h
new apocrita filter ...
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@302 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -1,0 +1,262 @@
+/*=========================================================================
+
+  Program:   Insight Segmentation & Registration Toolkit
+  Module:    $RCSfile: itkApocritaSegmentationImageFilter.h,v $
+  Language:  C++
+  Date:      $Date: $
+  Version:   $Revision: $
+
+  Copyright (c) Insight Software Consortium. All rights reserved.
+  See ITKCopyright.txt or http://www.itk.org/HTML/Copyright.htm for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkApocritaSegmentationImageFilter_h
+#define __itkApocritaSegmentationImageFilter_h
+
+#include "itkImageToImageFilter.h"
+
+#include "itkArray.h"
+#include "itkBSplineScatteredDataPointSetToImageFilter.h"
+#include "itkFixedArray.h"
+#include "itkPointSet.h"
+#include "itkVector.h"
+
+#include <vector>
+#include <map>
+#include <pair.h>
+
+namespace itk
+{
+
+/** \class ApocritaSegmentationImageFilter
+ * \brief Apocrita:  A Priori Classification with Registration Initialized
+ *  Template Assistance
+ *
+ * This filter provides an Expectation-Maximization framework for statistical
+ * segmentation where the intensity profile of each class is modeled as a
+ * Gaussian (a gaussian mixture model (GMM)) and spatial smoothness is
+ * enforced by an MRF prior.
+ *
+ * Initial labeling can be performed by otsu thresholding, kmeans clustering,
+ * a set of user-specified prior probability images, or a prior label image.
+ * If specified, the latter two initialization options are also used as
+ * priors in the MRF update step.
+ *
+ * The assumed labeling is such that classes are assigned consecutive
+ * indices 1, 2, 3, etc.  Label 0 is reserved for the background when a
+ * mask is specified.
+ *
+ */
+
+template<class TInputImage, class TMaskImage
+  = Image<unsigned char,::itk::GetImageDimension<TInputImage>::ImageDimension>,
+  class TClassifiedImage = TMaskImage>
+class ITK_EXPORT ApocritaSegmentationImageFilter :
[…]

### 2010-02-05 Documentation/Tex/ants.tex
aff description
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@307 fd2da0a4-d968-48df-9cb5-d8c496287cbf
ants aff explanation
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@308 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 These are text files recording the matrix and offset for the affine transform {\bf FIXME NEED MORE DETAIL on definition and geometry of affine matrix}.
+The interpretation depends on the header but generally it breaks down like this:
+
+\#Insight Transform File V1.0
+
+\# Transform 0
+
+Transform: MatrixOffsetTransformBase\_double\_3\_3
+
+Parameters:  $ a~b~c~d~e~f~g~h~i~j~k~l $
+
+FixedParameters: $ M~N~O $
+
+The meaning of these different terms:
+
+The matrix component (rotation, scaling, shearing all composed together):
+
+$ \begin{matrix}
+a & b & c \\
+d & e & f  \\
+g & h & i \\ 
+\end{matrix} $
+
+is followed by  $~j~k~l~$ --- this is the translation.
+
+$L~M~N$  is the center of rotation .
+
+You can use the --initial-affine  and    --continue-affine  options to control how this is used in registration.
+
 \subsubsection{Deformation Field File}

### 2010-02-05 ImageRegistration/itkCrossCorrelationRegistrationFunction.h
a stronger, faster, better ... correlation function
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@309 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -3,8 +3,8 @@
   Program:   Insight Segmentation & Registration Toolkit
   Module:    $RCSfile: itkCrossCorrelationRegistrationFunction.h,v $
   Language:  C++
-  Date:      $Date: 2009/01/05 20:09:47 $
-  Version:   $Revision: 1.19 $
+  Date:      $Date: 2008/11/15 23:46:06 $
+  Version:   $Revision: 1.18 $
 
   Copyright (c) 2002 Insight Consortium. All rights reserved.
   See ITKCopyright.txt or http://www.itk.org/HTML/Copyright.htm for details.
@@ -160,125 +160,6 @@ public:
   /** Set the object's state before each iteration. */
   virtual void InitializeIteration();
 
-
-  virtual VectorType  OpticalFlowUpdate(const NeighborhoodType &neighborhood, bool returninv=false)
-  {
-  // Get fixed image related information
-  IndexType index=neighborhood.GetIndex();    
-  typename TDeformationField::PixelType vec;
-  vec.Fill(0);
-  VectorType update;
-  update.Fill(0.0);
-  double fixedValue;
-  CovariantVectorType fixedGradient,movingGradient;
-  double fixedGradientSquaredMagnitude = 0;
-  double movingGradientSquaredMagnitude = 0;
-  fixedValue = (double) this->finitediffimages[0]->GetPixel( index );
-  double movingValue = (double) this->finitediffimages[1]->GetPixel( index );
-
-  if (!returninv)
-    {
-      fixedGradient = m_FixedImageGradientCalculator->EvaluateAtIndex( index );
-      for( unsigned int j = 0; j < ImageDimension; j++ )
-       {
-         fixedGradientSquaredMagnitude += vnl_math_sqr( fixedGradient[j] );
-      } 
-
-      double speedValue = fixedValue - movingValue;
-      if (fabs(speedValue) < this->m_RobustnessParameter) speedValue=0;
-      double denominator = vnl_math_sqr( speedValue ) / m_Normalizer + 
-	fixedGradientSquaredMagnitude;
-      double m_DenominatorThreshold = 1e-9;
-      double m_IntensityDifferenceThreshold = 0.001;  
-      if ( vnl_math_abs(speedValue) < m_IntensityDifferenceThreshold || 
-	   denominator < m_DenominatorThreshold )
-	{
-	  for(unsigned int j = 0; j < ImageDimension; j++ )
-	    {
-	      update[j] = 0.0;
-	    }
[…]

### 2010-02-06 ImageRegistration/itkANTSImageRegistrationOptimizer.h
updates for speed and consistency with past and for the CC metric
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@312 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -1351,6 +1351,13 @@ PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer mo
 		    float toler=0.1, int maxiter=20, bool print = false)
 {
   
+  float mytoler=toler;
+  unsigned int mymaxiter=maxiter; 
+  typename ParserType::OptionType::Pointer thicknessOption 
+      = this->m_Parser->GetOption( "go-faster" );  
+  if( thicknessOption->GetValue() == "true" ||  thicknessOption->GetValue() == "1" ) 
+    { mytoler=0.5; maxiter=12; }
+
   VectorType zero; zero.Fill(0);
   //  if (this->GetElapsedIterations() < 2 ) maxiter=10;
 
@@ -1435,7 +1442,7 @@ PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer mo
     float epsilon = (float)size[0]/256;
     if (epsilon > 1) epsilon = 1;
     
-    while ( difmag > toler && ct < maxiter && meandif > 0.001)
+    while ( difmag > mytoler && ct < mymaxiter && meandif > 0.001)
   {
     denergy=laste-difmag;//meandif;
     denergy2=laste-meandif;

### 2010-02-06 ImageRegistration/itkAvantsMutualInformationRegistrationFunction.h
switching computeupdate and computeupdateinv and mult by -1
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@315 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -589,7 +589,7 @@ public:
 
 
 
-  virtual VectorType ComputeUpdate(const NeighborhoodType &neighborhood,
+  virtual VectorType ComputeUpdateInv(const NeighborhoodType &neighborhood,
                                    void *globalData,
                                    const FloatOffsetType &offset = FloatOffsetType(0.0))
   {
@@ -601,7 +601,6 @@ public:
     if (!img) return update;
     typename FixedImageType::SpacingType spacing=img->GetSpacing();
     typename FixedImageType::SizeType imagesize=img->GetLargestPossibleRegion().GetSize();
-    //bool inimage=true;
     
     for (unsigned int dd=0; dd<ImageDimension; dd++)
       {
@@ -611,92 +610,35 @@ public:
       }    
     
     CovariantVectorType fixedGradient;
-    //    ImageDerivativesType fixedGradient;
-    CovariantVectorType fixedGradientNorm;
-    //std::cout << " grad " << std::endl;
-
     double loce=0.0;
-//    double nccp1=0,nccm1=0;
     ParametersType fdvec1(ImageDimension);
     ParametersType fdvec2(ImageDimension);    
     fdvec1.Fill(0);
     fdvec2.Fill(0);
-
-	fixedGradient = m_FixedImageGradientCalculator->EvaluateAtIndex( oindex ); 
-	float mag1=0;
-	for (int imd=0; imd<ImageDimension; imd++) 
-	  {
-	    mag1+=fixedGradient[imd]*fixedGradient[imd];
-	    fixedGradientNorm[imd]=fixedGradient[imd];
-	  }
-	mag1=sqrt(mag1);
-	if (mag1 > 1.e-5 ) fixedGradientNorm/=mag1;  
-	
-	double nccm1=0;
-	loce=this->GetValueAndDerivative(oindex,nccm1,fdvec1,fdvec2); 
-	float sign=(1.) * loce;//nccp1-nccm1; 
-	
-        double denominator = mag1;
-	if (denominator < 1.e-12)  denominator = 1.0;
-	denominator=1.;
-	for (int imd=0; imd<ImageDimension; imd++) update[imd]=sign*fixedGradient[imd]/denominator*spacing[imd];
-
-//    	if (this->m_MetricImage) this->m_MetricImage->SetPixel(oindex,(loce));//+this->m_MetricImage->GetPixel(oindex));
[…]

### 2010-03-30 Utilities/itkSurfaceCurvatureBase.h
take out rpoly include
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@455 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -21,7 +21,7 @@
 #include <vnl/algo/vnl_svd.h>
 #include <vnl/algo/vnl_symmetric_eigensystem.h>
 #include <vnl/vnl_cross.h>
- #include <vnl/algo/vnl_rpoly_roots.h>
+// #include <vnl/algo/vnl_rpoly_roots.h>
 #include "itkObject.h"
 #include "itkProcessObject.h"
 

### 2010-03-30 Documentation/Tex/ants.tex
update with some warp info
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@456 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 
-$L~M~N$  is the center of rotation .
+$M~N~O$  is the center of rotation .
 
@@
 one refers to the warp file as  ABCD.nii.gz , dropping the $*$vec from the name. 
+The displacement for each voxel is defined in physical image space.  If you want to 
+see the details of how a warp is achieved, look at the code in itkWarpImageMultiTransform.txx. 
+One issue to be aware of is that if you are directly indexing a vector component image in 
+voxel coordinates, then you need to apply the image's direction matrix to the vector.  
+An example of this is in ANTSJacobian.cxx, the TransformVector function.
+
 \subsubsection{Labeled Point Sets}

### 2010-08-14 ImageSegmentation/antsAtroposSegmentationImageFilter.h
ENH:  Continuation of better memory management refactoring.  This change was minor which shouldn't change output.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@626 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -290,7 +290,7 @@ public:
     unsigned int, unsigned int );
   typename RealImageType::Pointer
     GetDistancePriorProbabilityImageFromPriorLabelImage( unsigned int );
-  typename SampleType::Pointer GetScalarSamples( unsigned int );
+  typename SampleType::Pointer GetScalarSamples();
 
 #ifdef ITK_USE_CONCEPT_CHECKING
   /** Begin concept checking */

### 2010-08-16 Documentation/Tex/atropos.tex
initial version of atropos paper
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@628 fd2da0a4-d968-48df-9cb5-d8c496287cbf

% pdflatex !!!  pdf2ps !!! a2ps

%\documentclass[12pt]{article}
\documentclass[preprint,authoryear,12pt]{elsarticle}
%\documentclass[10pt,twocolumn]{article}
\usepackage{times,verbatim}
\pagestyle{empty}

% Set dimensions of columns, gap between columns, and paragraph indent 
\setlength{\textheight}{9in}
\setlength{\textwidth}{7in}
\setlength{\topmargin}{0in}
\setlength{\headheight}{0in}
\setlength{\headsep}{0in}
\setlength{\parindent}{1pc}
\setlength{\oddsidemargin}{-.1875in}  % Centers text.
\setlength{\evensidemargin}{-.1875in}

\newcommand{\FIGUREDIR}{/data/Avants/}
\newcommand{\FTDNUM}{25~}
\newcommand{\ADNUM}{24~}
\newcommand{\ELDNUM}{23~}
\newcommand{\x}{{\bf x}}
\newcommand{\y}{{\bf y}}
\newcommand{\I}{{\bar I}}
\newcommand {\alg}{{\em Atropos}}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{epsfig}
\usepackage{graphics}
\usepackage{soul}
\usepackage{rotating}
\usepackage{multirow,booktabs,ctable,array}
\usepackage{listing}						
\usepackage{listings}					
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}

\begin{document}

\date{}
\begin{frontmatter}
%\title{Evaluation of a Prior-Constrained Registration Initialized Template Assisted Segmentation Algorithm for $N$-Classes}
\title{NIREP and LPBA Evaluation of an Open-Source Framework for Cortical Parcellation}
\author{Nick Tustison, Brian Avants\footnote{First two authors equal co-authorship}, Philip A. Cook and James C. Gee\\  
  Penn Image Computing and Science Laboratory 
	\\ University of Pennsylvania \\ Philadelphia, PA~~19104-6389
	}

\begin{abstract} 
Neuroanatomical coordinate systems are essential for the
interpretation of structural and functional imaging studies.  This
work proposes a fully image-based, open-source approach to label-based
cortical parcellation that uses existing training data to identify the
same structures in new datasets.  The driving engine underlying this
system is a new N-class EM-MRF algorithm, \alg (distributed with the
Advanced Normalization Tools (ANTs) software), that incorporates
spatial priors to guide the segmentation.  \alg efficiently implements
a variety of regularization and statistical models for segmentation
and uses an [ EM or ICA ] energy minimization approach.  \alg's
efficient implementation allows one to solve {\em many class}
segmentation problems (here up to 56 classes are used) while using
both spatial probability maps and either geodesic or Euclidean
distance priors to constrain the segmentation.  This work evaluates
\alg performance on the parcellation problem given two different
training datasets, the LPBA40 dataset from UCLA and the NIREP dataset
from University of Iowa.  We evaluate three aspects of the problem: (1) the quality
of automatic cortical extraction from raw data; (2) the quality of
parcellation, given a ground truth cortical extraction; (3) the
quality of automated cortical extraction followed by automated
cortical parcellation.  Component (3) corresponds to the most
realistic clinical case, while component (2) eliminates the confound
of defining the cortex itself and focuses only on the parcellation
problem.  We also quantify the effect of using either a parametric or
non-parametric appearance model in the segmentation.  Finally, we
distribute all of the scripts and code from this evaluation.
\end{abstract}
\begin{keyword}
segmentation \sep expectation maximization \sep spatial prior \sep brain 
\end{keyword}
\end{frontmatter}

\section{Introduction}
Review segmentation methods , in particular Warfield 2009, FAST and
other methods that use spatial priors.  Review evaluation of
segmentation.

The expectation maximization and markov random field framework
(EM-MRF) for segmentation is reliable and efficient for large-scale
neuroimage processing in a variety of conditions.  However, EM-MRF
methods are sensitive to initial conditions which may lead to
performance instability across clinical subjects and in longitudinal
studies.  This paper proposes an open-source $N$-class extension to
the EM-MRF model, \alg, that includes strong use of template-based
information in both the initialization and as spatial priors in order
to guide the method into a consistent local minimum.

\begin{comment}{
  We evaluate this
method on BrainWeb 20 data and show that by relying upon accurate,
high-resolution diffeomorphic image registration, \alg~is capable of
reliably segmenting the BW20 whole head data into separate tissue
classes for muscle, skin, skull and bone marrow as well as the
standard three tissues, cerebrospinal fluid and gray and white matter.
\alg's three-tissue performance compares favorably with the standard
EM-MRF algorithm, FAST.  We also use the OASIS dataset to show that
\alg~increases segmentation repeatability relative to FAST.  The
evaluation data, the data-derived templates, the code and application
itself are publicly available.
}
\end{comment}

One advantage of the ANTS-Atropos pipeline is that we parcellate
cortex entirely in the image space, thus avoiding the difficulty of
transferring labels from the mesh space back to the image space---a
problem that is a confound of surface-based methods\cite{Klein2010}.

\section{Methods} \alg~encodes a family of segmentation techniques
that may be instantiated for different applications.  We describe the
general theory, algorithm and implementation that form \alg~and then
specify the particular parameters used for the applications of
interest to this research.

\subsection{Notation} An image, $I$, maps a domain, $\Omega$ into the
positive real numbers, such that $I \colon \Omega \rightarrow
\mathbb{R}^+$.  The goal of segmentation, in general, is to define the
spatial distribution of a finite set of labels over this domain.  We
denote the segmentation itself as $\eta \colon \Omega \rightarrow L$
where $L = \{ L_1 = 1 , \cdots , L_N=N \}$, a set of integer indexed
segmentation labels.  Note that $\eta$ may be formed as $\eta =
\sum_{i=1}^{i=N} L_i \eta_i $ where $\eta_i$ is the binary
segmentation for label $L_i$.  A prior estimate for the label image
$\eta_i$ is here denoted $\eta^s_i$ with a complete set of priors
denoted $\eta^s$.  The boundary of the binary segmentation -- where a
$0/1$ transition edge exists -- is denoted $\partial \eta^s_i$, for
the prior, and $\partial \eta_i$ for the label image.

\subsection{Apocrita Theory} A general maximum a posteriori criterion
for segmentation seeks,
\begin{equation} {\hat \eta} = \argmax _{\eta} \Pr( \eta | I )(\x) =
\Pr( I | \eta )(\x) \Pr(\eta)(\x),
\end{equation} where $I$ is the input image, ${\eta}$ represents the
label set configuration taken from the set $L$, $\Pr$ is the
probability, $\x$ is the spatial index and the optimal solution is
$\hat \eta$.  The input image $I$, here, is an unlabeled T1 MRI
indexed by the value $\x \in \Omega$ where $\Omega$ is the image's
spatial domain.  This probability is composed of the likelihood (first
term) and the prior (second term).  \alg~ uses a spatially varying
likelihood term and a two-component prior term that takes into account
both spatial distribution and label smoothness, the latter via a
standard MRF prior.

The likelihood term for a single label value $L_i \in L$ is,
\begin{eqnarray} \Pr( I | \eta_i )(\x)=\frac{1}{Z_i} \exp(- \| I(\x) -
\mu_i(\x) \|^2 / \sigma_i^2 ),
\end{eqnarray} where $Z_i$ is a normalizing constant, $\mu(\x)$ is a
spatially varying estimate of the tissue mean and $\sigma$ is a
standard deviation.  The prior term is given by,
\begin{eqnarray} \Pr(\eta_i)(\x) = \frac{1}{Z^\prime_i} \exp( -f( \x -
\y_{\partial \eta^s_i} ) / \sigma_{\eta^s_i}^2 ) p(\eta_i |
\eta^\aleph_i), \\ \notag f( \x - \y_{\partial \eta^s_i} ) = (1 -
\eta^s_i (\x) ) \| \x - \y_{\partial \eta^s_i} \|,
\end{eqnarray} where $p(\eta_i | \eta^N_i)$ is the MRF smoothness
probability based on the local neighborhood $\aleph$, the $Z$ is a
normalizing constant and $\y_{\partial \eta^s_i}$ is the nearest point
to $\x$ on the boundary of this labeling. \alg~requires a user or
template-defined $\eta^s_i$ and the standard deviation
$\sigma_{\eta^s_i}$ if a non-unity spatial prior component is desired
for that label.  The free parameters, that must be estimated
iteratively, are therefore $\mu_i, \sigma_i$ and the label set itself
$\hat \eta$, which defines the (locally) optimal spatial distribution
of $L$ through $\Omega$.  Figure~\ref{fig:spatp} shows the
distribution of the spatial prior as a function of the distance from
prior-defined object boundary.  Note that $f$ may easily be varied for
other applications or that fixed probability images may also be
substituted here.  This choice of $f$ is motivated by the fact that it
allows compressed storage of the priors in a single image, $\eta^s$,
while also maintaining the ability to manipulate -- for each
$\eta^s_i$ -- the spatial influence of the prior via
$\sigma_{\eta^s_i}$.  Practically, this is especially valuable when,
$N$, the number of labels, is large.

\subsection{Apocrita EM Algorithm}
The EM approach.

\subsection{Apocrita Implementation}
The command line and practical example, ITK, etc. 
Template-based initialization.

\subsection{The LPBA40 Dataset}  
\label{sec:lpba}
The LPBA40 dataset \cite{Shattuck2008} was collected at the North Shore Long
Island Jewish Health System imaging center and is maintained at UCLA.
LPBA40 contains 40 images (20 male $+$ 20 female) from normal, healthy
ethnically diverse volunteers with average age of 29.2 $\pm$ 6.3
years.  Each subject underwent 3D SPGR MRI on a 1.5T GE system
resulting in $0.86 \times 0.86 \times 1.5 mm^3$ images.  Each MRI in
the LPBA40 dataset was manually labeled with 56 independent structures
at the UCLA Laboratory of Neuro Imaging (LONI).  The test-retest
reliability of the labeling, across raters, was reported as a minimum
Jaccard ratio of $0.697$ in the supramarginal gyrus to a maximum of
$0.966$ in the gyrus rectus.  A single labeling of each image is made
available to the public and used, here, as silver-standard data 
for both training and testing in our cross-validation scheme.

\subsection{The NIREP Dataset}  
The non-rigid image registration evaluation project (NIREP
http://www.nirep.org/) is a resource of 16 high quality labeled brain
images at 1$mm^3$.  Each brain was labeled with 32 cortical regions
(16 on each hemisphere) and an additional class of other gray matter
tissue.  Regions vary in size from large (inferior temporal region) to
small (temporal pole, insula gyrus, frontal pole).  The main drawback
is a lack of inter-rater reliability numbers -- in particular because
visual inspection and comparison of labelings reveals a degree of
inconsistency in labeling of particular regions across subjects.
Nevertheless, the NIREP dataset is perhaps the highest quality
evaluation dataset currently available for the cortex.

\begin{figure}
%\center{ \includegraphics[width=3in]{figures/spatialprior.pdf} }
\caption{The spatial prior.
\label{fig:spatp} }
\vspace{-0.1in}
\end{figure}

\section{Results}
% \subsection{BWeb 20 Results} $N$-class segmentation and compare to FAST.

% \subsection{Oasis Results} Review stability across repeat images vs. FAST.

\section{Discussion}

\section{Conclusion}
To our knowledge, this is the first demonstration of a consistent and

\paragraph{Acknowledgments}
{This work was supported in part by NIH ... }

\bibliographystyle{elsarticle-harv.bst}
\bibliography{avantsAll,references}

\end{document}

### 2010-09-12 Documentation/atropos/atropos.tex
Initial commit.  Bulk of Atropos documentation.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@639 fd2da0a4-d968-48df-9cb5-d8c496287cbf

%\documentclass[11pt,english]{article}
%
%% Set page margins correctly
%\usepackage{geometry}
%%\geometry{letterpaper,top=0.5in,left=0.5in,bottom=0.5in,top=0.5in,headsep=6pt,footskip=18pt}
%\geometry{letterpaper,top=0.5in,left=0.5in,bottom=0.5in,top=0.5in,headsep=6pt,footskip=18pt}
%
%% Use fancy header style
%\usepackage{fancyhdr}
%\pagestyle{fancy}
%\renewcommand{\headrulewidth}{0.75pt}
%\renewcommand{\footrulewidth}{0.75pt}
%\lhead{\footnotesize Principal Investigator/Program Director (Last, First, Middle): }
%\rhead{Yushkevich, Paul A.} \lfoot{\footnotesize PHS 398/2590 (Rev. 04/06)} \rfoot{\footnotesize
%\textbf Continuation Format Page} \cfoot{\footnotesize Page \underline{~~\thepage~~}}
%
%% Use pslatex fonts
%\usepackage{pslatex}
%\renewcommand{\familydefault}{\sfdefault}
%\renewcommand{\baselinestretch}{.9}
%

% START JUNK
\documentclass[11pt,english]{article}

% Set page margins correctly
\usepackage{geometry}
\geometry{letterpaper,top=1.0in,left=1.0in,bottom=1.0in,top=1.0in,headsep=6pt,footskip=18pt}

\usepackage{lscape}
\usepackage[round,semicolon,authoryear,]{natbib}

% Use fancy header style
\usepackage{fancyhdr}
\pagestyle{empty}
\renewcommand{\headrulewidth}{0.75pt}
\renewcommand{\footrulewidth}{0.75pt}
%\lhead{\footnotesize Principal Investigator/Program Director (Last, First, Middle): }
%\rhead{Gee, James C.} \lfoot{\footnotesize PHS398 (06/07)} \rfoot{\footnotesize
%\textbf Continuation Format Page} \cfoot{\footnotesize Page \underline{~~\thepage~~}}

\usepackage{setspace}
\usepackage{listings}
\usepackage{float}

\floatstyle{plain}
\newfloat{command}{thp}{lop}
\floatname{command}{Command}

\doublespacing

% Use pslatex fonts
%\usepackage[T1]{fontenc}
%\usepackage{mathptmx}

% END JUNK

% All other packages
\usepackage{boxedminipage}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{babel}
\usepackage{enumerate}
%\usepackage{dsfont}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{sectsty}
\usepackage[compact]{titlesec}
\usepackage[usenames]{color}
\usepackage{ulem}
\usepackage{multirow,booktabs,ctable,array}

%GATHER{../../../shared/bibtex/biblio.bib}
%GATHER{../../../shared/bibtex/brainmri.bib}
%GATHER{../../../shared/bibtex/registration.bib}
\graphicspath{{./Figures/}
                          }

%\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\argmax}{\operatornamewithlimits{argmax}}
\newcommand{\argmin}{\operatornamewithlimits{argmin}}

\long\def\symbolfootnote[#1]#2{\begingroup%
\def\thefootnote{\fnsymbol{footnote}}\footnote[#1]{#2}\endgroup}

    \usepackage{color}

    \definecolor{listcomment}{rgb}{0.0,0.5,0.0}
    \definecolor{listkeyword}{rgb}{0.0,0.0,0.5}
    \definecolor{listnumbers}{gray}{0.65}
    \definecolor{listlightgray}{gray}{0.955}
    \definecolor{listwhite}{gray}{1.0}

\newcommand{\lstsetcpp}
{
\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listlightgray}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = none,
        numbersep = 6pt,
        numberstyle={\ttfamily\color{listnumbers}},
        tabsize = 2,
        language=[ANSI]C++,
        floatplacement=!h,
        caption={\small \baselineskip 12pt Atropos short command line menu which is invoked using the `{\ttfamily -h}' option. 
        The expanded menu, which provides additional details regarding the possible parameters and usage 
        options, is elicited using the `{\ttfamily {-}{-}help}' option.},
        captionpos=b,
        label=listing:command
        }
}

\newcommand{\lstsetcppnfour}
{
\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listlightgray}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = none,
        numbersep = 6pt,
        numberstyle={\ttfamily\color{listnumbers}},
        tabsize = 2,
        language=[ANSI]C++,
        floatplacement=!h,
        caption={\small \baselineskip 12pt N4 short command line menu which is invoked using the `{\ttfamily -h}' option. 
        The expanded menu, which provides additional details regarding the possible parameters and usage 
        options, is elicited using the `{\ttfamily {-}{-}help}' option.  },
        captionpos=b,
        label=listing:n4
        }
}

%\renewcommand{\topfraction}{0.85}
%\renewcommand{\textfraction}{0.1}
%\renewcommand{\floatpagefraction}{0.75}

% Different font in captions
%\newcommand{\captionfonts}{\small}
%\makeatletter  % Allow the use of @ in command names
%\long\def\@makecaption#1#2{%
%  \vskip\abovecaptionskip
%  \sbox\@tempboxa{{\captionfonts #1: #2}}%
%  \ifdim \wd\@tempboxa >\hsize
%    {\captionfonts #1: #2\par}
%  \else
%    \hbox to\hsize{\hfil\box\@tempboxa\hfil}%
%  \fi
%  \vskip\belowcaptionskip}
%\makeatother   % Cancel the effect of \makeatletter

%Automated Segmentation of Ventilation Defects on $^3$He Lung MRI}
%\date{}
%\author{ Nicholas J. Tustison, DSc,$^{1*}$ Talissa A. Altes, MD,$^2$ Eduard E. de Lange, MD,$^2$
%  Brian B. Avants, PhD, $^1$ John P. Mugler III, PhD,$^2$ and James C. Gee, PhD,$^1$ \\
%  $^1$ Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, Pennsylvania,  USA.\\
%  $^2$ Department of Radiology, University of Virginia, Charlottesville, Virginia, USA}
%\address{3600 Market Street, Suite 370 \\
%                 Philadelphia, PA  19104\\
%                 tustison@picsl.upenn.edu}
%\advisor{James Gee}
\begin{document}

\normalem

\vspace*{5cm}

\begin{center}
{\Large \bf An Open Source Framework for Multivariate $n$-Tissue Segmentation and Volumetric Cortical Parcellation:  Evaluation Using the BrainWeb, NIREP, and LPBA Public Data Sets} \\
\vspace*{0.5cm}
{\normalsize Brian B.~Avants$^{1*}$,  Nicholas J.~Tustison$^2$%
\symbolfootnote[1]{
  The first two authors contributed equally to this work.
}, 
Philip A.~Cook$^1$, Jue Woo$^1$, and James C.~Gee$^1$} \\
\begin{singlespace} 
{\scriptsize  $^1$ Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, Pennsylvania,  USA.\\
  $^2$ Department of Radiology, University of Virginia, Charlottesville, Virginia, USA}
\end{singlespace}
\end{center}

\vfill

\begin{singlespace} 
\scriptsize
\flushleft
%\line(1, 0){250} \\
{\bf Atropos:  $n$-Tissue Segmentation and Cortical Parcellation}\\
Corresponding author: \\
Brian B. Avants\\
3600 Market Street, Suite 370\\
Philadelphia, PA  19104\\
avants@picsl.upenn.edu\\
\end{singlespace}

%\clearpage
%
%\vspace*{7cm}
%
%\begin{center}
%{\Large \bf An Open Source Framework for Multivariate $n$-Tissue Segmentation and Volumetric Cortical Parcellation:  Evaluation Using the BrainWeb, NIREP, and LPBA Data Sets}\\
%\end{center}

\clearpage

\begin{abstract} 
Neuroanatomical coordinate systems are essential for the interpretation of 
structural and functional imaging studies.  However, manual delineation of the 
sulco-gyral complex is time consuming and prone to variability given cortical 
complexity.  This work describes an open source, image-based algorithmic approach to 
cortical parcellation which uses training data to propagate structural labelings to 
individual images.  Along with other neuroinformatics tools publicly distributed in our
ANTs (Advanced Normalization Tools)%
\footnote{
http://www.picsl.upenn.edu/ANTs
}
package, {\em Atropos} is an ITK-based multivariate
$n$-class Expectation Maximization (EM) segmentation algorithm which uses 
either parametric or non-parametric finite mixture modelings (FMM) of the class intensities.  
It is capable of incorporating prior labelings, prior probability maps, and/or 
Markov Random Field (MRF) modeling assumptions as spatial priors and geodesically 
propagating segmentation information across neuroanatomy to ensure a 
dense labeling.  Atropos has also been efficiently implemented to handle 
large quantities of possible labelings (in the experimental section, we use
up to 56 classes) with a minimal memory footprint.
In addition to describing the technical and implementation aspects of Atropos, 
this work evaluates its performance on the parcellation problem given two different
training datasets, the LPBA40 dataset from the University of California, Los Angeles
and the NIREP dataset
from University of Iowa.  We evaluate three aspects of the problem: (1) the quality
of automatic cortical extraction from raw data; (2) the quality of
parcellation, given a ground truth cortical extraction; (3) the
quality of automated cortical extraction followed by automated
cortical parcellation.

%Component (3) corresponds to the most
%realistic clinical case, while component (2) eliminates the confound
%of defining the cortex itself and focuses only on the parcellation
%problem.  Consistent with the open source dissemination of ANTs, we
%make publicly available all scripts and code from the evaluation.

\end{abstract}

%\begin{keyword}
%segmentation \sep expectation maximization \sep spatial prior \sep brain 
%\end{keyword}

\clearpage

\section{Introduction}

Establishment of anatomical and functional correspondence, both inter- and intra-subjectively, is
vital in neurodevelopmental and neurological research.  With advancements in image acquisition
 technology, significant investment has been made towards development of
 classification techniques for the cerebrum beginning 
with the seminal work of \cite{Vannier1985} in which NASA satellite image processing software was 
appropriated for statistical classification of the various head tissues in 2-D MR images.  Since
then, a proliferation of techniques has ensued with increasing sophistication.  A chronology of 
this development can be inferred from the relevant reviews of the contemporaneous state-of-the-art
\citep{Bezdek1993,Pal1993,Clarke1995,Pham2000,Viergever2001,Suri2002,Duncan2004,Balafar2010}.
Concomitant with the distillation of technical matters found in the various papers and reviews are
 important evaluation studies of these segmentation methods \citep[e.g.][]{Cuadra2005,Zaidi2006,Klauschen2009,Boer2010}.

Motivations for such algorithmic development are historically rooted in delineating the white and grey matters and the cerebrospinal fluid.
Following the early statistical approach of \cite{Vannier1985},  other researchers followed 
suit in adopting statistical methods for $n$-tissue anatomical brain segmentation.
Given the ``missing data'' aspect of this problem, a natural optimization strategy is the 
use of the Expectation-Maximization (EM) framework \citep{Dempster1977}.  The work described in \cite{Wells1996} 
was one  of the first to incorporate EM to find a locally optimal solution by iterating between
bias field estimation and tissue segmentation.  A core component of this work was the 
explicit modeling of the tissue intensity values as normal distributions \citep{Cline1990} 
for both 2-D univariate simulated data and T1 coronal images which continues to find utility in contemporary developments.   A secondary component was an 
extended nonparametric approach, also influenced by earlier work \citep{Kikinis1992}, where 
Parzen windowing is used to model the tissue intensity distribution omitting consideration of the underlying bias
field.  Although technically not an EM-based algorithm, the robustness of the latter approach has motivated its continued use even in more recent approaches \citep[e.g.][]{Weisenfeld2009}.

Subsequent development included the use of Markov Random Field (MRF)  modeling \citep{Geman1984}
to regularize the classification results \citep{Held1997} with later work adding heuristics concerning neuroanatomy to prevent 
over-regularization and the resulting loss of fine structural details \citep{Leemput1999,Leemput1999a}.  
A more formalized integration of generic MRF spatial priors was employed in the work of \cite{Zhang2001}, 
commonly referred to as FAST (FMRIB's Automated Segmentation Tool), which is in widespread use
given its public availability and good performance.  More recently, a uniform distribution of local MRFs within the brain volume and their subsequent integration into a global solution has been proposed obviating the need for an 
explicit bias correction solution \citep{Scherrer2009}.

Given the local optimization susceptibility characteristic of EM algorithms, various initialization schemes have been proposed.  Common low-level initialization strategies include uniform
probability assignment \citep{Wells1996}, Otsu thresholding \citep{Zhang2001}, and K-means clustering \citep{Pappas1992}.  More sophisticated low-level initialization schemes include that of \cite{Greenspan2006} in which a dense spatial distribution of Gaussians is used to capture the complex neuroanatomical layout with subsequent processing used to conjoin subsets of Gaussians belonging to the same tissue classes.  Recently, anatomically specific strategies for incorporating prior knowledge  into the segmentation solution have included the use of spatial prior probability maps  of the various structures of interest \citep{Leemput1999a,Marroquin2002,Ashburner2005}.  These spatial prior probability maps can also be used to provide an
initial segmentation.   Related technological developments encompasses those driven by partial volume modeling for increased accuracy in brain segmentation \citep{Ruan2000,Ballester2002,Leemput2003}.

Observations regarding a general trend towards a more integrative view 
of neuroanatomical image processing led to the work described in \cite{Ashburner2005} which is
publicly available as SPM---a large-scale Matlab module in which registration, segmentation, and bias field 
correction can be simultaneously modeled within a single optimization scheme. The roots of this very popular 
software package stem back to early seminal work by Karl Friston in which the basis for statistical parametric mapping was developed \citep{Friston1990} and subsequently provided to the research community. A similar holistic approach
to brain processing was provided in \cite{Pohl2006} in which segmentation
and registration parameters were optimized simultaneously while casting the inhomogeneity model parameters
of \cite{Wells1996} as nuisance variables.  Continued work involved recursive parcellation of the brain volume by considering sub-structures in a hierarchical manner \citep{Pohl2007}.  An implementation is provided in  3D slicer \cite{Pieper2006}---an open source package with developmental contributions from multiple agencies including both private and academic institutions.

Related neuroanatomical research concerns the selection of geometric features of the cortex \citep[e.g.][]{Goualher1999} with aims at understanding the functional-anatomical relationship of the human brain. Recent endeavors produce a dense cortical labeling in which every point of the cortex is classified, i.e. a cortical parcellation \citep{Fischl2004,Heckemann2006,Destrieux2010}.  Given the extreme time-consumption and tedium associated with manual effort towards the latter, various algorithms have been proposed of which a small subset has been publicly availed to the research community such as the popular software package known as Freesurfer \citep{Dale1999,Fischl1999,Fischl2004}.  In contrast to the volumetric approach detailed in this work, Freesurfer is primarily a surface-based technique in which the brain structures such as the grey-white matter interface and pial surfaces are processed, analyzed, and displayed as tessellated surfaces  \cite{Dale1999,Fischl1999}.  Advantages of surface-based approaches include the ability to map processed neuroanatomy to simple geometric primitives such as spheres or planes and the ease of including topological constraints in the analysis workflow.

{\color{red}{{\em Needs work}:  Mindboggle, Heckemann  One advantage of the pipeline we propose is that we parcellate
cortex entirely in the image space, thus avoiding the difficulty of
transferring labels from the mesh space back to the image space---a
problem that is a confound of surface-based methods\citep{Klein2010}.}}

Our open source segmentation tool, which we have dubbed {\em Atropos},%
\footnote{
Atropos is one of the three Fates from Greek mythology characterized by her dreaded shears used to decide the destiny of each mortal.  Also, consistent with the entomological motif of our ANTs toolkit, {\it Acherontia atropos} is a species of large moth known for the skull-like pattern visible on its thorax.
}
takes advantage of the $n$-tissue paradigm for volumetric image segmentation and harnesses its generalization for template-based cortical parcellation.
Atropos is an EM-based algorithm which contains many of the features of its predecessors for performing
$n$-tissue segmentation including imposition of prior information in the form of MRFs and template-based prior probability maps.  The latter is extended in this work through the formulation of a distance template-based prior used for EM label-propagation and refinement performed in an anatomically meaningful way.

Atropos addresses the bias field issue by incorporating the recently developed N4 bias correction algorithm \citep{Tustison2010} in an adaptive manner.  Coupled with the registration \citep{Avants2010b} and template building strategies \citep{Avants2010} already included in the ANTs toolkit,   
Atropos is a versatile and powerful software tool which touches multiple aspects of our brain processing pipeline from brain extraction \citep{Avants2010a} to grey matter/white matter/cerebrospinal fluid segmentation to label fusion and propagation for achieving a refined cortical parcellation based on both prior information and subject-specific image data.

Despite its versatility, we limit our Atropos evaluation concentrating on the latter task of image-based cortical parcellation.  Consistent with our advocacy of open science (not to mention the facilitation of analysis due to accessibility) we perform our evaluation analysis using public data sets.  Organization of this work is as follows:  we first describe the theory behind the various algorithmic components of Atropos while acknowledging that theoretical discussion is available elsewhere, e.g. in previously cited literature.  This is followed by a thorough discussion of implementation which is something not typically discussed but yet is of immense practical utility.  We then showcase Atropos' performance on public data sets in comparison with select algorithms previously mentioned.  Finally, we provide a discussion of our work both in the context of the remainder of this paper and in the larger context of previous and future work.

%\begin{comment}{

%  We evaluate this
%method on BrainWeb 20 data and show that by relying upon accurate,
%high-resolution diffeomorphic image registration, Atropos ~is capable of
%reliably segmenting the BW20 whole head data into separate tissue
%classes for muscle, skin, skull and bone marrow as well as the
%standard three tissues, cerebrospinal fluid and gray and white matter.
%Atropos 's three-tissue performance compares favorably with the standard
%EM-MRF algorithm, FAST.  We also use the OASIS dataset to show that
%Atropos ~increases segmentation repeatability relative to FAST.  The
%evaluation data, the data-derived templates, the code and application
%itself are publicly available.
%}
%\end{comment}

\section{Theoretical Foundations for Atropos Segmentation}

Atropos encodes a family of Bayesian segmentation techniques that may be configured in an application-specific manner.  The theory underlying Atropos dates back 20+ years and is representative of some of the most innovative work in the field.  Although we synthesize some of the theoretical work in this section, we recommend to the interested reader consultation with the literature cited previously for further insight.

Bayes' theorem provides a powerful mechanism for making inductive inferences assuming the availability of quantities defining the relevant conditional probabilities, specifically the likelihood and prior probability terms.  A Bayesian paradigm for brain image segmentation utilizes a user-selected observation model defining the likelihood term and one or more prior probability terms whose product is proportional to the posterior probability.  Whereas the likelihood term has been previously defined parametrically (e.g. a Gaussian model) or non-parametrically (e.g. Parzen windowing of the sample histogram), the prior term, as given in the literature, has been formed inclusively either as MRF-based or template-based.  An image segmentation solution in this context is an assignment of one label to each voxel%
\footnote{
In the classic 3-tissue segmentation case, each voxel in the brain region is assigned a label of `cerebrospinal fluid (csf)', `gray matter (gm)', or `white matter (wm)'. 
}
such that the posterior probability is maximized.  Following introductory notation, the following sections provide a formal description of these three essential components, viz.
\begin{itemize}
  \item the likelihood or observation model(s),
  \item the prior probability quantities derived from a generalized MRF, template-based prior terms, and a novel distance prior used for label propagation, and 
  \item the optimization framework for maximizing the posterior probability.
\end{itemize}

\subsection{Notation}

Assume a field, $\mathcal{F}$, whose values are known at discrete locations, i.e. sites, within a regular voxel lattice, $\mathcal{I}$.  Note that $\mathcal{F}$ can be a scalar field in the case of unimodal data (e.g. T1 image only) or a vector field in the case of multimodal data (e.g. T1, T2, and proton density images).
A specific set of observed values, denoted by $\mathbf{y}$, are indexed at $N$ discrete locations in $\mathcal{I}$ by $i \in \{1, 2, \ldots, N\}$.  It is assumed that the complete set of observations is a random field $Y = \{y_1, y_2, \ldots, y_N \}$.    A labeling of $\mathcal{I}$, 
also known as a hard segmentation of $\mathcal{I}$,
assigns to each site in $\mathcal{I}$ one of $K$ labels from the finite set $\mathcal{L} = \{l_1, l_2, \ldots, l_K\}$.  Also considered a random field, we denote this discrete labeling by $X = \{x_1, x_2, \ldots, x_N\}$.

%Furthermore, it is assumed that $\mathcal{F}$ is a random field whose true state, $\mathcal{S}$, is `hidden' beneath its observed values but fully represented by $\mathcal{L}$.%
%\footnote{
%For example, in the classic 3-tissue segmentation problem $\mathcal{L}$ is defined as the set $ \{\mathrm{csf}, \mathrm{gm}, \mathrm{wm}\}$ and each voxel in the brain region is assigned one and only one of these three labels.  
%}

\subsection{Likelihood or Observation Models}

To each of the $K$ labels corresponds a single probabilistic model describing the variation of $\mathcal{F}$ over $\mathcal{I}$.  We denote this set of $K$ likelihood models as $\Phi = \{p_1, p_2, \ldots, p_K\}$.  Using the standard notation, $\mathrm{Pr}(S=s) = p(s)$, $\mathrm{Pr}(S=s|T=t) = p(s|t)$, we can define these voxelwise probabilities, $\mathrm{Pr}_k( Y_i = y_i | X_i = l_k ) = p_k(y_i|l_k)$, in either parametric or non-parametric terms.   Given its simplicity and good performance, in the parametric case, $p_k$ is typically defined as a normalized distribution, i.e.
\begin{align}\label{eq:param}
  p_k\left(y_i|l_k\right) &= G\left(\mu_k;\sigma_k\right) \nonumber \\
                    &= \frac{1}{\sqrt{2\pi \sigma_k^2}}\exp\left( \frac{ -(y_i - \mu_k)^2 }{2\sigma_k^2} \right)
\end{align}
where the parameters $\mu_k$ and $\sigma_k^2$ respectively represent the mean and variance of the $k^{th}$ model.  In the case of multi-modal data, where $y_i$ is a vector quantity, a mulitvariate Gaussian is used defined by the mean vector, $\boldsymbol{\mu}_k$, and covariance matrix, $\Sigma_k$.  An initial estimate for these model parameters are tabulated during the initialization process, e.g. K-means or Otsu thresholding.

A common technique for the non-parametric variant is to define $p_k$ using Parzen windowing of the sample observation histogram of $\mathbf{y}$, i.e.
\begin{align} \label{eq:nonparam}
  p_k\left(y_i|l_k\right) &= \frac{1}{N_B} \sum_{j=1}^{N_B} \frac{1}{\sqrt{2\pi \sigma_h^2}}\exp\left( \frac{ -(y_i - c_j)^2 }{2\sigma_h^2} \right)
\end{align}
where $N_B$ is the number of bins used to define the histogram of the sample observations (in Atropos the default is $N_B = 32$) and $c_j$ is the center of $j^{th}$ bin in the histogram.  $\sigma_h$ is the width of each of the $N_B$ Gaussian kernels.  For multi-modal data in which the number of components of $y_i$ is greater than one, a Parzen window function is constructed for each component.  The likelihood value is determined by the joint probability given by their product.

In order to calculate the probability associated with the entire set of observations, $\mathbf{y}$, we assume independency between voxels which is done primarily for facilitating quantification as opposed to modeling realistic assumptions.  Spatial interdependency between voxels is modeled by the prior probabilities discussed in the next section.  Marginalizing over the set of possible labels, $\mathcal{L}$, leads to the following probabilistic formulation
\begin{align}\label{eq:likelihood}
  p(\mathbf{y}) = \prod_{i=1}^N \left(      
                                                        \sum_{k=1}^K \gamma_k p_k(y_i|l_k)
                                                        \right)
\end{align}
where $\gamma_k$ is the mixing parameter.

\subsection{Prior Probability Models}

By modeling $\mathcal{F}$ via the set of observation models $\Phi$, this so called finite-mixture model could be used to produce a labeling or segmentation \citep[e.g.][]{Wells1996}.  However, as pointed out by \cite{Zhang2001}, the exclusive use of the observation models in modeling the intensity profile produces a less than optimal solution as spatial contextual considerations are completely removed.  This has been remedied by the introduction of a host of  prior probability models including those characterized by their use of MRF theory and template-based information.  For example, in the works of \cite{Leemput1999a} and \cite{Weisenfeld2009}, the original global prior term given in \cite{Wells1996} is replaced by the product of the template-based and the MRF-based prior terms.  In addition to their descriptions below, we discuss our introduction of a third possible prior term in the form of an extension of template-based Euclidean and geodesic distance label propagation.

\subsubsection{Generalized MRF Prior}
In the labeling of a single voxel, an appealing strategy for incorporating spatial coherence into the segmentation solution is to consider the labeling of the voxel neighborhood and favoring labeling configurations in which voxel neighborhoods tended towards homogeneity. This intuition is formally described by MRF theory in which spatial interactions in voxel neighborhoods can be modeled \citep{Li2001}.

In order to apply MRF theory,  we assume the random field introduced earlier, $X$, is an MRF and is, therefore, characterized by a neighborhood system, $\mathcal{N}_i$, on the lattice, $\mathcal{I}$, composed of the neighboring sites of $i$.  This neighborhood system is both noninclusive, i.e. $i \notin \mathcal{N}_i$, and reciprocating, i.e. $i \in \mathcal{N}_j \Leftrightarrow j \in \mathcal{N}_i$.   As an MRF, $X$ also satisfies the positivity and locality conditions, i.e., respectively
\begin{align} \label{eq:mrf}
  p(\mathbf{x}) > 0, \,\, \forall\mathbf{x}
\end{align}
where $\mathbf{x}$ is any particular labeling configuration on $X$ (in other words, any labeling permutation on $X$ is {\em a priori} possible) and 
\begin{align}
  p\left(x_i | x_{\mathcal{I}-\{i\}}\right) = p\left(x_i | x_{\mathcal{N}_i}\right)
\end{align}
where $x_{\mathcal{I}-\{i\}}$ is the labeling of the entire image lattice except at site $i$ and  $x_{\mathcal{N}_i}$ is the labeling of $\mathcal{N}_i$.  This latter property enforces solely local considerations based on the neighborhood system in calculating the probability of the particular configuration, $\mathbf{x}$.  Due to these two assumption, the Hammersley--Clifford theorem provides the basis for treating the MRF distribution (cf. Eqn. (\ref{eq:mrf})) as a Gibbs distribution \citep{Besag1974,Geman1984}, i.e.
\begin{align} \label{eq:gibbs}
p(\mathbf{x}) = Z^{-1} \exp\left(-U(\mathbf{x})\right).
\end{align}
$Z$ is a normalization factor known as the {\em partition function} and $U(\mathbf{x})$ is the {\em energy function} which can take several forms \citep{Li2001}. In Atropos, as is the case with many other segmentation algorithms of the same family, we choose $U(\mathbf{x})$ such that it is only composed of pairwise interactions between neighboring sites,%
\footnote{
Using a more expansive definition of  $U(\mathbf{x})$, 
\begin{align}
U(\mathbf{x}) = \sum_{i = 1}^N \left( V_i(x_i) + \beta \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j ) \right), 
\end{align}
would permit casting the other prior terms inside the definition of $U(\mathbf{x})$ in the form of the external field $V_i(x_i)$ but, for clarity purposes, we consider them separately. 
}
i.e.
\begin{align}\label{eq:U}
U(\mathbf{x}) = \beta \sum_{i = 1}^N \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j )
\end{align}
where $V_{ij}$ is typically defined in terms of the Kronecker delta, $\delta_{ij}$, based on the Ising potential or Potts model \citep{Besag1974}
\begin{align}
V_{ij}(x_i, x_j) &= \delta_{ij} \nonumber \\
                          &= \left\{
                          \begin{array}{ll}
                            1 & \text{if } x_i = x_j \\
                            0 & \text{otherwise}
                          \end{array}
                         \right.   
\end{align}
and $\beta$ is a granularity term which weights the contribution of the MRF prior on the segmentation solution.
Since Atropos allows for neighborhood systems in which not just the immediate face-connected neighbors are considered, we use the modified function also used in \cite{Noe2001}, which weights the interaction term by the Euclidean distance, $d_{ij}$,  between interacting sites $i$ and $j$ such that 
\begin{align}
  V_{ij} = \frac{\delta_{ij}}{d_{ij}}
\end{align}
so that sites in the neighborhood closer to $i$ are weighted more heavily than sites further away.

\subsubsection{Template-Based Priors}
A number of researchers have used template-based strategies to both ensure spatial coherence and incorporate prior knowledge in obtaining a sutiable segmentation.  A common technique is to select labeled subjects from a population from which a template is constructed \citep[e.g.][which is also available in the ANTs toolkit]{Avants2010}.  Each labeling can then be warped to the template where the synthesis of warped labeled regions produces a prior probability map or prior label map encoding the spatial distribution of labeled anatomy which can be harnessed in joint segmentation/registration or, specifically, Atropos/ANTs strategies involving unlabeled subjects.

We employ the strategy given in \cite{Ashburner2005} in which the stationary mixing proportions, $\mathrm{Pr}(x_i = l_k) = \gamma_k$, describing the prior probability that $l_k$ corresponds to a particular voxel, regardless of intensity, are replaced by the following spatially varying mixing proportions
\begin{align}
\mathrm{Pr}(x_i = l_k) = \frac{\gamma_k t_{ik}}{\sum_{j=1}^K\gamma_j t_{ij}}
\end{align}
where $t_{ik}$ is the prior probability value due to the template-based prior information at site $i$.

\subsubsection{Distance Prior for Label Propagation}

In order to provide a dense segmentation even in regions where the template-based prior probability is 0, we use the template-based prior label images or probability maps described previously to formulate an optional distance prior probability.
Consider a subset of $\mathcal{I}$, which we denote $R_k$, defined by $\mathrm{Pr}(x_i = l_k) > \mathrm{Pr}(x_i = l_j) \,\,\forall j \in K, j \neq k$ and $\mathrm{Pr}(x_i = l_k) > 0$. 
We also denote the boundary of $R_k$ as $\partial R_k$ and the `peak' as $\rho_{R_k}$ where $i$ and $j$ are sites in $R_k$:
\begin{align}
  \rho_{R_k} = \argmax_{i \in R_k} \left\{ \min_{j \in \partial {R_k}} d\left(i,j\right)\right\}.
\end{align}
In other words, the peak is an interior site (or multiple interior sites) in $R_k$ characterized as having the maximum distance of all sites to any corresponding closest point on the boundary of $R_k$.  We denote this max-min distance as $\Delta_k$.  This allows us to define the distance prior probability for label propagation as follows:
\begin{align}\label{eq:prop}
  \mathrm{Pr}(x_i = l_k) = \left\{
                                               \begin{array}{ll}
                                                \alpha_k \exp \left( -\frac{d(i, \partial R_k)}{\sigma_k} \right) & \text{if } i \text{ is outside the region } R_k \\
                                                \alpha_k + d\left(i, \partial R_k\right)\left(\frac{1.0 - \alpha_k}{\Delta_k}\right) & \text{otherwise}
                                                \end{array}
                                                \right.
\end{align}
The probability calculations are illustrated in Fig.~\ref{fig:distancePrior} and were designed so that the most interior point, or set of points, of region $R_k$ would have a distance prior probability of 1.0 which would linearly decrease within the region to the boundary with a user-selected probability value of $\alpha_k$.  Outside $R_k$, the distance prior exponentially decays with a decay rate governed by another user-selected parameter $\sigma_k$.  This permits a dense segmentation throughout the segmentation region of $\mathcal{I}$ even if the template prior probabilities are zero.

\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[width=150mm]{distancePrior.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small Graphical representation of the distance prior probability calculation associated with label $l_k$.  This extension of the popular template-based prior probability strategy permits a dense labeling with the user specified mask. }
\label{fig:distancePrior}
\end{center}
\end{figure}

A crucial matter, particularly for propagating labels in the cortical regions, is the choice of distance function.  One of the advantages to surface-based methods is that distances can be calculated along the surface mesh, thus discriminating between proximal cortical surface points and those points which are situated close together due to cortical folding.  In Atropos, we provide two possibilities---(1) a geodesic distance function using a fast marching construction \citep{Osher1988} of the distance function over the image and (2) a Euclidean distance function using the algorithm described in \cite{Maurer2003}.  The differences in label propagation results are illustrated in Fig.~\ref{fig:label_propagation}.

\begin{figure}
\begin{center}
\begin{tabular}{ccc}
\includegraphics[width=50mm]{sparse_labels.png} &
\includegraphics[width=50mm]{prop_euclidean.png} &
\includegraphics[width=50mm]{prop_geodesic.png} \\
(a) & (b) & (c) \\
\end{tabular}
\caption{\baselineskip 12pt \small  Illustration of label propagation differences between the Euclidean and geodesic approaches which
are particularly acute in sinuous structures such as the cortex.  (a) Given a mask, represented in white, and a sparse labeling,
Atropos can be used to propagate the labels in a dense manner throughout the masked region.  (b) This propagation can occur using Euclidean distances in the image space via an Euclidean distance transform which can cause labels to jump meaningful anatomical boundaries.   (c) Given the cortical geometry, a more intuitive approach would be to propagate the labelings in a geodesic manner using a product of the fast marching paradigm which can be used in a manner that respects anatomical boundaries.  }
\label{fig:label_propagation}
\end{center}
\end{figure}

\subsection{Optimization}
Given the likelihood models and prior probabilities whose product is proportional to the posterior probability, the Bayesian segmentation solution is the labeling $\hat{\mathbf{x}}$ which maximizes the posterior probability, i.e.
\begin{align}
  \hat{\mathbf{x}} = \argmax_{\mathbf{x}} \left\{p(\mathbf{y}|\mathbf{x})p(\mathbf{x})\right\}.
\end{align}
Similar to its algorithmic predecessors, Atropos employs the EM framework \citep{Dempster1977} which, after initial estimation of the likelihood model parameters, iterates between calculation of the missing $\hat{\mathbf{x}}$ and subsequent re-estimation of the model parameters by maximizing the expectation of the complete data log-likelihood (cf. Eqn. (\ref{eq:likelihood})) 
%\begin{align}
%\log( p(\mathbf{y}) ) = \sum_{i=1}^N \log \left( \sum_{k=1}^K \gamma_k  p(\mathbf{y}|l_k)  \right)
%\end{align}
with the maximization of its expectation derived in various publications including \cite{Zhang2001} which yields the optimal mean and variance (or covariance).  The derivation of $\gamma_k$ at each iteration follows that provided in \cite{Ashburner2005}.%
\footnote{
Due to the lack of parameters in the non-parametric approach, it is not technically an EM algorithm (as described in \cite{Wells1996}).  However, the same iterative strategy is applicable and is quite robust in practice as evidenced by the number of researchers employing non-parametric strategies (see the Introduction).
}

With spatial coherence constraints given by the MRF prior, the optimal segmentation solution becomes intractable.%
\footnote{
Consider $N$ sites each with a possible $K$ labels for a total of $N^K$ possible labeling configurations.  Due to the large $K$ associated with cortical labeling problems, exact optimization is even more problematic than for the traditional 3-tissue scenario.
}
Although many optimization techniques exist \citep[see the introduction in][for a concise summary of the myriad optimization possibilities]{Marroquin2002} each with their characteristic advantages and disadvantages in terms of computational complexity and accuracy,  Atropos utilizes the well-known greedy strategy known as {\em Iterated Conditional Modes} (ICM)  \citep{Besag1986} which is computationally efficient and provides good performance.

%\begin{align}
%P(l_k | y_i, t_i, m_i,d_i) &= \frac{P(y_i|l_k ,t_i,m_i,d_i)   P(l_k|t_i,m_i,d_i)   }{P(y_i|t_i,m_i,d_i)}
%\end{align}

%An image, $I$, maps a domain, $\Omega$ into the
%positive real numbers, such that $I \colon \Omega \rightarrow
%\mathbb{R}^+$.  The goal of segmentation, in general, is to define the
%spatial distribution of a finite set of labels over this domain.  We
%denote the segmentation itself as $\eta \colon \Omega \rightarrow L$
%where $L = \{ L_1 = 1 , \cdots , L_N=N \}$, a set of integer indexed
%segmentation labels.  Note that $\eta$ may be formed as $\eta =
%\sum_{i=1}^{i=N} L_i \eta_i $ where $\eta_i$ is the binary
%segmentation for label $L_i$.  A prior estimate for the label image
%$\eta_i$ is here denoted $\eta^s_i$ with a complete set of priors
%denoted $\eta^s$.  The boundary of the binary segmentation -- where a
%$0/1$ transition edge exists -- is denoted $\partial \eta^s_i$, for
%the prior, and $\partial \eta_i$ for the label image.
%
%\subsection{Apocrita Theory} A general maximum a posteriori criterion
%for segmentation seeks,
%\begin{equation} {\hat \eta} = \argmax _{\eta} \Pr( \eta | I )(\x) =
%\Pr( I | \eta )(\x) \Pr(\eta)(\x),
%\end{equation} 
%where $I$ is the input image, ${\eta}$ represents the
%label set configuration taken from the set $L$, $\Pr$ is the
%probability, $\x$ is the spatial index and the optimal solution is
%$\hat \eta$.  The input image $I$, here, is an unlabeled T1 MRI
%indexed by the value $\x \in \Omega$ where $\Omega$ is the image's
%spatial domain.  This probability is composed of the likelihood (first
%term) and the prior (second term).  Atropos ~ uses a spatially varying
%likelihood term and a two-component prior term that takes into account
%both spatial distribution and label smoothness, the latter via a
%standard MRF prior.
%
%The likelihood term for a single label value $L_i \in L$ is,
%\begin{eqnarray} \Pr( I | \eta_i )(\x)=\frac{1}{Z_i} \exp(- \| I(\x) -
%\mu_i(\x) \|^2 / \sigma_i^2 ),
%\end{eqnarray} where $Z_i$ is a normalizing constant, $\mu(\x)$ is a
%spatially varying estimate of the tissue mean and $\sigma$ is a
%standard deviation.  The prior term is given by,
%\begin{eqnarray} \Pr(\eta_i)(\x) = \frac{1}{Z^\prime_i} \exp( -f( \x -
%\y_{\partial \eta^s_i} ) / \sigma_{\eta^s_i}^2 ) p(\eta_i |
%\eta^\aleph_i), \\ \notag f( \x - \y_{\partial \eta^s_i} ) = (1 -
%\eta^s_i (\x) ) \| \x - \y_{\partial \eta^s_i} \|,
%\end{eqnarray} where $p(\eta_i | \eta^N_i)$ is the MRF smoothness
%probability based on the local neighborhood $\aleph$, the $Z$ is a
%normalizing constant and $\y_{\partial \eta^s_i}$ is the nearest point
%to $\x$ on the boundary of this labeling. Atropos ~requires a user or
%template-defined $\eta^s_i$ and the standard deviation
%$\sigma_{\eta^s_i}$ if a non-unity spatial prior component is desired
%for that label.  The free parameters, that must be estimated
%iteratively, are therefore $\mu_i, \sigma_i$ and the label set itself
%$\hat \eta$, which defines the (locally) optimal spatial distribution
%of $L$ through $\Omega$.  Figure~\ref{fig:spatp} shows the
%distribution of the spatial prior as a function of the distance from
%prior-defined object boundary.  Note that $f$ may easily be varied for
%other applications or that fixed probability images may also be
%substituted here.  This choice of $f$ is motivated by the fact that it
%allows compressed storage of the priors in a single image, $\eta^s$,
%while also maintaining the ability to manipulate -- for each
%$\eta^s_i$ -- the spatial influence of the prior via
%$\sigma_{\eta^s_i}$.  Practically, this is especially valuable when,
%$N$, the number of labels, is large.

\section{Implementation}

\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[width=160mm]{AtroposFlowchart.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small Flowchart illustrating Atropos usage typically beginning with bias correction using the N4 algorithm.  
Initialization provides an estimate before the iterative segmentation portion of the algorithm in which the 
likelihood models for each class are tabulated from the current estimate followed by a recalculation of the 
posterior probabilities associated with each class.  Label propagation ensues to ensure a dense 
classification throughout the entire mask.   The multiple options associated with the different algorithmic 
components are indicated by the colored rounded rectangles connected to their respective core Atropos 
processes via curved dashed lines.  }
\label{fig:flowchart}
\end{center}
\end{figure}

As with other classes that comprise ANTs, Atropos uses the Insight Toolkit as a developmental foundation.  
This allows us to take advantage of the mature portions of ITK (e.g. image IO) and ensures the 
integrity of the ancillary processes such as those facilitated by the underlying statistical framework.  Although Atropos is 
publicly distributed with the rest of the ANTs package, we plan to contribute the Atropos segmentation
framework to the Insight Toolkit where it can be vetted and further improved by other interested researchers.

An overview of the core components of Atropos can be gleaned, in part, from the flowchart depicted in 
Fig.~\ref{fig:flowchart}.  To provide a more intuitive interface without the overhead costs of a graphical user interface, a set of unique command line parsing classes were developed which can also provide insight to the functionality of Atropos.  
The short version of the command line help menu is given in Listing \ref{listing:command} which is invoked by typing `{\ttfamily Atropos -h}' at the command prompt.  Both short and long option flags are available and each option has its own set of possible values and parameters introduced in a more formal way in both previous discussion and related papers cited in the introduction.  Here we  describe these options from the unique perspective of  implementation.

\singlespacing 
\begin{command}
\lstsetcpp
\begin{lstlisting}
COMMAND: 
     Atropos

OPTIONS: 
     -d, --image-dimensionality 2/3/4
     -a, --intensity-image [intensityImage,<adaptiveSmoothingWeight>]
     -b, --bspline [<numberOfLevels=6>,<initialMeshResolution=1x1x...>,<splineOrder=3>]
     -i, --initialization Random[numberOfClasses]
                          KMeans[numberOfClasses]
                          Otsu[numberOfClasses]
                          PriorProbabilityImages[numberOfClasses,
                            fileSeriesFormat(index=1 to numberOfClasses) or vectorImage,
                            priorWeighting,<priorProbabilityThreshold>]
                          PriorLabelImage[numberOfClasses,labelImage,priorWeighting]
     -x, --mask-image maskImageFilename
     -c, --convergence [<numberOfIterations=5>,<convergenceThreshold=0.001>]
     -k, --likelihood-model Gaussian
                            HistogramParzenWindows[<sigma=1.0>,<numberOfBins=32>]
     -m, --mrf [<smoothingFactor=0.3>,<radius=1x1x...>]
     -o, --output [classifiedImage,<posteriorProbabilityImageFileNameFormat>]
     -u, --minimize-memory-usage (0)/1
     -w, --winsorize-outliers BoxPlot[<lowerPercentile=0.25>,<upperPercentile=0.75>,
                               <whiskerLength=1.5>]
                              GrubbsRosner[<significanceLevel=0.05>,<winsorizingLevel=0.10>]
     -e, --use-euclidean-distance (0)/1
     -l, --label-propagation whichLabel[sigma=0.0,<boundaryProbability=1.0>]
     -h 
     --help 
\end{lstlisting} 
\end{command}
\doublespacing

\subsection{Integration of N4 Bias Correction}

The typical processing pipeline begins with an intensity normalization/bias correction
step using the recently developed N4 algorithm \citep{Tustison2010} which is based on the popular nonparametric 
nonuniformity intensity normalization (N3) method introduced in \cite{Sled1998}.    It was previously mentioned that several methods proposed in the literature have taken an integrative view of the segmentation problem by incorporating an intrinsic bias correction step into the actual algorithmic workflow.   The advancements
introduced with the N4 algorithm permit such an adaptive integration with Atropos.  Given recent 
demonstrations suggesting that improved white matter segmentation produces better gain field estimates using N3 
\citep{Boyes2008}, when performing 3-tissue segmentation, we use the posterior probability map of white matter
at the current  iteration as a weighted mask for input to N4.  This is done by setting the `{\ttfamily --weight-image}' option on the N4 command line call (see Listing \ref{listing:n4}) to the posterior probability image corresponding to the white matter produced as output in the Atropos call, i.e. `{\ttfamily Atropos --output}'.

\singlespacing 
\begin{command}
\lstsetcppnfour
\begin{lstlisting}
COMMAND: 
     N4BiasFieldCorrection

OPTIONS: 
     -d, --image-dimensionality 2/3/4
     -i, --input-image inputImageFilename
     -x, --mask-image maskImageFilename
     -w, --weight-image weightImageFilename
     -s, --shrink-factor 1/2/3/4/...
     -c, --convergence [<numberOfIterations=50>,<convergenceThreshold=0.001>]
     -b, --bspline-fitting [splineDistance,<splineOrder=3>,<sigmoidAlpha=0.0>,
                           <sigmoidBeta=0.5>]
                           [initialMeshResolution,<splineOrder=3>,<sigmoidAlpha=0.0>,
                           <sigmoidBeta=0.5>]
     -t, --histogram-sharpening [<FWHM=0.15>,<wienerNoise=0.01>,<numberOfHistogramBins=200>]
     -o, --output [correctedImage,<biasField>]
     -h 
     --help 
\end{lstlisting} 
\end{command}
\doublespacing

\subsection{From Segmentation Theory to Atropos Implementation}

\subsubsection{Initialization}
Atropos was designed to be dimensionally generic in the sense that the same code can be used to handle 2-, 3-, and 
even 4-dimensional image data.  Higher dimensions are possible although we have not yet encountered such 
application-specific need.  Multiple images (assumed to be of the same dimension, size, origin, etc.) can be specified 
on the command line using the `{\ttfamily -a/--intensity-image}' option for each image.  The first image specified on the 
command line is used to initialize the Random, Otsu, or K-means labeling with the latter initialization further
refined by incorporating the additional intensity images, i.e. an initial univariate K-means clustering is determined 
from the first intensity image which, along with the other images, provides the starting multivariate cluster centers for a 
follow-up multivariate K-means labeling.

Much of the ANTs-based processing in our lab (including Atropos) takes advantage of template-building strategies 
\citep{Avants2010} which are also offered  in ANTs.  Since aligned prior probability images and prior label maps are often 
associated with such templates, Atropos can be initialized with these data with their influence regulated by a prior probability
weighting term.  Although prior label maps can be specified as a single multi-label image, prior probability data are often represented as
multiple scalar images with a single image corresponding to a particular label.  For relatively small classifications, such as 
the standard 3-tissue segmentation (i.e. white matter, gray matter, and cerebrospinal fluid), this does not typically present
computational complexities using modern hardware.  However, when considering dense cortical parcellations where the number
of labels can range upwards of 74 per hemisphere \citep{Destrieux2010}, the memory load can be prohibitive if all label images 
are loaded into run-time memory simultaneously.  A major part of minimizing memory usage in Atropos,  which corresponds to the boolean `{\ttfamily -u/--minimize-memory-usage}' option, is the sparse representation of each of the prior probability images.
Motivated by the observation that these spatial prior probability maps tend to be highly localized for large quantities of  
cortical labels, a threshold is specified on the command line (default = 0.0) and only those probability values which exceed 
that threshold are stored in the sparse representation.  During the course of optimization, the prior probability image for a
given label is reconstructed on the fly as needed.  For the NIREP data used in this manuscript where the images are 
on the order of $300 \times 300 \times 256$ and 32 cortical labels, this typically shrinks run-time memory usage from a 
peak of 10+ GB to approximately 1.5 GB.

 \subsubsection{Likelihood or Observation Models}
As mentioned previously in the introduction, different groups have opted for different likelihood models which have included
either parametric, specifically Gaussian, or nonparametric variations.  However, these approaches are similar in that 
they require a list sample of intensity data from the input image(s) and a list of weighting values for each observation 
of the list sample from which the model is constructed.

Once the model constructed, a measurement
observation at a given pixel consisting of a single intensity (for univariate segmentation) or multiple intensities (for multivariate
segmentation), is used to query the model value.  These similarities permitted the creation of a generic plug-in architecture
where classes describing both parametric and nonparametric  observational models are all derived from an abstract 
list sample function class.   Three likelihood classes have been developed, one parametric and two nonparametric,
and are available for usage although one of the nonparametric classes is still in experimental development.  The plug-in architecture even permits mixing likelihood models with different classes during the same run for a hybrid parametric/non-parametric approach although this possibility has yet to be fully explored.

If the Gaussian likelihood model is chosen, the list sample of intensity values and 
corresponding weights comprised of the posterior probabilities are used to estimate the Gaussian model parameters, i.e.
the mean and variance.  For the non-parametric model, the list sample and posteriors are used
in a Parzen windowing scheme on a weighted histogram to estimate the observational model.

\subsubsection{Prior Probability Models}
Consistent with our previous discussion, we offer both an MRF-based prior probability for modeling spatial coherence and the possibility of specifying a set of prior probability maps or a prior label map with the latter extendable to creating a dense labeling.  To invoke the MRF `{\ttfamily -m/--mrf}' option, one specifies the smoothing factor (or the granularity parameter, $\beta$, given in Eqn. (\ref{eq:U}), and the radius of the neighborhood system using the vector notation `{\ttfamily 1x1x1}' for a neighborhood radius of 1 in all 3 dimensions.  Note that this radius prescribes an $L_\infty$ distance and thus includes not just the face-connected neighboring voxels.

To propagate labels with the specification of prior probability images or a prior label image, one needs to select the distance function (set by the `{\ttfamily -e/--use-euclidean-distance}' boolean option) which, by default, is determined by the geodesic distance.  The user also needs to set the label propagation parameters for one or more of the classes.  This  includes both the boundary probability and the exponential decay parameter, respectively $\alpha_k$ and $\sigma_k$ in Eqn. (\ref{eq:prop}).

\section{Comparative Evaluation}

{\color{red}{\em Brian, John}}

\subsection{The LPBA40 Dataset}  
\label{sec:lpba}
The LONI Probabilistic Brain Atlas data, LPBA40, \citep{Shattuck2008} was collected at the North Shore Long
Island Jewish Health System imaging center and is maintained at UCLA.
LPBA40 contains 40 images (20 male $+$ 20 female) from normal, healthy
ethnically diverse volunteers with average age of 29.2 $\pm$ 6.3
years.  Each subject underwent 3D SPGR MRI on a 1.5T GE system
resulting in $0.86 \times 0.86 \times 1.5 mm^3$ images.  Each MRI in
the LPBA40 dataset was manually labeled with 56 independent structures
at the UCLA Laboratory of Neuro Imaging (LONI).  The test-retest
reliability of the labeling, across raters, was reported as a minimum
Jaccard ratio of $0.697$ in the supramarginal gyrus to a maximum of
$0.966$ in the gyrus rectus.  A single labeling of each image is made
available to the public and used, here, as silver-standard data 
for both training and testing in our cross-validation scheme.

\subsection{The NIREP Dataset}  
The non-rigid image registration evaluation project (NIREP
http://www.nirep.org/) is a resource of 16 high quality labeled brain
images at 1$mm^3$.  Each brain was labeled with 32 cortical regions
(16 on each hemisphere) and an additional class of other gray matter
tissue.  Regions vary in size from large (inferior temporal region) to
small (temporal pole, insula gyrus, frontal pole).  The main drawback
is a lack of inter-rater reliability numbers -- in particular because
visual inspection and comparison of labelings reveals a degree of
inconsistency in labeling of particular regions across subjects.
Nevertheless, the NIREP dataset is perhaps the highest quality
evaluation dataset currently available for the cortex.

\section{Discussion}

{\color{red}{{\em Elaborate after including the results.}

A more general advantage which extends beyond the scope of the experimental 
evaluation section of this paper is the flexibility of the Atropos algorithm.   This includes 
not only $n$-tissue segmentation and dense volumetric cortical parcellation,
as reported in this work, but Atropos is also used in conjunction with our ANTs registration tools for robust
brain extraction which has reported good performance in comparison with other
popular, publicly available brain extraction tools \citep{Avants2010a}.  An additional successful
application area is that of ventilation-based segmentation of hyperpolarized helium-3 MRI
\citep{Tustison2010a}.}}

%\paragraph{Acknowledgments}
%{This work was supported in part by NIH ... }

\newpage

\bibliographystyle{neuroinformatics}
\bibliography{atropos}

\end{document}

### 2010-09-28 Documentation/atropos/atropos.tex
initial updates to atropos.tex
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@649 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 }
-package, {\em Atropos} is an ITK-based multivariate
-$n$-class Expectation Maximization (EM) segmentation algorithm which uses 
-either parametric or non-parametric finite mixture modelings (FMM) of the class intensities.  
-It is capable of incorporating prior labelings, prior probability maps, and/or 
-Markov Random Field (MRF) modeling assumptions as spatial priors and geodesically 
-propagating segmentation information across neuroanatomy to ensure a 
-dense labeling.  Atropos has also been efficiently implemented to handle 
-large quantities of possible labelings (in the experimental section, we use
-up to 56 classes) with a minimal memory footprint.
-In addition to describing the technical and implementation aspects of Atropos, 
-this work evaluates its performance on the parcellation problem given two different
-training datasets, the LPBA40 dataset from the University of California, Los Angeles
-and the NIREP dataset
-from University of Iowa.  We evaluate three aspects of the problem: (1) the quality
-of automatic cortical extraction from raw data; (2) the quality of
-parcellation, given a ground truth cortical extraction; (3) the
-quality of automated cortical extraction followed by automated
-cortical parcellation.
+package, {\em Atropos} is an ITK-based multivariate $n$-class
+Expectation Maximization (EM) segmentation algorithm which uses either
+parametric or non-parametric finite mixture modelings (FMM) of the
+class intensities.  It is capable of incorporating prior labelings,
+prior probability maps, and/or Markov Random Field (MRF) modeling
+assumptions as spatial priors and geodesically propagating
+segmentation information across neuroanatomy to ensure a dense
+labeling.  Atropos has also been efficiently implemented to handle
+large quantities of possible labelings (in the experimental section,
+we use up to 68 classes) with a minimal memory footprint.  In addition
+to describing the technical and implementation aspects of Atropos,
+this work evaluates its performance on the parcellation problem given
+two different training datasets, the BrainWeb dataset from Montreal
+Neurological Institute and the Hammers atlas from UCL [FIXME].  We use
+the datasets to evaluate three uses of the Atropos segmentation method: (1) K-means
+segmentation without use of template data; (2) segmentation with prior
+probability maps derived from a group template; (3) N-class EM
+segmentation for parcellating the brain into 68 classes based on
+training data.  These evaluation studies, combined with illustrative
+examples that excercise Atropos options, demonstrate both performance
+and wide applicability of this new segmentation tool.
 
@@
 
-\section{Introduction}
+\section{Introduction} Neurological and neurodevelopmental research
+rely on anatomical and functional correspondence both within and
+between subjects.  As image acquisition technology has advanced,
+concomitant investment has been made towards image classification
+techniques for the cerebrum beginning with the seminal work of
+\cite{Vannier1985} in which NASA satellite image processing software
+was appropriated for statistical classification of head tissues in 2-D
+MR images.  Since then, a proliferation of techniques has ensued with
+increasing sophistication.  The chronology of progress in segmentation
+may be tracked through both technical reviews
+\citep{Bezdek1993,Pal1993,Clarke1995,Pham2000,Viergever2001,Suri2002,Duncan2004,Balafar2010}
+and evaluation studies
+\citep[e.g.][]{Cuadra2005,Zaidi2006,Klauschen2009,Boer2010}.
 
-Establishment of anatomical and functional correspondence, both inter- and intra-subjectively, is
-vital in neurodevelopmental and neurological research.  With advancements in image acquisition
- technology, significant investment has been made towards development of
- classification techniques for the cerebrum beginning 
-with the seminal work of \cite{Vannier1985} in which NASA satellite image processing software was 
-appropriated for statistical classification of the various head tissues in 2-D MR images.  Since
-then, a proliferation of techniques has ensued with increasing sophistication.  A chronology of 
-this development can be inferred from the relevant reviews of the contemporaneous state-of-the-art
-\citep{Bezdek1993,Pal1993,Clarke1995,Pham2000,Viergever2001,Suri2002,Duncan2004,Balafar2010}.
-Concomitant with the distillation of technical matters found in the various papers and reviews are
- important evaluation studies of these segmentation methods \citep[e.g.][]{Cuadra2005,Zaidi2006,Klauschen2009,Boer2010}.
-
-Motivations for such algorithmic development are historically rooted in delineating the white and grey matters and the cerebrospinal fluid.
-Following the early statistical approach of \cite{Vannier1985},  other researchers followed 
-suit in adopting statistical methods for $n$-tissue anatomical brain segmentation.
-Given the ``missing data'' aspect of this problem, a natural optimization strategy is the 
-use of the Expectation-Maximization (EM) framework \citep{Dempster1977}.  The work described in \cite{Wells1996} 
-was one  of the first to incorporate EM to find a locally optimal solution by iterating between
-bias field estimation and tissue segmentation.  A core component of this work was the 
-explicit modeling of the tissue intensity values as normal distributions \citep{Cline1990} 
-for both 2-D univariate simulated data and T1 coronal images which continues to find utility in contemporary developments.   A secondary component was an 
-extended nonparametric approach, also influenced by earlier work \citep{Kikinis1992}, where 
-Parzen windowing is used to model the tissue intensity distribution omitting consideration of the underlying bias
-field.  Although technically not an EM-based algorithm, the robustness of the latter approach has motivated its continued use even in more recent approaches \citep[e.g.][]{Weisenfeld2009}.
+The problem of accurately delineating the white matter, grey matter
+and cerebrospinal fluid of the primate brain continuously spurs
+technical development in segmentation.  Following \cite{Vannier1985},
+many researchers adopted statistical methods for $n$-tissue anatomical
+brain segmentation.  Given the ``missing data'' aspect of this
+problem, a natural optimization strategy is the use of the
+Expectation-Maximization (EM) framework \citep{Dempster1977}.  The
+work described in \cite{Wells1996} was one of the first to use EM for
+finding a locally optimal solution by iterating between bias field
+estimation and tissue segmentation.  A core component of this work was
+explicit modeling of the tissue intensity values as normal
+distributions \citep{Cline1990} for both 2-D univariate simulated data
+and T1 coronal images, which continues to find utility in contemporary
+developments.  A secondary component was an extended nonparametric
+approach, also influenced by earlier work \citep{Kikinis1992}, where
+Parzen windowing is used to model the tissue intensity distribution
+omitting consideration of the underlying bias field.  Although
+technically not an EM-based algorithm, the robustness of the latter
+approach has motivated its continued use even more recently \citep[e.g.][]{Weisenfeld2009}.
 
@@
 
-Despite its versatility, we limit our Atropos evaluation concentrating on the latter task of image-based cortical parcellation.
-Consistent with our advocacy of open science (not to mention the facilitation of analysis due to accessibility) we perform our evaluation analysis using public data sets.
-Organization of this work is as follows:  we first describe the theory behind the various algorithmic components of Atropos while acknowledging that theoretical discussion is available elsewhere, e.g.
-in previously cited literature.
-This is followed by a thorough discussion of implementation which is something not typically discussed but yet is of immense practical utility.
-We then showcase Atropos' performance on public data sets in comparison with select algorithms previously mentioned.
-Finally, we provide a discussion of our work both in the context of the remainder of this paper and in the larger context of previous and future work.
+While Atropos may be applied to multivariate data from arbitrary
+modalities, we limit our Atropos evaluation to tissue-classification
+in T1 neuroimaging in part due to the abundance of ``gold-standard''
+data for this modality.  Consistent with our advocacy of open science
+(not to mention the facilitation of analysis due to accessibility) we
+also only use publicly available data sets.  Organization of this work
+is as follows: we first describe the theory behind the various
+algorithmic components of Atropos while acknowledging that theoretical
+discussion is available elsewhere, e.g. in previously cited
+literature.  This is followed by a thorough discussion of
+implementation which is something not typically discussed but yet is
+of immense practical utility.  Finally, we provide a discussion of our work both in the
+context of the remainder of this paper and in the larger context of
+previous and future work.
 
@@
 \end{align}
-where the parameters $\mu_k$ and $\sigma_k^2$ respectively represent the mean and variance of the $k^{th}$ model.
-In the case of multi-modal data, where $y_i$ is a vector quantity, a mulitvariate Gaussian is used defined by the mean vector, $\boldsymbol{\mu}_k$, and covariance matrix, $\Sigma_k$.
-An initial estimate for these model parameters are tabulated during the initialization process, e.g.
-K-means or Otsu thresholding.
+where the parameters $\mu_k$ and $\sigma_k^2$ respectively represent
+the mean and variance of the $k^{th}$ model.  In the case of
+multi-modal data, where $y_i$ is a vector quantity, a mulitvariate
+Gaussian is defined by the mean vector, $\boldsymbol{\mu}_k$, and
+covariance matrix, $\Sigma_k$.  An initial estimate for these model
+parameters are tabulated during the initialization process,
+e.g. K-means or Otsu thresholding.
 
@@
 \begin{align} \label{eq:nonparam}
-  p_k\left(y_i|l_k\right) &= \frac{1}{N_B} \sum_{j=1}^{N_B} \frac{1}{\sqrt{2\pi \sigma_h^2}}\exp\left( \frac{ -(y_i - c_j)^2 }{2\sigma_h^2} \right)
+  p_k\left(y_i|l_k\right) &= \frac{1}{N_B} \sum_{j=1}^{N_B} \frac{1}{\sqrt{2\pi \sigma_j^2}}\exp\left( \frac{ -(y_i - c_j)^2 }{2\sigma_j^2} \right)
 \end{align}
 where $N_B$ is the number of bins used to define the histogram of the sample observations (in Atropos the default is $N_B = 32$) and $c_j$ is the center of $j^{th}$ bin in the histogram.
-$\sigma_h$ is the width of each of the $N_B$ Gaussian kernels.
+$\sigma_j$ is the width of each of the $N_B$ Gaussian kernels.
 For multi-modal data in which the number of components of $y_i$ is greater than one, a Parzen window function is constructed for each component.
@@
 \end{align}
-where $\gamma_k$ is the mixing parameter.
+where $\gamma_k$ is the mixing parameter \cite{Ashburner2005}.
 
@@
 \end{align}
-where $\mathbf{x}$ is any particular labeling configuration on $X$ (in other words, any labeling permutation on $X$ is {\em a priori} possible) and 
+where $\mathbf{x}$ is any particular labeling configuration on $X$ (in
+other words, any labeling permutation on $X$ is {\em a priori}
+possible).  The MRF locality condition is 
 \begin{align}
-  p\left(x_i | x_{\mathcal{I}-\{i\}}\right) = p\left(x_i | x_{\mathcal{N}_i}\right)
+  p\left(x_i | x_{\mathcal{I}-\{i\}}\right) = p\left(x_i | x_{\mathcal{N}_i}\right),
 \end{align}
 where $x_{\mathcal{I}-\{i\}}$ is the labeling of the entire image lattice except at site $i$ and  $x_{\mathcal{N}_i}$ is the labeling of $\mathcal{N}_i$.
-This latter property enforces solely local considerations based on the neighborhood system in calculating the probability of the particular configuration, $\mathbf{x}$.
-Due to these two assumption, the Hammersley--Clifford theorem provides the basis for treating the MRF distribution (cf.
+This locality property enforces solely local considerations based on the neighborhood system in calculating the probability of the particular configuration, $\mathbf{x}$.
+Following these two assumptions, the Hammersley--Clifford theorem provides the basis for treating the MRF distribution (cf.
 Eqn.
@@
 \end{align}
-$Z$ is a normalization factor known as the {\em partition function} and $U(\mathbf{x})$ is the {\em energy function} which can take several forms \citep{Li2001}.
-In Atropos, as is the case with many other segmentation algorithms of the same family, we choose $U(\mathbf{x})$ such that it is only composed of pairwise interactions between neighboring sites,%
+$Z$ is a normalization factor known as the {\em partition function}
+and $U(\mathbf{x})$ is the {\em energy function} which can take
+several forms \citep{Li2001}. In Atropos, as is the case with many
+other segmentation algorithms of the same family, we choose
+$U(\mathbf{x})$ such that it is only composed of a sum over pairwise
+interactions between neighboring sites across the image,%
 \footnote{
@@
 \end{align}
-where $V_{ij}$ is typically defined in terms of the Kronecker delta, $\delta_{ij}$, based on the Ising potential or Potts model \citep{Besag1974}
+where $V_{ij}$ is typically defined in terms of the Kronecker delta,
+$\delta_{ij}$, based on the Ising potential (also known as a Potts model) \citep{Besag1974}
 \begin{align}
@@
 and $\beta$ is a granularity term which weights the contribution of the MRF prior on the segmentation solution.
-Since Atropos allows for neighborhood systems in which not just the immediate face-connected neighbors are considered, we use the modified function also used in \cite{Noe2001}, which weights the interaction term by the Euclidean distance, $d_{ij}$,  between interacting sites $i$ and $j$ such that
+Since Atropos allows for non-uniform neighborhood systems and systems in which not just the immediate face-connected neighbors are considered, we use the modified function also used in \cite{Noe2001}, which weights the interaction term by the Euclidean distance, $d_{ij}$,  between interacting sites $i$ and $j$ such that
 \begin{align}
@@
 
-We employ the strategy given in \cite{Ashburner2005} in which the stationary mixing proportions, $\mathrm{Pr}(x_i = l_k) = \gamma_k$, describing the prior probability that $l_k$ corresponds to a particular voxel, regardless of intensity, are replaced by the following spatially varying mixing proportions
+We employ the strategy given in \cite{Ashburner2005} in which the
+stationary mixing proportions, $\mathrm{Pr}(x_i = l_k) = \gamma_k$,
+describing the prior probability that label $l_k$ corresponds to a particular voxel, regardless of intensity, are replaced by the following spatially varying mixing proportions
 \begin{align}
@@
 \end{align}
-where $t_{ik}$ is the prior probability value due to the template-based prior information at site $i$.
+where $t_{ik}$ is the prior probability value due to the
+template-based prior information at site $i$.
+{\textcolor{red}{FIXME --- need explanation of heuristic ...  Also, we
+  allow a weighted combination defined by the ComputeLocalPosterior
+  function in Atropos --- how do we justify / explain this?   clearly,
+it's a balance of the global solution (no local prior) and the local
+one with spatial prior ... }}
 
@@
 Outside $R_k$, the distance prior exponentially decays with a decay rate governed by another user-selected parameter $\sigma_k$.
-This permits a dense segmentation throughout the segmentation region of $\mathcal{I}$ even if the template prior probabilities are zero.
+This permits a dense segmentation throughout the segmentation region of $\mathcal{I}$ even where the template prior probabilities are zero.
 
@@
 
-A crucial matter, particularly for propagating labels in the cortical regions, is the choice of distance function.
-One of the advantages to surface-based methods is that distances can be calculated along the surface mesh, thus discriminating between proximal cortical surface points and those points which are situated close together due to cortical folding.
-In Atropos, we provide two possibilities---(1) a geodesic distance function using a fast marching construction \citep{Osher1988} of the distance function over the image and (2) a Euclidean distance function using the algorithm described in \cite{Maurer2003}.
+A crucial matter, particularly for propagating labels in the cortical
+regions, is the choice of distance function.  One of the advantages to
+surface-based methods is that distances can be calculated along the
+surface mesh, thus discriminating between proximal cortical surface
+points and those points which are situated close together due to
+cortical folding.  In Atropos, we provide two possibilities---(1) a
+geodesic distance function using a fast marching construction
+\citep{Osher1988} of the distance function with Atropos's masked
+region of interest and (2) a Euclidean distance function using the algorithm described in \cite{Maurer2003}.
 The differences in label propagation results are illustrated in Fig.~\ref{fig:label_propagation}.
@@
 \end{align}
-Similar to its algorithmic predecessors, Atropos employs the EM framework \citep{Dempster1977} which, after initial estimation of the likelihood model parameters, iterates between calculation of the missing $\hat{\mathbf{x}}$ and subsequent re-estimation of the model parameters by maximizing the expectation of the complete data log-likelihood (cf.
+Similar to its algorithmic predecessors, Atropos employs the EM
+framework \citep{Dempster1977}.  After initial estimation of the
+likelihood model parameters, EM iterates between calculation of the missing $\hat{\mathbf{x}}$ and subsequent re-estimation of the model parameters by maximizing the expectation of the complete data log-likelihood (cf.
 Eqn.
-(\ref{eq:likelihood}))
+(\ref{eq:likelihood})).
 %\begin{align}
@@
 %\end{align}
-with the maximization of its expectation derived in various publications including \cite{Zhang2001} which yields the optimal mean and variance (or covariance).
-The derivation of $\gamma_k$ at each iteration follows that provided in \cite{Ashburner2005}.%
+The expectation maximization procedure is derived in various
+publications including \cite{Zhang2001} which yields the optimal mean
+and variance (or covariance), but sets the mixing parameter $\gamma_k$
+as a constant.  The Atropos implementation estimates $\gamma_k$ at
+each iteration as in \cite{Ashburner2005}.
 \footnote{
@@
 }  
-
-With spatial coherence constraints given by the MRF prior, the optimal segmentation solution becomes intractable.%
+When spatial coherence constraints are included as an MRF prior in Atropos, the optimal segmentation solution becomes intractable.%
 \footnote{
@@
 }
-Although many optimization techniques exist \citep[see the introduction in][for a concise summary of the myriad optimization possibilities]{Marroquin2002} each with their characteristic advantages and disadvantages in terms of computational complexity and accuracy,  Atropos utilizes the well-known greedy strategy known as {\em Iterated Conditional Modes} (ICM)  \citep{Besag1986} which is computationally efficient and provides good performance.
+Although many optimization techniques exist \citep[see the
+introduction in][for a concise summary of the myriad optimization
+possibilities]{Marroquin2002}---each with their characteristic
+advantages and disadvantages in terms of computational complexity and
+accuracy---Atropos uses the well-known {\em Iterated Conditional
+  Modes} (ICM)  \citep{Besag1986} which is greedy, computationally efficient and provides good performance.
 
@@
 publicly distributed with the rest of the ANTs package, we plan to contribute the Atropos segmentation
-framework to the Insight Toolkit where it can be vetted and further improved by other interested researchers.
+framework to the Insight Toolkit where it can be vetted and further
+improved by other interested researchers. 
+{\textcolor{red}{FIXME ---- should we save this for a deliverable in
+    some grant?}}
 
@@
 The advancements
-introduced with the N4 algorithm permit such an adaptive integration with Atropos.  Given recent 
-demonstrations suggesting that improved white matter segmentation produces better gain field estimates using N3 
-\citep{Boyes2008}, when performing 3-tissue segmentation, we use the posterior probability map of white matter
+introduced with the N4 algorithm permit such an adaptive integration with Atropos.  
+Recent demonstrations suggest improved white matter segmentation produces better gain field estimates using N3 
+\citep{Boyes2008}.  Thus, when performing 3-tissue segmentation, we use the posterior probability map of white matter
 at the current  iteration as a weighted mask for input to N4.
@@
 that threshold are stored in the sparse representation.  During the course of optimization, the prior probability image for a
-given label is reconstructed on the fly as needed.  For the NIREP data used in this manuscript where the images are 
+given label is reconstructed on the fly as needed.  {\textcolor{red}For the NIREP data used in this manuscript where the images are 
 on the order of $300 \times 300 \times 256$ and 32 cortical labels, this typically shrinks run-time memory usage from a 
-peak of 10+ GB to approximately 1.5 GB.
+peak of 10+ GB to approximately 1.5 GB. --- FIXME }
 
@@
 of the list sample from which the model is constructed.  
-
-Once the model constructed, a measurement
-observation at a given pixel consisting of a single intensity (for univariate segmentation) or multiple intensities (for multivariate
-segmentation), is used to query the model value.  These similarities permitted the creation of a generic plug-in architecture
+In general, one may query model probabilities by passing a given pixel's single intensity (for univariate segmentation) or multiple intensities (for multivariate
+segmentation) to the modeling function, regardless of whether the
+function is parameter or non-parametric.  These similarities permitted the creation of a generic plug-in architecture
 where classes describing both parametric and nonparametric  observational models are all derived from an abstract 
@@
 the mean and variance.  For the non-parametric model, the list sample and posteriors are used
-in a Parzen windowing scheme on a weighted histogram to estimate the observational model.
+in a Parzen windowing scheme on a weighted histogram to estimate the
+observational model  {\textcolor{red} cite Suyash}.
 
 \subsubsection{Prior Probability Models}
-Consistent with our previous discussion, we offer both an MRF-based prior probability for modeling spatial coherence and the possibility of specifying a set of prior probability maps or a prior label map with the latter extendable to creating a dense labeling.
-To invoke the MRF `{\ttfamily -m/--mrf}' option, one specifies the smoothing factor (or the granularity parameter, $\beta$, given in Eqn.
-(\ref{eq:U}), and the radius of the neighborhood system using the vector notation `{\ttfamily 1x1x1}' for a neighborhood radius of 1 in all 3 dimensions.
-Note that this radius prescribes an $L_\infty$ distance and thus includes not just the face-connected neighboring voxels.
+Consistent with our previous discussion, we offer both an MRF-based
+prior probability for modeling spatial coherence and the possibility
+of specifying a set of prior probability maps or a prior label map
+with the latter extendable to creating a dense labeling.  To invoke
+the MRF `{\ttfamily -m/--mrf}' option, one specifies the smoothing
+factor (or the granularity parameter, $\beta$, given in
+Eqn. (\ref{eq:U}), and the radius (in voxels) of the neighborhood
+system using the vector notation `{\ttfamily 1x1x1}' for a
+neighborhood radius of 1 in all 3 dimensions.  Note that this radius
+prescribes an $L_\infty$  {\textcolor{red} FIXME should this be $L_2$?} distance and thus includes not just the face-connected neighboring voxels.
 

### 2010-10-06 Documentation/atropos/atropos.tex
made some points about significant contributions
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@651 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 \usepackage{geometry}
+\usepackage{url}
 \geometry{letterpaper,top=1.0in,left=1.0in,bottom=1.0in,top=1.0in,headsep=6pt,footskip=18pt}
@@
 These spatial prior probability maps can also be used to provide an
-initial segmentation.   Related technological developments encompasses those driven by partial volume modeling for increased accuracy in brain segmentation \citep{Ruan2000,Ballester2002,Leemput2003}.
+initial segmentation.   Related technological developments encompasses
+those driven by partial volume modeling for increased accuracy in
+brain segmentation \citep{Ruan2000,Ballester2002,Leemput2003}.
 
@@
 
-While Atropos may be applied to multivariate data from arbitrary
-modalities, we limit our Atropos evaluation to tissue-classification
-in T1 neuroimaging in part due to the abundance of ``gold-standard''
-data for this modality.  Consistent with our advocacy of open science
-(not to mention the facilitation of analysis due to accessibility) we
-also only use publicly available data sets.  Organization of this work
-is as follows: we first describe the theory behind the various
-algorithmic components of Atropos while acknowledging that theoretical
-discussion is available elsewhere, e.g. in previously cited
-literature.  This is followed by a thorough discussion of
-implementation which is something not typically discussed but yet is
-of immense practical utility.  Finally, we provide a discussion of our work both in the
-context of the remainder of this paper and in the larger context of
-previous and future work.
+The major contribution of the Atropos algorithm is a flexible, memory
+efficient implementation of core expectation maximization machinery
+that is useful across many segmentation problems.  To highlight the
+value of this open-source, free contribution, we performed a search of
+software attributes on NITRC and found that as of {\textcolor{red}
+INSERT DATE NO STAND-ALONE N-CLASS EM methods are currently listed}.
+Atropos's memory efficiency enables EM-segmentation of many classes at
+once, a capability that is due to a sparse internal representation of
+prior probability images that minimizes the need for random access
+memory.  We evaluate EM-based classification of $1 mm^3$ T1 MR
+neuroimages into 68 neuronatomical regions to illustrate the practical
+value of the low-memory implementation within this paper.  Atropos
+also allows multivariate image features to drive the data term.  Such
+an approach is valuable when more than one view of anatomy aids
+segmentation, as in neonatal brain tissue classification
+\cite{Weisenfeld2009}.  Although Atropos may be applied to
+multivariate data from arbitrary modalities, we limit our evaluation
+to tissue-classification in T1 neuroimaging in part due to the
+abundance of ``gold-standard'' data for this modality.  Consistent
+with our advocacy of open science (not to mention the facilitation of
+analysis due to accessibility) we also only use publicly available
+data sets.
 
-%\begin{comment}{
+Organization of this work is as follows: we first describe the theory
+behind the various algorithmic components of Atropos while
+acknowledging that theoretical discussion is available elsewhere,
+e.g. in previously cited literature.  This is followed by a thorough
+discussion of implementation which, though often overlooked, is of
+immense practical utility.  We then report evaluation results on
+public T1 neuroimaging datasets that have reference gold/silver
+standard labelings.  Finally, we provide a discussion of our results
+and our open-source contribution in the context of the remainder of
+this paper and of previous and future work.
 
@@
 Atropos encodes a family of Bayesian segmentation techniques that may be configured in an application-specific manner.
-The theory underlying Atropos dates back 20+ years and is representative of some of the most innovative work in the field.
+The theory underlying Atropos dates back 20$+$ years and is representative of some of the most innovative work in the field.
 Although we synthesize some of the theoretical work in this section, we recommend to the interested reader consultation with the literature cited previously for further insight.
@@
 
+Figure~\ref{fig:neonate} illustrates the benefit of geodesic distance
+priors in segmenting T2 MRI of the 
+neonatal brain.  Geodesic distance priors may be used to reduce ambiguity at
+tissue interfaces, where partial voluming leads to mixed intensity.  
+In neonatal brain segmentation from T2 MRI, the
+interface of gray matter (low intensity) and cerebrospinal fluid
+(bright intensity) often matches that of white matter (medium intensity). 
+However, the white matter class should never appear near the surface
+of the brain.  Thus, the distance from the brain surface may be used
+to disambiguate voxels such as these.  
+In this case, the distance of a voxel from a known
+or expected tissue location serves to differentiate it from classes
+that have similar intensity signature.  The geodesic distance is therefore an
+alternative to explicit models of partial volume \citep{Ruan2000,Ballester2002,Leemput2003}.
+
 \begin{figure}
@@
 \label{fig:label_propagation}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.9\textwidth]{sparse_labels.png} 
+\caption{\baselineskip 12pt \small  Neonatal brain segmentation.  }
+\label{fig:neonate}
 \end{center}
@@
 
-\section{Comparative Evaluation}
+\section{Evaluation}
 
-{\color{red}{\em Brian, John}}
-
-\subsection{The LPBA40 Dataset}  
-\label{sec:lpba}
-The LONI Probabilistic Brain Atlas data, LPBA40, \citep{Shattuck2008} was collected at the North Shore Long
-Island Jewish Health System imaging center and is maintained at UCLA.
-LPBA40 contains 40 images (20 male $+$ 20 female) from normal, healthy
-ethnically diverse volunteers with average age of 29.2 $\pm$ 6.3
-years.  Each subject underwent 3D SPGR MRI on a 1.5T GE system
-resulting in $0.86 \times 0.86 \times 1.5 mm^3$ images.  Each MRI in
-the LPBA40 dataset was manually labeled with 56 independent structures
-at the UCLA Laboratory of Neuro Imaging (LONI).  The test-retest
-reliability of the labeling, across raters, was reported as a minimum
-Jaccard ratio of $0.697$ in the supramarginal gyrus to a maximum of
-$0.966$ in the gyrus rectus.  A single labeling of each image is made
-available to the public and used, here, as silver-standard data 
-for both training and testing in our cross-validation scheme.
-
-\subsection{The NIREP Dataset}  
-The non-rigid image registration evaluation project (NIREP
-http://www.nirep.org/) is a resource of 16 high quality labeled brain
-images at 1$mm^3$.  Each brain was labeled with 32 cortical regions
-(16 on each hemisphere) and an additional class of other gray matter
-tissue.  Regions vary in size from large (inferior temporal region) to
-small (temporal pole, insula gyrus, frontal pole).  The main drawback
-is a lack of inter-rater reliability numbers -- in particular because
-visual inspection and comparison of labelings reveals a degree of
-inconsistency in labeling of particular regions across subjects.
-Nevertheless, the NIREP dataset is perhaps the highest quality
-evaluation dataset currently available for the cortex.
+\input{evaluation_data_description.tex}
 

### 2010-12-07 ImageSegmentation/antsAtroposSegmentationImageFilter.h
ENH: Improving/extending implementation of fixed-segmentation values.      For prior label images , the output (when using Plato) will be      a segmentation image with values at the labeled voxels that match      the input label image.  Other voxels will have a value that is      determined statistically.      For prior probability images, a voxel in the spatial prior with      probability 1 will try to 'fix' the outcome as with prior label      images.  This capability is untested.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@688 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -88,6 +88,7 @@ public:
   /** Typedef support of input types. */
   typedef TInputImage                                 ImageType;
   typedef typename ImageType::PixelType               PixelType;
+  typedef typename ImageType::IndexType               IndexType;
   typedef TMaskImage                                  MaskImageType;
   typedef typename MaskImageType::PixelType           MaskLabelType;
   typedef TClassifiedImage                            ClassifiedImageType;
@@ -308,8 +309,8 @@ public:
    */
   typename RealImageType::Pointer GetLikelihoodImage( unsigned int );
 
-  RealType CalculateLocalPosteriorProbability( RealType, RealType,
-    RealType, RealType, RealType );
+  RealType CalculateLocalPosteriorProbability( RealType, RealType, RealType, 
+    RealType, RealType , IndexType index , unsigned int whichClass );
   typename RealImageType::Pointer GetPosteriorProbabilityImage( unsigned int );
   typename RealImageType::Pointer
     GetSmoothIntensityImageFromPriorImage( unsigned int, unsigned int );

### 2010-12-07 Documentation/atropos/atropos.tex
Updates to introduction and some FIXMEs
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@689 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 \begin{center}
-{\Large \bf An Open Source Framework for Multivariate $n$-Tissue Segmentation and Volumetric Cortical Parcellation:  Evaluation Using the BrainWeb, NIREP, and LPBA Public Data Sets} \\
+{\Large \bf An Open Source Framework for Multivariate $n$-Tissue
+  Segmentation and Evaluation Using Open Data Sets} \\
 \vspace*{0.5cm}
@@
 large quantities of possible labelings (in the experimental section,
-we use up to 68 classes) with a minimal memory footprint.  In addition
-to describing the technical and implementation aspects of Atropos,
-this work evaluates its performance on the parcellation problem given
-two different training datasets, the BrainWeb dataset from Montreal
-Neurological Institute and the Hammers atlas from UCL [FIXME].  We use
-the datasets to evaluate three uses of the Atropos segmentation method: (1) K-means
-segmentation without use of template data; (2) segmentation with prior
-probability maps derived from a group template; (3) N-class EM
-segmentation for parcellating the brain into 68 classes based on
-training data.  These evaluation studies, combined with illustrative
-examples that excercise Atropos options, demonstrate both performance
-and wide applicability of this new segmentation tool.
-
+we use up to 68 classes) with a minimal memory footprint.  This work
+describes the technical and implementation aspects of Atropos and
+evaluates its performance on two different ground-truth datasets.
+First, we use the BrainWeb dataset from Montreal Neurological
+Institute to evaluate three tissue segmentation performance via (1)
+K-means segmentation without use of template data; (2) Markov Random
+Field Segmentation with initialization by prior probability maps
+derived from a group template; (3) Prior-based Segmentation with use
+of spatial prior probability maps derived from a group template.  We
+also evaluate Atropos performance by using spatial priors to drive a 68 class EM segmentation
+problem defined by the Hammers atlas from University College London
+{\bf FIXME proper name}.  These evaluation studies, combined with
+illustrative examples that excercise Atropos options, demonstrate both
+performance and wide applicability of this new platform-independent 
+open source segmentation tool.
 %Component (3) corresponds to the most
@@
 %make publicly available all scripts and code from the evaluation.
-
 \end{abstract}
@@
 concomitant investment has been made towards image classification
-techniques for the cerebrum beginning with the seminal work of
-\cite{Vannier1985} in which NASA satellite image processing software
-was appropriated for statistical classification of head tissues in 2-D
-MR images.  Since then, a proliferation of techniques has ensued with
-increasing sophistication.  The chronology of progress in segmentation
-may be tracked through both technical reviews
+techniques for the cerebrum beginning the appropriation of NASA
+satellite image processing software for statistical classification of
+head tissues in 2-D MR images \cite{Vannier1985}.  A proliferation of
+techniques has ensued with increasing sophistication in both core
+methodology and degree of refinement for specific problems.  The
+chronology of progress in segmentation may be tracked through both
+technical reviews
 \citep{Bezdek1993,Pal1993,Clarke1995,Pham2000,Viergever2001,Suri2002,Duncan2004,Balafar2010}
@@
 Observations regarding a general trend towards a more integrative view
-of neuroanatomical image processing led to the work described in \cite{Ashburner2005} which is
-publicly available as SPM---a large-scale Matlab module in which registration, segmentation, and bias field 
-correction can be simultaneously modeled within a single optimization scheme. The roots of this very popular 
-software package stem back to early seminal work by Karl Friston in which the basis for statistical parametric mapping was developed \citep{Friston1990} and subsequently provided to the research community.
-A similar holistic approach
-to brain processing was provided in \cite{Pohl2006} in which segmentation
-and registration parameters were optimized simultaneously while casting the inhomogeneity model parameters
-of \cite{Wells1996} as nuisance variables.
-Continued work involved recursive parcellation of the brain volume by considering sub-structures in a hierarchical manner \citep{Pohl2007}.
-An implementation is provided in  3D slicer \cite{Pieper2006}---an open source package with developmental contributions from multiple agencies including both private and academic institutions.
+of neuroanatomical image processing led to the work described in
+\cite{Ashburner2005} which is publicly available as SPM---a
+large-scale Matlab module in which registration, segmentation, and
+bias field correction can be simultaneously modeled within a single
+optimization scheme. The roots of this very popular software package
+stem back to early seminal work by Karl Friston in which the basis for
+statistical parametric mapping was developed \citep{Friston1990} and
+subsequently provided to the research community. A similar holistic
+approach to brain processing was provided in \cite{Pohl2006} in which
+segmentation and registration parameters were optimized simultaneously
+while casting the inhomogeneity model parameters of \cite{Wells1996}
+as nuisance variables.  Continued work involved recursive parcellation
+of the brain volume by considering sub-structures in a hierarchical
+manner \citep{Pohl2007}.  An implementation is provided in 3D slicer
+\cite{Pieper2006}---an open source package with developmental
+contributions from multiple agencies including both private and
+academic institutions.  {\bf FIXME references proper names SIENA stuff}
+Researchers in aging have also developed tools that have focused on
+accurately segmenting the T1 MRI of elderly controls and subjects
+suffering from neurodegeneration.  Recent comparative evaluation studies {\bf
+  FIXME cite a number of these} have shown that methods such as FAST,
+SPM Unified Segmentation, SIENA and others show different performance
+characteristics under different evaluation criteria.  That is, no
+single method performs best under every measurement, highlighting
+the need for segmentation tools that are tunable for different
+problems and research goals.
 
-Related neuroanatomical research concerns the selection of geometric features of the cortex \citep[e.g.][]{Goualher1999} with aims at understanding the functional-anatomical relationship of the human brain.
-Recent endeavors produce a dense cortical labeling in which every point of the cortex is classified, i.e.
-a cortical parcellation \citep{Fischl2004,Heckemann2006,Destrieux2010}.
-Given the extreme time-consumption and tedium associated with manual effort towards the latter, various algorithms have been proposed of which a small subset has been publicly availed to the research community such as the popular software package known as Freesurfer \citep{Dale1999,Fischl1999,Fischl2004}.
-In contrast to the volumetric approach detailed in this work, Freesurfer is primarily a surface-based technique in which the brain structures such as the grey-white matter interface and pial surfaces are processed, analyzed, and displayed as tessellated surfaces  \cite{Dale1999,Fischl1999}.
-Advantages of surface-based approaches include the ability to map processed neuroanatomy to simple geometric primitives such as spheres or planes and the ease of including topological constraints in the analysis workflow.
+Related neuroanatomical research concerns the selection of geometric
+features of the cortex \citep[e.g.][]{Goualher1999} with aims at
+understanding the functional-anatomical relationship of the human
+brain. Recent endeavors produce a dense cortical labeling in which
+every point of the cortex is classified, i.e. a cortical parcellation
+\citep{Fischl2004,Heckemann2006,Destrieux2010}.  Given the extreme
+time-consumption and tedium associated with manual effort towards the
+latter, various algorithms have been proposed of which a small subset
+has been publicly availed to the research community such as the
+popular software package known as Freesurfer
+\citep{Dale1999,Fischl1999,Fischl2004}.  In contrast to the volumetric
+approach detailed in this work, Freesurfer is primarily a
+surface-based technique in which the brain structures such as the
+grey-white matter interface and pial surfaces are processed, analyzed,
+and displayed as tessellated surfaces  \cite{Dale1999,Fischl1999}.
+Advantages of surface-based approaches include the ability to map
+processed neuroanatomy to simple geometric primitives such as spheres
+or planes and the ease of including topological constraints in the
+analysis workflow.  These types of methods, as well as Klein's
+Mindboggle \cite{Klein2003}, would usually follow an initial segmentation by a volumetric method such as Atropos.
 
@@
 
-Our open source segmentation tool, which we have dubbed {\em Atropos},%
-\footnote{
-Atropos is one of the three Fates from Greek mythology characterized by her dreaded shears used to decide the destiny of each mortal.
-Also, consistent with the entomological motif of our ANTs toolkit, {\it Acherontia atropos} is a species of large moth known for the skull-like pattern visible on its thorax.
-}
-takes advantage of the $n$-tissue paradigm for volumetric image segmentation and harnesses its generalization for template-based cortical parcellation.
-Atropos is an EM-based algorithm which contains many of the features of its predecessors for performing
-$n$-tissue segmentation including imposition of prior information in the form of MRFs and template-based prior probability maps.
-The latter is extended in this work through the formulation of a distance template-based prior used for EM label-propagation and refinement performed in an anatomically meaningful way.
+Our open source segmentation tool, which we have dubbed {\em
+Atropos}, 
+\footnote{ Atropos is one of the three Fates from Greek
+mythology characterized by her dreaded shears used to decide the
+destiny of each mortal.  Also, consistent with the entomological motif
+of our ANTs toolkit, {\it Acherontia atropos} is a species of large
+moth known for the skull-like pattern visible on its thorax.  }
+efficiently and flexibly implements the $n$-tissue paradigm for
+volumetric image segmentation.  Atropos allows users to harness its
+generalized EM algorithm for standard tissue classification of
+the brain into gray matter, white matter and cerebrospinal fluid.
+Atropos equally allows incarnations that use EM to simultaneously
+maximize the posterior probabilities of many classes, for instance,
+when parcellating the brain into hemispheres, cortical regions and
+deep brain structures such as amygdala, hippocampus and
+thalamus.  Atropos contains features of its predecessors for
+performing $n$-tissue segmentation including imposition of prior
+information in the form of MRFs and template-based spatial prior
+probability maps as well as weighted combinations of these terms.  We
+extend the use of spatial priors, in this work, to allow sparse priors
+that may then be extended by a geodesic distance-based label propagation
+approach.  In short, Atropos seeks to provide a toolbox of
+segmentation approaches that may be modified, tuned and refined by the user
+for different problems.
 
-Atropos addresses the bias field issue by incorporating the recently developed N4 bias correction algorithm \citep{Tustison2010} in an adaptive manner.
-Coupled with the registration \citep{Avants2010b} and template building strategies \citep{Avants2010} already included in the ANTs toolkit,
-Atropos is a versatile and powerful software tool which touches multiple aspects of our brain processing pipeline from brain extraction \citep{Avants2010a} to grey matter/white matter/cerebrospinal fluid segmentation to label fusion and propagation for achieving a refined cortical parcellation based on both prior information and subject-specific image data.
+We also allow Atropos to address the bias field issue by incorporating
+the recently developed N4 bias correction algorithm
+\citep{Tustison2010} in an adaptive manner.  Coupled with the
+registration \citep{Avants2010b} and template building strategies
+\citep{Avants2010} already included in the ANTs toolkit, Atropos is a
+versatile and powerful software tool which touches multiple aspects of
+our brain processing pipeline from brain extraction
+\citep{Avants2010a} to grey matter/white matter/cerebrospinal fluid
+segmentation to label fusion and propagation for achieving a refined
+cortical parcellation based on both prior information and
+subject-specific image data.
 
@@
 }
-such that the posterior probability is maximized.  Following introductory notation, the following sections provide a formal description of these three essential components, viz.
+such that the posterior probability is maximized.  Following
+introductory notation, the next sections provide a formal description of these three essential components, viz.
 \begin{itemize}
@@
 \end{itemize}
+These key components are common across most segmentation approaches.
 
@@
 of labels can range upwards of 74 per hemisphere \citep{Destrieux2010}, the memory load can be prohibitive if all label images 
-are loaded into run-time memory simultaneously.
-A major part of minimizing memory usage in Atropos,  which corresponds to the boolean `{\ttfamily -u/--minimize-memory-usage}' option, is the sparse representation of each of the prior probability images.
+are loaded into run-time memory simultaneously.  A major part of
+minimizing memory usage in Atropos,  which corresponds to the boolean
+`{\ttfamily -u/--minimize-memory-usage}' option, is the sparse
+representation of each of the prior probability images.  
 Motivated by the observation that these spatial prior probability maps tend to be highly localized for large quantities of  
@@
 
+{\bf FIXME Prior label image definition?}
+
 \section{Evaluation}

### 2010-12-08 Utilities/itkN4MRIBiasFieldCorrectionImageFilter.h
STYLE:  Incorporated suggestions for handling the internal pipeline in a more rigorous fashion.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@690 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -212,8 +212,8 @@ private:
   N4MRIBiasFieldCorrectionImageFilter( const Self& ); //purposely not implemented
   void operator=( const Self& ); //purposely not implemented
 
-  void SharpenImage( RealImageType *, RealImageType * );
-  void UpdateBiasFieldEstimate( RealImageType *, void * );
+  typename RealImageType::Pointer SharpenImage( RealImageType * );
+  typename RealImageType::Pointer UpdateBiasFieldEstimate( RealImageType * );
   RealType CalculateConvergenceMeasurement( RealImageType *, RealImageType * );
 
   MaskPixelType                               m_MaskLabel;

### 2010-12-08 Documentation/atropos/atropos.tex
update to bib & update to tex - intro + abstract 'done'
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@691 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 %\line(1, 0){250} \\
-{\bf Atropos:  $n$-Tissue Segmentation and Cortical Parcellation}\\
+{\bf Atropos:  $n$-Tissue Segmentation}\\
 Corresponding author: \\
@@
 \begin{abstract} 
-Neuroanatomical coordinate systems are essential for the interpretation of 
-structural and functional imaging studies.  However, manual delineation of the 
-sulco-gyral complex is time consuming and prone to variability given cortical 
-complexity.  This work describes an open source, image-based algorithmic approach to 
-cortical parcellation which uses training data to propagate structural labelings to 
-individual images.  Along with other neuroinformatics tools publicly distributed in our
-ANTs (Advanced Normalization Tools)%
+%Neuroanatomical coordinate systems are essential for the
+%interpretation of structural and functional imaging studies.
+%However, manual delineation of the sulco-gyral complex is time
+%consuming and prone to variability given cortical complexity.  This
+%work describes an open source, image-based algorithmic approach to
+%cortical parcellation which uses training data to propagate
+%structural labelings to individual images.  Along with other
+%neuroinformatics tools publicly distributed in our ANTs (Advanced Normalization Tools)%
+We introduce {\em Atropos}, an ITK-based multivariate $n$-class
+Expectation Maximization (EM) segmentation algorithm based on either parametric or non-parametric finite mixture modelings (FMM) of the
+class intensities.  Atropos, available in ANTs
 \footnote{
 http://www.picsl.upenn.edu/ANTs
-}
-package, {\em Atropos} is an ITK-based multivariate $n$-class
-Expectation Maximization (EM) segmentation algorithm which uses either
-parametric or non-parametric finite mixture modelings (FMM) of the
-class intensities.  It is capable of incorporating prior labelings,
-prior probability maps, and/or Markov Random Field (MRF) modeling
-assumptions as spatial priors and geodesically propagating
-segmentation information across neuroanatomy to ensure a dense
-labeling.  Atropos has also been efficiently implemented to handle
+}, is capable of incorporating spatial prior probability maps,
+(sparse) prior label maps and/or Markov Random Field (MRF) modeling
+assumptions.  Atropos has also been efficiently implemented to handle
 large quantities of possible labelings (in the experimental section,
@@
 also evaluate Atropos performance by using spatial priors to drive a 68 class EM segmentation
-problem defined by the Hammers atlas from University College London
-{\bf FIXME proper name}.  These evaluation studies, combined with
+problem defined by the Hammers atlas from University College London.  These evaluation studies, combined with
 illustrative examples that excercise Atropos options, demonstrate both
@@
 
-\section{Introduction} Neurological and neurodevelopmental research
-rely on anatomical and functional correspondence both within and
-between subjects.  As image acquisition technology has advanced,
-concomitant investment has been made towards image classification
-techniques for the cerebrum beginning the appropriation of NASA
+\section{Introduction} As medical image acquisition technology has advanced,
+concomitant investment has been made towards adapting classification
+techniques for neuroanatomy.  Some of the earliest work appropriated NASA
 satellite image processing software for statistical classification of
 head tissues in 2-D MR images \cite{Vannier1985}.  A proliferation of
-techniques has ensued with increasing sophistication in both core
+techniques ensued with increasing sophistication in both core
 methodology and degree of refinement for specific problems.  The
@@
 
-Given the local optimization susceptibility characteristic of EM algorithms, various initialization schemes have been proposed.  Common low-level initialization strategies include uniform
-probability assignment \citep{Wells1996}, Otsu thresholding \citep{Zhang2001}, and K-means clustering \citep{Pappas1992}.
-More sophisticated low-level initialization schemes include that of \cite{Greenspan2006} in which a dense spatial distribution of Gaussians is used to capture the complex neuroanatomical layout with subsequent processing used to conjoin subsets of Gaussians belonging to the same tissue classes.
-Recently, anatomically specific strategies for incorporating prior knowledge  into the segmentation solution have included the use of spatial prior probability maps  of the various structures of interest \citep{Leemput1999a,Marroquin2002,Ashburner2005}.
-These spatial prior probability maps can also be used to provide an
-initial segmentation.   Related technological developments encompasses
-those driven by partial volume modeling for increased accuracy in
-brain segmentation \citep{Ruan2000,Ballester2002,Leemput2003}.
-
-Observations regarding a general trend towards a more integrative view
-of neuroanatomical image processing led to the work described in
-\cite{Ashburner2005} which is publicly available as SPM---a
+EM algorithms are sensitive to local optima. Consequently, a good initialization
+strategy is critical.  Common low-level initialization strategies include uniform
+probability assignment \citep{Wells1996}, Otsu thresholding
+\citep{Zhang2001}, and K-means clustering \citep{Pappas1992}.  More
+sophisticated low-level initialization schemes include that of
+\cite{Greenspan2006} in which a dense spatial distribution of
+Gaussians is used to capture the complex neuroanatomical layout with
+subsequent processing used to conjoin subsets of Gaussians belonging
+to the same tissue classes.  Recently, anatomically specific
+strategies for incorporating prior knowledge into the segmentation
+solution have included the use of spatial prior probability maps of
+the various structures of interest
+\citep{Leemput1999a,Marroquin2002,Ashburner2005}.  These spatial prior
+probability maps can also be used to provide an initial segmentation.
+Related technological developments encompasses those driven by partial
+volume modeling for increased accuracy in brain segmentation
+\citep{Ruan2000,Ballester2002,Leemput2003}.
+
+A general trend towards more integrative neuroanatomical image processing led to the work described in
+\cite{Ashburner2005} which is publicly available within SPM5---a
 large-scale Matlab module in which registration, segmentation, and
@@
 optimization scheme. The roots of this very popular software package
-stem back to early seminal work by Karl Friston in which the basis for
+stem back to early work by Karl Friston in which the basis for
 statistical parametric mapping was developed \citep{Friston1990} and
@@
 manner \citep{Pohl2007}.  An implementation is provided in 3D slicer
-\cite{Pieper2006}---an open source package with developmental
+\citep{Pieper2006}---an open source package with developmental
 contributions from multiple agencies including both private and
-academic institutions.  {\bf FIXME references proper names SIENA stuff}
-Researchers in aging have also developed tools that have focused on
+academic institutions. Researchers in aging often focus on
 accurately segmenting the T1 MRI of elderly controls and subjects
-suffering from neurodegeneration.  Recent comparative evaluation studies {\bf
-  FIXME cite a number of these} have shown that methods such as FAST,
+suffering from neurodegeneration, for instance, via the SIENA algorithm
+\citep{Smith2007}.  Two recent comparative evaluation studies \citep{,Bresser2011} have shown that methods such as FAST,
 SPM Unified Segmentation, SIENA and others show different performance
@@
 analysis workflow.  These types of methods, as well as Klein's
-Mindboggle \cite{Klein2003}, would usually follow an initial segmentation by a volumetric method such as Atropos.
-
-{\color{red}{{\em Needs work}:  Mindboggle, Heckemann  One advantage of the pipeline we propose is that we parcellate
-cortex entirely in the image space, thus avoiding the difficulty of
-transferring labels from the mesh space back to the image space---a
-problem that is a confound of surface-based methods\citep{Klein2010}.}}
+Mindboggle \cite{Klein2005}, would usually follow an initial segmentation by a volumetric method such as Atropos.
+
+%{\color{red}{{\em Needs work}:  Mindboggle, Heckemann  One advantage
+%of the pipeline we propose is that we parcellate cortex entirely in
+%the image space, thus avoiding the difficulty of transferring labels
+%from the mesh space back to the image space---a problem that is a confound of surface-based methods\citep{Klein2010}.}}
 
@@
 moth known for the skull-like pattern visible on its thorax.  }
-efficiently and flexibly implements the $n$-tissue paradigm for
-volumetric image segmentation.  Atropos allows users to harness its
+efficiently and flexibly implements a $n$-tissue paradigm for
+voxel-based image segmentation.  Atropos allows users to harness its
 generalized EM algorithm for standard tissue classification of
@@
 
-The major contribution of the Atropos algorithm is a flexible, memory
-efficient implementation of core expectation maximization machinery
-that is useful across many segmentation problems.  To highlight the
+The major contribution of the Atropos algorithm is a flexible,
+well-designed, memory efficient implementation of core expectation maximization machinery
+that is useful across many segmentation problems.  Atropos options
+include different maximization criterion, an interactive mode, tunable
+priors and multivariate density estimators using both
+non-parametric and parametric functions.  To highlight the
 value of this open-source, free contribution, we performed a search of
-software attributes on NITRC and found that as of {\textcolor{red}
-INSERT DATE NO STAND-ALONE N-CLASS EM methods are currently listed}.
+software attributes on NITRC and found that as of November 2010 no
+stand-alone EM methods are currently listed.
 Atropos's memory efficiency enables EM-segmentation of many classes at
@@
 value of the low-memory implementation within this paper.  Atropos
-also allows multivariate image features to drive the data term.  Such
-an approach is valuable when more than one view of anatomy aids
+also allows multivariate image features to drive the data term, relevant when more than one view of anatomy aids
 segmentation, as in neonatal brain tissue classification
@@
 analysis due to accessibility) we also only use publicly available
-data sets.
+data sets.  For this reason, all results in this paper are
+reproducible with the caveat that users may require some guidance from
+the authors or other users in the community.
 
@@
 where $t_{ik}$ is the prior probability value due to the
-template-based prior information at site $i$.
+template-based prior information at site $i$.  The user may also
+choose mixing proportions equal to $\frac{1}{K}$ via the command line
+interface to the posterior formulation.
 {\textcolor{red}{FIXME --- need explanation of heuristic ...  Also, we

### 2010-12-09 Scripts/buildtemplateparallel.sh
ENH: adding multi template script -- a basic 'starter' script for multi-atlas labeling ENH: btp and shapeupdate enhancement s.t. they now work with vector.nii.gz files. Verification: the following test performed ok: ~/code/ANTS/Scripts/buildtemplateparallel.sh -d 2 -c 0 -o TEST r*slice.nii BUG: in MeasureMinMaxMean min/max computation for vectors.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@697 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -288,33 +288,14 @@ function shapeupdatetotemplate {
     echo " shapeupdatetotemplate 2"
     echo "--------------------------------------------------------------------------------------"
 
-    if [ $dim -eq 2  ]
-	then 
-	${ANTSPATH}AverageImages $dim ${templatename}warpxvec.nii.gz 0 `ls ${outputname}*Warpxvec.nii.gz | grep -v "InverseWarpxvec"`
-	${ANTSPATH}AverageImages $dim ${templatename}warpyvec.nii.gz 0 `ls ${outputname}*Warpyvec.nii.gz | grep -v "InverseWarpyvec"`
-	
-    elif [ $dim -eq 3  ]
-	then
-	${ANTSPATH}AverageImages $dim ${templatename}warpxvec.nii.gz 0 `ls ${outputname}*Warpxvec.nii.gz | grep -v "InverseWarpxvec"`
-	${ANTSPATH}AverageImages $dim ${templatename}warpyvec.nii.gz 0 `ls ${outputname}*Warpyvec.nii.gz | grep -v "InverseWarpyvec"`
-	${ANTSPATH}AverageImages $dim ${templatename}warpzvec.nii.gz 0 `ls ${outputname}*Warpzvec.nii.gz | grep -v "InverseWarpzvec"`
-    fi
+	${ANTSPATH}AverageImages $dim ${templatename}warp.nii.gz 0 `ls ${outputname}*Warp.nii.gz | grep -v "InverseWarp"`
 
     echo
     echo "--------------------------------------------------------------------------------------"
     echo " shapeupdatetotemplate 3"
     echo "--------------------------------------------------------------------------------------"
-    if [ $dim -eq 2  ]
-	then 
-	${ANTSPATH}MultiplyImages $dim ${templatename}warpxvec.nii.gz ${gradientstep} ${templatename}warpxvec.nii.gz 
-	${ANTSPATH}MultiplyImages $dim ${templatename}warpyvec.nii.gz ${gradientstep} ${templatename}warpyvec.nii.gz 
-
-    elif [ $dim -eq 3  ]
-	then
-	${ANTSPATH}MultiplyImages $dim ${templatename}warpxvec.nii.gz ${gradientstep} ${templatename}warpxvec.nii.gz 
-	${ANTSPATH}MultiplyImages $dim ${templatename}warpyvec.nii.gz ${gradientstep} ${templatename}warpyvec.nii.gz 
[…]

### 2010-12-09 Scripts/shapeupdatetotemplate.sh

@@ -31,34 +31,14 @@ function shapeupdatetotemplate {
     echo " shapeupdatetotemplate 2"
     echo "--------------------------------------------------------------------------------------"
 
-    if [ $dim -eq 2  ]
-	then 
-	${ANTSPATH}AverageImages $dim ${templatename}warpxvec.nii.gz 0 `ls ${outputname}*Warpxvec.nii.gz | grep -v "InverseWarpxvec"`
-	${ANTSPATH}AverageImages $dim ${templatename}warpyvec.nii.gz 0 `ls ${outputname}*Warpyvec.nii.gz | grep -v "InverseWarpyvec"`
-	
-    elif [ $dim -eq 3  ]
-	then
-	${ANTSPATH}AverageImages $dim ${templatename}warpxvec.nii.gz 0 `ls ${outputname}*Warpxvec.nii.gz | grep -v "InverseWarpxvec"`
-	${ANTSPATH}AverageImages $dim ${templatename}warpyvec.nii.gz 0 `ls ${outputname}*Warpyvec.nii.gz | grep -v "InverseWarpyvec"`
-	${ANTSPATH}AverageImages $dim ${templatename}warpzvec.nii.gz 0 `ls ${outputname}*Warpzvec.nii.gz | grep -v "InverseWarpzvec"`
-    fi
+	${ANTSPATH}AverageImages $dim ${templatename}warp.nii.gz 0 `ls ${outputname}*Warp.nii.gz | grep -v "InverseWarp"`
 
     echo
     echo "--------------------------------------------------------------------------------------"
     echo " shapeupdatetotemplate 3"
     echo "--------------------------------------------------------------------------------------"
-    if [ $dim -eq 2  ]
-	then 
-	${ANTSPATH}MultiplyImages $dim ${templatename}warpxvec.nii.gz ${gradientstep} ${templatename}warpxvec.nii.gz 
-	${ANTSPATH}MultiplyImages $dim ${templatename}warpyvec.nii.gz ${gradientstep} ${templatename}warpyvec.nii.gz 
-
-    elif [ $dim -eq 3  ]
-	then
-	${ANTSPATH}MultiplyImages $dim ${templatename}warpxvec.nii.gz ${gradientstep} ${templatename}warpxvec.nii.gz 
-	${ANTSPATH}MultiplyImages $dim ${templatename}warpyvec.nii.gz ${gradientstep} ${templatename}warpyvec.nii.gz 
[…]

### 2010-12-12 Documentation/atropos/atropos.tex
ENH: Updates to bib, tex and eval description --- next commit should reorganize the paper.  TODO: add parameters to theory section.  describe supervised initialization.  describe an application with r-slice data. provide script for this.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@703 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 \usepackage{amsmath,amsfonts}
-\usepackage{babel}
+\usepackage{babel,verbatim}
 \usepackage{enumerate}
@@
 
-Bayes' theorem provides a powerful mechanism for making inductive inferences assuming the availability of quantities defining the relevant conditional probabilities, specifically the likelihood and prior probability terms.
-A Bayesian paradigm for brain image segmentation utilizes a user-selected observation model defining the likelihood term and one or more prior probability terms whose product is proportional to the posterior probability.
-Whereas the likelihood term has been previously defined parametrically (e.g.
-a Gaussian model) or non-parametrically (e.g.
-Parzen windowing of the sample histogram), the prior term, as given in the literature, has been formed inclusively either as MRF-based or template-based.
-An image segmentation solution in this context is an assignment of one label to each voxel%
+Bayes' theorem provides a powerful mechanism for making inductive
+inferences assuming the availability of quantities defining the
+relevant conditional probabilities, specifically the likelihood and
+prior probability terms.  Bayesian paradigms for brain image
+segmentation utilize a user-selected observation model defining the
+likelihood term and one or more prior probability terms.  The produce
+of likelihood(s) and prior(s) is proportional to the posterior
+probability.  The likelihood term has been previously defined both 
+parametrically (e.g. a Gaussian model) and non-parametrically
+(e.g. Parzen windowing of the sample histogram).  The prior term, as
+given in the literature, has often been formed either as
+MRF-based or template-based.  An image segmentation solution in this
+context is an assignment of one label to each voxel%
 \footnote{
@@
 }
-such that the posterior probability is maximized.  Following
-introductory notation, the next sections provide a formal description of these three essential components, viz.
+such that the posterior probability is maximized.  The next sections
+introduce notation and provide a formal description of three essential
+components in Bayesian segmentation, viz.
 \begin{itemize}
   \item the likelihood or observation model(s),
-  \item the prior probability quantities derived from a generalized MRF, template-based prior terms, and a novel distance prior used for label propagation, and 
+  \item the prior probability quantities derived from a generalized
+    MRF and template-based prior terms, and 
+%, and a novel distance prior used for label propagation, and 
   \item the optimization framework for maximizing the posterior probability.
 \end{itemize}
-These key components are common across most segmentation approaches.
+These components are common across most EM segmentation approaches.
 
 \subsection{Notation}
-
-Assume a field, $\mathcal{F}$, whose values are known at discrete locations, i.e.
-sites, within a regular voxel lattice, $\mathcal{I}$.
-Note that $\mathcal{F}$ can be a scalar field in the case of unimodal data (e.g.
-T1 image only) or a vector field in the case of multimodal data (e.g.
-T1, T2, and proton density images).
-A specific set of observed values, denoted by $\mathbf{y}$, are indexed at $N$ discrete locations in $\mathcal{I}$ by $i \in \{1, 2, \ldots, N\}$.
-It is assumed that the complete set of observations is a random field $Y = \{y_1, y_2, \ldots, y_N \}$.
-A labeling of $\mathcal{I}$,
-also known as a hard segmentation of $\mathcal{I}$,
-assigns to each site in $\mathcal{I}$ one of $K$ labels from the finite set $\mathcal{L} = \{l_1, l_2, \ldots, l_K\}$.
-Also considered a random field, we denote this discrete labeling by $X = \{x_1, x_2, \ldots, x_N\}$.
+Assume a field, $\mathcal{F}$, whose values are known at discrete
+locations, i.e. sites, within a regular voxel lattice that makes up an
+image domain, $\mathcal{I}$.  Note that $\mathcal{F}$ can be a scalar
+field in the case of unimodal data (e.g. T1 image only) or a vector
+field in the case of multimodal data (e.g. T1, T2, and proton density
+images).  A specific set of observed values, denoted by $\mathbf{y}$,
+are indexed at $N$ discrete locations in $\mathcal{I}$ by $i \in \{1,
+2, \ldots, N\}$.  This random field, $Y = \{y_1, y_2, \ldots, y_N \}$,
+serves as a discrete representation of an observed image's intensities.  A labeling
+of this image, also known as a hard segmentation, assigns to each site
+in $\mathcal{I}$ one of $K$ labels from the finite set $\mathcal{L} =
+\{l_1, l_2, \ldots, l_K\}$.  Also considered a random field, this
+discrete labeling is $X = \{x_1, x_2, \ldots, x_N\}$ where each
+$x_i \in \mathcal{L}$.   We use $\mathbf{x}$ to denote a specific set of labels in
+$\mathcal{I}$ and a valid, though not necessarily optimal, solution to the segmentation problem.
 
@@
 
+\subsection{Segmentation Objective Function} Atropos optimizes a class
+of user-selectable maximization objective functions each of which may
+be represented in a generic Bayesian framework, as described by \cite{Sanjay-Gopal1998}.  This framework
+requires {\em likelihood models} and {\em prior models} which enter
+into Bayes' formula,
+\begin{equation}\label{eq:bayes}
+p(\mathbf{x}|\mathbf{y})=\underbrace{
+p(\mathbf{y}|\mathbf{x})}_{\text{Likelihood(s)}} \underbrace{
+p(\mathbf{x})}_{\text{Prior(s)}}\frac{1}{\mathbf{y}}, 
+\end{equation} 
+where the normalization term, $1/\mathbf{y}$, is a constant that does
+not affect the optimization \citep{Sanjay-Gopal1998}.
+Given choices for likelihood models and prior probabilities, the Bayesian
+segmentation solution is the labeling $\hat{\mathbf{x}}$ which
+maximizes the posterior probability, i.e.
+\begin{align} \hat{\mathbf{x}} = \argmax_{\mathbf{x}}
+\left\{p(\mathbf{y}|\mathbf{x})p(\mathbf{x})\right\}.
+\end{align} Similar to its algorithmic predecessors, Atropos employs
+the EM framework \citep{Dempster1977} to find maximum likelihood
+solutions to this problem.  The following sections detail
+Atropos choices for the likelihood(s), prior(s) and EM strategy.
+
 \subsection{Likelihood or Observation Models}
-
 To each of the $K$ labels corresponds a single probabilistic model describing the variation of $\mathcal{F}$ over $\mathcal{I}$.
@@
 Using the standard notation, $\mathrm{Pr}(S=s) = p(s)$, $\mathrm{Pr}(S=s|T=t) = p(s|t)$, we can define these voxelwise probabilities, $\mathrm{Pr}_k( Y_i = y_i | X_i = l_k ) = p_k(y_i|l_k)$, in either parametric or non-parametric terms.
-Given its simplicity and good performance, in the parametric case, $p_k$ is typically defined as a normalized distribution, i.e.
+Given its simplicity and good performance, in the parametric case, $p_k$ is typically defined as a normal distribution, i.e.
 \begin{align}\label{eq:param}
@@
 Gaussian is defined by the mean vector, $\boldsymbol{\mu}_k$, and
-covariance matrix, $\Sigma_k$.  An initial estimate for these model
+covariance matrix, $\Sigma_k$.  Multivariate probabilities are derived from the
+Mahalanobis distance.  An initial estimate for these model
 parameters are tabulated during the initialization process,
@@
 
+Atropos enables two choices for the likelihood formulation.  One is
+based on the classical finite mixture model (FMM) and the second is based on
+the spatially varying mixture model (SVMM) defined in \citep{Sanjay-Gopal1998}.
 In order to calculate the probability associated with the entire set of observations, $\mathbf{y}$, we assume independency between voxels which is done primarily for facilitating quantification as opposed to modeling realistic assumptions.
@@
 \begin{align}\label{eq:likelihood}
-  p(\mathbf{y}) = \prod_{i=1}^N \left(      
+  p(\mathbf{y}|\mathbf{x}) = \prod_{i=1}^N \left(      
                                                         \sum_{k=1}^K \gamma_k p_k(y_i|l_k)
@@
 \end{align}
-where $\gamma_k$ is the mixing parameter \cite{Ashburner2005}.
+where $\gamma_k$ is the mixing parameter \citep{Ashburner2005}.
 
 \subsection{Prior Probability Models}
-
-By modeling $\mathcal{F}$ via the set of observation models $\Phi$, this so called finite-mixture model could be used to produce a labeling or segmentation \citep[e.g.][]{Wells1996}.
-However, as pointed out by \cite{Zhang2001}, the exclusive use of the observation models in modeling the intensity profile produces a less than optimal solution as spatial contextual considerations are completely removed.
-This has been remedied by the introduction of a host of  prior probability models including those characterized by their use of MRF theory and template-based information.
-For example, in the works of \cite{Leemput1999a} and \cite{Weisenfeld2009}, the original global prior term given in \cite{Wells1996} is replaced by the product of the template-based and the MRF-based prior terms.
-In addition to their descriptions below, we discuss our introduction of a third possible prior term in the form of an extension of template-based Euclidean and geodesic distance label propagation.
+By modeling $\mathcal{F}$ via the set of observation models $\Phi$,
+this so called finite-mixture model could be used to produce a
+labeling or segmentation \citep[e.g.][]{Wells1996}.  However, as
+pointed out by \cite{Zhang2001}, the exclusive use of the observation
+models in modeling the intensity profile produces a less than optimal
+solution because spatial contextual considerations are ignored.  This
+has been remedied by the introduction of a host of prior probability
+models including those characterized by use of MRF theory and
+template-based information.  For example, in the works of
+\cite{Leemput1999a} and \cite{Weisenfeld2009}, the original global
+prior term given in \cite{Wells1996} is replaced by the product of the
+template-based and the MRF-based prior terms.  In addition to their
+descriptions below, we discuss our introduction of a third possible
+prior/objective combination in the form of a (sparse) prior labeling which fixes specific points
+of the segmentation and uses EM to propagate this information
+elsewhere in the image.  
+% extension of template-based Euclidean and geodesic distance label propagation.
 
 \subsubsection{Generalized MRF Prior}
-In the labeling of a single voxel, an appealing strategy for incorporating spatial coherence into the segmentation solution is to consider the labeling of the voxel neighborhood and favoring labeling configurations in which voxel neighborhoods tended towards homogeneity.
-This intuition is formally described by MRF theory in which spatial interactions in voxel neighborhoods can be modeled \citep{Li2001}.
-
-In order to apply MRF theory,  we assume the random field introduced earlier, $X$, is an MRF and is, therefore, characterized by a neighborhood system, $\mathcal{N}_i$, on the lattice, $\mathcal{I}$, composed of the neighboring sites of $i$.
+One may incorporate spatial coherence into the segmentation by
+favoring labeling configurations in which voxel neighborhoods tend
+towards homogeneity. This intuition is formally described by MRF
+theory in which spatial interactions in voxel neighborhoods can be
+modeled \citep{Li2001}.
+
+We assume the random field introduced earlier, $X$, is an MRF characterized by a neighborhood system, $\mathcal{N}_i$, on the lattice, $\mathcal{I}$, composed of the neighboring sites of $i$.
 This neighborhood system is both noninclusive, i.e.
@@
 $i \in \mathcal{N}_j \Leftrightarrow j \in \mathcal{N}_i$.
-As an MRF, $X$ also satisfies the positivity and locality conditions, i.e., respectively
+As an MRF, $X$ also satisfies the positivity and locality conditions, i.e.,
  \begin{align} \label{eq:mrf}
@@
  \end{align}
-where $\mathbf{x}$ is any particular labeling configuration on $X$ (in
+and where $\mathbf{x}$ is any particular labeling configuration on $X$ (in
 other words, any labeling permutation on $X$ is {\em a priori}
-possible).  The MRF locality condition is 
+possible).  The MRF locality condition is then, 
 \begin{align}
@@
 where $V_{ij}$ is typically defined in terms of the Kronecker delta,
-$\delta_{ij}$, based on the Ising potential (also known as a Potts model) \citep{Besag1974}
+$\delta_{ij}$, based on the classical Ising potential (also known as a Potts model) \citep{Besag1974}
 \begin{align}
@@
 \subsubsection{Template-Based Priors}
-A number of researchers have used template-based strategies to both ensure spatial coherence and incorporate prior knowledge in obtaining a sutiable segmentation.
+A number of researchers have used template-based strategies to both ensure spatial coherence and incorporate prior knowledge in segmentation.
 A common technique is to select labeled subjects from a population from which a template is constructed \citep[e.g.][which is also available in the ANTs toolkit]{Avants2010}.
@@
 stationary mixing proportions, $\mathrm{Pr}(x_i = l_k) = \gamma_k$,
+FIXME DEFINE THE GAMMA --- SHOULD USE SVMM FROM HEBERT ?
 describing the prior probability that label $l_k$ corresponds to a particular voxel, regardless of intensity, are replaced by the following spatially varying mixing proportions
@@
 interface to the posterior formulation.
-{\textcolor{red}{FIXME --- need explanation of heuristic ...  Also, we
-  allow a weighted combination defined by the ComputeLocalPosterior
-  function in Atropos --- how do we justify / explain this?   clearly,
-it's a balance of the global solution (no local prior) and the local
-one with spatial prior ... }}
-
+
+% {\textcolor{red}{FIXME --- need explanation of heuristic ...  Also, we
+
+%  allow a weighted combination defined by the ComputeLocalPosterior
+
+%  function in Atropos --- how do we justify / explain this?   clearly,
+
+% it's a balance of the global solution (no local prior) and the local
+
+% one with spatial prior ... }}
+
+\subsubsection{Supervised Initialization to the EM Algorithm}
+PriorLabelImage
+
+\begin{comment}
+{
 \subsubsection{Distance Prior for Label Propagation}
@@
 
+}
+\end{comment}
+
 \subsection{Optimization}
-Given the likelihood models and prior probabilities whose product is proportional to the posterior probability, the Bayesian segmentation solution is the labeling $\hat{\mathbf{x}}$ which maximizes the posterior probability, i.e.
-\begin{align}
-  \hat{\mathbf{x}} = \argmax_{\mathbf{x}} \left\{p(\mathbf{y}|\mathbf{x})p(\mathbf{x})\right\}.
-\end{align}
-Similar to its algorithmic predecessors, Atropos employs the EM
-framework \citep{Dempster1977}.  After initial estimation of the
-likelihood model parameters, EM iterates between calculation of the missing $\hat{\mathbf{x}}$ and subsequent re-estimation of the model parameters by maximizing the expectation of the complete data log-likelihood (cf.
-Eqn.
-(\ref{eq:likelihood})).
+Atropos uses expectation maximization to find a locally optimal
+solution for the user-selected version of the Bayesian segmentation
+problem (cf. Eqn. (\ref{eq:bayes})).
+After initial estimation of the
+likelihood model parameters, EM iterates between calculation of the
+missing $\hat{\mathbf{x}}$ and subsequent re-estimation of the model
+parameters by maximizing the expectation of the complete data
+log-likelihood (cf. Eqn. (\ref{eq:likelihood})).  
 %\begin{align}
@@
 Consider $N$ sites each with a possible $K$ labels for a total of $N^K$ possible labeling configurations.
-Due to the large $K$ associated with cortical labeling problems, exact optimization is even more problematic than for the traditional 3-tissue scenario.
+Due to the large $K$ associated with $K>>1$ problems, exact optimization is even more problematic than for the traditional 3-tissue scenario.
 }
@@
 accuracy---Atropos uses the well-known {\em Iterated Conditional
-  Modes} (ICM)  \citep{Besag1986} which is greedy, computationally efficient and provides good performance.
+  Modes} (ICM)  \citep{Besag1986} which is greedy, computationally
+efficient and provides good performance.  The EM algorithm employed in
+Atropos may therefore be written
+\begin{eqnarray}
+X \\ 
+Y \\
+Z
+\end{eqnarray}
+The solution will be a local maxima.
 
@@
 improved by other interested researchers.
-{\textcolor{red}{FIXME ---- should we save this for a deliverable in
-    some grant?}}
 
@@
 even 4-dimensional image data.  Higher dimensions are possible although we have not yet encountered such 
-application-specific need.  Multiple images (assumed to be of the same dimension, size, origin, etc.) can be specified 
+application-specific need.  Multiple images (assumed to be of the same
+dimension, size, origin, etc.), which enable multivariate likelihoods, can be specified 
 on the command line using the `{\ttfamily -a/--intensity-image}' option for each image.  The first image specified on the 
@@
 in a Parzen windowing scheme on a weighted histogram to estimate the
-observational model  {\textcolor{red} cite Suyash}.
+observational model  \citep{Awate2006}.
 
@@
 system using the vector notation `{\ttfamily 1x1x1}' for a
-neighborhood radius of 1 in all 3 dimensions.  Note that this radius
-prescribes an $L_\infty$  {\textcolor{red} FIXME should this be $L_2$?} distance and thus includes not just the face-connected neighboring voxels.
+neighborhood radius of 1 in all 3 dimensions.  This radius is defined
+such that voxels including but not limited to those that are face
+connected will influence the MRF.
 

### 2010-12-20 Documentation/atropos/atropos.tex
ENH: Update to introduction, theory and experiment sections.  Adding Hammers figure.  BWeb results next.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@707 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 
-\doublespacing
+% \doublespacing
 
@@
 We introduce {\em Atropos}, an ITK-based multivariate $n$-class
-Expectation Maximization (EM) segmentation algorithm based on either parametric or non-parametric finite mixture modelings (FMM) of the
+Expectation Maximization (EM) segmentation algorithm based on either parametric or non-parametric finite mixture modeling (FMM) of the
 class intensities.  Atropos, available in ANTs
@@
 }, is capable of incorporating spatial prior probability maps,
-(sparse) prior label maps and/or Markov Random Field (MRF) modeling
-assumptions.  Atropos has also been efficiently implemented to handle
+(sparse) prior label maps and/or Markov Random Field (MRF) modeling.  Atropos has also been efficiently implemented to handle
 large quantities of possible labelings (in the experimental section,
@@
 Institute to evaluate three tissue segmentation performance via (1)
-K-means segmentation without use of template data; (2) Markov Random
-Field Segmentation with initialization by prior probability maps
-derived from a group template; (3) Prior-based Segmentation with use
+K-means segmentation without use of template data; (2) MRF segmentation with initialization by prior probability maps
+derived from a group template; (3) Prior-based segmentation with use
 of spatial prior probability maps derived from a group template.  We
@@
 satellite image processing software for statistical classification of
-head tissues in 2-D MR images \cite{Vannier1985}.  A proliferation of
+head tissues in 2-D MR images \citep{Vannier1985}.  A proliferation of
 techniques ensued with increasing sophistication in both core
@@
 probability maps can also be used to provide an initial segmentation.
-Related technological developments encompasses those driven by partial
-volume modeling for increased accuracy in brain segmentation
+Related technological developments model partial
+volume effects for increased accuracy in brain segmentation
 \citep{Ruan2000,Ballester2002,Leemput2003}.
@@
 A general trend towards more integrative neuroanatomical image processing led to the work described in
-\cite{Ashburner2005} which is publicly available within SPM5---a
+\cite{Ashburner2005} which is publicly available within SPM5, a
 large-scale Matlab module in which registration, segmentation, and
@@
 suffering from neurodegeneration, for instance, via the SIENA algorithm
-\citep{Smith2007}.  Two recent comparative evaluation studies \citep{,Bresser2011} have shown that methods such as FAST,
-SPM Unified Segmentation, SIENA and others show different performance
-characteristics under different evaluation criteria.  That is, no
-single method performs best under every measurement, highlighting
-the need for segmentation tools that are tunable for different
+\citep{Smith2007}.  A recent evaluation study 
+compared kNN segmentation, SPM Unified Segmentation and SIENA and
+found different performance
+characteristics under different evaluation criteria
+\citep{Bresser2011}.  \cite{Klauschen2009} had similar findings when
+comparing SPM5, FSL and FreeSurfer.  These studies suggest that no
+single method performs best under every measurement and, along with the No
+Free Lunch theorem, highlight the need for segmentation tools that are tunable for different
 problems and research goals.
@@
 probability maps as well as weighted combinations of these terms.  We
-extend the use of spatial priors, in this work, to allow sparse priors
-that may then be extended by a geodesic distance-based label propagation
-approach.  In short, Atropos seeks to provide a toolbox of
-segmentation approaches that may be modified, tuned and refined by the user
-for different problems.
+also borrow an idea from \cite{Boykov2004} and use sparse spatial
+priors to provide initialization and boundary conditions for Atropos
+EM segmentation in a semi-interactive manner.  In short, Atropos seeks
+to provide a toolbox of segmentation approaches that may be modified,
+tuned and refined by the user for different problems.
 
@@
 
-The major contribution of the Atropos algorithm is a flexible,
-well-designed, memory efficient implementation of core expectation maximization machinery
-that is useful across many segmentation problems.  Atropos options
-include different maximization criterion, an interactive mode, tunable
-priors and multivariate density estimators using both
-non-parametric and parametric functions.  To highlight the
-value of this open-source, free contribution, we performed a search of
+The Atropos algorithm is multi-platform, free, stand-alone and able to
+manage a wide range of scalar and vector image datatypes.  To highlight the
+value of this open-source contribution, we performed a search of
 software attributes on NITRC and found that as of November 2010 no
@@
 prior probability terms.  Bayesian paradigms for brain image
-segmentation utilize a user-selected observation model defining the
-likelihood term and one or more prior probability terms.  The produce
+segmentation employ a user-selected observation model defining the
+likelihood term and one or more prior probability terms.  The product
 of likelihood(s) and prior(s) is proportional to the posterior
@@
 MRF-based or template-based.  An image segmentation solution in this
-context is an assignment of one label to each voxel%
+context is an assignment of one label to each voxel
 \footnote{
@@
 where the parameters $\mu_k$ and $\sigma_k^2$ respectively represent
-the mean and variance of the $k^{th}$ model.  In the case of
-multi-modal data, where $y_i$ is a vector quantity, a mulitvariate
-Gaussian is defined by the mean vector, $\boldsymbol{\mu}_k$, and
-covariance matrix, $\Sigma_k$.  Multivariate probabilities are derived from the
-Mahalanobis distance.  An initial estimate for these model
-parameters are tabulated during the initialization process,
-e.g. K-means or Otsu thresholding.
+the mean and variance of the $k^{th}$ model.  When $y_i$ is a vector
+quantity, we replace the Euclidean distance by by Mahalanobis distance
+and define multivariate Gaussian parameters via a mean vector,
+$\boldsymbol{\mu}_k$, and covariance matrix, ${\bf \Sigma_k}$.
 
@@
 
-Atropos enables two choices for the likelihood formulation.  One is
-based on the classical finite mixture model (FMM) and the second is based on
-the spatially varying mixture model (SVMM) defined in \citep{Sanjay-Gopal1998}.
-In order to calculate the probability associated with the entire set of observations, $\mathbf{y}$, we assume independency between voxels which is done primarily for facilitating quantification as opposed to modeling realistic assumptions.
-Spatial interdependency between voxels is modeled by the prior probabilities discussed in the next section.
-Marginalizing over the set of possible labels, $\mathcal{L}$, leads to the following probabilistic formulation
+Atropos segmentation likelihood estimates are based on the classical finite mixture model (FMM).
+FMM assumes independency between voxels to calculate the probability
+associated with the entire set of observations, $\mathbf{y}$.  Spatial
+interdependency between voxels is modeled by the prior probabilities
+discussed in the next section.  Marginalizing over the set of possible labels, $\mathcal{L}$, leads to the following probabilistic formulation
 \begin{align}\label{eq:likelihood}
@@
 labeling or segmentation \citep[e.g.][]{Wells1996}.  However, as
-pointed out by \cite{Zhang2001}, the exclusive use of the observation
-models in modeling the intensity profile produces a less than optimal
+pointed out by \cite{Zhang2001}, exclusive use of the intensity profile produces a less than optimal
 solution because spatial contextual considerations are ignored.  This
@@
 template-based and the MRF-based prior terms.  In addition to their
-descriptions below, we discuss our introduction of a third possible
+descriptions below, we discuss a third possible
 prior/objective combination in the form of a (sparse) prior labeling which fixes specific points
@@
 We employ the strategy given in \cite{Ashburner2005} in which the
-stationary mixing proportions, $\mathrm{Pr}(x_i = l_k) = \gamma_k$,
-FIXME DEFINE THE GAMMA --- SHOULD USE SVMM FROM HEBERT ?
-describing the prior probability that label $l_k$ corresponds to a particular voxel, regardless of intensity, are replaced by the following spatially varying mixing proportions
+stationary mixing proportions, $\mathrm{Pr}(x_i = l_k) = \gamma_k$
+(cf. Eqn. (\ref{eq:likelihood})), 
+describing the prior probability that label $l_k$ corresponds to a particular voxel, regardless of intensity, are replaced by the following spatially varying mixing proportions,
 \begin{align}
-\mathrm{Pr}(x_i = l_k) = \frac{\gamma_k t_{ik}}{\sum_{j=1}^K\gamma_j t_{ij}}
+\mathrm{Pr}(x_i = l_k) = \frac{\gamma_k t_{ik}}{\sum_{j=1}^K\gamma_j t_{ij}}.
 \end{align}
-where $t_{ik}$ is the prior probability value due to the
-template-based prior information at site $i$.  The user may also
-choose mixing proportions equal to $\frac{1}{K}$ via the command line
-interface to the posterior formulation.
-
-% {\textcolor{red}{FIXME --- need explanation of heuristic ...  Also, we
-
-%  allow a weighted combination defined by the ComputeLocalPosterior
-
-%  function in Atropos --- how do we justify / explain this?   clearly,
-
-% it's a balance of the global solution (no local prior) and the local
-
-% one with spatial prior ... }}
-
-\subsubsection{Supervised Initialization to the EM Algorithm}
-PriorLabelImage
+The $t_{ik}$ is the prior probability value at site $i$ which was
+mapped, typically by image registration, to the local image
+from a template data set.  The user may also
+choose mixing proportions equal to
+\begin{align}
+\mathrm{Pr}(x_i = l_k) = \frac{t_{ik}}{\sum_{j=1}^K t_{ij}}.
+\end{align}
+via the command line interface to the posterior formulation.
+{\textcolor{red}{FIXME --- need explanation of heuristic ... how do we
+  estimate the $\gamma$ without the $t_{ij}$ present?}
+
+\subsubsection{Supervised Semi-Interactive Segmentation}
+Brain segmentation methods have relied on user interaction for many
+years \citep{Lim1989,Julin1997,Freeborough1997a,Yushkevich2006}.
+Atropos is capable of benefitting from user knowledge via an
+initialization and optimization strategy that depends upon a spatially
+varying prior label image passed as input.  Rapid, sparse
+labeling---with visualization provided by ITK-SNAP
+(www.itksnap.org)---enables an interaction and execution processing
+loop that can be critical to solving segmentation problems with
+challenging clinical data in which automated approaches fail.  This
+part of Atropos design is inspired by the interactive graph cut methods
+pioneered by \cite{Boykov2001} and which has spawned many follow-up
+applications.  The Atropos prior label
+image pre-specificies the segmentation results at a subset of the
+spatial domain by fixing the priors and likelihood (and, thus, the
+posterior) at a subset of $\mathcal{I}$ to be $1$ for the known label
+and $0$ for each other label at the same site.  The user-input
+therefore not only initializes the algorithm, but also gives boundary conditions
+that influence the EM solution outside of the known sites.  While the
+graph-based min-cut max-flow solution is globally optimal for two
+labels, only locally optimal approachs are available for 3 or more
+classes.  Thus, in most practical applications, the EM approach is a
+reasonable and efficient alterative to Boykov's solution.
+Furthermore, as with similar algorithms, one may automate the
+initialization process.
 
@@
 likelihood model parameters, EM iterates between calculation of the
-missing $\hat{\mathbf{x}}$ and subsequent re-estimation of the model
+missing optimal labels $\hat{\mathbf{x}}$ and subsequent re-estimation of the model
 parameters by maximizing the expectation of the complete data
@@
 efficient and provides good performance.  The EM algorithm employed in
-Atropos may therefore be written
-\begin{eqnarray}
-X \\ 
-Y \\
-Z
-\end{eqnarray}
-The solution will be a local maxima.
-
+Atropos may therefore be written as a series of steps:
+\begin{description}
+\item[Initialization:] In all cases, the user defines the number of
+classes to segment.  The simplest initialization is by the classic
+K-means algorithm with only the number of classes as a prior.
+Otherwise, the user must provide prior information for each class in
+the form of either a single $n$-ary prior label image or a series of
+prior probability images, one for each class.
+\item[E-Step: ICM Label Update:] Given the initialization and fixed
+model parameters, Atropos will evaluate the posterior probability for
+each possible label and assign, to each voxel, the label that
+maximizes the posterior.  {\bf \textcolor{red}{ FIXME more specific?}}.
+\item[M-Step: Parameter Update:] We describe the parametric update.  Given fixed labels, we update the
+  parameters by the FMM-defined likelihood.  We compute, for each of
+  $K$ labels, the mean,
+$$ \mu^{new}_k \leftarrow \frac{ \sum_{i=1}^N  y_i \gamma_k p_k(y_i|l_k)}{ \sum_{i=1}^N \gamma_k p_k(y_i|l_k) },
+$$
+variance {\bf \textcolor{red}{ FIXME should this be the full
+    covariance matrix?}},
+$$ \sigma^{new}_k \leftarrow  \frac{ \sum_{i=1}^N  (y_i - \mu_k ) \gamma_k p_k(y_i|l_k)}{ \sum_{i=1}^N  \gamma_k p_k(y_i|l_k) },
+$$
+and mixture parameters, 
+$$ \gamma^{new}_k \leftarrow  \textcolor{red}{ FIXME .}
+$$
+This type of update strategy is known as {\em soft} EM.  Hard EM, in
+contrast, only uses sites containing label $l_k$ to update model $k$ parameters.
+A similar strategy is used in the multivariate and non-parametric cases.
+\end{description}
+The EM algorithm will iterate toward a local maximum.  We track
+convergence of the algorithm by summing up the maximum posterior
+probability at each site over the segmentation domain.  
 %\begin{align}
@@
 %\end{align}
-
+%
+%
 %An image, $I$, maps a domain, $\Omega$ into the
@@
 \end{figure}
-
+\subsection{The Atropos User Interface}
 As with other classes that comprise ANTs, Atropos uses the Insight Toolkit as a developmental foundation.  
@@
 \end{command}
-\doublespacing
-
-\subsection{Integration of N4 Bias Correction}
-
-The typical processing pipeline begins with an intensity normalization/bias correction
-step using the recently developed N4 algorithm \citep{Tustison2010} which is based on the popular nonparametric 
-nonuniformity intensity normalization (N3) method introduced in \cite{Sled1998}.
-It was previously mentioned that several methods proposed in the literature have taken an integrative view of the segmentation problem by incorporating an intrinsic bias correction step into the actual algorithmic workflow.
-The advancements
-introduced with the N4 algorithm permit such an adaptive integration with Atropos.  
-Recent demonstrations suggest improved white matter segmentation produces better gain field estimates using N3 
-\citep{Boyes2008}.  Thus, when performing 3-tissue segmentation, we use the posterior probability map of white matter
-at the current  iteration as a weighted mask for input to N4.
-This is done by setting the `{\ttfamily --weight-image}' option on the N4 command line call (see Listing \ref{listing:n4}) to the posterior probability image corresponding to the white matter produced as output in the Atropos call, i.e.
-`{\ttfamily Atropos --output}'.
+
+\subsection{Initializing the Atropos Objective}
+The Atropos algorithm has a number of parameters defined within
+Listing \ref{listing:n4} and will function on 2, 3 or 4 dimensional
+data.  However, the majority of the time, users will be concerned with
+a smaller set of input parameters.  Here, we list the recommended input and
+an example definition for each parameter:
+\begin{description}
+\item[Input images to be segmented:] If more than
+  one input image is passed, then a multivariate model will be
+  instantiated.  E.g.  {\ttfamily -a Image.nii.gz} for one image and
+  {\ttfamily -a Image1.nii.gz -a Image2.nii.gz} for multiple images.
+\item[Input image mask:] This binary image
+  will define the spatial segmentation domain.  Regions outside the
+  mask receive the label $0$.  E.g.  {\ttfamily -x mask.nii.gz}. 
+\item[Convergence criteria:] The algorithm terminates if it reaches max-iterations or
+  produces a change less than the minimum change in the posterior.
+  E.g. {\ttfamily -c [5,1.e-5]}. 
+\item[MRF prior:] The key parameter to increase or decrease the
+  spatial smoothness of the label map is $\beta$.  A useful range of
+  $\beta$ values is $0$ to $0.5$ where we usually use $0.05, 0.1, 0.2$
+  in brain segmentation.  E.g.  {\ttfamily -m [ 0.1, 1x1x1 ]} would
+  define $\beta=0.1$ with a MRF radius of one voxel in each of three dimensions.
+\item[Initialization:] The initialization options include (where the
+  first parameter defines $K$, here 3 for each below),
+\begin{itemize}
+\item  {\ttfamily -i  kmeans[3]} standard kmeans initialization for
+  three classes.
+\item  {\ttfamily -i  PriorLabelImage[3, label\_image.nii.gz , w ]}
+  where scalar $w \in [0,1]$ is unused \textcolor{red}{ FIXME .} 
+\item  {\ttfamily -i  PriorProbabilityImages[3, label\_prob\%02d.nii.gz ,
+    w ]} where $w=0$~(use the prior probability images only for
+  initialization) or $w=0.5$ (use the prior probability images
+  throughout the optimization).  If one chooses $0 < w < 0.5$ then one
+  will increase (from zero) the weight on the priors.  These images,
+  like the PriorLabelImage, should be defined with the same domain as
+  the input images to be segmented.  
+\end{itemize}
+\item[Posterior formulation:]  The user may choose to estimate the
+  mixture proportions (or not) by setting  {\ttfamily -p Socrates[1]}
+  or {\ttfamily -p Socrates[0] }.  Fixed label boundary conditions may be employed by
+  selecting the PriorLabelImage initialization and   {\ttfamily -p Plato[0] }.
+\item[Output:] Atropos will output the hard segmentation and the
+  probability image for each model.  E.g.  {\ttfamily -o
+    [segmentation.nii.gz , seg\_prob\%02d.nii.gz] } will write
+  out the hard segmentation in the first output parameter and a
+  probability image for each class named, here, seg\_prob01.nii.gz ,
+  seg\_prob02.nii.gz, etc. 
+\end{description}
+Higher dimensions than 4 are possible although we have not yet encountered such 
+application-specific need.  Multiple images (assumed to be of the same
+dimension, size, origin, etc.), will automatically enable multivariate
+likelihoods.  In that case, the first image specified on the 
+command line is used to initialize the Random, Otsu, or K-means labeling with the latter initialization further
+refined by incorporating the additional intensity images, i.e. an initial univariate K-means clustering is determined 
+from the first intensity image which, along with the other images, provides the starting multivariate cluster centers for a 
+follow-up multivariate K-means labeling.    More details on each of
+the key implementation options are given below.
+
+\subsection{Likelihood Implementation}
+As mentioned previously in the introduction, different groups have
+opted for different likelihood models which have included either
+parametric, specifically Gaussian, or nonparametric variations.
+However, these approaches are similar in that they require a list
+sample of intensity data from the input image(s) and a list of
+weighting values for each observation of the list sample from which
+the model is constructed.  In general, one may query model
+probabilities by passing a given pixel's single intensity (for
+univariate segmentation) or multiple intensities (for multivariate
+segmentation) to the modeling function, regardless of whether the
+function is parameter or non-parametric.  These similarities permitted
+the creation of a generic plug-in architecture where classes
+describing both parametric and nonparametric observational models are
+all derived from an abstract list sample function class.  Three
+likelihood classes have been developed, one parametric and two
+nonparametric, and are available for usage although one of the
+nonparametric classes is still in experimental development.  The
+plug-in architecture even permits mixing likelihood models with
+different classes during the same run for a hybrid
+parametric/non-parametric approach although this possibility has yet
+to be fully explored.
+
+If the Gaussian likelihood model is chosen, the list sample of
+intensity values and corresponding weights comprised of the posterior
+probabilities are used to estimate the Gaussian model parameters, i.e.
+the mean and variance.  For the non-parametric model, the list sample
+and posteriors are used in a Parzen windowing scheme on a weighted
+histogram to estimate the observational model \citep{Awate2006}.
+
+\subsection{Prior Probability Models}
+Consistent with our previous discussion, we offer both an MRF-based
+prior probability for modeling spatial coherence and the possibility
+of specifying a set of prior probability maps or a prior label map
+with the latter extendable to creating a dense labeling.  To invoke
+the MRF `{\ttfamily -m/--mrf}' option, one specifies the smoothing
+factor (or the granularity parameter, $\beta$, given in
+Eqn. (\ref{eq:U}), and the radius (in voxels) of the neighborhood
+system using the vector notation `{\ttfamily 1x1x1}' for a
+neighborhood radius of 1 in all 3 dimensions.  This radius is defined
+such that voxels including but not limited to those that are face
+connected will influence the MRF.
+
+\begin{comment}{
+The user also has an option to propagate initial labels, set by the
+prior options, by a distance map tool.  
+To propagate labels with the specification of prior probability images
+or a prior label image, one needs to select the distance function (set
+by the `{\ttfamily -e/--use-euclidean-distance}' boolean option)
+which, by default, is determined by the geodesic distance.  The user
+also needs to set the label propagation parameters for one or more of
+the classes.   This  includes both the boundary probability and the exponential
+ decay parameter, respectively $\alpha_k$ and $\sigma_k$ in Eqn. (\ref{eq:prop}).  
+}\end{comment}
+
+\subsection{Integration with Registration and N4 Bias Correction} 
+Much of the image processing in our lab takes advantage of multiple
+open source tools.  Most critically, these involve registration and
+bias correction, in addition to segmentation. 
+\subsubsection{Registration of Probability Maps} 
+Registration approaches enable the ability to transform information
+between spatial domains which may aid in both segmentation and bias
+correction.  We rely heavily on template-building strategies
+\citep{Avants2010,Avants2010a} which are also offered in ANTs.  Since
+aligned prior probability images and prior label maps are often
+associated with such templates, Atropos can be initialized with these
+data with their influence regulated by a prior probability weighting
+term.  Although prior label maps can be specified as a single
+multi-label image, prior probability data are often represented as
+multiple scalar images with a single image corresponding to a
+particular label.  For relatively small classifications, such as the
+standard 3-tissue segmentation (i.e. white matter, gray matter, and
+cerebrospinal fluid), this does not typically present computational
+complexities using modern hardware.  However, when considering dense
+cortical parcellations where the number of labels can range upwards of
+74 per hemisphere \citep{Destrieux2010}, the memory load can be
+prohibitive if all label images are loaded into run-time memory
+simultaneously.  A major part of minimizing memory usage in Atropos,
+which corresponds to the boolean `{\ttfamily
+-u/--minimize-memory-usage}' option, is the sparse representation of
+each of the prior probability images.  Motivated by the observation
+that these spatial prior probability maps tend to be highly localized
+for large quantities of cortical labels, a threshold is specified on
+the command line (default = 0.0) and only those probability values
+which exceed that threshold are stored in the sparse representation.
+During the course of optimization, the prior probability image for a
+given label is reconstructed on the fly as needed.  For instance, the
+NIREP (www.nirep.org) evaluation images are on the order of $300
+\times 300 \times 256$ with 32 cortical labels.  Our novel memory
+minimizing strategy typically shrinks run-time memory usage from a
+peak of 10+ GB to approximately 1.5 GB and enable these datasets to be
+used for training/prior-based cortical parcellation.
 
@@
 \end{command}
-\doublespacing
-
-\subsection{From Segmentation Theory to Atropos Implementation}
-
-\subsubsection{Initialization}
-Atropos was designed to be dimensionally generic in the sense that the same code can be used to handle 2-, 3-, and 
-even 4-dimensional image data.  Higher dimensions are possible although we have not yet encountered such 
-application-specific need.  Multiple images (assumed to be of the same
-dimension, size, origin, etc.), which enable multivariate likelihoods, can be specified 
-on the command line using the `{\ttfamily -a/--intensity-image}' option for each image.  The first image specified on the 
-command line is used to initialize the Random, Otsu, or K-means labeling with the latter initialization further
-refined by incorporating the additional intensity images, i.e. an initial univariate K-means clustering is determined 
-from the first intensity image which, along with the other images, provides the starting multivariate cluster centers for a 
-follow-up multivariate K-means labeling.
-
-Much of the ANTs-based processing in our lab (including Atropos) takes advantage of template-building strategies 
-\citep{Avants2010} which are also offered  in ANTs.  Since aligned prior probability images and prior label maps are often 
-associated with such templates, Atropos can be initialized with these data with their influence regulated by a prior probability
-weighting term.  Although prior label maps can be specified as a single multi-label image, prior probability data are often represented as
-multiple scalar images with a single image corresponding to a particular label.  For relatively small classifications, such as 
-the standard 3-tissue segmentation (i.e. white matter, gray matter, and cerebrospinal fluid), this does not typically present
-computational complexities using modern hardware.  However, when considering dense cortical parcellations where the number
-of labels can range upwards of 74 per hemisphere \citep{Destrieux2010}, the memory load can be prohibitive if all label images 
-are loaded into run-time memory simultaneously.  A major part of
-minimizing memory usage in Atropos,  which corresponds to the boolean
-`{\ttfamily -u/--minimize-memory-usage}' option, is the sparse
-representation of each of the prior probability images.  
-Motivated by the observation that these spatial prior probability maps tend to be highly localized for large quantities of  
-cortical labels, a threshold is specified on the command line (default = 0.0) and only those probability values which exceed 
-that threshold are stored in the sparse representation.  During the course of optimization, the prior probability image for a
-given label is reconstructed on the fly as needed.  {\textcolor{red}For the NIREP data used in this manuscript where the images are 
-on the order of $300 \times 300 \times 256$ and 32 cortical labels, this typically shrinks run-time memory usage from a 
-peak of 10+ GB to approximately 1.5 GB. --- FIXME }
-
- \subsubsection{Likelihood or Observation Models}
-As mentioned previously in the introduction, different groups have opted for different likelihood models which have included
-either parametric, specifically Gaussian, or nonparametric variations.  However, these approaches are similar in that 
-they require a list sample of intensity data from the input image(s) and a list of weighting values for each observation 
-of the list sample from which the model is constructed.  
-In general, one may query model probabilities by passing a given pixel's single intensity (for univariate segmentation) or multiple intensities (for multivariate
-segmentation) to the modeling function, regardless of whether the
-function is parameter or non-parametric.  These similarities permitted the creation of a generic plug-in architecture
-where classes describing both parametric and nonparametric  observational models are all derived from an abstract 
-list sample function class.   Three likelihood classes have been developed, one parametric and two nonparametric,
-and are available for usage although one of the nonparametric classes is still in experimental development.
-The plug-in architecture even permits mixing likelihood models with different classes during the same run for a hybrid parametric/non-parametric approach although this possibility has yet to be fully explored.
-
-If the Gaussian likelihood model is chosen, the list sample of intensity values and 
-corresponding weights comprised of the posterior probabilities are used to estimate the Gaussian model parameters, i.e.
-the mean and variance.  For the non-parametric model, the list sample and posteriors are used
-in a Parzen windowing scheme on a weighted histogram to estimate the
-observational model  \citep{Awate2006}.
-
-\subsubsection{Prior Probability Models}
-Consistent with our previous discussion, we offer both an MRF-based
-prior probability for modeling spatial coherence and the possibility
-of specifying a set of prior probability maps or a prior label map
-with the latter extendable to creating a dense labeling.  To invoke
-the MRF `{\ttfamily -m/--mrf}' option, one specifies the smoothing
-factor (or the granularity parameter, $\beta$, given in
-Eqn. (\ref{eq:U}), and the radius (in voxels) of the neighborhood
-system using the vector notation `{\ttfamily 1x1x1}' for a
-neighborhood radius of 1 in all 3 dimensions.  This radius is defined
-such that voxels including but not limited to those that are face
-connected will influence the MRF.
-
-To propagate labels with the specification of prior probability images or a prior label image, one needs to select the distance function (set by the `{\ttfamily -e/--use-euclidean-distance}' boolean option) which, by default, is determined by the geodesic distance.
-The user also needs to set the label propagation parameters for one or more of the classes.
-This  includes both the boundary probability and the exponential decay parameter, respectively $\alpha_k$ and $\sigma_k$ in Eqn.
-(\ref{eq:prop}).
-
-{\bf FIXME Prior label image definition?}
+\subsubsection{Integration of N4 Bias Correction}
+The typical segmentation processing pipeline begins with an intensity
+normalization/bias correction step using the recently developed N4
+algorithm \citep{Tustison2010} which is based on the popular
+nonparametric nonuniformity intensity normalization (N3) method
+introduced in \cite{Sled1998}.  It was previously mentioned that
+several methods proposed in the literature have taken an integrative
+view of the segmentation problem by incorporating an intrinsic bias
+correction step into the actual algorithmic workflow.  The
+advancements introduced with the N4 algorithm permit such an adaptive
+integration with Atropos.  Recent demonstrations suggest improved
+white matter segmentation produces better gain field estimates using
+N3 \citep{Boyes2008}.  Thus, when performing 3-tissue segmentation, we
+use the posterior probability map of white matter at the current
+iteration as a weighted mask for input to N4.  This is done by setting
+the `{\ttfamily --weight-image}' option on the N4 command line call
+(see Listing \ref{listing:n4}) to the posterior probability image
+corresponding to the white matter produced as output in the Atropos
+call, i.e. `{\ttfamily Atropos --output}'.  The N4 algorithm is
+currently being added as a standard part of the Insight ToolKit.  The
+evaluation section will illustrate inclusion of Atropos, N4 and ANTs
+in a brain processing pipeline.
 
 \section{Evaluation}
-
-\input{evaluation_data_description.tex}
+Atropos~encodes a family of segmentation techniques that may be
+instantiated for different applications but here we evaluate only two
+of the many possibilites. First, we perform an evaluation on the
+BrainWeb dataset using both the standard T1 image with multiple bias
+and noise levels and also the BrainWeb20 data
+\citep{Aubert-Broche2006a,Battaglini2008}.  In combination, this data
+allows one to vary not only noise and bias but also the underlying
+anatomy.  Second, we evaluate the use of Atropos in improving whole
+brain parcellation and excercise its ability to efficiently solve {\em
+many class} expectation maximization problem.  We choose this
+evaluation problem in part to illustrate the flexibility of Atropos
+and also the benefits of the novel, efficient implementation that
+allows many-class problems to be solved with low memory usage ($<$2GB
+for a 68 class model on 1mm$^3$ brain data).
+
+\subsection{BrainWeb Evaluation}
+\label{sec:bweb} The BrainWeb data is freely available at
+\url{http://mouldy.bic.mni.mcgill.ca/brainweb/}.  We employ both the
+individual subject data and the BrainWeb20 data in this evaluation.
+\subsubsection{Single-Subject Evaluation} We use the single-subject
+data with 3\% noise and bias 0, 20 and 40 \% RF inhomogeneity.  We
+study the effect of the MRF prior term and initialization strategy on
+the Dice overlap between ground truth and the segmentation result for
+each tissue.  We test both K-means and prior label image
+initialization with MRF $\beta \in \{ 0.00 , 0.05 , 0.10 , 0.15 , 0.20
+, 0.25 , 0.30 \}$ at each bias field.  We also feed the white matter
+probability map derived from K-means into N4 to guide the bias
+correction.  Segmentation is then repeated, with the same parameters,
+but with the N4 corrected image as input.  Thus, we are able to
+compare the impact of bias on our N4 and Atropos algorithm and also
+the validity of our prior label image initialization.  The prior label
+image manual initialization was based on a crude ITK-SNAP-based labeling of a
+small set of voxels for each tissue.  
+\begin{figure}
+\begin{center}
+\begin{tabular}{c}
+% \includegraphics[height=8in]{Figures/final_results_majority_vote_vs_atropos.pdf}
+\end{tabular}
+\caption{\baselineskip 12pt \small Brainweb single-subject results for
+  each tissue.  The results show that the N4 bias correction
+  algorithm, combined with Atropos, results in a minimal effect of
+  bias, even at the 40\% level.  The optimal $\beta$ for the MRF term
+  appears to be between $0.1$ and $0.2$.  The prior label image
+  initialization is competitive with but inferior to K-means.}
+\label{fig:hammers}
+\end{center}
+\end{figure}
+\subsubsection{20 Subject Evaluation}
+
+\subsection{The Hammers Dataset Evaluation}
+\label{sec:hammers}
+We evaluate the ability to improve multi-template
+labeling results by converting the group labels to probability maps
+and using them to drive many-class EM segmentation.  The procedure we
+use here is similar to that described in \citep{Avants2010a}.  The
+ground truth labels cover 68 classes and much of the brain.  Some
+unlabeled regions remain which we assign to label 69 such that all
+brain parenchyma contains a unique label.  The initialization of our
+evaluation applies the script {\verb ants_multitemplate_labeling.sh }
+(available in the ANTs toolkit) to the 19 Hammers evaluation
+datasets located at \url{http://www.brain-development.org/} \citep{Hammers2003,Heckemann2006}.
+These initial majority voting results are competitive with prior work
+\citep{Heckemann2006,Heckemann2010} and serve as a baseline against
+which we compare.
+
+We first convert each of the 69 labels within the original evaluation
+dataset to an individual image.  The remainder of the procedure,
+summarized briefly, is the same for each of the 19 subjects.  We
+select one subject as an unlabeled target.  The other 18 datasets are
+then mapped (as in the script above) to that subject.  We then deform,
+individually, the 69 $\times$ 18 label images to the unlabeled
+subject.  The label probability map is gained by averaging the 18
+deformed images associated with each label.  We repeat this for each
+subject.  The following parameters are the most relevant to this discussion:
+{\ttfamily  -i  PriorProbabilityImages[3, label\_prob\%02d.nii.gz , 0.5
+  ] -m [0.1,1x1x1] -c [5,0] -p Socrates[1] }.  Results, in terms of
+Dice overlap, are shown in figure~\ref{fig:hammers}.
+\begin{figure}
+\begin{center}
+\begin{tabular}{c}
+\includegraphics[height=8in]{Figures/final_results_majority_vote_vs_atropos.pdf}
+\end{tabular}
+\caption{\baselineskip 12pt \small The figure compares the Dice
+overlap results from Atropos versus the raw results from majority
+voting for each of 68 neuroanatomical regions and, in addition, the
+unlabeled portions of the brain from the Hammers evaluation dataset.
+We evaluated Atropos via N-fold cross-validation and employed
+PriorProbabilityImages for each class where probabilities are gained
+by averaging mapped subject labels. The color coding highlights those
+regions that have the highest (yellow) and lowest (pink) improvement.
+The significance of the improvement, measured by pairwise T-test, is
+also shown as is a trinary coding of that improvement as: $+$
+significant improvement, $-$ performance reduction, $\sim$ no change.}
+\label{fig:hammers}
+\end{center}
+\end{figure}
 
@@
 
-{\color{red}{{\em Elaborate after including the results.}
+{\textcolor{red}{ Elaborate after including the results.} }
 
@@
 application area is that of ventilation-based segmentation of hyperpolarized helium-3 MRI
-\citep{Tustison2010a}.}}
+\citep{Tustison2010a}.
 

### 2010-12-20 Documentation/atropos/atropos.tex
ENH: Bweb evaluation results added.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@708 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 correction.  Segmentation is then repeated, with the same parameters,
-but with the N4 corrected image as input.  Thus, we are able to
-compare the impact of bias on our N4 and Atropos algorithm and also
-the validity of our prior label image initialization.  The prior label
-image manual initialization was based on a crude ITK-SNAP-based labeling of a
-small set of voxels for each tissue.  
+but with the N4 corrected image as input.  The resulting algorithm is
+similar to those methods that fix segmentation parameters while
+estimating bias and fix bias while estimating segmentation parameters.
+Thus, with this simple evaluation, we are able to compare the impact
+of bias on our N4 and Atropos algorithm and also the validity of our
+prior label image initialization.  The prior label image manual
+initialization was based on a crude ITK-SNAP labeling of a small
+set of voxels for each tissue.  Results, in terms of
+Dice overlap, are shown in figure~\ref{fig:bweb1}.
 \begin{figure}
@@
 \begin{tabular}{c}
-% \includegraphics[height=8in]{Figures/final_results_majority_vote_vs_atropos.pdf}
+\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF0.pdf}
+\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF20.pdf}
+\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF40.pdf}\\
+\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF0.pdf}
+\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF20.pdf}
+\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF40.pdf}\\
+\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF0.pdf}
+\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF20.pdf}
+\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF40.pdf}
 \end{tabular}
@@
   initialization is competitive with but inferior to K-means.}
-\label{fig:hammers}
+\label{fig:bweb1}
 \end{center}

### 2010-12-23 Documentation/atropos/atropos.tex
ENH: adding Jue Wu results, some discussion, fixes to number of labels.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@709 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 }, 
-Philip A.~Cook$^1$, Jue Woo$^1$, and James C.~Gee$^1$} \\
+Jue Wu$^1$, Philip A.~Cook$^1$, and James C.~Gee$^1$} \\
 \begin{singlespace} 
@@
 large quantities of possible labelings (in the experimental section,
-we use up to 68 classes) with a minimal memory footprint.  This work
+we use up to 69 classes) with a minimal memory footprint.  This work
 describes the technical and implementation aspects of Atropos and
@@
 of spatial prior probability maps derived from a group template.  We
-also evaluate Atropos performance by using spatial priors to drive a 68 class EM segmentation
+also evaluate Atropos performance by using spatial priors to drive a 69 class EM segmentation
 problem defined by the Hammers atlas from University College London.  These evaluation studies, combined with
@@
 memory.  We evaluate EM-based classification of $1 mm^3$ T1 MR
-neuroimages into 68 neuronatomical regions to illustrate the practical
+neuroimages into 69 neuronatomical regions to illustrate the practical
 value of the low-memory implementation within this paper.  Atropos
@@
 allows many-class problems to be solved with low memory usage ($<$2GB
-for a 68 class model on 1mm$^3$ brain data).
+for a 69 class model on 1mm$^3$ brain data).
 
@@
 \end{figure}
-\subsubsection{20 Subject Evaluation}
+\subsubsection{20 Subject Evaluation} The single-subject brainweb
+study in the previous section tested the basic Atropos options and the
+benefit of N4 for segmentation in the presence of bias.  The 20
+subject brainweb data allows us to use 2-fold cross-validation to test
+our ability to segment different individuals reliably.  In this study,
+we divide the 20 subjects equally into training and testing groups.
+We then exploit the ground-truth labeling of the training data to
+build both a group template \citep{Avants2010} and also prior
+probability maps for each of the three major tissues in the cerebrum.
+Each prior probability map is gained by deforming the ground truth
+labels from each of the 10 training subjects to their template and
+averaging component by component.  We then deform the template---and
+priors---to the ten testing subjects and run Atropos with not only
+{\ttfamily KMeans[3]} initialization but also {\ttfamily
+  PriorProbabilityMap[3,priors\%02d.nii.gz,$w$] } where $w \in
+\{0.0,0.5\}$.  We then switch the roles of testing and training sets to
+gain 3-tissue segmentation for each of the twenty subjects.  
+When $w=0$, the priors are only used in initializing
+the model parameters but not during subsequent EM iterations.  
+When $w=1$, the priors are only maintained in the product with the
+likelihood during all EM iterations.  Results, in terms of bar
+plots for Dice overlap mean and standard deviation, are shown in figure~\ref{fig:bweb20}.
+\begin{figure}
+\begin{center}
+\begin{tabular}{c}
+\includegraphics[width=2.2in]{Figures/bw20_kmeans.png}
+\includegraphics[width=2.2in]{Figures/bw20_priorIni.png}
+\includegraphics[width=2.2in]{Figures/bw20_priorSeg.png}
+\end{tabular}
+\caption{\baselineskip 12pt \small Brainweb 20-subject results for
+  each tissue as a function of MRF-$\beta$ parameter where MRF-$\beta$
+  is in $\{0,0.05,0.1,0.15,0.2\}$ and increases left to right.  The results show that the PriorProbabilityMaps with
+  $w=0.5$ (far right) gives the best performance for all tissues.}
+\label{fig:bweb20}
+\end{center}
+\end{figure}
 
@@
 use here is similar to that described in \citep{Avants2010a}.  The
-ground truth labels cover 68 classes and much of the brain.  Some
+ground truth labels cover 69 classes and much of the brain.  Some
 unlabeled regions remain which we assign to label 69 such that all
@@
 
+\subsection{Reproducibility of this Evaluation}
+The brainweb data is freely available.  We used single-subject
+brainweb data as is.  The 20 subject data, however, required excluding
+non-cerebrum tissue classes.  The Hammers data was also used as is.
+The scripts for the single-subject brainweb study is available in
+supplementary material.  The initialization procedure for the brainweb 20 and
+the Hammers evaluation data is based on freely available scripts
+included with the ANTs toolkit.  {\bf \textcolor{red}{ FIXME}}.
+
 \section{Discussion}
-
-{\textcolor{red}{ Elaborate after including the results.} }
-
+We introduced the Atropos algorithm, the theory and implementation
+details and documented its performance in a variety of use cases.  We
+also showed evidence that the openly available N4 bias correction
+method can easily used with Atropos to improve segmentation.
+Furthermore, we used multiple subject brainweb data to build dataset
+specific priors that provided the most consistent segmentation
+performance across tissues.  Finally, we used majority voting to
+initialize an Atropos EM solution to a 69 class brain parcellation
+problem.  Significant improvements were gained in multiple brain
+regions, in particular in temporal lobe cortex, the hippocampi and
+amygdalae and the lateral ventricles.  This work, in summary, proves
+the applicability of Atropos in both basic and extended use cases.
+
+\subsection{Performance on Single-Subject Brainweb Data}
+How does Atropos performance compare to other methods that have been
+evaluated on the basic brainweb data?   {\bf {\textcolor{red}{ FIXME}}}.
+
+\subsection{Performance on Twenty-Subject Brainweb Data}
+How does Atropos performance compare to other methods that have been
+evaluated on the brainweb twenty data?   {\bf {\textcolor{red}{ FIXME}}}.
+
+\subsection{Performance on Hammers Data}
+Our prior work, \citep{Avants2010b}, showed that the majority vote
+initialization provided to Atropos by ANTs template mapping is
+competitive with \cite{Heckemann2006}.  The Atropos EM extension of
+improved these results further.  However, the Atropos EM segmentation
+performed significantly worse.  This is not surprising, in that
+Atropos EM assumes that signal from the likelihood and MRF term is
+valuable in improving the segmentation.  This assumption held for
+amygdala and lateral ventricles among other areas.  However, in
+pallidum and corpus callosum (the most significant areas with loss of
+performance), this is not true.  We believe the explanation is that
+the intensity varies within these structures and that a more complex
+intensity model (or finer parcellation) would be needed here.  An
+alternative solution would be to use boundary conditions for these
+structures, as in the PriorLabelImage Atropos option.
+
+\subsection{Other Thoughts} {\bf {\textcolor{red}{ FIXME}}}
 A more general advantage which extends beyond the scope of the experimental 
@@
 
-%\paragraph{Acknowledgments}
-%{This work was supported in part by NIH ... }
+\section{Conclusion}
+The Atropos algorithm is freely available to the public.  We release
+this code not only to make it available to clinical researchers but
+with the hope that other researchers in segmentation will provide
+feedback about the implementation decisions that we made.  EM
+segmentation is non-trivial and there are numerous design alternatives
+available not only in the models selected but also in the ICM coding
+strategy, alternatives to ICM and the specific method in which prior
+and likelihood are combined.  Due to the flexibility of Atropos, we
+also hope that some of its capabilities, though not evaluated here,
+are explored by the segmentation or clinical community.
+
+\paragraph{Acknowledgments}
+{This work was supported in part by NIH  {\bf {\textcolor{red}{ FIXME}}} }
 

### 2010-12-24 Documentation/atropos/atropos.tex
ENH: Bweb discussion and associated references
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@710 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
   appears to be between $0.1$ and $0.2$.  The prior label image
-  initialization is competitive with but inferior to K-means.}
+  initialization is competitive with but inferior to K-means.  The
+  legend is in the same position in each graph, allowing a visual
+  comparison of the results.  As one may see, the N4 assisted overlap
+  values are consistent across bias field/RF inhomogeneity.}
 \label{fig:bweb1}
@@
 
-\subsection{Performance on Single-Subject Brainweb Data}
-How does Atropos performance compare to other methods that have been
-evaluated on the basic brainweb data?   {\bf {\textcolor{red}{ FIXME}}}.
-
-\subsection{Performance on Twenty-Subject Brainweb Data}
-How does Atropos performance compare to other methods that have been
-evaluated on the brainweb twenty data?   {\bf {\textcolor{red}{ FIXME}}}.
+\subsection{Performance on Brainweb Data}
+Atropos results are competitive with the state of the art.  For
+instance, \cite{Ashburner2005} (SPM5) evaluated on 0\% RF (bias field) 3\%
+noise braineweb single subject data finding 0.932 (GM) and .961 (WM)
+Dice. Results on 40\% RF were .934 (GM) and .961 (WM). SPM5 exhibits
+insensitivity to bias similar to our own best results on the 40\%RF, 3
+\% noise case (MRF-$\beta$=0.2, Kmeans $+$ N4) with Dice for GM is
+0.950737 and for WM is 0.962586.  \cite{Nakamura2009} gave GM Dice
+results (brainweb single 3\% noise) of 0.962 (0\%RF), 0.964(20\%RF)
+and .956(40\%RF) which are slightly higher than either SPM5 or Atropos
+results.  However, \cite{Nakamura2009} do not report WM or CSF numbers
+for comparison.  Topology preserving approaches also perform well.
+\cite{Shiee2010} achieved Dice results for 3\% noise 20\% RF brainweb
+single subject with 0.927, 0.912 and 0.900 for WM, GM and CSF,
+respectively.  These are excellent numbers given the additional
+constraint applied to the segmentation.  
+\cite{Bazin2007c} proposed the TOADS method and, estimating from the
+paper's graph, showed that the average Dice accuracy for 3\% noise various RF was
+.95-.96 for WM, .93-.95 for GM and .92-.94 for CSF.
+Perhaps the most recent balanced evaluation was performed in
+\citep{Klauschen2009}, which reports confusion matrix numbers, rather
+than Dice overlap.  Because the absolute true number of GM and WM
+voxels for brainweb are known, we can convert the confusion matrix to
+Dice.  In that case, the SPM5 Dice for brainweb single GM and WM is .8847
+and .9085, respectively, while FreeSurfer and FSL's accuracy is
+lower. The best GM result for 20 subjects is obtained by SPM5: 0.9295;
+best WM Dice is from FSL: 0.9499.  We note that \cite{Klauschen2009}
+used a comprehensive evaluation where quality of brain extraction also
+contributed to the outcome.  Thus, the results must be interpreted
+slightly differently than those from other papers.
 
@@
 
-\subsection{Other Thoughts} {\bf {\textcolor{red}{ FIXME}}}
+\subsection{Clinically-Related Evaluation}
+While specifying performance on brainweb is highly
+valuable, clinical validation is a second important aspect of
+segmentation evaluation.  For instance, 
+\citep{Freeborough1997,Westlye2009,Sanchez-Benavides2010,Chou2009,Bresser2011}
+are only a few of the papers that evaluate segmentation performance with
+respect to a known neurobiological outcome measure.  Atropos is
+currently used in clinical studies and a number of clinically-focused,
+application-specific evaluations are ongoing and will constitute
+future work.  One early example of a clinically-focused Atropos
+neuroimaging application is in \citep{Avants2010c}.  An additional successful
+application area is that of ventilation-based segmentation of hyperpolarized helium-3 MRI
+\citep{Tustison2010a} which also used the open-source  Glamorous Glue
+algorithm to impose topology constraints.  Thus, future work may
+incorporate topology more closely into the Atropos methodology.
+
+\section{Conclusion}
 A more general advantage which extends beyond the scope of the experimental 
@@
 brain extraction which has reported good performance in comparison with other
-popular, publicly available brain extraction tools \citep{Avants2010a}.  An additional successful
-application area is that of ventilation-based segmentation of hyperpolarized helium-3 MRI
-\citep{Tustison2010a}.
-
-\section{Conclusion}
+popular, publicly available brain extraction tools \citep{Avants2010a}.
+
 The Atropos algorithm is freely available to the public.  We release

### 2010-12-28 Documentation/atropos/atropos.tex
ENH: spell-checked (probably 1 error that i missed) and added a new figure.  Removed over use of words : procedure, algorithm, etc.  Also added a basic evaluation example script that forms the basis of the paper's evaluation.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@711 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 %consuming and prone to variability given cortical complexity.  This
-%work describes an open source, image-based algorithmic approach to
+%work describes an open source, image-based approach to
 %cortical parcellation which uses training data to propagate
@@
 We introduce {\em Atropos}, an ITK-based multivariate $n$-class
-Expectation Maximization (EM) segmentation algorithm based on either parametric or non-parametric finite mixture modeling (FMM) of the
+Expectation Maximization (EM) segmentation tool based on either parametric or non-parametric finite mixture modeling (FMM) of the
 class intensities.  Atropos, available in ANTs
@@
 problem defined by the Hammers atlas from University College London.  These evaluation studies, combined with
-illustrative examples that excercise Atropos options, demonstrate both
+illustrative examples that exercise Atropos options, demonstrate both
 performance and wide applicability of this new platform-independent 
@@
 brain segmentation.  Given the ``missing data'' aspect of this
-problem, a natural optimization strategy is the use of the
-Expectation-Maximization (EM) framework \citep{Dempster1977}.  The
+problem, use of the Expectation-Maximization (EM) framework is natural\citep{Dempster1977}.  The
 work described in \cite{Wells1996} was one of the first to use EM for
@@
 developments.  A secondary component was an extended nonparametric
-approach, also influenced by earlier work \citep{Kikinis1992}, where
+probability model, also influenced by earlier work \citep{Kikinis1992}, where
 Parzen windowing is used to model the tissue intensity distribution
 omitting consideration of the underlying bias field.  Although
-technically not an EM-based algorithm, the robustness of the latter
-approach has motivated its continued use even more recently \citep[e.g.][]{Weisenfeld2009}.
+technically not an EM-based algorithm, the robustness of the latter has motivated its continued use even more recently \citep[e.g.][]{Weisenfeld2009}.
 
@@
 
-EM algorithms are sensitive to local optima. Consequently, a good initialization
-strategy is critical.  Common low-level initialization strategies include uniform
-probability assignment \citep{Wells1996}, Otsu thresholding
-\citep{Zhang2001}, and K-means clustering \citep{Pappas1992}.  More
-sophisticated low-level initialization schemes include that of
-\cite{Greenspan2006} in which a dense spatial distribution of
-Gaussians is used to capture the complex neuroanatomical layout with
-subsequent processing used to conjoin subsets of Gaussians belonging
-to the same tissue classes.  Recently, anatomically specific
-strategies for incorporating prior knowledge into the segmentation
-solution have included the use of spatial prior probability maps of
-the various structures of interest
+EM may easily be caught in local optima. Consequently, a good
+initialization is critical.  Common low-level initialization steps 
+include uniform probability assignment \citep{Wells1996},
+Otsu thresholding \citep{Zhang2001}, and K-means clustering
+\citep{Pappas1992}.  More sophisticated low-level initialization
+schemes include that of \cite{Greenspan2006} in which a dense spatial
+distribution of Gaussians is used to capture the complex
+neuroanatomical layout with subsequent processing used to conjoin
+subsets of Gaussians belonging to the same tissue classes.  Recently,
+anatomically specific strategies for incorporating prior knowledge
+into the segmentation solution have included the use of spatial prior
+probability maps of the various structures of interest
 \citep{Leemput1999a,Marroquin2002,Ashburner2005}.  These spatial prior
 probability maps can also be used to provide an initial segmentation.
-Related technological developments model partial
-volume effects for increased accuracy in brain segmentation
+Related technological developments model partial volume effects for
+increased accuracy in brain segmentation
 \citep{Ruan2000,Ballester2002,Leemput2003}.
@@
 statistical parametric mapping was developed \citep{Friston1990} and
-subsequently provided to the research community. A similar holistic
-approach to brain processing was provided in \cite{Pohl2006} in which
+subsequently provided to the research community.  Similar integrative brain processing was provided in \cite{Pohl2006} in which
 segmentation and registration parameters were optimized simultaneously
@@
 accurately segmenting the T1 MRI of elderly controls and subjects
-suffering from neurodegeneration, for instance, via the SIENA algorithm
-\citep{Smith2007}.  A recent evaluation study 
+suffering from neurodegeneration, for instance, via SIENA \citep{Smith2007}.  A recent evaluation study 
 compared kNN segmentation, SPM Unified Segmentation and SIENA and
@@
 time-consumption and tedium associated with manual effort towards the
-latter, various algorithms have been proposed of which a small subset
+latter, various techniques have been proposed of which a small subset
 has been publicly availed to the research community such as the
@@
 and displayed as tessellated surfaces  \cite{Dale1999,Fischl1999}.
-Advantages of surface-based approaches include the ability to map
+Advantages of surface representations include the ability to map
 processed neuroanatomy to simple geometric primitives such as spheres
@@
 EM segmentation in a semi-interactive manner.  In short, Atropos seeks
-to provide a toolbox of segmentation approaches that may be modified,
+to provide a segmentation toolbox that may be modified,
 tuned and refined by the user for different problems.
@@
 We also allow Atropos to address the bias field issue by incorporating
-the recently developed N4 bias correction algorithm
-\citep{Tustison2010} in an adaptive manner.  Coupled with the
-registration \citep{Avants2010b} and template building strategies
+the recently developed N4 bias correction software \citep{Tustison2010} in an adaptive manner.  Coupled with the
+registration \citep{Avants2010b} and template building 
 \citep{Avants2010} already included in the ANTs toolkit, Atropos is a
@@
 
-The Atropos algorithm is multi-platform, free, stand-alone and able to
+Atropos is multi-platform, free, stand-alone and able to
 manage a wide range of scalar and vector image datatypes.  To highlight the
@@
 memory.  We evaluate EM-based classification of $1 mm^3$ T1 MR
-neuroimages into 69 neuronatomical regions to illustrate the practical
+neuroimages into 69 neuroanatomical regions to illustrate the practical
 value of the low-memory implementation within this paper.  Atropos
@@
 Organization of this work is as follows: we first describe the theory
-behind the various algorithmic components of Atropos while
+behind the various components of Atropos while
 acknowledging that theoretical discussion is available elsewhere,
@@
 this paper and of previous and future work.
-
-%  We evaluate this
-%method on BrainWeb 20 data and show that by relying upon accurate,
-%high-resolution diffeomorphic image registration, Atropos ~is capable of
-%reliably segmenting the BW20 whole head data into separate tissue
-%classes for muscle, skin, skull and bone marrow as well as the
-%standard three tissues, cerebrospinal fluid and gray and white matter.
-%Atropos 's three-tissue performance compares favorably with the standard
-%EM-MRF algorithm, FAST.  We also use the OASIS dataset to show that
-%Atropos ~increases segmentation repeatability relative to FAST.  The
-%evaluation data, the data-derived templates, the code and application
-%itself are publicly available.
-%}
-%\end{comment}
 
@@
 \end{itemize}
-These components are common across most EM segmentation approaches.
+These components are common across most EM segmentation algorithms.
 
@@
 \left\{p(\mathbf{y}|\mathbf{x})p(\mathbf{x})\right\}.
-\end{align} Similar to its algorithmic predecessors, Atropos employs
+\end{align} Similar to its predecessors, Atropos employs
 the EM framework \citep{Dempster1977} to find maximum likelihood
-solutions to this problem.  The following sections detail
-Atropos choices for the likelihood(s), prior(s) and EM strategy.
+solutions to this problem.  The following sections detail the 
+Atropos EM along with choices for the likelihood(s), prior(s).
 
@@
 \subsubsection{Template-Based Priors}
-A number of researchers have used template-based strategies to both ensure spatial coherence and incorporate prior knowledge in segmentation.
-A common technique is to select labeled subjects from a population from which a template is constructed \citep[e.g.][which is also available in the ANTs toolkit]{Avants2010}.
-Each labeling can then be warped to the template where the synthesis of warped labeled regions produces a prior probability map or prior label map encoding the spatial distribution of labeled anatomy which can be harnessed in joint segmentation/registration or, specifically, Atropos/ANTs strategies involving unlabeled subjects.
+A number of researchers have used templates to both
+ensure spatial coherence and incorporate prior knowledge in
+segmentation.  A common technique is to select labeled subjects from a
+population from which a template is constructed \citep[e.g.][which is
+also available in the ANTs toolkit]{Avants2010}.  Each labeling can
+then be warped to the template where the synthesis of warped labeled
+regions produces a prior probability map or prior label map encoding
+the spatial distribution of labeled anatomy which can be harnessed in
+joint segmentation/registration or, specifically, Atropos/ANTs hybrids involving unlabeled subjects.
 
@@
 via the command line interface to the posterior formulation.
-{\textcolor{red}{FIXME --- need explanation of heuristic ... how do we
-  estimate the $\gamma$ without the $t_{ij}$ present?}
 
@@
 Atropos is capable of benefitting from user knowledge via an
-initialization and optimization strategy that depends upon a spatially
+initialization and optimization that depends upon a spatially
 varying prior label image passed as input.  Rapid, sparse
@@
 challenging clinical data in which automated approaches fail.  This
-part of Atropos design is inspired by the interactive graph cut methods
+part of Atropos design is inspired by the interactive graph cuts
 pioneered by \cite{Boykov2001} and which has spawned many follow-up
@@
 and $0$ for each other label at the same site.  The user-input
-therefore not only initializes the algorithm, but also gives boundary conditions
+therefore not only initializes the optimization, but also gives boundary conditions
 that influence the EM solution outside of the known sites.  While the
 graph-based min-cut max-flow solution is globally optimal for two
-labels, only locally optimal approachs are available for 3 or more
-classes.  Thus, in most practical applications, the EM approach is a
+labels, only locally optimal optimizers are available for 3 or more
+classes.  Thus, in most practical applications, EM is a
 reasonable and efficient alterative to Boykov's solution.
-Furthermore, as with similar algorithms, one may automate the
+Furthermore, one may automate the
 initialization process.
@@
 \caption{\baselineskip 12pt \small Graphical representation of the distance prior probability calculation associated with label $l_k$.
-This extension of the popular template-based prior probability strategy permits a dense labeling with the user specified mask.
+This extension of the popular template-based prior probability permits a dense labeling with the user specified mask.
 }
@@
 \end{tabular}
-\caption{\baselineskip 12pt \small  Illustration of label propagation differences between the Euclidean and geodesic approaches which
+\caption{\baselineskip 12pt \small  Illustration of label propagation differences between the Euclidean and geodesic distances which
 are particularly acute in sinuous structures such as the cortex.  (a) Given a mask, represented in white, and a sparse labeling,
@@
 Due to the lack of parameters in the non-parametric approach, it is not technically an EM algorithm (as described in \cite{Wells1996}).
-However, the same iterative strategy is applicable and is quite robust in practice as evidenced by the number of researchers employing non-parametric strategies (see the Introduction).
+However, the same iterative maximization is applicable and is quite robust in practice as evidenced by the number of researchers employing non-parametric models (see the Introduction).
 }  
@@
   Modes} (ICM)  \citep{Besag1986} which is greedy, computationally
-efficient and provides good performance.  The EM algorithm employed in
+efficient and provides good performance.  The EM employed in
 Atropos may therefore be written as a series of steps:
@@
 maximizes the posterior.  {\bf \textcolor{red}{ FIXME more specific?}}.
-\item[M-Step: Parameter Update:] We describe the parametric update.  Given fixed labels, we update the
-  parameters by the FMM-defined likelihood.  We compute, for each of
-  $K$ labels, the mean,
-$$ \mu^{new}_k \leftarrow \frac{ \sum_{i=1}^N  y_i \gamma_k p_k(y_i|l_k)}{ \sum_{i=1}^N \gamma_k p_k(y_i|l_k) },
+\item[M-Step: Parameter Update:] 
+The posteriors used in this step are called $p^{old}$ to make it
+clear that we use the previous parameter estimates.
+We use a common and elementary estimate of the mixing parameters:
+\begin{align}
+\gamma^{new}_k \leftarrow  \sum_i \frac{p^{old}_{ik}}{\sum_{j=1}^K p^{old}_{ij}}. 
+\end{align}
+For parametric model update, for each of
+  $K$ labels, we compute the mean,
+$$ \mu^{new}_k \leftarrow \frac{ \sum_{i=1}^N  y_i p^{old}_k(l_k|y_i)}{ \sum_{i=1}^N p^{old}_k(l_k|y_i) },
 $$
-variance {\bf \textcolor{red}{ FIXME should this be the full
+and variance {\bf \textcolor{red}{ FIXME should this be the full
     covariance matrix?}},
-$$ \sigma^{new}_k \leftarrow  \frac{ \sum_{i=1}^N  (y_i - \mu_k ) \gamma_k p_k(y_i|l_k)}{ \sum_{i=1}^N  \gamma_k p_k(y_i|l_k) },
+$$ \sigma^{new}_k \leftarrow  \frac{ \sum_{i=1}^N  (y_i - \mu_k ) p^{old}_k(l_k|y_i)}{ \sum_{i=1}^N p^{old}_k(l_k|y_i) }.
 $$
-and mixture parameters, 
-$$ \gamma^{new}_k \leftarrow  \textcolor{red}{ FIXME .}
-$$
-This type of update strategy is known as {\em soft} EM.  Hard EM, in
+This type of update is known as {\em soft} EM.  Hard EM, in
 contrast, only uses sites containing label $l_k$ to update model $k$ parameters.
-A similar strategy is used in the multivariate and non-parametric cases.
+A similar pattern is used in the multivariate and non-parametric cases.
 \end{description}
-The EM algorithm will iterate toward a local maximum.  We track
-convergence of the algorithm by summing up the maximum posterior
+EM will iterate toward a local maximum.  We track
+convergence by summing up the maximum posterior
 probability at each site over the segmentation domain.  
@@
 \end{tabular}
-\caption{\baselineskip 12pt \small Flowchart illustrating Atropos usage typically beginning with bias correction using the N4 algorithm.  
-Initialization provides an estimate before the iterative segmentation portion of the algorithm in which the 
+\caption{\baselineskip 12pt \small Flowchart illustrating Atropos usage typically beginning with bias correction via N4.  
+Initialization provides an estimate before the iterative optimization in which the 
 likelihood models for each class are tabulated from the current estimate followed by a recalculation of the 
-posterior probabilities associated with each class.  Label propagation ensues to ensure a dense 
-classification throughout the entire mask.   The multiple options associated with the different algorithmic 
+posterior probabilities associated with each class.  The multiple options associated with the different algorithmic 
 components are indicated by the colored rounded rectangles connected to their respective core Atropos 
@@
 \subsection{The Atropos User Interface}
-As with other classes that comprise ANTs, Atropos uses the Insight Toolkit as a developmental foundation.  
-This allows us to take advantage of the mature portions of ITK (e.g. image IO) and ensures the 
-integrity of the ancillary processes such as those facilitated by the underlying statistical framework.  Although Atropos is 
-publicly distributed with the rest of the ANTs package, we plan to contribute the Atropos segmentation
-framework to the Insight Toolkit where it can be vetted and further
+As with other classes that comprise ANTs, Atropos uses the Insight
+Toolkit as a developmental foundation.  This allows us to take
+advantage of the mature portions of ITK (e.g. image IO) and ensures
+the integrity of the ancillary processes such as those facilitated by
+the underlying statistical framework.  Although Atropos is publicly
+distributed with the rest of the ANTs package, we plan to contribute
+the Atropos to the Insight Toolkit where it can be vetted and further
 improved by other interested researchers.
@@
 \subsection{Initializing the Atropos Objective}
-The Atropos algorithm has a number of parameters defined within
+Atropos has a number of parameters defined within
 Listing \ref{listing:n4} and will function on 2, 3 or 4 dimensional
@@
 \item  {\ttfamily -i  PriorLabelImage[3, label\_image.nii.gz , w ]}
-  where scalar $w \in [0,1]$ is unused \textcolor{red}{ FIXME .} 
+  where scalar $w \in [0,1]$ is unused. 
 \item  {\ttfamily -i  PriorProbabilityImages[3, label\_prob\%02d.nii.gz ,
@@
 different classes during the same run for a hybrid
-parametric/non-parametric approach although this possibility has yet
+parametric/non-parametric model although this possibility has yet
 to be fully explored.
@@
 \subsubsection{Registration of Probability Maps} 
-Registration approaches enable the ability to transform information
+Image registration enables the ability to transform information
 between spatial domains which may aid in both segmentation and bias
@@
 \times 300 \times 256$ with 32 cortical labels.  Our novel memory
-minimizing strategy typically shrinks run-time memory usage from a
+minimizing image representation typically shrinks run-time memory usage from a
 peak of 10+ GB to approximately 1.5 GB and enable these datasets to be
@@
 algorithm \citep{Tustison2010} which is based on the popular
-nonparametric nonuniformity intensity normalization (N3) method
-introduced in \cite{Sled1998}.  It was previously mentioned that
+nonparametric nonuniformity intensity normalization (N3) introduced in \cite{Sled1998}.  It was previously mentioned that
 several methods proposed in the literature have taken an integrative
 view of the segmentation problem by incorporating an intrinsic bias
-correction step into the actual algorithmic workflow.  The
-advancements introduced with the N4 algorithm permit such an adaptive
+correction step into the actual workflow.  The
+advancements introduced with N4 permit such an adaptive
 integration with Atropos.  Recent demonstrations suggest improved
@@
 N3 \citep{Boyes2008}.  Thus, when performing 3-tissue segmentation, we
-use the posterior probability map of white matter at the current
+may opt to use, for instance, the posterior probability map of white matter at the current
 iteration as a weighted mask for input to N4.  This is done by setting
@@
 corresponding to the white matter produced as output in the Atropos
-call, i.e. `{\ttfamily Atropos --output}'.  The N4 algorithm is
+call, i.e. `{\ttfamily Atropos --output}'.  N4 is
 currently being added as a standard part of the Insight ToolKit.  The
 evaluation section will illustrate inclusion of Atropos, N4 and ANTs
-in a brain processing pipeline.
+in a brain processing pipeline.  An example of this procedure, using brainweb data with
+40\% RF bias, is in figure~\ref{fig:bwebrf40}.  We supply the
+information necessary to repeat this example in the script entitled
+`{\ttfamily  atroposBwebRF40FigureExample.sh}' which is available in the ANTs toolkit's Atropos
+documentation folder as of svn commit 711.
+\begin{figure}
+\begin{center}
+\begin{tabular}{c}
+\includegraphics[width=6in]{Figures/bwebN4ex.pdf}
+\end{tabular}
+\caption{\baselineskip 12pt \small We combine N4 and Atropos by simple
+sequential processing and apply to brainweb single-subject data with
+40\% RF bias and 3\% noise.  The $\beta$ for the MRF term is, here,
+$0.2$.  Slice 71 of the input data is in (a).  The initial K-means
+($K=3$) segmentation is in panel (b).  We use the brain mask to guide
+N4 bias correction and produce the image, shown in (c).  We repeat the
+K-means segmentation, but with the N4 corrected image as input, and
+produce the segmentation in (d).  The average 3-tissue Dice overlap of
+result (b) is 0.906 while the average overlap for (d) is 0.954.
+Arrows highlight a region of large before-after segmentation
+discrepancy.}
+\label{fig:bwebrf40}
+\end{center}
+\end{figure}
 
@@
 anatomy.  Second, we evaluate the use of Atropos in improving whole
-brain parcellation and excercise its ability to efficiently solve {\em
+brain parcellation and exercise its ability to efficiently solve {\em
 many class} expectation maximization problem.  We choose this
@@
 data with 3\% noise and bias 0, 20 and 40 \% RF inhomogeneity.  We
-study the effect of the MRF prior term and initialization strategy on
+study the effect of the MRF prior term and initialization on
 the Dice overlap between ground truth and the segmentation result for
@@
 but with the N4 corrected image as input.  The resulting algorithm is
-similar to those methods that fix segmentation parameters while
+similar to those that fix segmentation parameters while
 estimating bias and fix bias while estimating segmentation parameters.
 Thus, with this simple evaluation, we are able to compare the impact
-of bias on our N4 and Atropos algorithm and also the validity of our
+of bias on our N4 and Atropos and also the validity of our
 prior label image initialization.  The prior label image manual
@@
 set of voxels for each tissue.  Results, in terms of
-Dice overlap, are shown in figure~\ref{fig:bweb1}.
+Dice overlap, are shown in figure~\ref{fig:bweb1}  Because overlap
+ratios with N4 bias correction approximate those of the zero bias data, we may
+conclude that simple N4 pre-processing is adequate to correct even the 40\%
+bias level.
 \begin{figure}
@@
 \caption{\baselineskip 12pt \small Brainweb single-subject results for
-  each tissue.  The results show that the N4 bias correction
-  algorithm, combined with Atropos, results in a minimal effect of
+  each tissue.  The results show that N4 bias correction, combined with Atropos, results in a minimal effect of
   bias, even at the 40\% level.  The optimal $\beta$ for the MRF term
@@
 \label{sec:hammers}
-We evaluate the ability to improve multi-template
-labeling results by converting the group labels to probability maps
-and using them to drive many-class EM segmentation.  The procedure we
-use here is similar to that described in \citep{Avants2010a}.  The
-ground truth labels cover 69 classes and much of the brain.  Some
-unlabeled regions remain which we assign to label 69 such that all
-brain parenchyma contains a unique label.  The initialization of our
+We evaluate the ability to improve multi-template labeling results by
+converting the group labels to probability maps and using them to
+drive many-class EM segmentation.  The ground truth labels cover 69
+classes and much of the brain.  Some unlabeled regions remain which we
+assign to label 69 such that all brain parenchyma contains a unique
+label.  Following \cite{Avants2010a}, the initialization of our
 evaluation applies the script {\verb ants_multitemplate_labeling.sh }
-(available in the ANTs toolkit) to the 19 Hammers evaluation
-datasets located at \url{http://www.brain-development.org/} \citep{Hammers2003,Heckemann2006}.
-These initial majority voting results are competitive with prior work
+(available in the ANTs toolkit) to the 19 Hammers evaluation datasets
+located at \url{http://www.brain-development.org/}
+\citep{Hammers2003,Heckemann2006}.  These initial majority voting
+results are competitive with prior work
 \citep{Heckemann2006,Heckemann2010} and serve as a baseline against
@@
 We first convert each of the 69 labels within the original evaluation
-dataset to an individual image.  The remainder of the procedure,
-summarized briefly, is the same for each of the 19 subjects.  We
+dataset to an individual image.  The remaining steps,
+summarized briefly, are the same for each of the 19 subjects.  We
 select one subject as an unlabeled target.  The other 18 datasets are
@@
 the Hammers evaluation data is based on freely available scripts
-included with the ANTs toolkit.  {\bf \textcolor{red}{ FIXME}}.
+included with the ANTs toolkit, specifically  {\verb ants_multitemplate_labeling.sh } and  {\verb atroposBwebRF40FigureExample.sh }.
 
 \section{Discussion}
-We introduced the Atropos algorithm, the theory and implementation
+We introduced Atropos, the theory and implementation
 details and documented its performance in a variety of use cases.  We
-also showed evidence that the openly available N4 bias correction
-method can easily used with Atropos to improve segmentation.
+also showed evidence that the openly available N4 bias correction can easily used with Atropos to improve segmentation.
 Furthermore, we used multiple subject brainweb data to build dataset
@@
 instance, \cite{Ashburner2005} (SPM5) evaluated on 0\% RF (bias field) 3\%
-noise braineweb single subject data finding 0.932 (GM) and .961 (WM)
+noise brainweb single subject data finding 0.932 (GM) and .961 (WM)
 Dice. Results on 40\% RF were .934 (GM) and .961 (WM). SPM5 exhibits
@@
 results.  However, \cite{Nakamura2009} do not report WM or CSF numbers
-for comparison.  Topology preserving approaches also perform well.
+for comparison.  Topology preserving methods also perform well.
 \cite{Shiee2010} achieved Dice results for 3\% noise 20\% RF brainweb
@@
 constraint applied to the segmentation.  
-\cite{Bazin2007c} proposed the TOADS method and, estimating from the
+\cite{Bazin2007c} proposed TOADS and, estimating from the
 paper's graph, showed that the average Dice accuracy for 3\% noise various RF was
@@
 
-\section{Conclusion}
 A more general advantage which extends beyond the scope of the experimental 
-evaluation section of this paper is the flexibility of the Atropos algorithm.   This includes 
+evaluation section of this paper is the flexibility of Atropos.   This includes 
 not only $n$-tissue segmentation and dense volumetric cortical parcellation,
@@
 
-The Atropos algorithm is freely available to the public.  We release
+\section{Conclusion}
+The Atropos software is freely available to the public.  We release
 this code not only to make it available to clinical researchers but
@@
 segmentation is non-trivial and there are numerous design alternatives
-available not only in the models selected but also in the ICM coding
-strategy, alternatives to ICM and the specific method in which prior
+available not only in the models selected but also in the ICM coding, alternatives to ICM and the specific method in which prior
 and likelihood are combined.  Due to the flexibility of Atropos, we

### 2010-12-28 Documentation/atropos/atroposBwebRF40FigureExample.sh

@@ -1,0 +1,43 @@
+# input image 
+IMG=T1_ICBM_1mm_3perc_40_bias/t1_icbm_normal_1mm_pn3_rf40.mhd
+if [ ! -s $IMG ] ; then 
+ echo no image named $IMG exists --- exiting.   you should be able to download this image from the brainweb website and use the header defined below to make it readable by Atropos. 
+ exit
+fi
+# number of classes
+NC=3
+# convergence criterion 
+CONV=[3,0]
+# MRF-beta parameter 
+MRF=0.20
+# socrates parameter --- estimates mixture model
+SOC=1
+# kmeans with N4
+NAMING=bwebRF40example
+# binary image used as a weight and mask image for N4 and segmentation mask for Atropos 
+# you can get this by thresholding the ground_truth labeling 
+WTIM=mask.nii.gz
+$EX  -d 3 -a  $IMG -i KMeans[$NC] -o [test_init.nii.gz,test_init_prob%02d.nii.gz] -m [${MRF},1x1x1] -x mask.nii.gz -p Socrates[${SOC}]  -c $CONV 
+BS="N4BiasFieldCorrection -d 3  -h 0 " ; ITS=20 ; ITS2=3
+$BS  -i $IMG   -o  n4.nii.gz -s 2 -b [200] -w mask.nii.gz -c [${ITS}x${ITS}x${ITS}x${ITS}x${ITS2},0.0001] -x mask.nii.gz 
+Atropos -d 3 -a  n4.nii.gz -i KMeans[$NC] -o [test.nii.gz,test_prob%02d.nii.gz] -m [${MRF},1x1x1] -x mask.nii.gz -p Socrates[${SOC}]  -c $CONV 
+# evaluation numbers
+if [ -s ground_truth.nii.gz ] ; then 
+ImageMath 3 dice_out.txt DiceAndMinDistSum test.nii.gz ground_truth.nii.gz 
+fi
+exit
+
+# below this line is a header file that can be used for the brain web raw data:
+ObjectType = Image
+NDims = 3
+BinaryData = True
+BinaryDataByteOrderMSB = False
+CompressedData = False
+TransformMatrix = 1 0 0 0 1 0 0 0 1
+Offset = 0 0 0
+CenterOfRotation = 0 0 0
+AnatomicalOrientation = RAI
+ElementSpacing = 1 1 1
+DimSize = 181 217 181
+ElementType = MET_UCHAR
+ElementDataFile = t1_icbm_normal_1mm_pn3_rf40.rawb

### 2010-12-28 Documentation/atropos/atropos.tex
ENH:  A few minor edits.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@712 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 {\Large \bf An Open Source Framework for Multivariate $n$-Tissue
-  Segmentation and Evaluation Using Open Data Sets} \\
+  Segmentation with Evaluation Using Open Data Sets} \\
 \vspace*{0.5cm}
@@
 We introduce {\em Atropos}, an ITK-based multivariate $n$-class
-Expectation Maximization (EM) segmentation tool based on either parametric or non-parametric finite mixture modeling (FMM) of the
-class intensities.  Atropos, available in ANTs
+open source segmentation algorithm distributed
+with ANTs.%
 \footnote{
 http://www.picsl.upenn.edu/ANTs
-}, is capable of incorporating spatial prior probability maps,
+}  The Bayesian formulation of the segmentation problem is 
+solved using the Expectation Maximization (EM) algorithm
+with the modeling of the class intensities 
+ based on either parametric or non-parametric finite mixtures.  
+Atropos is capable of incorporating spatial prior probability maps,
 (sparse) prior label maps and/or Markov Random Field (MRF) modeling.  Atropos has also been efficiently implemented to handle
@@
 also evaluate Atropos performance by using spatial priors to drive a 69 class EM segmentation
-problem defined by the Hammers atlas from University College London.  These evaluation studies, combined with
+problem derived from the Hammers atlas from University College London.  These evaluation studies, combined with
 illustrative examples that exercise Atropos options, demonstrate both
@@
 \section{Introduction} As medical image acquisition technology has advanced,
-concomitant investment has been made towards adapting classification
-techniques for neuroanatomy.  Some of the earliest work appropriated NASA
+significant investment has been made towards adapting classification
+techniques for neuroanatomy.  Early work appropriated NASA
 satellite image processing software for statistical classification of
@@
 The problem of accurately delineating the white matter, grey matter
-and cerebrospinal fluid of the primate brain continuously spurs
+and cerebrospinal fluid (and further subdivisions) of the human brain continuously spurs
 technical development in segmentation.  Following \cite{Vannier1985},
@@
 brain segmentation.  Given the ``missing data'' aspect of this
-problem, use of the Expectation-Maximization (EM) framework is natural\citep{Dempster1977}.  The
+problem, use of the Expectation-Maximization (EM) framework is natural \citep{Dempster1977}.  The
 work described in \cite{Wells1996} was one of the first to use EM for
@@
 
-EM may easily be caught in local optima. Consequently, a good
-initialization is critical.  Common low-level initialization steps 
+Several initialization strategies have been proposed to overcome the characteristic
+susceptibility of EM algorithms to local optima.   Common low-level initialization steps 
 include uniform probability assignment \citep{Wells1996},
@@
 neuroanatomical layout with subsequent processing used to conjoin
-subsets of Gaussians belonging to the same tissue classes.  Recently,
+subsets of such Gaussians belonging to the same tissue classes.  Recently,
 anatomically specific strategies for incorporating prior knowledge
@@
 optimization scheme. The roots of this very popular software package
-stem back to early work by Karl Friston in which the basis for
-statistical parametric mapping was developed \citep{Friston1990} and
-subsequently provided to the research community.  Similar integrative brain processing was provided in \cite{Pohl2006} in which
+stem back to early work by Karl Friston which laid the basis for
+statistical parametric mapping \citep{Friston1990}. 
+ Similar integrative brain processing was provided in \cite{Pohl2006} in which
 segmentation and registration parameters were optimized simultaneously
@@
 manner \citep{Pohl2007}.  An implementation is provided in 3D slicer
-\citep{Pieper2006}---an open source package with developmental
+\citep{Pieper2006}---an open source medical image computation and
+visualization package with developmental
 contributions from multiple agencies including both private and
-academic institutions. Researchers in aging often focus on
-accurately segmenting the T1 MRI of elderly controls and subjects
-suffering from neurodegeneration, for instance, via SIENA \citep{Smith2007}.  A recent evaluation study 
-compared kNN segmentation, SPM Unified Segmentation and SIENA and
-found different performance
-characteristics under different evaluation criteria
-\citep{Bresser2011}.  \cite{Klauschen2009} had similar findings when
-comparing SPM5, FSL and FreeSurfer.  These studies suggest that no
-single method performs best under every measurement and, along with the No
-Free Lunch theorem, highlight the need for segmentation tools that are tunable for different
-problems and research goals.
+academic institutions.
 
@@
 grey-white matter interface and pial surfaces are processed, analyzed,
-and displayed as tessellated surfaces  \cite{Dale1999,Fischl1999}.
+and displayed as tessellated surfaces  \citep{Dale1999,Fischl1999}.
 Advantages of surface representations include the ability to map
@@
 or planes and the ease of including topological constraints in the
-analysis workflow.  These types of methods, as well as Klein's
-Mindboggle \cite{Klein2005}, would usually follow an initial segmentation by a volumetric method such as Atropos.
+analysis workflow.  These types of methods, including Klein's
+Mindboggle \citep{Klein2005}, would usually follow an initial segmentation by a 
+volumetric method such as Atropos.
 
@@
 
+Researchers in aging often focus on
+accurately segmenting the T1 MRI of elderly controls and subjects
+suffering from neurodegeneration, for instance, via SIENA \citep{Smith2007}.  A recent evaluation study 
+compared kNN segmentation, SPM Unified Segmentation and SIENA and
+found different performance
+characteristics under different evaluation criteria
+\citep{Bresser2011}.  \cite{Klauschen2009} had similar findings when
+comparing SPM5, FSL and FreeSurfer.  These studies suggest that no
+single method performs best under every measurement and, along with the No
+Free Lunch theorem \citep{Wolpert1997}, highlight the need for segmentation tools that are tunable for different
+problems and research goals.
+
 Our open source segmentation tool, which we have dubbed {\em
-Atropos}, 
-\footnote{ Atropos is one of the three Fates from Greek
+Atropos},%
+\footnote{Atropos is one of the three Fates from Greek
 mythology characterized by her dreaded shears used to decide the
@@
 generalized EM algorithm for standard tissue classification of
-the brain into gray matter, white matter and cerebrospinal fluid.
+the brain into gray matter, white matter and cerebrospinal fluid even
+in cases of multivariate image data---relevant when more than one 
+view of anatomy aids segmentation, as in neonatal brain tissue classification
+\citep[e.g.][]{Prastawa2005,Weisenfeld2009}.  
 Atropos equally allows incarnations that use EM to simultaneously
-maximize the posterior probabilities of many classes, for instance,
+maximize the posterior probabilities of many classes with minimal
+random access memory requirements, for instance,
 when parcellating the brain into hemispheres, cortical regions and
@@
 to provide a segmentation toolbox that may be modified,
-tuned and refined by the user for different problems.
+tuned and refined for different use scenarios.
 
@@
 
-Atropos is multi-platform, free, stand-alone and able to
-manage a wide range of scalar and vector image datatypes.  To highlight the
+%Atropos is multi-platform, free, /stand-alone and able to
+%manage a wide range of scalar and vector image datatypes.
+
+To highlight the
 value of this open-source contribution, we performed a search of
@@
 stand-alone EM methods are currently listed.
-Atropos's memory efficiency enables EM-segmentation of many classes at
-once, a capability that is due to a sparse internal representation of
-prior probability images that minimizes the need for random access
-memory.  We evaluate EM-based classification of $1 mm^3$ T1 MR
+%Atropos's memory efficiency enables EM-segmentation of many classes at
+%once, a capability that is due to a sparse internal representation of
+%prior probability images that minimizes the need for random access
+%memory.  
+We evaluate EM-based classification of $1 mm^3$ T1 MR
 neuroimages into 69 neuroanatomical regions to illustrate the practical
-value of the low-memory implementation within this paper.  Atropos
-also allows multivariate image features to drive the data term, relevant when more than one view of anatomy aids
-segmentation, as in neonatal brain tissue classification
-\cite{Weisenfeld2009}.  Although Atropos may be applied to
+value of the low-memory implementation within this paper.  
+Although Atropos may be applied to
 multivariate data from arbitrary modalities, we limit our evaluation
@@
 behind the various components of Atropos while
-acknowledging that theoretical discussion is available elsewhere,
-e.g. in previously cited literature.  This is followed by a thorough
+acknowledging that additional theoretical discussion is available elsewhere.  
+This is followed by a thorough
 discussion of implementation which, though often overlooked, is of
@@
 standard labelings.  Finally, we provide a discussion of our results
-and our open-source contribution in the context of the remainder of
+and our open source contribution in the context of the remainder of
 this paper and of previous and future work.
@@
 given in the literature, has often been formed either as
-MRF-based or template-based.  An image segmentation solution in this
-context is an assignment of one label to each voxel
+MRF-based and/or template-based.  An image segmentation solution in this
+context is an assignment of one label to each voxel%
 \footnote{
@@
 
-\subsection{Segmentation Objective Function} Atropos optimizes a class
-of user-selectable maximization objective functions each of which may
+\subsection{Segmentation Objective Function}
+
+ Atropos optimizes a class
+of user-selectable objective functions each of which may
 be represented in a generic Bayesian framework, as described by \cite{Sanjay-Gopal1998}.  This framework
@@
 solutions to this problem.  The following sections detail the 
-Atropos EM along with choices for the likelihood(s), prior(s).
+Atropos EM along with choices for the likelihood and prior terms.
 
@@
 and define multivariate Gaussian parameters via a mean vector,
-$\boldsymbol{\mu}_k$, and covariance matrix, ${\bf \Sigma_k}$.
+$\boldsymbol{\mu}_k$, and covariance matrix, ${\bf \Sigma}_k$.
 
@@
 as a constant.  The Atropos implementation estimates $\gamma_k$ at
-each iteration as in \cite{Ashburner2005}.
+each iteration as in \cite{Ashburner2005}.%
 \footnote{
@@
 Consider $N$ sites each with a possible $K$ labels for a total of $N^K$ possible labeling configurations.
-Due to the large $K$ associated with $K>>1$ problems, exact optimization is even more problematic than for the traditional 3-tissue scenario.
+Due to the large $K$ associated with $K>>1$ problems, exact optimization is even more intractable than for the traditional 3-tissue scenario.
 }
@@
 prior probability images, one for each class.
-\item[E-Step: ICM Label Update:] Given the initialization and fixed
+\item[ICM Label Update (E-Step):] Given the initialization and fixed
 model parameters, Atropos will evaluate the posterior probability for
@@
 maximizes the posterior.  {\bf \textcolor{red}{ FIXME more specific?}}.
-\item[M-Step: Parameter Update:] 
-The posteriors used in this step are called $p^{old}$ to make it
-clear that we use the previous parameter estimates.
+\item[Parameter Update (M-Step):] 
+Note that the posteriors used in the previous iteration are used to 
+estimate the parameters at the current iteration.  
 We use a common and elementary estimate of the mixing parameters:
 \begin{align}
-\gamma^{new}_k \leftarrow  \sum_i \frac{p^{old}_{ik}}{\sum_{j=1}^K p^{old}_{ij}}. 
+\gamma_k \leftarrow \frac{1}{N} \sum_{i=1}^N  p_k(l_k|y_i)
+%\gamma^{new}_k \leftarrow  \sum_i \frac{p^{old}_{ik}}{\sum_{j=1}^K p^{old}_{ij}}. 
 \end{align}
@@
   $K$ labels, we compute the mean,
-$$ \mu^{new}_k \leftarrow \frac{ \sum_{i=1}^N  y_i p^{old}_k(l_k|y_i)}{ \sum_{i=1}^N p^{old}_k(l_k|y_i) },
-$$
-and variance {\bf \textcolor{red}{ FIXME should this be the full
-    covariance matrix?}},
-$$ \sigma^{new}_k \leftarrow  \frac{ \sum_{i=1}^N  (y_i - \mu_k ) p^{old}_k(l_k|y_i)}{ \sum_{i=1}^N p^{old}_k(l_k|y_i) }.
-$$
+\begin{align}  
+ \mu_k \leftarrow \frac{ \sum_{i=1}^N  y_i p_k(l_k|y_i)}{ \sum_{i=1}^N p_k(l_k|y_i) },
+\end{align}
+and variance,
+\begin{align}  
+ \sigma^2_k \leftarrow  \frac{ \sum_{i=1}^N  (y_i - \mu_k ) p_k(l_k|y_i)}{ \sum_{i=1}^N p_k(l_k|y_i) }.
+\end{align}
+The latter two quantities are modified, respectively, in the case of multivariate data as follows:
+\begin{align}
+ \boldsymbol{\mu}_k \leftarrow \frac{ \sum_{i=1}^N  \mathbf{y}_i p_k(l_k|\mathbf{y}_i)}{ \sum_{i=1}^N p_k(l_k|\mathbf{y}_i) },
+\end{align}
+and the $(m,n)^{th}$ element of the covariance matrix, $\boldsymbol{\Sigma}$, is calculated from
+\begin{align}
+ \boldsymbol{\Sigma}_{mn,k} \leftarrow \frac{ \sum_{i=1}^N  ( \mathbf{y}_i - \boldsymbol{\mu}_k ) p_k(l_k|\mathbf{y}_i)}{1 - \sum_{i=1}^N p^2_k(l_k|\mathbf{y}_i) }.
+\end{align}
+
 This type of update is known as {\em soft} EM.  Hard EM, in

### 2010-12-28 Documentation/atropos/atropos.tex
ENH: dealt with all fixmes, grant support, afaik ready to be submitted modulo a 2nd reading
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@714 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 each possible label and assign, to each voxel, the label that
-maximizes the posterior.  {\bf \textcolor{red}{ FIXME more specific?}}.
+maximizes the posterior. 
+% {\bf \textcolor{red}{ FIXME more specific?}}.
 \item[Parameter Update (M-Step):] 
@@
 \gamma_k \leftarrow \frac{1}{N} \sum_{i=1}^N  p_k(l_k|y_i)
-%\gamma^{new}_k \leftarrow  \sum_i \frac{p^{old}_{ik}}{\sum_{j=1}^K p^{old}_{ij}}. 
-\end{align}
-For parametric model update, for each of
-  $K$ labels, we compute the mean,
+\end{align}
+We update the parametric model parameters by computing, for each of
+  $K$ labels, the mean,
 \begin{align}  
@@
 \begin{align}  
- \sigma^2_k \leftarrow  \frac{ \sum_{i=1}^N  (y_i - \mu_k ) p_k(l_k|y_i)}{ \sum_{i=1}^N p_k(l_k|y_i) }.
+ \sigma^2_k \leftarrow  \frac{ \sum_{i=1}^N  (y_i - \mu_k )^T
+   p_k(l_k|y_i) (y_i - \mu_k )}{ \sum_{i=1}^N p_k(l_k|y_i) }.
 \end{align}
@@
 \begin{align}
- \boldsymbol{\Sigma}_{mn,k} \leftarrow \frac{ \sum_{i=1}^N  ( \mathbf{y}_i - \boldsymbol{\mu}_k ) p_k(l_k|\mathbf{y}_i)}{1 - \sum_{i=1}^N p^2_k(l_k|\mathbf{y}_i) }.
+ \boldsymbol{\Sigma}_{mn,k} \leftarrow \frac{ \sum_{i=1}^N  ( \mathbf{y}_i - \boldsymbol{\mu}_k )^T p_k(l_k|\mathbf{y}_i) ( \mathbf{y}_i - \boldsymbol{\mu}_k )}{1 - \sum_{i=1}^N p^2_k(l_k|\mathbf{y}_i) }.
 \end{align}
@@
 contrast, only uses sites containing label $l_k$ to update model $k$ parameters.
-A similar pattern is used in the multivariate and non-parametric cases.
+A similar pattern is used in non-parametric cases.
 \end{description}
@@
 The brainweb data is freely available.  We used single-subject
-brainweb data as is.  The 20 subject data, however, required excluding
+brainweb data as is but added a meta format data header to the raw
+binary files.   An example copy of this header is contained in {\verb atroposBwebRF40FigureExample.sh }.  The 20 subject data, however, required excluding
 non-cerebrum tissue classes.  The Hammers data was also used as is.
@@
 \paragraph{Acknowledgments}
-{This work was supported in part by NIH  {\bf {\textcolor{red}{ FIXME}}} }
+{This work was supported in part by NIH (AG17586, AG15116, NS44266, and
+NS53488).}
 

### 2010-12-28 Documentation/atropos/atropos.tex
ENH: title change
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@715 fd2da0a4-d968-48df-9cb5-d8c496287cbf
ENH:  A few minor edits.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@716 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 \begin{center}
-{\Large \bf An Open Source Framework for Multivariate $n$-Tissue
-  Segmentation with Evaluation Using Open Data Sets} \\
+{\Large \bf An Open-Source Multivariate Framework for $n$-Tissue
+  Segmentation with Evaluation on Public Data} \\
 \vspace*{0.5cm}
@@
 \end{align}
-and the $(m,n)^{th}$ element of the covariance matrix, $\boldsymbol{\Sigma}$, is calculated from
+and the $k^{th}$ covariance matrix, $\boldsymbol{\Sigma}_k$, is calculated from
 \begin{align}
- \boldsymbol{\Sigma}_{mn,k} \leftarrow \frac{ \sum_{i=1}^N  ( \mathbf{y}_i - \boldsymbol{\mu}_k )^T p_k(l_k|\mathbf{y}_i) ( \mathbf{y}_i - \boldsymbol{\mu}_k )}{1 - \sum_{i=1}^N p^2_k(l_k|\mathbf{y}_i) }.
+ \boldsymbol{\Sigma}_{k} \leftarrow \frac{ \sum_{i=1}^N  p_k(l_k|\mathbf{y}_i) ( \mathbf{y}_{i} - \boldsymbol{\mu}_{k} )^{\mathrm{T}}  ( \mathbf{y}_{i} - \boldsymbol{\mu}_{k} )}{1 - \sum_{i=1}^N p^2_k(l_k|\mathbf{y}_{i}) }.
 \end{align}
@@
 all derived from an abstract list sample function class.  Three
-likelihood classes have been developed, one parametric and two
+likelihood classes have been developed, one parametric and one
 nonparametric, and are available for usage although one of the

### 2010-12-28 Documentation/atropos/atropos.tex
ENH:  A few minor edits.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@717 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 Consider $N$ sites each with a possible $K$ labels for a total of $N^K$ possible labeling configurations.
-Due to the large $K$ associated with $K>>1$ problems, exact optimization is even more intractable than for the traditional 3-tissue scenario.
+For large $K \gg 3$, exact optimization is even more intractable than for the traditional 3-tissue scenario.
 }
@@
 \item[Input image mask:] This binary image
-  will define the spatial segmentation domain.  Regions outside the
-  mask receive the label $0$.  E.g.  {\ttfamily -x mask.nii.gz}. 
-\item[Convergence criteria:] The algorithm terminates if it reaches max-iterations or
-  produces a change less than the minimum change in the posterior.
+  defines the spatial segmentation domain.  Voxels outside the
+  masked region are designated with the label $0$.  E.g.  {\ttfamily -x mask.nii.gz}. 
+\item[Convergence criteria:] The algorithm terminates if it reaches  the maximum number of iterations or
+  produces a change less than the minimum threshold change in the posterior.
   E.g. {\ttfamily -c [5,1.e-5]}. 
@@
 \begin{itemize}
-\item  {\ttfamily -i  kmeans[3]} standard kmeans initialization for
-  three classes.
-\item  {\ttfamily -i  PriorLabelImage[3, label\_image.nii.gz , w ]}
-  where scalar $w \in [0,1]$ is unused. 
-\item  {\ttfamily -i  PriorProbabilityImages[3, label\_prob\%02d.nii.gz ,
-    w ]} where $w=0$~(use the prior probability images only for
+\item  {\ttfamily -i  Kmeans[3]} standard K-means initialization for
+  three classes,
+\item  {\ttfamily -i  PriorLabelImage[3,label\_image.nii.gz]} and
+%  where scalar $w \in [0,1]$ is unused. 
+\item  {\ttfamily -i  PriorProbabilityImages[3,label\_prob\%02d.nii.gz,w]} 
+where $w=0$~(use the prior probability images only for
   initialization) or $w=0.5$ (use the prior probability images
@@
   probability image for each model.  E.g.  {\ttfamily -o
-    [segmentation.nii.gz , seg\_prob\%02d.nii.gz] } will write
+    [segmentation.nii.gz,seg\_prob\%02d.nii.gz] } will write
   out the hard segmentation in the first output parameter and a

### 2010-12-28 Documentation/atropos/atropos.tex
ENH:  A few minor edits.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@718 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 where $w=0$~(use the prior probability images only for
-  initialization) or $w=0.5$ (use the prior probability images
-  throughout the optimization).  If one chooses $0 < w < 0.5$ then one
+  initialization) or $w>0.0$ (use the prior probability images
+  throughout the optimization).  If one chooses $0 < w < 1.0$ then one
   will increase (from zero) the weight on the priors.  These images,
-  like the PriorLabelImage, should be defined with the same domain as
+  like the {\ttfamily PriorLabelImage}, should be defined with the same domain as
   the input images to be segmented.  
@@
   or {\ttfamily -p Socrates[0] }.  Fixed label boundary conditions may be employed by
-  selecting the PriorLabelImage initialization and   {\ttfamily -p Plato[0] }.
+  selecting the {\ttfamily PriorLabelImage} initialization and   {\ttfamily -p Plato[0]}.
 \item[Output:] Atropos will output the hard segmentation and the
   probability image for each model.  E.g.  {\ttfamily -o
-    [segmentation.nii.gz,seg\_prob\%02d.nii.gz] } will write
+    [segmentation.nii.gz,seg\_prob\%02d.nii.gz]} will write
   out the hard segmentation in the first output parameter and a
-  probability image for each class named, here, seg\_prob01.nii.gz ,
+  probability image for each class named, here, seg\_prob01.nii.gz,
   seg\_prob02.nii.gz, etc. 

### 2010-12-28 Documentation/atropos/atropos.tex
ENH:  A few minor edits.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@719 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
   spatial smoothness of the label map is $\beta$.  A useful range of
-  $\beta$ values is $0$ to $0.5$ where we usually use $0.05, 0.1, 0.2$
-  in brain segmentation.  E.g.  {\ttfamily -m [ 0.1, 1x1x1 ]} would
+  $\beta$ values is $0$ to $0.5$ where we usually use 0.05, 0.1 or 0.2
+  in brain segmentation.  E.g.  {\ttfamily -m [0.1, 1x1x1]} would
   define $\beta=0.1$ with a MRF radius of one voxel in each of three dimensions.

### 2010-12-28 Documentation/atropos/atropos.tex
ENH:  A few minor edits.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@720 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 and T1 coronal images, which continues to find utility in contemporary
-developments.  A secondary component was an extended nonparametric
+developments.  A secondary component was an extended non-parametric
 probability model, also influenced by earlier work \citep{Kikinis1992}, where
@@
 opted for different likelihood models which have included either
-parametric, specifically Gaussian, or nonparametric variations.
+parametric, specifically Gaussian, or non-parametric variations.
 However, these approaches are similar in that they require a list
@@
 segmentation) to the modeling function, regardless of whether the
-function is parameter or non-parametric.  These similarities permitted
+function is parametric or non-parametric.  These similarities permitted
 the creation of a generic plug-in architecture where classes
-describing both parametric and nonparametric observational models are
+describing both parametric and non-parametric observational models are
 all derived from an abstract list sample function class.  Three
 likelihood classes have been developed, one parametric and one
-nonparametric, and are available for usage although one of the
-nonparametric classes is still in experimental development.  The
+non-parametric, and are available for usage although one of the
+non-parametric classes is still in experimental development.  The
 plug-in architecture even permits mixing likelihood models with
@@
 algorithm \citep{Tustison2010} which is based on the popular
-nonparametric nonuniformity intensity normalization (N3) introduced in \cite{Sled1998}.  It was previously mentioned that
+non-parametric nonuniformity intensity normalization (N3) introduced in \cite{Sled1998}.  It was previously mentioned that
 several methods proposed in the literature have taken an integrative

### 2010-12-28 Documentation/atropos/atropos.tex
ENH:  A few minor edits.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@721 fd2da0a4-d968-48df-9cb5-d8c496287cbf
ENH:  A few minor edits.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@722 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 call, i.e. `{\ttfamily Atropos --output}'.  N4 is
-currently being added as a standard part of the Insight ToolKit.  The
+currently being added as a standard part of the Insight Toolkit.  The
 evaluation section will illustrate inclusion of Atropos, N4 and ANTs
@@
 each tissue.  We test both K-means and prior label image
-initialization with MRF $\beta \in \{ 0.00 , 0.05 , 0.10 , 0.15 , 0.20
-, 0.25 , 0.30 \}$ at each bias field.  We also feed the white matter
+initialization with MRF $\beta \in \{ 0.00 , 0.05 , 0.10 , 0.15 , 0.20, 
+0.25 , 0.30 \}$ at each bias field.  We also feed the white matter
 probability map derived from K-means into N4 to guide the bias
-correction.  Segmentation is then repeated, with the same parameters,
+correction.%
+\footnote{
+A comprehensive evaluation of N4 reported in \cite{Tustison2010} used the BrainWeb data set
+to compare performance with the original N3 algorithm \cite{Sled1998}.
+}
+Segmentation is then repeated, with the same parameters,
 but with the N4 corrected image as input.  The resulting algorithm is

### 2010-12-28 Documentation/atropos/atropos.tex
ENH:  A few minor edits.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@723 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 subject.  The following parameters are the most relevant to this discussion:
-{\ttfamily  -i  PriorProbabilityImages[3, label\_prob\%02d.nii.gz , 0.5
-  ] -m [0.1,1x1x1] -c [5,0] -p Socrates[1] }.  Results, in terms of
+{\ttfamily  -i  PriorProbabilityImages[3, label\_prob\%02d.nii.gz,0.5] 
+-m [0.1,1x1x1] -c [5,0] -p Socrates[1] }.  Results, in terms of
 Dice overlap, are shown in figure~\ref{fig:hammers}.

### 2010-12-28 Documentation/atropos/atropos.tex
ENH:  A few minor edits.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@724 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 the recently developed N4 bias correction software \citep{Tustison2010} in an adaptive manner.  Coupled with the
-registration \citep{Avants2010b} and template building 
+registration \citep{Avants2011} and template building 
 \citep{Avants2010} already included in the ANTs toolkit, Atropos is a
@@
 \subsection{Performance on Hammers Data}
-Our prior work, \citep{Avants2010b}, showed that the majority vote
+Our prior work, \citep{Avants2011}, showed that the majority vote
 initialization provided to Atropos by ANTs template mapping is

### 2010-12-28 Documentation/atropos/atropos.tex
ENH:  A few minor edits.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@727 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 \end{tabular}
-\caption{\baselineskip 12pt \small Brainweb single-subject results for
+\caption{\baselineskip 12pt \small BrainWeb single-subject results for
   each tissue.  The results show that N4 bias correction, combined with Atropos, results in a minimal effect of
@@
 \end{tabular}
-\caption{\baselineskip 12pt \small Brainweb 20-subject results for
+\caption{\baselineskip 12pt \small BrainWeb 20-subject results for
   each tissue as a function of MRF-$\beta$ parameter where MRF-$\beta$
@@
 subject.  The following parameters are the most relevant to this discussion:
-{\ttfamily  -i  PriorProbabilityImages[3, label\_prob\%02d.nii.gz,0.5] 
+{\ttfamily  -i  PriorProbabilityImages[3,label\_prob\%02d.nii.gz,0.5] 
 -m [0.1,1x1x1] -c [5,0] -p Socrates[1] }.  Results, in terms of
@@
 
-\subsection{Performance on Brainweb Data}
+\subsection{Performance on BrainWeb Data}
 Atropos results are competitive with the state of the art.  For
@@
 noise brainweb single subject data finding 0.932 (GM) and .961 (WM)
-Dice. Results on 40\% RF were .934 (GM) and .961 (WM). SPM5 exhibits
-insensitivity to bias similar to our own best results on the 40\%RF, 3
-\% noise case (MRF-$\beta$=0.2, Kmeans $+$ N4) with Dice for GM is
+Dice. Results on 40\% RF were 0.934 (GM) and 0.961 (WM). SPM5 exhibits
+insensitivity to bias similar to our own best results on the 40\%RF, 3\% 
+noise case (MRF-$\beta$=0.2, K-means $+$ N4) with Dice for GM is
 0.950737 and for WM is 0.962586.  \cite{Nakamura2009} gave GM Dice
 results (brainweb single 3\% noise) of 0.962 (0\%RF), 0.964(20\%RF)
-and .956(40\%RF) which are slightly higher than either SPM5 or Atropos
+and 0.956(40\%RF) which are slightly higher than either SPM5 or Atropos
 results.  However, \cite{Nakamura2009} do not report WM or CSF numbers
@@
 paper's graph, showed that the average Dice accuracy for 3\% noise various RF was
-.95-.96 for WM, .93-.95 for GM and .92-.94 for CSF.
+0.95-0.96 for WM, 0.93-0.95 for GM and 0.92-0.94 for CSF.
 Perhaps the most recent balanced evaluation was performed in
@@
 voxels for brainweb are known, we can convert the confusion matrix to
-Dice.  In that case, the SPM5 Dice for brainweb single GM and WM is .8847
-and .9085, respectively, while FreeSurfer and FSL's accuracy is
+Dice.  In that case, the SPM5 Dice for brainweb single GM and WM is 0.8847
+and 0.9085, respectively, while FreeSurfer and FSL's accuracy is
 lower. The best GM result for 20 subjects is obtained by SPM5: 0.9295;

### 2010-12-28 Documentation/atropos/atropos.tex
ENH:  A few minor edits.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@728 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 evaluation section will illustrate inclusion of Atropos, N4 and ANTs
-in a brain processing pipeline.  An example of this procedure, using brainweb data with
+in a brain processing pipeline.  An example of this procedure, using BrainWeb data with
 40\% RF bias, is in figure~\ref{fig:bwebrf40}.  We supply the
@@
 \caption{\baselineskip 12pt \small We combine N4 and Atropos by simple
-sequential processing and apply to brainweb single-subject data with
+sequential processing and apply to BrainWeb single-subject data with
 40\% RF bias and 3\% noise.  The $\beta$ for the MRF term is, here,
@@
 \label{sec:bweb} The BrainWeb data is freely available at
-\url{http://mouldy.bic.mni.mcgill.ca/brainweb/}.  We employ both the
+\url{http://mouldy.bic.mni.mcgill.ca/BrainWeb/}.  We employ both the
 individual subject data and the BrainWeb20 data in this evaluation.
@@
 \end{figure}
-\subsubsection{20 Subject Evaluation} The single-subject brainweb
+\subsubsection{20 Subject Evaluation} The single-subject BrainWeb
 study in the previous section tested the basic Atropos options and the
 benefit of N4 for segmentation in the presence of bias.  The 20
-subject brainweb data allows us to use 2-fold cross-validation to test
+subject BrainWeb data allows us to use 2-fold cross-validation to test
 our ability to segment different individuals reliably.  In this study,
@@
 \subsection{Reproducibility of this Evaluation}
-The brainweb data is freely available.  We used single-subject
-brainweb data as is but added a meta format data header to the raw
+The BrainWeb data is freely available.  We used single-subject
+BrainWeb data as is but added a meta format data header to the raw
 binary files.   An example copy of this header is contained in {\verb atroposBwebRF40FigureExample.sh }.  The 20 subject data, however, required excluding
 non-cerebrum tissue classes.  The Hammers data was also used as is.
-The scripts for the single-subject brainweb study is available in
-supplementary material.  The initialization procedure for the brainweb 20 and
+The scripts for the single-subject BrainWeb study is available in
+supplementary material.  The initialization procedure for the BrainWeb 20 and
 the Hammers evaluation data is based on freely available scripts
@@
 also showed evidence that the openly available N4 bias correction can easily used with Atropos to improve segmentation.
-Furthermore, we used multiple subject brainweb data to build dataset
+Furthermore, we used multiple subject BrainWeb data to build dataset
 specific priors that provided the most consistent segmentation
@@
 instance, \cite{Ashburner2005} (SPM5) evaluated on 0\% RF (bias field) 3\%
-noise brainweb single subject data finding 0.932 (GM) and .961 (WM)
+noise BrainWeb single subject data finding 0.932 (GM) and .961 (WM)
 Dice. Results on 40\% RF were 0.934 (GM) and 0.961 (WM). SPM5 exhibits
-insensitivity to bias similar to our own best results on the 40\%RF, 3\% 
+insensitivity to bias similar to our own best results on the 40\% RF, 3\% 
 noise case (MRF-$\beta$=0.2, K-means $+$ N4) with Dice for GM is
 0.950737 and for WM is 0.962586.  \cite{Nakamura2009} gave GM Dice
-results (brainweb single 3\% noise) of 0.962 (0\%RF), 0.964(20\%RF)
-and 0.956(40\%RF) which are slightly higher than either SPM5 or Atropos
+results (BrainWeb single 3\% noise) of 0.962 (0\%cRF), 0.964 (20\% RF)
+and 0.956 (40\%RF) which are slightly higher than either SPM5 or Atropos
 results.  However, \cite{Nakamura2009} do not report WM or CSF numbers
 for comparison.  Topology preserving methods also perform well.
-\cite{Shiee2010} achieved Dice results for 3\% noise 20\% RF brainweb
+\cite{Shiee2010} achieved Dice results for 3\% noise 20\% RF BrainWeb
 single subject with 0.927, 0.912 and 0.900 for WM, GM and CSF,
@@
 paper's graph, showed that the average Dice accuracy for 3\% noise various RF was
-0.95-0.96 for WM, 0.93-0.95 for GM and 0.92-0.94 for CSF.
+0.95--0.96 for WM, 0.93--0.95 for GM and 0.92--0.94 for CSF.
 Perhaps the most recent balanced evaluation was performed in
@@
 than Dice overlap.  Because the absolute true number of GM and WM
-voxels for brainweb are known, we can convert the confusion matrix to
-Dice.  In that case, the SPM5 Dice for brainweb single GM and WM is 0.8847
+voxels for BrainWeb are known, we can convert the confusion matrix to
+Dice.  In that case, the SPM5 Dice for BrainWeb single GM and WM is 0.8847
 and 0.9085, respectively, while FreeSurfer and FSL's accuracy is
@@
 \subsection{Clinically-Related Evaluation}
-While specifying performance on brainweb is highly
+While specifying performance on BrainWeb is highly
 valuable, clinical validation is a second important aspect of
@@
 application area is that of ventilation-based segmentation of hyperpolarized helium-3 MRI
-\citep{Tustison2010a} which also used the open-source  Glamorous Glue
-algorithm to impose topology constraints.  Thus, future work may
+\citep{Tustison2010a} which also used the open source Glamorous Glue
+algorithm to impose topology constraints \cite{Tustison2010b}.  Thus, future work may
 incorporate topology more closely into the Atropos methodology.

### 2010-12-28 Documentation/atropos/atropos.tex
ENH:  A few minor edits.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@730 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 To highlight the
-value of this open-source contribution, we performed a search of
+value of this open source contribution, we performed a search of
 software attributes on NITRC and found that as of November 2010 no
@@
 alternative solution would be to use boundary conditions for these
-structures, as in the PriorLabelImage Atropos option.
+structures, as in the {\ttfamily PriorLabelImage} Atropos initialization option.
 

### 2010-12-29 Documentation/atropos/atropos.tex
ENH: multivariate atropos example
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@731 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 
-% \doublespacing
+\doublespacing
 
@@
 \begin{center}
-{\Large \bf An Open-Source Multivariate Framework for $n$-Tissue
+{\Large \bf An Open Source Multivariate Framework for $n$-Tissue
   Segmentation with Evaluation on Public Data} \\
@@
 
+\doublespacing
 \subsection{Initializing the Atropos Objective}
@@
 
+\doublespacing
 \subsection{Integration with Registration and N4 Bias Correction} 
@@
 \end{command}
-\subsubsection{Integration of N4 Bias Correction}
+\doublespacing
+\subsubsection{Integrating N4 Bias Correction}
 The typical segmentation processing pipeline begins with an intensity
@@
 evaluation section will illustrate inclusion of Atropos, N4 and ANTs
-in a brain processing pipeline.  An example of this procedure, using BrainWeb data with
-40\% RF bias, is in figure~\ref{fig:bwebrf40}.  We supply the
-information necessary to repeat this example in the script entitled
-`{\ttfamily  atroposBwebRF40FigureExample.sh}' which is available in the ANTs toolkit's Atropos
-documentation folder as of svn commit 711.
-\begin{figure}
-\begin{center}
-\begin{tabular}{c}
-\includegraphics[width=6in]{Figures/bwebN4ex.pdf}
-\end{tabular}
-\caption{\baselineskip 12pt \small We combine N4 and Atropos by simple
-sequential processing and apply to BrainWeb single-subject data with
-40\% RF bias and 3\% noise.  The $\beta$ for the MRF term is, here,
-$0.2$.  Slice 71 of the input data is in (a).  The initial K-means
-($K=3$) segmentation is in panel (b).  We use the brain mask to guide
-N4 bias correction and produce the image, shown in (c).  We repeat the
-K-means segmentation, but with the N4 corrected image as input, and
-produce the segmentation in (d).  The average 3-tissue Dice overlap of
-result (b) is 0.906 while the average overlap for (d) is 0.954.
-Arrows highlight a region of large before-after segmentation
-discrepancy.}
-\label{fig:bwebrf40}
-\end{center}
-\end{figure}
+in a brain processing pipeline.
 
@@
 conclude that simple N4 pre-processing is adequate to correct even the 40\%
-bias level.
+bias level.  An example of this procedure, using BrainWeb data with 40\% RF bias,
+is in figure~\ref{fig:bwebrf40}.  We supply the information necessary
+to repeat the results in this figure in the script entitled
+`{\ttfamily atroposBwebRF40FigureExample.sh}' which is available in
+the ANTs toolkit's Atropos documentation folder as of svn commit 711.
+The script may be easily modified to run the whole evaluation.  In
+addition, Figure~\ref{fig:bwebrf40} shows the results of simultaneously using proton density
+and T1 weighted BrainWeb data to perform the segmentation.  This
+multivariate input data outperforms the univariate T1 weighted data
+alone. 
+\begin{figure}
+\begin{center}
+\begin{tabular}{c}
+\includegraphics[width=6in]{Figures/bwebN4ex.pdf}
+\end{tabular}
+\caption{\baselineskip 12pt \small We combine N4 and Atropos by simple
+sequential processing and apply to BrainWeb T1-weighted single-subject
+data with 40\% RF bias and 3\% noise.  The $\beta$ for the MRF term
+is, here, $0.2$.  Slice 71 of the input data is in (a).  The initial
+K-means ($K=3$) segmentation is in panel (b).  We use the brain mask
+to guide N4 bias correction and produce the image in (c).  We
+repeat the K-means segmentation, but with the N4 corrected image as
+input and produce the segmentation in (d).  The average 3-tissue Dice
+overlap of result (b) is 0.906 while the average overlap for (d) is
+0.954.  Arrows highlight a region of large before-after segmentation
+discrepancy.  In (e) we see the BrainWeb proton density image with no
+inhomogeneity and 3\% noise.  Its segmentation is in (f) with average
+3-tissue Dice overlap of 0.895.  In (g) we use both proton density
+data and T1 data as multiple modality input to Atropos.  The segmentation
+of this two-modality input data, using a multivariate Gaussian model,
+produces average 3-tissue Dice overlap of 0.958, which exceeds
+the univariate solution.  An arrow highlights one region where there
+is small, visually recognizable improvement in sulcal segmentation
+relative to the result from T1 data alone.  A second area of
+improvement is the putamen segmentation.  The ground truth
+segmentation is in (h).  The multivariate segmentation result, in
+combination with the low PD segmentation performance, suggests PD and
+T1 provide complementary information that may improve 3-tissue
+segmentation and serves to validate the multivariate Atropos
+implementation.
+}
+\label{fig:bwebrf40}
+\end{center}
+\end{figure}
+
 \begin{figure}
@@
 label.  Following \cite{Avants2010a}, the initialization of our
-evaluation applies the script {\verb ants_multitemplate_labeling.sh }
+evaluation applies the script {\ttfamily ants\_multitemplate\_labeling.sh }
 (available in the ANTs toolkit) to the 19 Hammers evaluation datasets
@@
 \subsection{Reproducibility of this Evaluation}
-The BrainWeb data is freely available.  We used single-subject
-BrainWeb data as is but added a meta format data header to the raw
-binary files.   An example copy of this header is contained in {\verb atroposBwebRF40FigureExample.sh }.  The 20 subject data, however, required excluding
-non-cerebrum tissue classes.  The Hammers data was also used as is.
-The scripts for the single-subject BrainWeb study is available in
-supplementary material.  The initialization procedure for the BrainWeb 20 and
-the Hammers evaluation data is based on freely available scripts
-included with the ANTs toolkit, specifically  {\verb ants_multitemplate_labeling.sh } and  {\verb atroposBwebRF40FigureExample.sh }.
+\noindent{\bf Data:} The BrainWeb data is freely available.  We used
+single-subject BrainWeb data as is but added a meta format data header
+to the raw binary files.  An example copy of this header is contained
+in {\ttfamily atroposBwebRF40FigureExample.sh }.  The 20 subject data,
+however, required excluding non-cerebrum tissue classes.  The Hammers
+data was also used as is (\url{http://www.brain-development.org/}).
+
+\noindent{\bf Software:} The ANTs software is available at \url{http://www.picsl.upenn.edu/ANTs} with download and compilation
+instructions at \url{http://www.picsl.upenn.edu/ANTS/download.php}.
+Svn release 711 was used for the examples and evaluations performed in
+this paper.  Some components of ANTs depend on the Insight ToolKit.
+The most critical dependency, for Atropos, is the ITK statistics framework
+used to implement the univariate and multivariate parametric models.
+We linked to the git version of ITK current as of Dec. 1, 2010.  See
+\url{http://www.itk.org/Wiki/ITK/Git/Download} for instructions on git ITK.
+
+\noindent{\bf Scripts:} The complete script for the single-subject
+BrainWeb study is based on generalizing {\ttfamily atroposBwebRF40FigureExample.sh }, which is available in the ANTs
+toolkit (svn release 711 or greater) and which reproduces Figure 2
+results.  The template-based normalization procedure for the BrainWeb
+20 and the Hammers evaluation data is based on freely available
+scripts included with the ANTs toolkit, specifically {\ttfamily
+  ants\_multitemplate\_labeling.sh } and {\ttfamily
+  buildtemplateparallel.sh }.  A release version of ANTs---with a final version of Atropos---will
+be prepared with the final version of this manuscript.
 
@@
 \citep{Tustison2010a} which also used the open source Glamorous Glue
-algorithm to impose topology constraints \cite{Tustison2010b}.  Thus, future work may
+algorithm to impose topology constraints \citep{Tustison2010b}.  Thus, future work may
 incorporate topology more closely into the Atropos methodology.
@@
 \end{document}
+
+%# T1_RF40_3%_N4 reference results 
+ Label 1 DICE 0.941891  RO 0.890165 TP1 0.904786 TP2 0.98217
+ Label 3 DICE 0.966776  RO 0.935688 TP1 0.962814 TP2 0.97077
+ Label 2 DICE 0.953862  RO 0.911794 TP1 0.971043 TP2 0.937279
+ AvgDice  : 0.954176
+%# T1_RF40_3%_N4 + T2_RF0_3% noise reference results 
+Label 1 DICE 0.937508  RO 0.882367 TP1 0.889406 TP2 0.99111
+ Label 3 DICE 0.96701  RO 0.936128 TP1 0.958398 TP2 0.975778
+ Label 2 DICE 0.951639  RO 0.90774 TP1 0.977645 TP2 0.926981
+ AvgDice  : 0.952052
+%# T1_RF40_3%_N4 + PD_RF0_3% noise reference results 
+ Label 1 DICE 0.944732  RO 0.895254 TP1 0.912413 TP2 0.979425
+ Label 3 DICE 0.970348  RO 0.942405 TP1 0.969169 TP2 0.971531
+ Label 2 DICE 0.957976  RO 0.919341 TP1 0.97094 TP2 0.945353
+ AvgDice  : 0.957685
+%# PD_RF0_3% noise reference results 
+ Label 2 DICE 0.905959  RO 0.828085 TP1 0.892396 TP2 0.919941
+ Label 3 DICE 0.942477  RO 0.891212 TP1 0.943874 TP2 0.941084
+ Label 1 DICE 0.837847  RO 0.720944 TP1 0.87262 TP2 0.805739
+ AvgDice  : 0.895428
+%# T2 
+ Label 1 DICE 0.931616  RO 0.871986 TP1 0.984176 TP2 0.884385
+ Label 2 DICE 0.882997  RO 0.790505 TP1 0.79822 TP2 0.987921
+ Label 3 DICE 0.901592  RO 0.820817 TP1 0.99405 TP2 0.82487
+ AvgDice  : 0.905402

### 2010-12-29 Documentation/atropos/atroposBwebRF40FigureExample.sh

@@ -1,5 +1,7 @@
 # input image 
 IMG=T1_ICBM_1mm_3perc_40_bias/t1_icbm_normal_1mm_pn3_rf40.mhd
+PDG=PD_ICBM_1mm_3perc_0_bias/pd_icbm_normal_1mm_pn3_rf0.mhd
+IT2=T2_ICBM_1mm_3perc_0_bias/t2_icbm_normal_1mm_pn3_rf0.mhd
 if [ ! -s $IMG ] ; then 
  echo no image named $IMG exists --- exiting.   you should be able to download this image from the brainweb website and use the header defined below to make it readable by Atropos. 
  exit
@@ -17,15 +19,40 @@ NAMING=bwebRF40example
 # binary image used as a weight and mask image for N4 and segmentation mask for Atropos 
 # you can get this by thresholding the ground_truth labeling 
 WTIM=mask.nii.gz
-$EX  -d 3 -a  $IMG -i KMeans[$NC] -o [test_init.nii.gz,test_init_prob%02d.nii.gz] -m [${MRF},1x1x1] -x mask.nii.gz -p Socrates[${SOC}]  -c $CONV 
 BS="N4BiasFieldCorrection -d 3  -h 0 " ; ITS=20 ; ITS2=3
-$BS  -i $IMG   -o  n4.nii.gz -s 2 -b [200] -w mask.nii.gz -c [${ITS}x${ITS}x${ITS}x${ITS}x${ITS2},0.0001] -x mask.nii.gz 
-Atropos -d 3 -a  n4.nii.gz -i KMeans[$NC] -o [test.nii.gz,test_prob%02d.nii.gz] -m [${MRF},1x1x1] -x mask.nii.gz -p Socrates[${SOC}]  -c $CONV 
+# $BS  -i $IMG   -o  n4.nii.gz -s 2 -b [200] -w mask.nii.gz -c [${ITS}x${ITS}x${ITS}x${ITS}x${ITS2},0.0001] -x mask.nii.gz 
+Atropos  -d 3 -a n4.nii.gz -a $PDG -i KMeans[$NC] -o [test.nii.gz,test_prob%02d.nii.gz] -m [${MRF},1x1x1] -x mask.nii.gz -p Socrates[${SOC}]  -c $CONV 
 # evaluation numbers
 if [ -s ground_truth.nii.gz ] ; then 
[…]

### 2010-12-29 Documentation/atropos/atropos.tex
ENH:  Minor edit.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@732 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 %memory.  
-We evaluate EM-based classification of $1 mm^3$ T1 MR
+We evaluate EM-based classification of 1 mm$^3$ T1 MR
 neuroimages into 69 neuroanatomical regions to illustrate the practical
@@
 allows many-class problems to be solved with low memory usage ($<$2GB
-for a 69 class model on 1mm$^3$ brain data).
+for a 69 class model on 1 mm$^3$ brain data).
 
@@
 A comprehensive evaluation of N4 reported in \cite{Tustison2010} used the BrainWeb data set
-to compare performance with the original N3 algorithm \cite{Sled1998}.
+to compare performance with the original N3 algorithm \citep{Sled1998}.
 }

### 2010-12-29 Documentation/atropos/atropos.tex
ENH: submitted version of atropos paper.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@733 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 \begin{align}
-\mathrm{Pr}(x_i = l_k) = \frac{t_{ik}}{\sum_{j=1}^K t_{ij}}.
+\mathrm{Pr}(x_i = l_k) = \frac{t_{ik}}{\sum_{j=1}^K t_{ij}},
 \end{align}
@@
 reasonable and efficient alterative to Boykov's solution.
-Furthermore, one may automate the
-initialization process.
+Furthermore, one may automate the initialization process.  We compare
+our sparse prior labeling strategy to fully automated methods in the
+evaluation section.
 
@@
 \end{description}
-EM will iterate toward a local maximum.  We track
-convergence by summing up the maximum posterior
-probability at each site over the segmentation domain.  
+EM will iterate toward a local maximum.  We track convergence by
+summing up the maximum posterior probability at each site over the
+segmentation domain.  The E-step, above, depends upon the selected
+coding strategy \citep{Besag1986}.  Atropos may use either a
+classical, sequential checkerboard update or a synchronous update of
+the labels, the latter of which is commonly used in practice.
+Synchronous update does not guarantee convergence but we employ it by
+default due to its intrinsic parallelism and speed.  The user may
+alternatively select checkerboard update if he or she desires
+theoretical convergence guarantees.  However, we have not identified
+performance differences, relative to ground truth, that convince us of
+the absolute superiority of one approach over the other.
 %\begin{align}
@@
 distributed with the rest of the ANTs package, we plan to contribute
-the Atropos to the Insight Toolkit where it can be vetted and further
+its core elements to the Insight Toolkit where it can be vetted and further
 improved by other interested researchers.
 
-An overview of the core components of Atropos can be gleaned, in part, from the flowchart depicted in 
+An overview of Atropos components can be gleaned, in part, from the flowchart depicted in 
 Fig.~\ref{fig:flowchart}.
@@
 The typical segmentation processing pipeline begins with an intensity
-normalization/bias correction step using the recently developed N4
+normalization/bias correction step using a method such as the recently developed N4
 algorithm \citep{Tustison2010} which is based on the popular
@@
 \subsubsection{Single-Subject Evaluation} We use the single-subject
-data with 3\% noise and bias 0, 20 and 40 \% RF inhomogeneity.  We
+data with 3\% noise and three levels of bias referred to as 0, 20 and 40 \% RF inhomogeneity.  We
 study the effect of the MRF prior term and initialization on
@@
 Thus, with this simple evaluation, we are able to compare the impact
-of bias on our N4 and Atropos and also the validity of our
+of bias on $($N4 $+$ Atropos$)$ and also the validity of our
 prior label image initialization.  The prior label image manual
@@
 set of voxels for each tissue.  Results, in terms of
-Dice overlap, are shown in figure~\ref{fig:bweb1}  Because overlap
+Dice overlap, are shown in Figure~\ref{fig:bweb1}. Because overlap
 ratios with N4 bias correction approximate those of the zero bias data, we may
@@
 bias level.  An example of this procedure, using BrainWeb data with 40\% RF bias,
-is in figure~\ref{fig:bwebrf40}.  We supply the information necessary
+is in Figure~\ref{fig:bwebrf40}.  We supply the information necessary
 to repeat the results in this figure in the script entitled
@@
   appears to be between $0.1$ and $0.2$.  The prior label image
-  initialization is competitive with but inferior to K-means.  The
+  initialization is competitive with but inferior to K-means.  This is
+  sensible in that the prior labels image outcome is sensitive to the
+  user choices made.  In this study, we labeled the brain web template
+  very sparsely, quickly and crudely and did not expect highly accurate results.  The
   legend is in the same position in each graph, allowing a visual
@@
 the model parameters but not during subsequent EM iterations.  
-When $w=1$, the priors are only maintained in the product with the
+When $w=0.5$, the priors are maintained in the product with the
 likelihood during all EM iterations.  Results, in terms of bar
-plots for Dice overlap mean and standard deviation, are shown in figure~\ref{fig:bweb20}.
+plots for Dice overlap mean and standard deviation, are shown in Figure~\ref{fig:bweb20}.
 \begin{figure}
@@
 -m [0.1,1x1x1] -c [5,0] -p Socrates[1] }.  Results, in terms of
-Dice overlap, are shown in figure~\ref{fig:hammers}.
+Dice overlap, are shown in Figure~\ref{fig:hammers}.
 \begin{figure}
@@
 details and documented its performance in a variety of use cases.  We
-also showed evidence that the openly available N4 bias correction can easily used with Atropos to improve segmentation.
+also showed evidence that the openly available N4 bias correction can
+easily be used with Atropos to improve segmentation.
 Furthermore, we used multiple subject BrainWeb data to build dataset
@@
 0.950737 and for WM is 0.962586.  \cite{Nakamura2009} gave GM Dice
-results (BrainWeb single 3\% noise) of 0.962 (0\%cRF), 0.964 (20\% RF)
+results (BrainWeb single 3\% noise) of 0.962 (0\% RF), 0.964 (20\% RF)
 and 0.956 (40\%RF) which are slightly higher than either SPM5 or Atropos
@@
 constraint applied to the segmentation.  
-\cite{Bazin2007c} proposed TOADS and, estimating from the
-paper's graph, showed that the average Dice accuracy for 3\% noise various RF was
-0.95--0.96 for WM, 0.93--0.95 for GM and 0.92--0.94 for CSF.
+\cite{Bazin2007c} proposed TOADS and, estimating from the paper's
+graph, showed that the average Dice accuracy for 3\% noise various RF
+was 0.95--0.96 for WM, 0.93--0.95 for GM and 0.92--0.94 for CSF.
 Perhaps the most recent balanced evaluation was performed in
@@
 voxels for BrainWeb are known, we can convert the confusion matrix to
-Dice.  In that case, the SPM5 Dice for BrainWeb single GM and WM is 0.8847
-and 0.9085, respectively, while FreeSurfer and FSL's accuracy is
-lower. The best GM result for 20 subjects is obtained by SPM5: 0.9295;
-best WM Dice is from FSL: 0.9499.  We note that \cite{Klauschen2009}
-used a comprehensive evaluation where quality of brain extraction also
-contributed to the outcome.  Thus, the results must be interpreted
-slightly differently than those from other papers.
+Dice.  In that case, the SPM5 Dice for BrainWeb single GM and WM is
+0.8847 and 0.9085, respectively, while FreeSurfer and FSL's accuracy
+is lower. The best GM result for 20 subjects is obtained by SPM5:
+0.9295; best WM Dice is from FSL: 0.9499.  We note that
+\cite{Klauschen2009} used a comprehensive evaluation where quality of
+brain extraction also contributed to the outcome.  Thus, the results
+must be interpreted slightly differently than those from other papers.
+Finally, we note that in our evaluation of 20 subject BrainWeb data, the prior
+probability models performed best of all the models used.  Compared to
+the KMeans based segmentation, the prior based segmentation
+performance also peaked at lower values of the MRF-$\beta$ term (0.0
+and 0.05).  This is reasonable in that the spatial priors themselves
+impose a degree of regularity to the segmentation, as in SPM5.
 
@@
 initialization provided to Atropos by ANTs template mapping is
-competitive with \cite{Heckemann2006}.  The Atropos EM extension of
-improved these results further.  However, the Atropos EM segmentation
+competitive with \cite{Heckemann2006}.  Overall, the Atropos EM
+extension improved these results further.  However, in a few regions
+of the mid-brain, the Atropos EM segmentation
 performed significantly worse.  This is not surprising, in that

### 2010-12-29 Documentation/atropos/atroposBwebRF40FigureExample.sh
ENH: providing reference results for brain web evaluation and implementation details in the script
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@734 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -20,8 +20,8 @@ NAMING=bwebRF40example
 # you can get this by thresholding the ground_truth labeling 
 WTIM=mask.nii.gz
 BS="N4BiasFieldCorrection -d 3  -h 0 " ; ITS=20 ; ITS2=3
-# $BS  -i $IMG   -o  n4.nii.gz -s 2 -b [200] -w mask.nii.gz -c [${ITS}x${ITS}x${ITS}x${ITS}x${ITS2},0.0001] -x mask.nii.gz 
-Atropos  -d 3 -a n4.nii.gz -a $PDG -i KMeans[$NC] -o [test.nii.gz,test_prob%02d.nii.gz] -m [${MRF},1x1x1] -x mask.nii.gz -p Socrates[${SOC}]  -c $CONV 
+$BS  -i $IMG   -o  n4.nii.gz -s 2 -b [200] -w mask.nii.gz -c [${ITS}x${ITS}x${ITS}x${ITS}x${ITS2},0.0001] -x mask.nii.gz 
+Atropos  -d 3 -a n4.nii.gz  -i KMeans[$NC] -o [test.nii.gz,test_prob%02d.nii.gz] -m [${MRF},1x1x1] -x mask.nii.gz -p Socrates[${SOC}]  -c $CONV # -k HistogramParzenWindows[1,32]
 # evaluation numbers
 if [ -s ground_truth.nii.gz ] ; then 
 ImageMath 3 dice_out.txt DiceAndMinDistSum test.nii.gz ground_truth.nii.gz 

### 2011-01-04 Documentation/atropos/atropos.tex
ENH: this is the actual version submitted first round to neuroinformatics
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@735 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
         caption={\small \baselineskip 12pt Atropos short command line menu which is invoked using the `{\ttfamily -h}' option. 
-        The expanded menu, which provides additional details regarding the possible parameters and usage 
+        The expanded menu, which provides details regarding the possible parameters and usage 
         options, is elicited using the `{\ttfamily {-}{-}help}' option.},
@@
         caption={\small \baselineskip 12pt N4 short command line menu which is invoked using the `{\ttfamily -h}' option. 
-        The expanded menu, which provides additional details regarding the possible parameters and usage 
+        The expanded menu, which provides details regarding the possible parameters and usage 
         options, is elicited using the `{\ttfamily {-}{-}help}' option.  },
@@
 The problem of accurately delineating the white matter, grey matter
-and cerebrospinal fluid (and further subdivisions) of the human brain continuously spurs
+and cerebrospinal fluid (and subdivisions) of the human brain continuously spurs
 technical development in segmentation.  Following \cite{Vannier1985},
@@
 every point of the cortex is classified, i.e. a cortical parcellation
-\citep{Fischl2004,Heckemann2006,Destrieux2010}.  Given the extreme
-time-consumption and tedium associated with manual effort towards the
-latter, various techniques have been proposed of which a small subset
-has been publicly availed to the research community such as the
-popular software package known as Freesurfer
+\citep{Fischl2004,Heckemann2006,Destrieux2010}.  Various techniques
+have been proposed to reduce the manual effort required to densely
+label a high-resolution neuroimage; one example is the popular
+software package known as Freesurfer
 \citep{Dale1999,Fischl1999,Fischl2004}.  In contrast to the volumetric
@@
 analysis workflow.  These types of methods, including Klein's
-Mindboggle \citep{Klein2005}, would usually follow an initial segmentation by a 
-volumetric method such as Atropos.
+Mindboggle \citep{Klein2005}, would usually follow an initial
+segmentation by a volumetric method such as Atropos.
 
@@
 
-We also allow Atropos to address the bias field issue by incorporating
-the recently developed N4 bias correction software \citep{Tustison2010} in an adaptive manner.  Coupled with the
-registration \citep{Avants2011} and template building 
+Coupled with the registration \citep{Avants2011} and template building
 \citep{Avants2010} already included in the ANTs toolkit, Atropos is a
 versatile and powerful software tool which touches multiple aspects of
-our brain processing pipeline from brain extraction
-\citep{Avants2010a} to grey matter/white matter/cerebrospinal fluid
-segmentation to label fusion and propagation for achieving a refined
-cortical parcellation based on both prior information and
-subject-specific image data.
-
-%Atropos is multi-platform, free, /stand-alone and able to
-%manage a wide range of scalar and vector image datatypes.
-
-To highlight the
-value of this open source contribution, we performed a search of
-software attributes on NITRC and found that as of November 2010 no
-stand-alone EM methods are currently listed.
-%Atropos's memory efficiency enables EM-segmentation of many classes at
-%once, a capability that is due to a sparse internal representation of
-%prior probability images that minimizes the need for random access
-%memory.  
-We evaluate EM-based classification of 1 mm$^3$ T1 MR
-neuroimages into 69 neuroanatomical regions to illustrate the practical
-value of the low-memory implementation within this paper.  
-Although Atropos may be applied to
-multivariate data from arbitrary modalities, we limit our evaluation
-to tissue-classification in T1 neuroimaging in part due to the
-abundance of ``gold-standard'' data for this modality.  Consistent
+our brain processing pipeline.  We use Atropos to address brain
+extraction \citep{Avants2010a}, grey matter/white matter/cerebrospinal
+fluid segmentation, label fusion/propagation and cortical
+parcellation.  We also allow Atropos to interact with the recently
+developed N4 bias correction software \citep{Tustison2010} in an
+adaptive manner.  To further highlight the value of this open source
+contribution, we performed a search of software attributes on NITRC
+and found that as of November 2010 no stand-alone EM methods are
+currently listed.  We also evaluate Atropos performance on two brain
+MRI segmentation objectives.  First, three tissue classification.
+Second, we test our ability to parcellate the brain into 69
+neuroanatomical regions to illustrate the practical value of the
+low-memory implementation within this paper.  Although Atropos may be
+applied to multivariate data from arbitrary modalities, we limit our
+evaluation to tissue-classification in T1 neuroimaging in part due to
+the abundance of ``gold-standard'' data for this modality.  Consistent
 with our advocacy of open science (not to mention the facilitation of
@@
 behind the various components of Atropos while
-acknowledging that additional theoretical discussion is available elsewhere.  
+acknowledging that more theoretical discussion is available elsewhere.  
 This is followed by a thorough
@@
 
-Atropos encodes a family of Bayesian segmentation techniques that may be configured in an application-specific manner.
-The theory underlying Atropos dates back 20$+$ years and is representative of some of the most innovative work in the field.
-Although we synthesize some of the theoretical work in this section, we recommend to the interested reader consultation with the literature cited previously for further insight.
+Atropos encodes a family of Bayesian segmentation techniques that may
+be configured in an application-specific manner.  The theory
+underlying Atropos dates back 20$+$ years and is representative of
+some of the most innovative work in the field.  Although we synthesize
+some of the theoretical work in this section, we recommend to the
+interested reader consultation with the literature cited previously
+for additional perspective.
 
@@
 \end{align}
-so that sites in the neighborhood closer to $i$ are weighted more heavily than sites further away.
+so that sites in the neighborhood closer to $i$ are weighted more
+heavily than distant sites.
 
@@
 classes.  Thus, in most practical applications, EM is a
-reasonable and efficient alterative to Boykov's solution.
+reasonable and efficient alternative to Boykov's solution.
 Furthermore, one may automate the initialization process.  We compare
@@
 as a constant.  The Atropos implementation estimates $\gamma_k$ at
-each iteration as in \cite{Ashburner2005}.%
+each iteration, similar to \cite{Ashburner2005}.%
 \footnote{
@@
 the form of either a single $n$-ary prior label image or a series of
-prior probability images, one for each class.
+prior probability images, one for each class.  The initialization also
+provides starter parameters.
 \item[ICM Label Update (E-Step):] Given the initialization and fixed
@@
 \begin{align}
-\gamma_k \leftarrow \frac{1}{N} \sum_{i=1}^N  p_k(l_k|y_i)
+\gamma_k \leftarrow \frac{1}{N} \sum_{i=1}^N  p_k(l_k|y_i).
 \end{align}
@@
 \section{Implementation} 
-
+Organization of the implementation section roughly follows that of the
+theory section.  
+%Thus, the reader can refer to, for instance, 2.2 for
+%the likelihood theory and 3.3 for likelihood implementation.
 \begin{figure}
@@
 distributed with the rest of the ANTs package, we plan to contribute
-its core elements to the Insight Toolkit where it can be vetted and further
-improved by other interested researchers.
+its core elements to the Insight Toolkit where it can be vetted and improved by other interested researchers.
 
@@
 dimension, size, origin, etc.), will automatically enable multivariate
-likelihoods.  In that case, the first image specified on the 
-command line is used to initialize the Random, Otsu, or K-means labeling with the latter initialization further
-refined by incorporating the additional intensity images, i.e. an initial univariate K-means clustering is determined 
-from the first intensity image which, along with the other images, provides the starting multivariate cluster centers for a 
-follow-up multivariate K-means labeling.    More details on each of
-the key implementation options are given below.
+likelihoods.  In that case, the first image specified on the command
+line is used to initialize the Random, Otsu, or K-means labeling with
+the latter initialization refined by incorporating the additional intensity images, i.e. an initial univariate K-means clustering is
+determined from the first intensity image which, along with the other
+images, provides the starting multivariate cluster centers for a
+follow-up multivariate K-means labeling.  More details on each of the
+key implementation options are given below.
 
@@
 \subsection{Prior Probability Models}
+\subsubsection{Label Regularity}
 Consistent with our previous discussion, we offer both an MRF-based
@@
 }\end{comment}
-
 \doublespacing
-\subsection{Integration with Registration and N4 Bias Correction} 
-Much of the image processing in our lab takes advantage of multiple
-open source tools.  Most critically, these involve registration and
-bias correction, in addition to segmentation. 
-\subsubsection{Registration of Probability Maps} 
+\subsubsection{Registration and Probability Maps} 
 Image registration enables the ability to transform information
@@
 \subsubsection{Integrating N4 Bias Correction}
-The typical segmentation processing pipeline begins with an intensity
+Assumptions about bias correction may be thought of as another prior
+model.  As such, the typical segmentation processing pipeline begins with an intensity
 normalization/bias correction step using a method such as the recently developed N4
-algorithm \citep{Tustison2010} which is based on the popular
+algorithm \citep{Tustison2010}.  N4 builds upon the popular
 non-parametric nonuniformity intensity normalization (N3) introduced in \cite{Sled1998}.  It was previously mentioned that
@@
 in a brain processing pipeline.
+
+\subsection{Running the Atropos Optimization} 
+The Atropos algorithm is cross-platform and compiles on, at minimum,
+modern osx, windows and linux-based operating systems.  The
+user-interface may be reached through the operating system's user
+terminal.  Because of its portability and low-level efficiency,
+Atropos can easily be called from within other packages, such as
+Matlab or Slicer, or, alternatively, integrated at compile time as a
+library.  A typical call to the algorithm, illustrated here with ANTs
+example data, is:
+{\ttfamily Atropos -d 2 -a r16slice.nii.gz -i kmeans[3] -c [5,0] -x
+mask.nii.gz -m [0.2,1x1] -o
+[r16\_seg.nii.gz,r16\_prob\_\%02d.nii.gz]}.  In this case, Atropos
+will output the segmentation image, the per-class probability images
+and a listing of the parameters used to set up the algorithm.  A
+useful feature is that one may re-initialize the Atropos EM via the
+{\ttfamily -i PriorProbabilityImages[...]} option.  This feature
+allows one to compute an initial segmentation via K-means, alter the
+output probabilities by externally computed
+functions (e.g. Gaussian smoothing, image similarity or edge maps) and re-estimate the
+segmentation with the modified priors.  Finally, the functionality
+that is available to parametric models is equally available to the
+non-parametric models enabled by Atropos.
 
@@
 Thus, with this simple evaluation, we are able to compare the impact
-of bias on $($N4 $+$ Atropos$)$ and also the validity of our
+of bias on the combination of N4 and Atropos and also the validity of our
 prior label image initialization.  The prior label image manual
 initialization was based on a crude ITK-SNAP labeling of a small
-set of voxels for each tissue.  Results, in terms of
+set of voxels for each tissue.  Results of these evaluation scenarios, in terms of
 Dice overlap, are shown in Figure~\ref{fig:bweb1}. Because overlap
@@
 the ANTs toolkit's Atropos documentation folder as of svn commit 711.
-The script may be easily modified to run the whole evaluation.  In
-addition, Figure~\ref{fig:bwebrf40} shows the results of simultaneously using proton density
-and T1 weighted BrainWeb data to perform the segmentation.  This
-multivariate input data outperforms the univariate T1 weighted data
-alone. 
+The script may be easily modified to run the whole
+evaluation.  Figure~\ref{fig:bwebrf40} shows the results of
+simultaneously using proton density and T1 weighted BrainWeb data to
+perform the segmentation.  This multivariate input data outperforms
+the univariate T1 weighted data alone.
 \begin{figure}
@@
 segmentation and serves to validate the multivariate Atropos
-implementation.
+implementation.  In this case, the benefit is likely to derive from
+the fact that the PD image has no bias.
 }
@@
 single subject with 0.927, 0.912 and 0.900 for WM, GM and CSF,
-respectively.  These are excellent numbers given the additional
-constraint applied to the segmentation.  
+respectively.  These are excellent numbers given the topological constraint applied to the segmentation.  
 \cite{Bazin2007c} proposed TOADS and, estimating from the paper's
@@
 future work.  One early example of a clinically-focused Atropos
-neuroimaging application is in \citep{Avants2010c}.  An additional successful
+neuroimaging application is in \citep{Avants2010c}.  A second successful
 application area is that of ventilation-based segmentation of hyperpolarized helium-3 MRI

### 2011-02-06 Documentation/atropos/atropos.tex
ENH: fixed 'bug' in equation (4)
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@738 fd2da0a4-d968-48df-9cb5-d8c496287cbf
ENH:  Added ICM section (in red) and updated Atropos command line listing.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@762 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 p(\mathbf{y}|\mathbf{x})}_{\text{Likelihood(s)}} \underbrace{
-p(\mathbf{x})}_{\text{Prior(s)}}\frac{1}{\mathbf{y}}, 
+p(\mathbf{x})}_{\text{Prior(s)}}\frac{1}{p(\mathbf{y})}, 
 \end{equation} 
@@
 provides starter parameters.
-\item[ICM Label Update (E-Step):] Given the initialization and fixed
-model parameters, Atropos will evaluate the posterior probability for
-each possible label and assign, to each voxel, the label that
-maximizes the posterior.  
+\item[Label Update (E-Step):] Given the initialization and fixed
+model parameters, {\color{red}{Atropos is capable of updating the current label 
+estimates using either a {\em synchronous} or {\em asynchronous} scheme.
+The former is characterized by iterating through the image and determining
+which label maximizes the posterior probability without updating any
+labels until all voxels in the mask have been visited at which point
+all the voxel labels are updated simultaneously (hence the descriptor 
+``synchronous'').  This option is specified with \texttt{--icm [0]}. 
+However, unlike asynchronous schemes characteristic of ICM, synchronous 
+updates lack convergence guarantees.  To determine the labeling 
+which maximizes the posterior probability for the asynchronous option, 
+an ``ICM code'' image is created once for all iterations by iterating 
+through the image and assigning an ICM code label to each voxel in the 
+mask such that each MRF neighborhood has a non-repeating code label set.  
+Thus each masked voxel in the ICM code image is assigned a value in the 
+range $\{1,\ldots,C\}$ where $C$ is the maximum code label.
+Such an image can be created and viewed with Atropos by assigning a valid
+filename in the \texttt{--icm [1]} set of options.  An example is given in 
+Figure \ref{fig:r16icm}.  For each iteration, a random ordering of the
+set $\{1,\ldots,C\}$ is created which prescribes the order in which the
+voxel labels are updated asynchronously. After the first pass through the
+set of code labels, additional passes can further increase the posterior
+probability until convergence (in $\sim$5 iterations).  One can specify
+a maximum number of these ``ICM iterations'' on the command line.}}
+
+\begin{figure}
+\begin{center}
+\begin{tabular}{cc}
+\includegraphics[width=70mm]{r16slice.png} &
+\includegraphics[width=70mm]{r16icm.png} \\
+(a) & (b) \\
+\end{tabular}
+\end{center}
+\caption{\baselineskip 12pt \small  (a) An adult brain image slice and (b) the corresponding ICM code image created using a $5\times5$ MRF neighborhood with a maximum ICM code label of `13'.  }
+\label{fig:r16icm}
+\end{figure}
+
 % {\bf \textcolor{red}{ FIXME more specific?}}.
@@
 \end{align}
-
 This type of update is known as {\em soft} EM.  Hard EM, in
@@
 \end{description}
+
 EM will iterate toward a local maximum.  We track convergence by
@@
                           PriorLabelImage[numberOfClasses,labelImage,priorWeighting]
+     -p, --posterior-formulation Socrates[<useMixtureModelProportions=1>,
+                                   <initialAnnealingTemperature=1>,<annealingRate=1>,
+                                   <minimumTemperature=0.1>]
+                                 Plato[<useMixtureModelProportions=1>,
+                                   <initialAnnealingTemperature=1>,<annealingRate=1>,
+                                   <minimumTemperature=0.1>]
      -x, --mask-image maskImageFilename
@@
      -m, --mrf [<smoothingFactor=0.3>,<radius=1x1x...>]
+     -g, --icm [<useAsynchronousUpdate=1>,<maximumNumberOfICMIterations=1>,
+                <icmCodeImage=`'>]     
      -o, --output [classifiedImage,<posteriorProbabilityImageFileNameFormat>]
@@
 corresponding to the white matter produced as output in the Atropos
-call, i.e. `{\ttfamily Atropos --output}'.  N4 is
-currently being added as a standard part of the Insight Toolkit.  The
+call, i.e. `{\ttfamily Atropos --output}'.  N4 was recently added  
+to the Insight Toolkit repository%
+\footnote{
+http://www.itk.org/Doxygen/html/classitk\_1\_1N4MRIBiasFieldCorrectionImageFilter.html
+} 
+where it is built and tested on
+multiple platforms nightly.  The
 evaluation section will illustrate inclusion of Atropos, N4 and ANTs

### 2011-02-07 Documentation/atropos/ResponseToReviewers.tex
ENH: Getting ready for resubmission to Neuroinformatics
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@763 fd2da0a4-d968-48df-9cb5-d8c496287cbf

\documentclass[11pt]{article}
\newcommand{\X}{{\bf X}}
\newcommand{\x}{{\bf x}}
\newcommand{\y}{{\bf y}}
\renewcommand{\v}{{\bf v}}
\renewcommand{\textfraction}{0.05}
\renewcommand{\topfraction}{0.95}
\renewcommand{\bottomfraction}{0.95}

\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\usepackage{verbatim}
\usepackage[usenames]{color}
\usepackage{listings}
\lstset{breaklines=true,basicstyle=\ttfamily,breakindent=0pt}
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\begin{document}
%\maketitle
\noindent
David Kennedy, Ph.D.,\\
Co-Editor-in-Chief, Neuroinformatics\\

\noindent
Dear Dr. Kennedy and Reviewers,\\

\noindent
My coauthors and I would like to thank the editors and reviewers for
handling our manuscript entitled ``An Open Source Multivariate Framework for $n$-Tissue
  Segmentation with Evaluation on Public Data.''  The reviews were extremely helpful in highlighting both
the strengths and weaknesses of the previous version of the paper.  In
summary, the review found the work to be worthy of publication, with
revisions that focused on improving clarity of some key points. We make every effort, in this response letter and the
associated revision, to clarify our work and improve the paper
according to the reviewers' suggestions. We believe that we carefully
revised the manuscript to address all concerns and provide responses
to the reviewers' comments, which you will find below.   We
also elected to add detail regarding the ICM strategy
available in Atropos. Thus, we would
like to resubmit the revised manuscript which we hope will be suitable
for publication. \\

\noindent
The revision of the paper includes:
\begin{enumerate}
\item  A new figure revision highlighting ICM and providing a simple
  starter example for Atropos usage. 
\item Fixes to two typos in the equations. 
\item  Many minor changes in the body of the text.  
\item  The references were altered according to reviewer suggestions.
\end{enumerate}
We express our sincere gratitude to
the editorial staff and to the reviewers for their efforts in
evaluating this paper. \\

\noindent
Sincerely,\\

\noindent
Brian Avants, Ph.D.\\
Penn Image Computing and Science Laboratory \\ 
University of Pennsylvania\\
3600 Market Street, Ste 370\\
Philadelphia, PA 19104
\newpage
\begin{center}
\LARGE{{\bf Response To Reviewers}}
\end{center}
\section{Reviewer 1}
We thank reviewer 1 for his/her careful reading for technical content
and for improving our work.\\
\newline
\newline
{\em 1) Page 5, 2nd line from bottom. What are silver standard
  labellings?}
\newline
\newline
We've changed this line to be more specific in the current revision.
\newline
\newline
Done.  
\newline
\newline
{\em 2) Page 10, Equations 8 - 11. These follow standard
  approaches. However, Equation 11 does not seem consistent with [8]
  and [10]. If delta(x\_i,x\_j)=1 for x\_i=x\_j, then U(x) as defined in
  [10] increases, and p(x) as defined in [8] decreases - hence
  labelling similar regions seems to be penalised. However, things
  seem consistent if equation 11 is changed as in the standard graph
  cuts approaches - e.g. see equation 3 Boykov \& Jolly ICCV 2001.}
\newline
\newline
We appreciate the reviewer catching this error in the equation.  If $x_i = x_j$ then $V=0$ and is
1 otherwise, as is standard in Besag's approach. 
\newline
\newline
{\em 3) Page 10, Equation 11. Computing the value of U(x) only
  considers if neighbours are the same or not. However, as this paper
  deals with segmenting a large number of labels why was an approach
  such as that taken in Noe \& Gee IPMI 2001 not considered -
  i.e. learning the probability of labels being neighbours and using
  these?}
\newline
\newline
This is an interesting point and we intend to investigate this
direction of research in future work.  In fact, some basic
functionality along these lines is available within the guts of the
code but additional comparative research into the myriad partial volume
estimation models that have been proposed is required to determine the
best approach.  Thus, we added comments in the discussion to indicate
our interest in pursuing partial volume models but elect not to pursue
this task for this resubmission. 
\newline
\newline
{\em 4) Page 12, lines 10-14. Authors comment that applying graph cuts
  to more than one label is not globally optimal and that EM is a
  reasonable and efficient alternative. Although this is true,another
  property of graph cuts is that it is fast. How does the EM approach
  compare in terms of speed?}
\newline
\newline
Fast is a relative term and we cannot comment on the speed of a method
for which we do not have an optimal implementation.  
The main computational challenge with graph cuts is memory.  Because one needs to
solve multiple shortest path problems---and store the solution of
each---the potential for memory bloat is high.  Even in EM, memory
concerns emerge when enough labels are used (e.g. in our examples).
In theory, though, EM computations are embarassingly parallel which is
not a property shared by graph-based methods.  As such, one would
expect even an optimally implemented multi-label graph cut method to
lag in speed behind an optimally implemented EM segmentation. 
\newline
\newline
{\em 5) Page 13, Equation 17. This looks like a multivariate equation to me. I'm also not familiar with the form of Equation 19 - why is it different from 17?'}
\newline
\newline
The difference between equations 17 and 19 is that the input to the
data term is a vector (rather than scalar) and the scalar variance
parameter is replaced by the covariance matrix.  This is a standard
extension from univariate to multivariate probability. 
\newline
\newline
{\em 6) Page 17. Penultimate line. "Three likelihood classes have been
  developed, one parametric and one ...". Should this be "Three
  likelihood classes have been developed, one parametric and [two]
  ..."?'}
\newline
\newline
Yes, thank you.  The fix is made.  
\newline
\newline
{\em 7) Page 26 Figure 5. In the sign column the signs for the Left
  and Right Caudate Nucleus and Thalamus should be -ve as according to
  the data in the table, Atropos performed significantly worse than
  the majority vote.}
\newline
\newline
Fixed.  
\newline
\newline
{\em 8) I could not verify the validity of all the references as they were not numbered neither were they in alphabetical order. }
\newline
\newline
{\bf FIXME NICK Standard neuroinformatics style?}
\newline
\newline
The reviewer also mentioned not being able to locate the 19 Hammers
datasets.  It is unfortunate that the brain-development website is not
well annotated.  We suspect the reviewer looked at the IXI data which
is not the Hammers data.  We suggest the reviewer look at the adult
atlas section of the website which, currently, contains the data we
used.  This is noted in the revised text. 
\newline
\newline
We also followed each of the reviewer's suggestions on the structure
and grammar of the text. 
\newline
\newline
We appreciate reviewer 1's time and effort.  
\newline
\newline
\section{Reviewer 2}
We thank reviewer 2 for his/her attention to the manuscript and overall 
positive comments.  We made the reviewer's suggested minor changes in this revision.  
The following comments were addressed: 
\newline
\newline
{\em I was very pleased with the clarity of most of the article,
 particularly with the careful introductory explanations,
 demonstrations of the command-line arguments for Atropos, and the
 excellent review of the literature. I was also happy to see the
 multivariate results in Figure 2, and would have liked to have seen
 accompanying quantitative results. If such results were to assert the
 significant improvement possible when including a complementary
 modality (i.e., without bias), that would be a very nice contribution
 in itself.}
\newline
\newline
We agree with the reviewer that establishing the benefit of
multivariate segmentation in the presence of bias would be
significant.  However, because we only have one multivariate dataset
on which to test this hypothesis, it is difficult to do much more than
provide suggestive results, as we've done here.
\newline
\newline
{\em I have only two concerns with the article. First, there is almost
  no explanation of any of the figures in the body of the text, which
  raises questions about the significance of the findings, such as
  those of Figure 4 ("The results show that the PriorProbabilityMaps
  with w = 0.5 (far right) gives the best performance for all
  tissues."). 
\newline
\newline
The revision adds explanatory material in the body of the text
associated with each figure.  {\bf FIXME}
\newline
\newline
{\em Second, why is it that "In this study, we labeled the brain web (sic) template very sparsely, quickly and crudely and did not expect highly accurate results."
The results depending on these manual prior labels aren't informative
-- they don't give good results, and don't tell us whether the results
would be better if better manual labels were used.}
\newline
\newline
We agree with the reviewer and removed this comment and the evaluation
results.  We added: {\color{red}{We provide this capability to allow the user to implement
    an interactive editing and segmentation loop.  The user may
    run Atropos with sparse manual label guidance, evaluate the
    results, update the manual labels and repeat until achieving the desired
    outcome.  This processing loop be performed easily with, e.g., ITK-SNAP.}
\newline
\newline
{\em On page 22: "The advancements introduced with N4 permit such an
  adaptive integration with Atropos." What relevant advances are
  introduced with respect to N3?}
\newline
\newline
{\bf FIXME NICK}
\newline
\newline
{\em Equations: Please remove punctuation following equations, because it came sometimes be distracting. For example, on pg.9, eq.1's comma might also appear to be a prime/apostrophe for the variable y.}
\newline
\newline
Done except in cases where the equation ends a sentence.   
\newline
\newline
Again, we thank each of the three reviewers for their efforts.  We
truly feel that their reading of the paper was excellent and greatly
improved the work.

% \bibliographystyle{elsarticle-harv.bst}
% \bibliography{ftdcbdlong,avantsAll}

\end{document}

### 2011-02-07 Documentation/atropos/atropos.tex
ENH: Getting ready for resubmission to Neuroinformatics
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@763 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 First, we use the BrainWeb dataset from Montreal Neurological
-Institute to evaluate three tissue segmentation performance via (1)
+Institute to evaluate three-tissue segmentation performance via (1)
 K-means segmentation without use of template data; (2) MRF segmentation with initialization by prior probability maps
@@
 of spatial prior probability maps derived from a group template.  We
-also evaluate Atropos performance by using spatial priors to drive a 69 class EM segmentation
+also evaluate Atropos performance by using spatial priors to drive a 69-class EM segmentation
 problem derived from the Hammers atlas from University College London.  These evaluation studies, combined with
@@
 many researchers adopted statistical methods for $n$-tissue anatomical
-brain segmentation.  Given the ``missing data'' aspect of this
-problem, use of the Expectation-Maximization (EM) framework is natural \citep{Dempster1977}.  The
-work described in \cite{Wells1996} was one of the first to use EM for
+brain segmentation.  The Expectation-Maximization (EM) framework is
+natural \citep{Dempster1977} given the ``missing data'' aspect of this
+problem.  The work described in \cite{Wells1996} was one of the first to use EM for
 finding a locally optimal solution by iterating between bias field
@@
 neuroanatomical layout with subsequent processing used to conjoin
-subsets of such Gaussians belonging to the same tissue classes.  Recently,
-anatomically specific strategies for incorporating prior knowledge
-into the segmentation solution have included the use of spatial prior
-probability maps of the various structures of interest
-\citep{Leemput1999a,Marroquin2002,Ashburner2005}.  These spatial prior
-probability maps can also be used to provide an initial segmentation.
+subsets of such Gaussians belonging to the same tissue classes.
+Recently, reseachers have begun to rely on spatial prior probability
+maps of anatomical structures of interest to encode domain knowledge\citep{Leemput1999a,Marroquin2002,Ashburner2005}.  These spatial prior
+probability maps may also provide an initial segmentation.
 Related technological developments model partial volume effects for
@@
 Related neuroanatomical research concerns the selection of geometric
-features of the cortex \citep[e.g.][]{Goualher1999} with aims at
+features of the cortex \citep[e.g.][]{Goualher1999} which aims at
 understanding the functional-anatomical relationship of the human
@@
 destiny of each mortal.  Also, consistent with the entomological motif
-of our ANTs toolkit, {\it Acherontia atropos} is a species of large
+of our ANTs, {\it Acherontia atropos} is a species of large
 moth known for the skull-like pattern visible on its thorax.  }
-efficiently and flexibly implements a $n$-tissue paradigm for
+efficiently and flexibly implements an $n$-tissue paradigm for
 voxel-based image segmentation.  Atropos allows users to harness its
@@
 Coupled with the registration \citep{Avants2011} and template building
-\citep{Avants2010} already included in the ANTs toolkit, Atropos is a
+\citep{Avants2010} already included in the ANTs, Atropos is a
 versatile and powerful software tool which touches multiple aspects of
@@
 currently listed.  We also evaluate Atropos performance on two brain
-MRI segmentation objectives.  First, three tissue classification.
+MRI segmentation objectives.  First, three-tissue classification.
 Second, we test our ability to parcellate the brain into 69
@@
 applied to multivariate data from arbitrary modalities, we limit our
-evaluation to tissue-classification in T1 neuroimaging in part due to
+evaluation to tissue classification in T1 neuroimaging in part due to
 the abundance of ``gold-standard'' data for this modality.  Consistent
@@
 Organization of this work is as follows: we first describe the theory
-behind the various components of Atropos while
-acknowledging that more theoretical discussion is available elsewhere.  
-This is followed by a thorough
-discussion of implementation which, though often overlooked, is of
-immense practical utility.  We then report evaluation results on
-public T1 neuroimaging datasets that have reference gold/silver
-standard labelings.  Finally, we provide a discussion of our results
-and our open source contribution in the context of the remainder of
-this paper and of previous and future work.
+behind the various components of Atropos while acknowledging that more
+theoretical discussion is available elsewhere.  This is followed by a
+thorough discussion of implementation which, though often overlooked,
+is of immense practical utility.  We then report results on the
+BrainWeb and Hammers dataset.  Finally, we
+provide a discussion of our results and our open source contribution
+in the context of the remainder of this paper and of previous and
+future work.
 
@@
 underlying Atropos dates back 20$+$ years and is representative of
-some of the most innovative work in the field.  Although we synthesize
-some of the theoretical work in this section, we recommend to the
-interested reader consultation with the literature cited previously
-for additional perspective.
+some of the most innovative work in the field.  Although we summarize
+some of the theoretical work in this section, we recommend that 
+the interested reader consult the deep literature in this field for
+additional perspective and proofs behind the major concepts.
 
@@
 prior probability terms.  Bayesian paradigms for brain image
-segmentation employ a user-selected observation model defining the
+segmentation employ a user selected observation model defining the
 likelihood term and one or more prior probability terms.  The product
@@
  Atropos optimizes a class
-of user-selectable objective functions each of which may
+of user selectable objective functions each of which may
 be represented in a generic Bayesian framework, as described by \cite{Sanjay-Gopal1998}.  This framework
@@
 p(\mathbf{y}|\mathbf{x})}_{\text{Likelihood(s)}} \underbrace{
-p(\mathbf{x})}_{\text{Prior(s)}}\frac{1}{p(\mathbf{y})}, 
+p(\mathbf{x})}_{\text{Prior(s)}}\frac{1}{p(\mathbf{y})} 
 \end{equation} 
@@
 the mean and variance of the $k^{th}$ model.  When $y_i$ is a vector
-quantity, we replace the Euclidean distance by by Mahalanobis distance
+quantity, we replace the Euclidean distance by Mahalanobis distance
 and define multivariate Gaussian parameters via a mean vector,
@@
 \end{align}
-where $N_B$ is the number of bins used to define the histogram of the sample observations (in Atropos the default is $N_B = 32$) and $c_j$ is the center of $j^{th}$ bin in the histogram.
+where $N_B$ is the number of bins used to define the histogram of the
+sample observations (in Atropos the default is $N_B = 32$) and $c_j$
+is the center of the $j^{th}$ bin in the histogram.
 $\sigma_j$ is the width of each of the $N_B$ Gaussian kernels.
@@
 \begin{align}
-  p\left(x_i | x_{\mathcal{I}-\{i\}}\right) = p\left(x_i | x_{\mathcal{N}_i}\right),
+  p\left(x_i | x_{\mathcal{I}-\{i\}}\right) = p\left(x_i |
+    x_{\mathcal{N}_i}\right) 
 \end{align}
@@
 \begin{align} \label{eq:gibbs}
-p(\mathbf{x}) = Z^{-1} \exp\left(-U(\mathbf{x})\right).
-\end{align}
-$Z$ is a normalization factor known as the {\em partition function}
-and $U(\mathbf{x})$ is the {\em energy function} which can take
+p(\mathbf{x}) = Z^{-1} \exp\left(-U(\mathbf{x})\right) 
+\end{align}
+with $Z$ a normalization factor known as the {\em partition function}
+and $U(\mathbf{x})$ the {\em energy function} which can take
 several forms \citep{Li2001}. In Atropos, as is the case with many
@@
 \begin{align}
-U(\mathbf{x}) = \sum_{i = 1}^N \left( V_i(x_i) + \beta \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j ) \right), 
+U(\mathbf{x}) = \sum_{i = 1}^N \left( V_i(x_i) + \beta \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j ) \right)
 \end{align}
@@
                           \begin{array}{ll}
-                            1 & \text{if } x_i = x_j \\
-                            0 & \text{otherwise}
+                            0 & \text{if } x_i = x_j \\
+                            1 & \text{otherwise}
                           \end{array}
@@
 \end{align}
-and $\beta$ is a granularity term which weights the contribution of the MRF prior on the segmentation solution.
+and $\beta$ is a granularity term which weights the contribution of
+the MRF prior on the segmentation solution.  
 Since Atropos allows for non-uniform neighborhood systems and systems in which not just the immediate face-connected neighbors are considered, we use the modified function also used in \cite{Noe2001}, which weights the interaction term by the Euclidean distance, $d_{ij}$,  between interacting sites $i$ and $j$ such that
@@
 population from which a template is constructed \citep[e.g.][which is
-also available in the ANTs toolkit]{Avants2010}.  Each labeling can
+also available in ANTs]{Avants2010}.  Each labeling can
 then be warped to the template where the synthesis of warped labeled
@@
 the spatial distribution of labeled anatomy which can be harnessed in
-joint segmentation/registration or, specifically, Atropos/ANTs hybrids involving unlabeled subjects.
+joint segmentation/registration or Atropos/ANTs hybrids involving unlabeled subjects.
 
@@
 \begin{align}
-\mathrm{Pr}(x_i = l_k) = \frac{t_{ik}}{\sum_{j=1}^K t_{ij}},
+\mathrm{Pr}(x_i = l_k) = \frac{t_{ik}}{\sum_{j=1}^K t_{ij}}
 \end{align}
@@
 posterior) at a subset of $\mathcal{I}$ to be $1$ for the known label
-and $0$ for each other label at the same site.  The user-input
+and $0$ for each other label at the same site.  The user input
 therefore not only initializes the optimization, but also gives boundary conditions
@@
 reasonable and efficient alternative to Boykov's solution.
-Furthermore, one may automate the initialization process.  We compare
-our sparse prior labeling strategy to fully automated methods in the
-evaluation section.
+Furthermore, one may automate the initialization process.  
+{\color{red}{We provide this capability to allow the user to implement
+    an interactive editing and segmentation loop.  The user may
+    run Atropos with sparse manual label guidance, evaluate the
+    results, update the manual labels and repeat until achieving the desired
+    outcome.  This processing loop be performed easily with, e.g., ITK-SNAP.}}
 
@@
 \end{align}
-The probability calculations are illustrated in Fig.~\ref{fig:distancePrior} and were designed so that the most interior point, or set of points, of region $R_k$ would have a distance prior probability of 1.0 which would linearly decrease within the region to the boundary with a user-selected probability value of $\alpha_k$.
-Outside $R_k$, the distance prior exponentially decays with a decay rate governed by another user-selected parameter $\sigma_k$.
+The probability calculations are illustrated in
+Fig.~\ref{fig:distancePrior} and were designed so that the most
+interior point, or set of points, of region $R_k$ would have a
+distance prior probability of 1.0 which would linearly decrease within
+the region to the boundary with a user selected probability value of
+$\alpha_k$.  Outside $R_k$, the distance prior exponentially decays
+with a decay rate governed by another user selected parameter $\sigma_k$.
 This permits a dense segmentation throughout the segmentation region of $\mathcal{I}$ even where the template prior probabilities are zero.
@@
 Atropos uses expectation maximization to find a locally optimal
-solution for the user-selected version of the Bayesian segmentation
+solution for the user selected version of the Bayesian segmentation
 problem (cf. Eqn. (\ref{eq:bayes})).
@@
 \begin{align}  
- \mu_k \leftarrow \frac{ \sum_{i=1}^N  y_i p_k(l_k|y_i)}{ \sum_{i=1}^N p_k(l_k|y_i) },
+ \mu_k \leftarrow \frac{ \sum_{i=1}^N  y_i p_k(l_k|y_i)}{ \sum_{i=1}^N p_k(l_k|y_i) }
 \end{align}
@@
 \begin{align}
- \boldsymbol{\mu}_k \leftarrow \frac{ \sum_{i=1}^N  \mathbf{y}_i p_k(l_k|\mathbf{y}_i)}{ \sum_{i=1}^N p_k(l_k|\mathbf{y}_i) },
+ \boldsymbol{\mu}_k \leftarrow \frac{ \sum_{i=1}^N  \mathbf{y}_i p_k(l_k|\mathbf{y}_i)}{ \sum_{i=1}^N p_k(l_k|\mathbf{y}_i) }
 \end{align}
@@
 This type of update is known as {\em soft} EM.  Hard EM, in
-contrast, only uses sites containing label $l_k$ to update model $k$ parameters.
+contrast, only uses sites containing label $l_k$ to update the 
+parameters for the $k^{th}$ model.
 A similar pattern is used in non-parametric cases.
@@
 To provide a more intuitive interface without the overhead costs of a graphical user interface, a set of unique command line parsing classes were developed which can also provide insight to the functionality of Atropos.
-The short version of the command line help menu is given in Listing \ref{listing:command} which is invoked by typing `{\ttfamily Atropos -h}' at the command prompt.
-Both short and long option flags are available and each option has its own set of possible values and parameters introduced in a more formal way in both previous discussion and related papers cited in the introduction.
-Here we  describe these options from the unique perspective of  implementation.
+The short version of the command line help menu is given in Listing
+\ref{listing:command} which is invoked by typing `{\ttfamily Atropos
+  -h}' at the command prompt.  Both short and long option flags are
+available and each option has its own set of possible values and
+parameters introduced in a more formal way in both the previous discussion and related papers cited in the introduction.  Here we describe these options from the unique perspective of  implementation.
 
@@
   mixture proportions (or not) by setting  {\ttfamily -p Socrates[1]}
-  or {\ttfamily -p Socrates[0] }.  Fixed label boundary conditions may be employed by
+  or {\ttfamily -p Socrates[0]}.  Fixed label boundary conditions may be employed by
   selecting the {\ttfamily PriorLabelImage} initialization and   {\ttfamily -p Plato[0]}.
@@
 Higher dimensions than 4 are possible although we have not yet encountered such 
-application-specific need.  Multiple images (assumed to be of the same
+an application-specific need.  Multiple images (assumed to be of the same
 dimension, size, origin, etc.), will automatically enable multivariate
@@
 opted for different likelihood models which have included either
-parametric, specifically Gaussian, or non-parametric variations.
+parametric (Gaussian) or non-parametric variations.
 However, these approaches are similar in that they require a list
@@
 all derived from an abstract list sample function class.  Three
-likelihood classes have been developed, one parametric and one
+likelihood classes have been developed, one parametric and two
 non-parametric, and are available for usage although one of the
@@
 neighborhood radius of 1 in all 3 dimensions.  This radius is defined
-such that voxels including but not limited to those that are face
-connected will influence the MRF.
+such that voxels including but not limited to those that are face-connected will influence the MRF.
 
@@
 \subsubsection{Registration and Probability Maps} 
-Image registration enables the ability to transform information
+Image registration enables one to transfer information
 between spatial domains which may aid in both segmentation and bias
@@
 correction step into the actual workflow.  The
-advancements introduced with N4 permit such an adaptive
+advances introduced with N4 permit such an adaptive
 integration with Atropos.  Recent demonstrations suggest improved
@@
 The Atropos algorithm is cross-platform and compiles on, at minimum,
-modern osx, windows and linux-based operating systems.  The
-user-interface may be reached through the operating system's user
+modern OSX, Windows and Linux-based operating systems.  The
+user interface may be reached through the operating system's user
 terminal.  Because of its portability and low-level efficiency,
@@
 instantiated for different applications but here we evaluate only two
-of the many possibilites. First, we perform an evaluation on the
+of the many possibilities. First, we perform an evaluation on the
 BrainWeb dataset using both the standard T1 image with multiple bias
 and noise levels and also the BrainWeb20 data
-\citep{Aubert-Broche2006a,Battaglini2008}.  In combination, this data
-allows one to vary not only noise and bias but also the underlying
-anatomy.  Second, we evaluate the use of Atropos in improving whole
-brain parcellation and exercise its ability to efficiently solve {\em
-many class} expectation maximization problem.  We choose this
+\citep{Aubert-Broche2006a,Battaglini2008}.  In combination, these data
+allow one to vary not only noise and bias but also the underlying
+anatomy.  Second, we evaluate the use of Atropos in improving whole-brain parcellation and exercise its ability to efficiently solve {\em
+many-class} expectation maximization problem.  We choose this
 evaluation problem in part to illustrate the flexibility of Atropos
@@
 allows many-class problems to be solved with low memory usage ($<$2GB
-for a 69 class model on 1 mm$^3$ brain data).
+for a 69-class model on 1 mm$^3$ brain data).
 
@@
 \subsubsection{Single-Subject Evaluation} We use the single-subject
-data with 3\% noise and three levels of bias referred to as 0, 20 and 40 \% RF inhomogeneity.  We
+data with 3\% noise and three levels of bias referred to as 0, 20 and 40\% RF inhomogeneity.  We
 study the effect of the MRF prior term and initialization on
@@
 Segmentation is then repeated, with the same parameters,
-but with the N4 corrected image as input.  The resulting algorithm is
+but with the N4-corrected image as input.  The resulting algorithm is
 similar to those that fix segmentation parameters while
@@
 of bias on the combination of N4 and Atropos and also the validity of our
-prior label image initialization.  The prior label image manual
-initialization was based on a crude ITK-SNAP labeling of a small
-set of voxels for each tissue.  Results of these evaluation scenarios, in terms of
+prior label image initialization.  
+% The prior label image manual initialization was based on a crude
+% ITK-SNAP labeling of a small set of voxels for each tissue.  
+Results of these evaluation scenarios, in terms of
 Dice overlap, are shown in Figure~\ref{fig:bweb1}. Because overlap
@@
 conclude that simple N4 pre-processing is adequate to correct even the 40\%
-bias level.  An example of this procedure, using BrainWeb data with 40\% RF bias,
+RF bias level.  An example of this procedure, using BrainWeb data with 40\% RF bias,
 is in Figure~\ref{fig:bwebrf40}.  We supply the information necessary
@@
 `{\ttfamily atroposBwebRF40FigureExample.sh}' which is available in
-the ANTs toolkit's Atropos documentation folder as of svn commit 711.
+the ANTs Atropos documentation folder as of SVN commit 711.
 The script may be easily modified to run the whole
 evaluation.  Figure~\ref{fig:bwebrf40} shows the results of
-simultaneously using proton density and T1 weighted BrainWeb data to
+simultaneously using proton density and T1-weighted BrainWeb data to
 perform the segmentation.  This multivariate input data outperforms
-the univariate T1 weighted data alone.
+the univariate T1-weighted data alone.
 \begin{figure}
@@
 to guide N4 bias correction and produce the image in (c).  We
-repeat the K-means segmentation, but with the N4 corrected image as
+repeat the K-means segmentation, but with the N4-corrected image as
 input and produce the segmentation in (d).  The average 3-tissue Dice
@@
 \includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF20.pdf}
-\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF40.pdf}\\
-\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF0.pdf}
-\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF20.pdf}
-\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF40.pdf}
+\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF40.pdf}
+% \includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF0.pdf}
+% \includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF20.pdf}
+% \includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF40.pdf}
 \end{tabular}
@@
   bias, even at the 40\% level.  The optimal $\beta$ for the MRF term
-  appears to be between $0.1$ and $0.2$.  The prior label image
-  initialization is competitive with but inferior to K-means.  This is
-  sensible in that the prior labels image outcome is sensitive to the
-  user choices made.  In this study, we labeled the brain web template
-  very sparsely, quickly and crudely and did not expect highly accurate results.  The
-  legend is in the same position in each graph, allowing a visual
-  comparison of the results.  As one may see, the N4 assisted overlap
+  appears to be between $0.1$ and $0.2$.  
+% The prior label image initialization is competitive with but inferior to K-means.  This is
+%  sensible in that the prior labels image outcome is sensitive to the
+%  user choices made.  In this study, we labeled the brain web template
+%  very sparsely, quickly and crudely and did not expect highly
+%  accurate results.  
+  The legend is in the same position in each graph, allowing a visual
+  comparison of the results.  As one may see, the N4-assisted overlap
   values are consistent across bias field/RF inhomogeneity.}
@@
 \end{figure}
-\subsubsection{20 Subject Evaluation} The single-subject BrainWeb
+\subsubsection{20-subject Evaluation} The single-subject BrainWeb
 study in the previous section tested the basic Atropos options and the
@@
 evaluation applies the script {\ttfamily ants\_multitemplate\_labeling.sh }
-(available in the ANTs toolkit) to the 19 Hammers evaluation datasets
-located at \url{http://www.brain-development.org/}
+(available in ANTs) to the 19 Hammers evaluation datasets
+located at \url{http://www.brain-development.org/} and currently under
+the adult atlases section 
 \citep{Hammers2003,Heckemann2006}.  These initial majority voting
@@
 {\ttfamily  -i  PriorProbabilityImages[3,label\_prob\%02d.nii.gz,0.5] 
--m [0.1,1x1x1] -c [5,0] -p Socrates[1] }.  Results, in terms of
+-m [0.1,1x1x1] -c [5,0] -p Socrates[1]}.  Results, in terms of
 Dice overlap, are shown in Figure~\ref{fig:hammers}.
@@
 \noindent{\bf Data:} The BrainWeb data is freely available.  We used
-single-subject BrainWeb data as is but added a meta format data header
+single-subject BrainWeb data as is but added a metaformat data header
 to the raw binary files.  An example copy of this header is contained
-in {\ttfamily atroposBwebRF40FigureExample.sh }.  The 20 subject data,
+in {\ttfamily atroposBwebRF40FigureExample.sh }.  The 20-subject data,
 however, required excluding non-cerebrum tissue classes.  The Hammers
@@
 instructions at \url{http://www.picsl.upenn.edu/ANTS/download.php}.
-Svn release 711 was used for the examples and evaluations performed in
+SVN release 711 was used for the examples and evaluations performed in
 this paper.  Some components of ANTs depend on the Insight ToolKit.
@@
 \noindent{\bf Scripts:} The complete script for the single-subject
-BrainWeb study is based on generalizing {\ttfamily atroposBwebRF40FigureExample.sh }, which is available in the ANTs
-toolkit (svn release 711 or greater) and which reproduces Figure 2
+BrainWeb study is based on generalizing {\ttfamily atroposBwebRF40FigureExample.sh}, which is available in the ANTs
+toolkit (SVN release 711 or greater) and which reproduces Figure 2
 results.  The template-based normalization procedure for the BrainWeb
 20 and the Hammers evaluation data is based on freely available
-scripts included with the ANTs toolkit, specifically {\ttfamily
-  ants\_multitemplate\_labeling.sh } and {\ttfamily
-  buildtemplateparallel.sh }.  A release version of ANTs---with a final version of Atropos---will
+scripts included with ANTs, {\ttfamily
+  ants\_multitemplate\_labeling.sh} and {\ttfamily
+  buildtemplateparallel.sh}.  A release version of ANTs---with a final version of Atropos---will
 be prepared with the final version of this manuscript.
@@
 easily be used with Atropos to improve segmentation.
-Furthermore, we used multiple subject BrainWeb data to build dataset
-specific priors that provided the most consistent segmentation
+Furthermore, we used multiple subject BrainWeb data to build dataset-specific priors that provided the most consistent segmentation
 performance across tissues.  Finally, we used majority voting to
-initialize an Atropos EM solution to a 69 class brain parcellation
+initialize an Atropos EM solution to a 69-class brain parcellation
 problem.  Significant improvements were gained in multiple brain
@@
 Atropos results are competitive with the state of the art.  For
-instance, \cite{Ashburner2005} (SPM5) evaluated on 0\% RF (bias field) 3\%
+instance, \cite{Ashburner2005} (SPM5) evaluated on 0\% RF bias field, 3\%
 noise BrainWeb single subject data finding 0.932 (GM) and .961 (WM)
-Dice. Results on 40\% RF were 0.934 (GM) and 0.961 (WM). SPM5 exhibits
-insensitivity to bias similar to our own best results on the 40\% RF, 3\% 
-noise case (MRF-$\beta$=0.2, K-means $+$ N4) with Dice for GM is
-0.950737 and for WM is 0.962586.  \cite{Nakamura2009} gave GM Dice
-results (BrainWeb single 3\% noise) of 0.962 (0\% RF), 0.964 (20\% RF)
-and 0.956 (40\%RF) which are slightly higher than either SPM5 or Atropos
+Dice overlap. Results on 40\% RF bias were 0.934 (GM) and 0.961 (WM). SPM5 exhibits
+insensitivity to bias similar to our own best results on the 40\% RF bias, 3\% 
+noise case (MRF-$\beta$=0.2, K-means $+$ N4) with Dice overlap for GM is
+0.951 and for WM is 0.963.  \cite{Nakamura2009} gave GM Dice overlap
+results (BrainWeb single 3\% noise) of 0.962 (0\% RF bias), 0.964
+(20\% RF bias)
+and 0.956 (40\% RF bias) which are slightly higher than either SPM5 or Atropos
 results.  However, \cite{Nakamura2009} do not report WM or CSF numbers
-for comparison.  Topology preserving methods also perform well.
-\cite{Shiee2010} achieved Dice results for 3\% noise 20\% RF BrainWeb
+for comparison.  Topology-preserving methods also perform well.
+\cite{Shiee2010} achieved Dice overlap for 3\% noise 20\% RF bias BrainWeb
 single subject with 0.927, 0.912 and 0.900 for WM, GM and CSF,
@@
 \cite{Bazin2007c} proposed TOADS and, estimating from the paper's
-graph, showed that the average Dice accuracy for 3\% noise various RF
-was 0.95--0.96 for WM, 0.93--0.95 for GM and 0.92--0.94 for CSF.
+graph, showed that the average Dice overlap accuracy for 3\% noise for various RF
+was 0.950--0.960 for WM, 0.930--0.950 for GM and 0.920--0.940 for CSF.
 Perhaps the most recent balanced evaluation was performed in
@@
 voxels for BrainWeb are known, we can convert the confusion matrix to
-Dice.  In that case, the SPM5 Dice for BrainWeb single GM and WM is
-0.8847 and 0.9085, respectively, while FreeSurfer and FSL's accuracy
+Dice overlap.  In that case, the SPM5 Dice overlap for BrainWeb single GM and WM is
+0.885 and 0.909, respectively, while FreeSurfer and FSL's accuracy
 is lower. The best GM result for 20 subjects is obtained by SPM5:
-0.9295; best WM Dice is from FSL: 0.9499.  We note that
+0.930; best WM Dice overlap is from FSL: 0.950.  We note that
 \cite{Klauschen2009} used a comprehensive evaluation where quality of
@@
 must be interpreted slightly differently than those from other papers.
-Finally, we note that in our evaluation of 20 subject BrainWeb data, the prior
+Finally, we note that in our evaluation of 20-subject BrainWeb data, the prior
 probability models performed best of all the models used.  Compared to
-the KMeans based segmentation, the prior based segmentation
+the K-Means based segmentation, the prior based segmentation
 performance also peaked at lower values of the MRF-$\beta$ term (0.0
 and 0.05).  This is reasonable in that the spatial priors themselves
-impose a degree of regularity to the segmentation, as in SPM5.
+impose a degree of regularity on the segmentation, as in SPM5.
 
@@
 respect to a known neurobiological outcome measure.  Atropos is
-currently used in clinical studies and a number of clinically-focused,
+currently used in clinical studies and a number of clinically focused,
 application-specific evaluations are ongoing and will constitute
@@
 segmentation is non-trivial and there are numerous design alternatives
-available not only in the models selected but also in the ICM coding, alternatives to ICM and the specific method in which prior
-and likelihood are combined.  Due to the flexibility of Atropos, we
-also hope that some of its capabilities, though not evaluated here,
-are explored by the segmentation or clinical community.
+available not only in the models selected but also in the ICM coding,
+alternatives to ICM and the method in which prior and likelihood are
+combined.  Due to the flexibility of Atropos, we also hope that some
+of its capabilities, though not evaluated here, are explored by the segmentation or clinical community.
+
+\paragraph{Information Sharing Statement}
+{\color{red}{Atropos software is available in ANTs
+    \url{http://www.picsl.upenn.edu/ANTs} 
+which depends on ITK \url{http://www.itk.org/Wiki/ITK/Git/Download}.
+The data used in this work is available in the ANTs software
+repository, BrainWeb 
+\url{http://mouldy.bic.mni.mcgill.ca/BrainWeb/} and at \url{www.brain-development.org}.
+We employed itk-SNAP for visualization \url{www.itksnap.org}.}
 

### 2011-02-07 Documentation/atropos/atropos.tex
Added some stuff.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@765 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 \usepackage{lscape}
-\usepackage[round,semicolon,authoryear,]{natbib}
+\usepackage[square,comma,numbers,sort&compress]{natbib}
 
@@
 Furthermore, one may automate the initialization process.  
-{\color{red}{We provide this capability to allow the user to implement
+{\color{blue}{We provide this capability to allow the user to implement
     an interactive editing and segmentation loop.  The user may
@@
 \item[Label Update (E-Step):] Given the initialization and fixed
-model parameters, {\color{red}{Atropos is capable of updating the current label 
+model parameters, {\color{blue}{Atropos is capable of updating the current label 
 estimates using either a {\em synchronous} or {\em asynchronous} scheme.
@@
 \begin{center}
-\begin{tabular}{cc}
-\includegraphics[width=70mm]{r16slice.png} &
-\includegraphics[width=70mm]{r16icm.png} \\
-(a) & (b) \\
+\begin{tabular}{c}
+\includegraphics[width=160mm]{ICMfigure.pdf}
+%\includegraphics[width=70mm]{r16slice.png} &
+%\includegraphics[width=70mm]{r16icm.png} \\
+%(a) & (b) \\
 \end{tabular}
 \end{center}
-\caption{\baselineskip 12pt \small  (a) An adult brain image slice and (b) the corresponding ICM code image created using a $5\times5$ MRF neighborhood with a maximum ICM code label of `13'.  }
+\caption{\baselineskip 12pt \small {\color{blue}{An adult brain image slice is shown 
+with its ICM code image corresponding to a $5\times5$ MRF neighborhood.  
+To the right of the ICM code image, we focus on a single neighborhood
+with a center voxel associated with the ICM code label of '10'.  Each 
+center voxel in a specified neighborhood exhibits a unique ICM code
+label which does not appear elsewhere in its neighborhood.  When performing
+the segmentation labeling update for ICM, we iterate through the set
+of ICM code labels and, for each code label, we iterate through the image
+and update only those voxels associated with the current code label.}}
+  }
 \label{fig:r16icm}
@@
 normalization/bias correction step using a method such as the recently developed N4
-algorithm \citep{Tustison2010}.  N4 builds upon the popular
-non-parametric nonuniformity intensity normalization (N3) introduced in \cite{Sled1998}.  It was previously mentioned that
-several methods proposed in the literature have taken an integrative
-view of the segmentation problem by incorporating an intrinsic bias
-correction step into the actual workflow.  The
-advances introduced with N4 permit such an adaptive
-integration with Atropos.  Recent demonstrations suggest improved
+algorithm \citep{Tustison2010}.  {\color{blue}{N4 extends the popular
+nonparametric nonuniform intensity normalization (N3) algorithm \citep{Sled1998}
+in two principal ways:
+\begin{itemize}
+  \item We replace the least squares B-spline fitting with a parallelizable alternative 
+  (which we also made publicly available in the Insight Toolkit)---
+  the advantages being that 1) computation is much faster 
+  and 2) smoothing is not susceptible to outliers as is characteristic with
+  standard least squares fitting algorithms.
+  \item The fitting algorithm permits a multi-resolution approach so whereas standard 
+  N3 practice is to select a single resolution at which bias correction occurs, the N4
+  framework permits a multi-resolution correction where a base resolution is chosen and
+  correction can then occur at multiple resolution levels each resolution being twice 
+  the resolution of the previous level.
+\end{itemize}  
+Specifically, with respect to segmentation, there exists a third advantage with N4 over
+N3 in that the former permits the specification of a probabilistic mask as opposed 
+to a binary mask.}}
+%It was previously mentioned that
+%several methods proposed in the literature have taken an integrative
+%view of the segmentation problem by incorporating an intrinsic bias
+%correction step into the actual workflow.  The
+%advances introduced with N4 permit such an adaptive
+%integration with Atropos.  
+Recent demonstrations suggest improved
 white matter segmentation produces better gain field estimates using
@@
 \paragraph{Information Sharing Statement}
-{\color{red}{Atropos software is available in ANTs
+{\color{blue}{Atropos software is available in ANTs
     \url{http://www.picsl.upenn.edu/ANTs} 
@@
 \url{http://mouldy.bic.mni.mcgill.ca/BrainWeb/} and at \url{www.brain-development.org}.
-We employed itk-SNAP for visualization \url{www.itksnap.org}.}
+We employed itk-SNAP for visualization \url{www.itksnap.org}.}}
 

### 2011-02-07 Documentation/atropos/ResponseToReviewers.tex
Added some stuff.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@766 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
   starter example for Atropos usage. 
-\item Fixes to two typos in the equations. 
+\item Typo fixes in the equations. 
 \item  Many minor changes in the body of the text.  
@@
 concerns emerge when enough labels are used (e.g. in our examples).
-In theory, though, EM computations are embarassingly parallel which is
+In theory, though, EM computations are embarrassingly parallel which is
 not a property shared by graph-based methods.  As such, one would
@@
 lag in speed behind an optimally implemented EM segmentation.
+
+{\color{blue}{ We would also
+like to point out that we have had prior experience with graph cuts (see,
+for example, our implementation available at http://hdl.handle.net/1926/1503
+and related publications:  Zhuang Song, Nicholas J. Tustison, Brian B. Avants, 
+James C. Gee: Integrated Graph Cuts for Brain MRI Segmentation. MICCAI (2) 2006: 831-838
+and Zhuang Song, Nicholas J. Tustison, Brian B. Avants, James C. Gee: 
+Integrated Graph Cuts for Brain MRI Segmentation. MICCAI (2) 2006: 831-838) and
+our preference for our current EM-based approach is certainly not restricted to 
+speed considerations.  As enthusiastic supporters of open source software,
+patent issues constrain the use of graph cuts where no such limitations exist
+for EM-based approaches.}}
 \newline
@@
 \newline
-{\em 6) Page 17. Penultimate line. "Three likelihood classes have been
-  developed, one parametric and one ...". Should this be "Three
+{\em 6) Page 17. Penultimate line. ``Three likelihood classes have been
+  developed, one parametric and one ...''. Should this be ``Three
   likelihood classes have been developed, one parametric and [two]
-  ..."?'}
+  ...''?'}
 \newline
@@
 \newline
-{\bf FIXME NICK Standard neuroinformatics style?}
+{\color{blue}{We apologize.  In preparing the original
+manuscript, we were unable to find a .bst for the
+Neuroinformatics journal so we created our own using
+natbib and did not use all the options to facilitate
+the reviewing process.  The references are now numbered
+in the manuscript and in the bibliography section and
+are ordered as they appear in the manuscript.   }}
 \newline
@@
   raises questions about the significance of the findings, such as
-  those of Figure 4 ("The results show that the PriorProbabilityMaps
+  those of Figure 4 (``The results show that the PriorProbabilityMaps
   with w = 0.5 (far right) gives the best performance for all
-  tissues."). 
+  tissues.''). }
 \newline
@@
 \newline
-{\em Second, why is it that "In this study, we labeled the brain web (sic) template very sparsely, quickly and crudely and did not expect highly accurate results."
+{\em Second, why is it that ``In this study, we labeled the brain web (sic) template very sparsely, quickly and crudely and did not expect highly accurate results.''
 The results depending on these manual prior labels aren't informative
@@
 We agree with the reviewer and removed this comment and the evaluation
-results.  We added: {\color{red}{We provide this capability to allow the user to implement
+results.  We added: 
+    \begin{quote}
+    We provide this capability to allow the user to implement
     an interactive editing and segmentation loop.  The user may
@@
     results, update the manual labels and repeat until achieving the desired
-    outcome.  This processing loop be performed easily with, e.g., ITK-SNAP.}
-\newline
-\newline
-{\em On page 22: "The advancements introduced with N4 permit such an
-  adaptive integration with Atropos." What relevant advances are
+    outcome.  This processing loop be performed easily with, e.g., ITK-SNAP.
+    \end{quote}
+{\em On page 22: ``The advancements introduced with N4 permit such an
+  adaptive integration with Atropos.'' What relevant advances are
   introduced with respect to N3?}
@@
 \newline
-{\bf FIXME NICK}
-\newline
-\newline
+{\color{blue}{We added the following elaboration on the benefits of N4 in section 3.4.3 (page 21):
+\begin{quote}
+N4 extends the popular
+nonparametric nonuniform intensity normalization (N3) algorithm [63]
+in two principal ways:
+\begin{itemize}
+  \item We replace the least squares B-spline fitting with a parallelizable alternative 
+  (which we also made publicly available in the Insight Toolkit)---
+  the advantages being that 1) computation is much faster 
+  and 2) smoothing is not susceptible to outliers as is characteristic with
+  standard least squares fitting algorithms.
+  \item The fitting algorithm permits a multi-resolution approach so whereas standard 
+  N3 practice is to select a single resolution at which bias correction occurs, the N4
+  framework permits a multi-resolution correction where a base resolution is chosen and
+  correction can then occur at multiple resolution levels each resolution being twice 
+  the resolution of the previous level.
+\end{itemize}  
+Specifically, with respect to segmentation, there exists a third advantage with N4 over
+N3 in that the former permits the specification of a probabilistic mask as opposed 
+to a binary mask.  Recent demonstrations suggest improved
+white matter segmentation produces better gain field estimates using
+N3 [64].  Thus, when performing 3-tissue segmentation, we
+may opt to use, for instance, the posterior probability map of white matter at the current
+iteration as a weighted mask for input to N4.
+\end{quote} }}
 {\em Equations: Please remove punctuation following equations, because it came sometimes be distracting.

### 2011-02-07 Documentation/atropos/atropos.tex
Added some stuff.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@767 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 }
-Although many optimization techniques exist \citep[see the
-introduction in][for a concise summary of the myriad optimization
-possibilities]{Marroquin2002}---each with their characteristic
+Although many optimization techniques exist (see the introduction in 
+\cite{Marroquin2002} for a concise summary of the myriad optimization
+possibilities)---each with their characteristic
 advantages and disadvantages in terms of computational complexity and
@@
 classes to segment.  The simplest initialization is by the classic
-K-means algorithm with only the number of classes as a prior.
+K-means or Otsu thresholding algorithms with only the number of classes 
+specified by the user.
 Otherwise, the user must provide prior information for each class in
@@
 Such an image can be created and viewed with Atropos by assigning a valid
-filename in the \texttt{--icm [1]} set of options.  An example is given in 
-Figure \ref{fig:r16icm}.  For each iteration, a random ordering of the
+filename in the \texttt{--icm [1]} set of options.  An example adult brain
+slice and the associated code image is given in 
+Figure \ref{fig:r16icm} for an MRF neighborhood of $5 \times 5$ pixels.
+This produces a maximum code label of `13'.
+For each iteration, a random ordering of the
 set $\{1,\ldots,C\}$ is created which prescribes the order in which the
@@
 probability until convergence (in $\sim$5 iterations).  One can specify
-a maximum number of these ``ICM iterations'' on the command line.}}
+a maximum number of these ``ICM iterations'' on the command line.  For
+our example in Figure \ref{fig:r16icm}, this means that for each
+ICM iteration, we would iterate through the image 13 times only updating
+those segmentation labels associated with the current ICM code.}}
 
@@
 Fig.~\ref{fig:flowchart}.
+{\color{blue}{Given a set of input images and a mask images, each is preprocessed using N4 to correct for intensity inhomogeneity.
+For our brain processing pipeline, the mask is usually obtained from the standard skull-stripping preprocessing step which also uses Atropos.
+Initialization can be performed in various ways using standard clustering techniques, such as K-means, to prior-based images.
+This initialization is used to provide the initial estimate of the parameters of the likelihood model for each class.
+These likelihoods combine with the current labeling to generate the current estimate of the posterior probabilities at each voxel for each class.
+At each iteration, one can also integrate N4 by using the current posterior probability estimation of the white matter to update the estimate of bias field.
+}}
+
 To provide a more intuitive interface without the overhead costs of a graphical user interface, a set of unique command line parsing classes were developed which can also provide insight to the functionality of Atropos.

### 2011-02-09 Documentation/atropos/atropos.tex
ENH: minor changes to response and atropos wrt wu's reading.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@772 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 possible).  The MRF locality condition is then, 
-\begin{align}
+\begin{align}\label{eq:cond}
   p\left(x_i | x_{\mathcal{I}-\{i\}}\right) = p\left(x_i |
@@
 \begin{align}
-U(\mathbf{x}) = \sum_{i = 1}^N \left( V_i(x_i) + \beta \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j ) \right)
+U(\mathbf{x}) = \sum_{i = 1}^N \left( V_i(x_i) + \beta \sum_{j \in
+    \mathcal{N}_i} V_{ij}( x_i, x_j ) \right) \nonumber
 \end{align}
@@
     results, update the manual labels and repeat until achieving the desired
-    outcome.  This processing loop be performed easily with, e.g., ITK-SNAP.}}
-
+    outcome.  This processing loop may be performed easily with, e.g., ITK-SNAP.}}
 \begin{comment}
@@
 This produces a maximum code label of `13'.
-For each iteration, a random ordering of the
-set $\{1,\ldots,C\}$ is created which prescribes the order in which the
+For each iteration, one has the option to permute the
+set $\{1,\ldots,C\}$ which prescribes the order in which the
 voxel labels are updated asynchronously. After the first pass through the
@@
 \begin{tabular}{c}
-\includegraphics[width=2.2in]{Figures/bw20_kmeans.png}
-\includegraphics[width=2.2in]{Figures/bw20_priorIni.png}
-\includegraphics[width=2.2in]{Figures/bw20_priorSeg.png}
+\includegraphics[width=2.2in]{Figures/bw20_kmeans.pdf}
+\includegraphics[width=2.2in]{Figures/bw20_priorIni.pdf}
+\includegraphics[width=2.2in]{Figures/bw20_priorSeg.pdf}
 \end{tabular}
@@
 subject.  The following parameters are the most relevant to this discussion:
-{\ttfamily  -i  PriorProbabilityImages[3,label\_prob\%02d.nii.gz,0.5] 
--m [0.1,1x1x1] -c [5,0] -p Socrates[1]}.  Results, in terms of
+{\ttfamily  -i  PriorProbabilityImages[69,label\_prob\%02d.nii.gz,0.5] 
+-m [0.2,1x1x1] -c [5,0] -p Socrates[1]}.  Results, in terms of
 Dice overlap, are shown in Figure~\ref{fig:hammers}.

### 2011-02-09 Documentation/Tex/ants.tex
ENH: updating BTP example
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@773 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 computationally distributed optimal template construction.
+You can also call  \verb ANTS \verb --help to get basic usage.
 
@@
 neighbor interpolation, otherwise linear interpolation is used. 
+You can call  \verb WarpImageMultiTransform \verb --help to get basic usage.
 
@@
 
-\subsection{The ImageMath Executable} This is a multi-purpose program that has the following syntax:  ``ImageMath ImageDimension outputfilename   Operation  InputFileName  parameters ``.
-Most basic scalar image operations -- and some tensor operations -- may be performed with this program.
-Some operations output text files, some output images and some output only to the terminal.
-ImageMath allows one to multiply images together (m), to negate images (Neg), to take an image to a power (pow), to test the invertibility of transformations (InvId), to compute the fractional anisotropy of an image (TensorFA) and to perform Markov Random Field segmentation with or without priors (Segment).
-Many other operations are available.
-Like all other ANTS programs, one may call ImageMath from the command line to see all of its options.
-ImageMath is used heavily in ANTS scripts.
+\subsection{The ImageMath Executable} This is a multi-purpose program
+that has the following syntax:  ``ImageMath ImageDimension
+outputfilename   Operation  InputFileName  parameters ``.  Most basic
+scalar image operations -- and some tensor operations -- may be
+performed with this program.  Some operations output text files, some
+output images and some output only to the terminal.   ImageMath allows
+one to multiply images together (m), to negate images (Neg), to take
+an image to a power (pow), to test the invertibility of
+transformations (InvId), to compute the fractional anisotropy of an
+image (TensorFA) and to perform Markov Random Field segmentation with
+or without priors (Segment).   Many other operations are available.
+Like all other ANTS programs, one may call ImageMath from the command
+line to see all of its options.   ImageMath is used heavily in ANTS
+scripts.   You can call it, in compination with grep or less, to attempt to
+find some of its capabilities.
 
@@
 \begin{verbatim}
-~/Code/ANTSSourceForgeSVN/bin/ANTS 2 -m PR\[r16slice.nii,r64slice.nii,1,2\]
+~/Code/ANTSSourceForgeSVN/bin/ANTS 2 -m CC\[r16slice.nii,r64slice.nii,1,2\]
 \end{verbatim}
@@
 \begin{verbatim}
-~/Code/ANTSSourceForgeSVN/bin/ANTS 2 -m PR[r16slice.nii,r64slice.nii,1,2]
+~/Code/ANTSSourceForgeSVN/bin/ANTS 2 -m CC[r16slice.nii,r64slice.nii,1,2]
 \end{verbatim}
@@
     \end{tabular}
-  \caption{Transformations and a subset of the similarity metrics available in ANTS.
-Similarity metric acronyms:  CC = fast cross correlation, PR = pure cross correlation (the preferred metric), MSQ = mean squared difference, MI = mutual information, PSE = point-set expectation \cite{Pluta2008}.
+  \caption{Transformations and a subset of the similarity metrics
+    available in ANTS.  Similarity metric acronyms:  CC = fast cross
+    correlation (the preferred metric), PR = cross correlation (deprecated), MSQ = mean squared difference, MI = mutual information, PSE = point-set expectation \cite{Pluta2008}.
 ANTS also provides the inverse of those transformations denoted by the `$\dagger$' symbol.
@@
 \begin{verbatim}
-ANTS 3 -m PR[AtlasHead.nii,crAnatomical.nii,1,4] -r Gauss[0,3]  
+ANTS 3 -m CC[AtlasHead.nii,crAnatomical.nii,1,4] -r Gauss[0,3]  
 -o ElasticTest.nii.gz -i 30x20x10  -t Elast[1.5] 
 \end{verbatim}
-Here, we use the PR metric (a cross-correlation implementation) with
+Here, we use the CC metric (a cross-correlation implementation) with
 window radius 4, weight 1 and gradient step length 1.5.  The
@@
 assume nii but other file types may be used. The standard naming
-extension is Warpxvec.nii for the x component of the deformation with
+extension is Warpxvec.nii.gz for the x component of the deformation with
 similar naming for y and z. Each component is stored in a separate
@@
 image is the physical space displacement from that voxel. Note that
-the inverse mapping is stored in InverseWarpxvec.nii and provides the
+the inverse mapping is stored in InverseWarpxvec.nii.gz and provides the
 mapping in the opposite direction.  Note that an inverse -- in the 
@@
 \begin{verbatim}
-ANTS 2 -m PR[A,B,1,2] -o OUTPUT 
+ANTS 2 -m CC[A,B,1,2] -o OUTPUT 
 \end{verbatim}
@@
 ITS=" -i 100x100x30 " # 3 optimization levels 
-TSYNWITHTIME=" -t SyN[1,2,0.05]  -r Gauss[3,0.5] "
+TSYNWITHTIME=" -t SyN[0.5,5,0.01]  -r Gauss[3,0] "
 TGREEDYSYN=" -t SyN[0.25]  -r Gauss[3,0] "
-TELAST=" -t Elast[1] -r Gauss[0,3] " 
-TEXP="  -t Exp[0.5,10] -r Gauss[0,3] " 
+TELAST=" -t Elast[1] -r Gauss[0.5,3] " 
+TEXP="  -t Exp[0.5,10] -r Gauss[0.5,3] " 
 LABELGUIDED=" -m PSE[${II},${JJ},${II},${JJ},0.75,0.1,25,0,10] "
-INTMSQ=" -m MSQ[${II},${JJ},1,0.1] "
+INTMSQ=" -m MSQ[${II},${JJ},1,0] "
 INTMI=" -m MI[${II},${JJ},1,32] "
-INTPR=" -m PR[${II},${JJ},1,4] "
+INTCC=" -m CC[${II},${JJ},1,4] "
 NAME=TEST
@@
 ANTS $DIM -o $NAME $ITS  $TRAN  $INT
-INVW=" -i ${NAME}Affine.txt ${NAME}InverseWarp.nii "
-FWDW=" ${NAME}Warp.nii ${NAME}Affine.txt "
-WarpImageMultiTransform  2 ${II} a.nii -R ${JJ} $INVW
-WarpImageMultiTransform  2 ${JJ} b.nii -R ${II} $FWDW
+INVW=" -i ${NAME}Affine.txt ${NAME}InverseWarp.nii.gz "
+FWDW=" ${NAME}Warp.nii.gz ${NAME}Affine.txt "
+WarpImageMultiTransform  2 ${II} a.nii.gz -R ${JJ} $INVW
+WarpImageMultiTransform  2 ${JJ} b.nii.gz -R ${II} $FWDW
 \end{verbatim}
@@
 \begin{verbatim}
-ANTS 3 -m PR[T1template.nii,T1subject.nii,1.25,4]
-       -m PR[FAtemplate.nii,FAsubject.nii,1,4] 
+ANTS 3 -m CC[T1template.nii,T1subject.nii,1.25,4]
+       -m CC[FAtemplate.nii,FAsubject.nii,1,4] 
 -o MultiVar -i 30x30x20  -r Gauss[3,0] -t SyN[0.25]
@@
 In this case, dense features within white matter are gained by using the DT component  
-in addition to the T1 component.   The T1 PR metric, here, is weighted slightly more 
-than the FA PR metric, 1.25 vs. 1.0.  {\em The convergence criterion for 
+in addition to the T1 component.   The T1 CC metric, here, is weighted slightly more 
+than the FA CC metric, 1.25 vs. 1.0.  {\em The convergence criterion for 
 the multivariate scenario is a slave to the {\bf last} metric you 
@@
 \item  On the command line, within that directory, run the following command: 
-\verb  sh ~ \verb buildtemplate.sh ~ to get usage. 
-\item If you follow the usage, you will then call the script something like this: 
-\verb sh ~ \verb buildtemplate.sh ~ \verb 2 ~ \verb B ~ \verb tiff ~ \verb OUT ~ \verb 4 ~ . 
-\item The script should exit and ask if the parameters are ok for your problem.  
-Check the path to the programs, the fact that all the images you want to use are being 
-listed and that the template output name is ok with you.  
-Once everything is prepared, then edit the script to remove the exit call at the end of the parameter check. 
-\item  Call the script as in step 4 and then wait for the output.  The template will 
-be called something like (in this example)  \verb OUTtemplate.nii .
+\verb  bash ~ \verb buildtemplateparallel.sh -h ~ to get usage. 
+\item Following the usage, you can write a script that contains a call like this: 
+\verb buildtemplateparallel.sh ~ \verb  -d ~ \verb  2 ~ \verb -o ~ \verb  BTP ~ \verb -c ~ \verb  0 ~ \verb  B*tiff
+% \verb bash ~ \verb ~ \verb 2 ~ \verb B ~ \verb tiff ~ \verb OUT ~ \verb 4 ~ . 
+\item  Call the script in step 4 and then wait for the output.  The template will 
+be called (in this example)  \verb BTPtemplate.nii.gz .
+\item If the script does not produce the desired output, check that
+  you defined \$ANTSPATH in your path.
+You may also check and or define the path to the ANTs programs
+explicitly by editing this variable in buildtemplateparallel and antsIntroduction.
 \end{enumerate}
-The script is fairly easy to alter so that you can send the whole thing to distributed computing. 
-Typically, one would use voxbo or the qsub program available in the Sun Grid Engine.  This 
-is what is done in \verb buildtemplateparallel.sh .  
+The script offers an option ( -c 1  for qsub , -c 2 for voxbo )  so that you can send the whole thing to distributed computing. 
+Typically, one would use voxbo or the qsub program available in the Sun Grid Engine.  
 Figure~\ref{fig:template} shows the data and a result that may be recreated by the user.
@@
 \begin{verbatim}
-ANTS 3 -m PR[template.nii.gz,subject.nii.gz,1,2] -i 10x50x50x20
- -o subjectmap.nii -t SyN[0.25] -r Gauss[3,0]
+ANTS 3 -m CC[template.nii.gz,subject.nii.gz,1,2] -i 10x50x50x20
+ -o subjectmap.nii.gz -t SyN[0.25] -r Gauss[3,0]
 \end{verbatim}
@@
 \begin{verbatim}
-ANTS 3 -m PR[template.nii.gz,subject.nii.gz,1,4] -i 100x100x100x20
- -o subjectmap.nii -t SyN[0.25] -r Gauss[3,0] 
+ANTS 3 -m CC[template.nii.gz,subject.nii.gz,1,4] -i 100x100x100x20
+ -o subjectmap.nii.gz -t SyN[0.25] -r Gauss[3,0] 
 \end{verbatim}
@@
 \subsection{Basic Segmentation}
-A basic MRF segmentation algorithm is available in ANTS ImageMath.  We apply 
+An advanced MRF segmentation algorithm is available in ANTS via the
+\verb Atropos tool, which is still under development.  We apply 
 the segmentation to the example data in: ANTS/Examples/Data/r16slice.nii.  
 \begin{verbatim}
-ImageMath 2 r16slice.nii Segment r16slice.nii 4 0 
-% output = r16slice_seg.nii, r16slice_prob_0.nii, r16slice_prob_1.nii, r16slice_prob_2.nii .  
-\end{verbatim}
-The parameter \verb 4 ~indicates 3 tissues plus background are sought.  The \verb 0 ~indicates 
-that there are no prior images available to guide the segmentation.  Output is shown 
+  ThresholdImage 2 r16slice.nii mask.nii.gz 0.001 Inf
+  Atropos -d 2 -a r16slice.nii -o [r16slice_seg.nii.gz,r16slice_prob_%02d.nii.gz] \
+      -c [10,0.001] -m [0.05,1] -i kmeans[3] -x mask.nii.gz
+\end{verbatim}
+The parameter \verb kmeans[3] ~indicates 3 tissues plus background are sought.  The \verb [0.05,1] ~indicates 
+that some MRF smoothing will be applied to the labels.  Other
+options are defined in the Atropos help.  Output is shown 
 in figure~\ref{fig:seg} and -- with priors -- in figure~\ref{fig:seg2}.
@@
 \begin{verbatim}
-ImageMath 2 r16sliceprior.nii Segment r16slice.nii 4 0.5 
-   csfprior.nii wmprior.nii  gmprior.nii 
-% output = r16sliceprior_seg.nii  r16sliceprior_prob_0.nii  
-% r16sliceprior_prob_1.nii  r16sliceprior_prob_2.nii .  
-\end{verbatim}
-The parameter \verb 4 indicates 3 tissues plus background are sought.  The \verb 0.5 indicates 
+  Atropos -d 2 -a r16slice.nii -o [r16slice_seg.nii.gz,r16slice_prob_%02d.nii.gz] \
+      -c [10,0.001] -m [0.05,1] \ 
+      -i  PriorProbabilityImages[3,r16slice_prob_%02d.nii.gz,0.5] -x mask.nii.gz
+\end{verbatim}
+The \verb PriorProbabilityImages ~parameter~ \verb 3 ~indicates 3 tissues
+plus background are sought.  The \verb PriorProbabilityImages ~term \verb 0.5 ~indicates 
 we weight the priors equally as the data term and use a spatially varying set of Gaussians 
@@
 \begin{verbatim}
-ANTS 3 -m PR[tp22_s1.nii,template.nii.gz,1,4] -i 50x20x0
-   -o tp22map -t SyN[0.5] -x mask.nii.gz -r Gauss[3,0] 
+ANTS 3 -m CC[tp22_s1.nii,template.nii.gz,1,4] -i 50x20x0
+   -o tp22map -t SyN[0.25] -x mask.nii.gz -r Gauss[3,0] 
 \end{verbatim}
@@
 \begin{enumerate}
-\item Preprocess images -- bias correction, segmentation, etc.  Potentially construct an optimal template. 
+\item Preprocess images -- bias correction (e.g. via the \verb
+  N4BiasFieldCorrection program), segmentation (via \verb Atropos ), etc.  Potentially construct an optimal template. 
 \item Normalize images -- run \verb ants.sh ~ to map a population of images to a template and store the deformations. 
-\item Derive any data from the deformation that may be necessary, e.g. the log-Jacobian.
+\item Derive any data from the deformation that may be necessary,
+  e.g. the log-Jacobian, via ANTSJacobian.
 \item Apply the warp to any images one may want to analyze in template space.  
@@
 \begin{verbatim}
-1.  ImageMath 2 r64slice.nii Segment r64slice.nii 4 0
-1.  ImageMath 2 r16slice.nii Segment r16slice.nii 4 0
-2.  sh ants.sh 2 r16slice.nii r64slice.nii SPM 100x100x100
-3.  SmoothImage 2 r64slice_prob_1.nii  1. r64slice_prob_1.nii  
-3.  WarpImageMultiTransform 2 r64slice_prob_1.nii  r64slice_prob_1_norm.nii
-     -R r16slice.nii SPMWarp.nii SPMAffine.txt 
-3.  ANTSJacobian 2 SPMWarp.nii SPM 1  # take the log
-4.  ImageMath 2 SPMlogjacobianmask.nii m SPMlogjacobian.nii r64slice_prob_1_norm.nii
-5.  Repeat for a population and run statistics on log jacobians.
+# 0.  Here, we use r16 as a template and r64 as a subject T1 image. 
+# 1. Segment the input images 
+  ThresholdImage 2 r16slice.nii mask.nii.gz 0.001 Inf
+  Atropos -d 2 -a r16slice.nii -o [r16slice_seg.nii.gz,r16slice_prob_%02d.nii.gz] \
+      -c [10,0.001] -m [0.05,1] -i kmeans[3] -x mask.nii.gz
+  ThresholdImage 2 r64slice.nii mask.nii.gz 0.001 Inf
+  Atropos -d 2 -a r64slice.nii -o [r64slice_seg.nii.gz,r64slice_prob_%02d.nii.gz] \
+       -c [10,0.001] -m [0.05,1] -i kmeans[3] -x mask.nii.gz
+# 2. Compute the deformation
+  sh ants.sh 2 r16slice_prob_02.nii.gz r64slice_prob_02.nii.gz SPM 100x100x100
+# 3. Smooth the segmented image 
+  SmoothImage 2 r64slice_prob_02.nii.gz  1. sr64slice_prob_02.nii.gz  
+# 3. Warp the segmented image by the 
+  WarpImageMultiTransform 2 sr64slice_prob_02.nii.gz  sr64slice_prob_02_norm.nii.gz \ 
+     -R r16slice.nii SPMWarp.nii.gz SPMAffine.txt 
+# 4. Compute the log Jacobian from the transformation.
+  ANTSJacobian 2 SPMWarp.nii.gz SPM 1  # take the log
+  gzip SPMlogjacobian.nii 
+# 5. Multiply the log-jacobian by the gray matter probability. 
+ ImageMath 2 SPMlogjacobianmask.nii.gz m SPMlogjacobian.nii.gz sr64slice_prob_02_norm.nii.gz
+# 6.  Estimate cortical thickness --- use svn revision 684 or greater (not in official release yet). 
+ KellySlater 2 r64slice_seg.nii.gz r64slice_prob_03.nii.gz  r64slice_prob_02.nii.gz \
+    r64slice_thickness.nii.gz 0.02 50 8 0 0.5
+# 7.  Warp to template space. 
+  WarpImageMultiTransform 2 r64slice_thickness.nii.gz  r64slice_thickness_norm.nii.gz  \
+    -R r16slice.nii SPMWarp.nii.gz SPMAffine.txt 
+# 8.  Repeat steps 1-7 for a population and run statistics.
 \end{verbatim}
@@
 \section{Annotated Bibliography}
-The original statement of the symmetric normalization and template
+The {\em de facto} ANTs reference is now
+\href{http://www.ncbi.nlm.nih.gov/pubmed/20851191}{``A reproducible
+  evaluation of ANTs similarity metric performance in brain image
+  registration.'' (this is a link).}  The original statement of the symmetric normalization and template
 construction methodology was given in \cite{Avants2004} with a more 

### 2011-02-11 Temporary/itkAutumnColormapFunctor.h
Removing redundant files which are in the itk repository.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@778 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -1,77 +1,0 @@
-/*=========================================================================
-
-  Program:   Advanced Normalization Tools
-  Module:    $RCSfile: itkAutumnColormapFunctor.h,v $
-  Language:  C++
-  Date:      $Date: 2009-05-15 19:22:31 $
-  Version:   $Revision: 1.3 $
-
-  Copyright (c) ConsortiumOfANTS. All rights reserved.
-  See accompanying COPYING.txt or 
- http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
-
-     This software is distributed WITHOUT ANY WARRANTY; without even 
-     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
-     PURPOSE.  See the above copyright notices for more information.
-
-=========================================================================*/
-#ifndef __itkAutumnColormapFunctor_h
-#define __itkAutumnColormapFunctor_h
-
-#include "itkColormapFunctor.h"
-
-namespace itk {
-
-namespace Functor {  
-
-/**
- * \class AutumnColormapFunctor
- * \brief Function object which maps a scalar value into an RGB colormap value.
- *
- * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and James C. Gee
- * 
- * This code was contributed in the Insight Journal paper:
- *
- * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
- * http://www.insight-journal.org/browse/publication/285
- * http://hdl.handle.net/1926/1452
- *
- */
-template< class TScalar, class TRGBPixel >
-class ITK_EXPORT AutumnColormapFunctor
-: public ColormapFunctor<TScalar, TRGBPixel>
-{
-public:
-
-  typedef AutumnColormapFunctor                  Self;
-  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
-  typedef SmartPointer<Self>                     Pointer;
-  typedef SmartPointer<const Self>               ConstPointer;
-
-  /** Method for creation through the object factory. */
-  itkNewMacro( Self );
-
-  typedef typename Superclass::RGBPixelType      RGBPixelType;
-  typedef typename Superclass::ScalarType        ScalarType;  
-  typedef typename Superclass::RealType          RealType;
-  
-  virtual RGBPixelType operator()( const TScalar & ) const;
-
-protected:
[…]

### 2011-02-11 Temporary/itkBlueColormapFunctor.h

@@ -1,78 +1,0 @@
-/*=========================================================================
-
-  Program:   Advanced Normalization Tools
-  Module:    $RCSfile: itkBlueColormapFunctor.h,v $
-  Language:  C++
-  Date:      $Date: 2009-05-15 19:08:09 $
-  Version:   $Revision: 1.2 $
-
-  Copyright (c) ConsortiumOfANTS. All rights reserved.
-  See accompanying COPYING.txt or 
- http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
-
-     This software is distributed WITHOUT ANY WARRANTY; without even 
-     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
-     PURPOSE.  See the above copyright notices for more information.
-
-=========================================================================*/
-#ifndef __itkBlueColormapFunctor_h
-#define __itkBlueColormapFunctor_h
-
-#include "itkColormapFunctor.h"
-
-namespace itk {
-
-namespace Functor {  
-
-/**
- * \class BlueColormapFunctor
- * \brief Function object which maps a scalar value into an RGB colormap value.
- *
- * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
- * 
- * This code was contributed in the Insight Journal paper:
- *
- * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
- * http://www.insight-journal.org/browse/publication/285
- * http://hdl.handle.net/1926/1452
- *
- */
-template< class TScalar, class TRGBPixel >
-class ITK_EXPORT BlueColormapFunctor
-: public ColormapFunctor<TScalar, TRGBPixel>
-{
-public:
-
-  typedef BlueColormapFunctor                    Self;
-  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
-  typedef SmartPointer<Self>                     Pointer;
-  typedef SmartPointer<const Self>               ConstPointer;
-
-  /** Method for creation through the object factory. */
-  itkNewMacro( Self );
-
-  typedef typename Superclass::RGBPixelType      RGBPixelType;
-  typedef typename Superclass::ScalarType        ScalarType;  
-  typedef typename Superclass::RealType          RealType;
-  
-  virtual RGBPixelType operator()( const TScalar & ) const;
-
-protected:
[…]

### 2011-02-11 Temporary/itkColormapFunctor.h

@@ -1,154 +1,0 @@
-/*=========================================================================
-
-  Program:   Advanced Normalization Tools
-  Module:    $RCSfile: itkColormapFunctor.h,v $
-  Language:  C++
-  Date:      $Date: 2009-05-21 20:16:56 $
-  Version:   $Revision: 1.6 $
-
-  Copyright (c) ConsortiumOfANTS. All rights reserved.
-  See accompanying COPYING.txt or 
- http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
-
-     This software is distributed WITHOUT ANY WARRANTY; without even
-     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
-     PURPOSE.  See the above copyright notices for more information.
-
-=========================================================================*/
-#ifndef __itkColormapFunctor_h
-#define __itkColormapFunctor_h
-
-#include "itkLightObject.h"
-#include "itkNumericTraits.h"
-#include "itkRGBPixel.h"
-
-namespace itk {
-
-namespace Functor {
-
-/**
- * \class ColormapFunctor
- * \brief Function object which maps a scalar value into an RGB colormap value.
- *
- * 
- * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
- * 
- * This code was contributed in the Insight Journal paper:
- *
- * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
- * http://www.insight-journal.org/browse/publication/285
- * http://hdl.handle.net/1926/1452
- *
- */
-template< class TScalar, class TRGBPixel >
-class ITK_EXPORT ColormapFunctor : public Object
-{
-public:
-
-  typedef ColormapFunctor                          Self;
-  typedef Object                                   Superclass;
-  typedef SmartPointer<Self>                       Pointer;
-  typedef SmartPointer<const Self>                 ConstPointer;
-
-  /** Run-time type information (and related methods). */
-  itkTypeMacro(ColormapFunctor, Object);
-
-  typedef TRGBPixel                                     RGBPixelType;
-  typedef typename TRGBPixel::ComponentType             RGBComponentType;
-  typedef TScalar                                       ScalarType;
[…]

### 2011-02-11 Temporary/itkCoolColormapFunctor.h

@@ -1,79 +1,0 @@
-/*=========================================================================
-
-  Program:   Advanced Normalization Tools
-  Module:    $RCSfile: itkCoolColormapFunctor.h,v $
-  Language:  C++
-  Date:      $Date: 2009-05-15 19:08:09 $
-  Version:   $Revision: 1.2 $
-
-  Copyright (c) ConsortiumOfANTS. All rights reserved.
-  See accompanying COPYING.txt or 
- http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
-
-     This software is distributed WITHOUT ANY WARRANTY; without even 
-     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
-     PURPOSE.  See the above copyright notices for more information.
-
-=========================================================================*/
-#ifndef __itkCoolColormapFunctor_h
-#define __itkCoolColormapFunctor_h
-
-#include "itkColormapFunctor.h"
-
-namespace itk {
-
-namespace Functor {  
-
-/**
- * \class CoolColormapFunctor
- * \brief Function object which maps a scalar value into an RGB colormap value.
- *
- * 
- * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
- * 
- * This code was contributed in the Insight Journal paper:
- *
- * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
- * http://www.insight-journal.org/browse/publication/285
- * http://hdl.handle.net/1926/1452
- *
- */
-template< class TScalar, class TRGBPixel >
-class ITK_EXPORT CoolColormapFunctor
-: public ColormapFunctor<TScalar, TRGBPixel>
-{
-public:
-
-  typedef CoolColormapFunctor                    Self;
-  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
-  typedef SmartPointer<Self>                     Pointer;
-  typedef SmartPointer<const Self>               ConstPointer;
-
-  /** Method for creation through the object factory. */
-  itkNewMacro( Self );
-
-  typedef typename Superclass::RGBPixelType      RGBPixelType;
-  typedef typename Superclass::ScalarType        ScalarType;  
-  typedef typename Superclass::RealType          RealType;
-  
-  virtual RGBPixelType operator()( const TScalar & ) const;
-
-protected:
[…]

### 2011-02-11 Temporary/itkCopperColormapFunctor.h

@@ -1,79 +1,0 @@
-/*=========================================================================
-
-  Program:   Advanced Normalization Tools
-  Module:    $RCSfile: itkCopperColormapFunctor.h,v $
-  Language:  C++
-  Date:      $Date: 2009-05-15 19:08:09 $
-  Version:   $Revision: 1.2 $
-
-  Copyright (c) ConsortiumOfANTS. All rights reserved.
-  See accompanying COPYING.txt or 
- http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
-
-     This software is distributed WITHOUT ANY WARRANTY; without even 
-     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
-     PURPOSE.  See the above copyright notices for more information.
-
-=========================================================================*/
-#ifndef __itkCopperColormapFunctor_h
-#define __itkCopperColormapFunctor_h
-
-#include "itkColormapFunctor.h"
-
-namespace itk {
-
-namespace Functor {  
-
-/**
- * \class CopperColormapFunctor
- * \brief Function object which maps a scalar value into an RGB colormap value.
- *
- *
- * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
- * 
- * This code was contributed in the Insight Journal paper:
- *
- * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
- * http://www.insight-journal.org/browse/publication/285
- * http://hdl.handle.net/1926/1452
- *
- */
-template< class TScalar, class TRGBPixel >
-class ITK_EXPORT CopperColormapFunctor
-: public ColormapFunctor<TScalar, TRGBPixel>
-{
-public:
-
-  typedef CopperColormapFunctor                  Self;
-  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
-  typedef SmartPointer<Self>                     Pointer;
-  typedef SmartPointer<const Self>               ConstPointer;
-
-  /** Method for creation through the object factory. */
-  itkNewMacro( Self );
-
-  typedef typename Superclass::RGBPixelType      RGBPixelType;
-  typedef typename Superclass::ScalarType        ScalarType;  
-  typedef typename Superclass::RealType          RealType;
-  
-  virtual RGBPixelType operator()( const TScalar & ) const;
-
-protected:
[…]

### 2011-02-11 Temporary/itkCustomColormapFunctor.h

@@ -1,113 +1,0 @@
-/*=========================================================================
-
-  Program:   Advanced Normalization Tools
-  Module:    $RCSfile: itkCustomColormapFunctor.h,v $
-  Language:  C++
-  Date:      $Date: 2009-05-15 19:22:31 $
-  Version:   $Revision: 1.2 $
-
-  Copyright (c) ConsortiumOfANTS. All rights reserved.
-  See accompanying COPYING.txt or 
- http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
-
-     This software is distributed WITHOUT ANY WARRANTY; without even
-     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
-     PURPOSE.  See the above copyright notices for more information.
-
-=========================================================================*/
-#ifndef __itkCustomColormapFunctor_h
-#define __itkCustomColormapFunctor_h
-
-#include "itkColormapFunctor.h"
-
-#include <vector>
-
-namespace itk {
-
-namespace Functor {
-
-/**
- * \class CustomColormapFunctor
- * \brief Function object which maps a scalar value into an RGB colormap value.
- *
- *
- * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
- * 
- * This code was contributed in the Insight Journal paper:
- *
- * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
- * http://www.insight-journal.org/browse/publication/285
- * http://hdl.handle.net/1926/1452
- *
- */
-template< class TScalar, class TRGBPixel >
-class ITK_EXPORT CustomColormapFunctor
-: public ColormapFunctor<TScalar, TRGBPixel>
-{
-public:
-
-  typedef CustomColormapFunctor                  Self;
-  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
-  typedef SmartPointer<Self>                     Pointer;
-  typedef SmartPointer<const Self>               ConstPointer;
-
-  /** Method for creation through the object factory. */
-  itkNewMacro( Self );
-
-  typedef typename Superclass::RGBPixelType      RGBPixelType;
-  typedef typename Superclass::ScalarType        ScalarType;
-  typedef typename Superclass::RealType          RealType;
-
[…]

### 2011-02-11 Temporary/itkGreenColormapFunctor.h

@@ -1,78 +1,0 @@
-/*=========================================================================
-
-  Program:   Advanced Normalization Tools
-  Module:    $RCSfile: itkGreenColormapFunctor.h,v $
-  Language:  C++
-  Date:      $Date: 2009-05-15 19:08:09 $
-  Version:   $Revision: 1.2 $
-
-  Copyright (c) ConsortiumOfANTS. All rights reserved.
-  See accompanying COPYING.txt or 
- http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
-
-     This software is distributed WITHOUT ANY WARRANTY; without even 
-     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
-     PURPOSE.  See the above copyright notices for more information.
-
-=========================================================================*/
-#ifndef __itkGreenColormapFunctor_h
-#define __itkGreenColormapFunctor_h
-
-#include "itkColormapFunctor.h"
-
-namespace itk {
-
-namespace Functor {  
-
-/**
- * \class GreenColormapFunctor
- * \brief Function object which maps a scalar value into an RGB colormap value.
- *
- * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
- * 
- * This code was contributed in the Insight Journal paper:
- *
- * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
- * http://www.insight-journal.org/browse/publication/285
- * http://hdl.handle.net/1926/1452
- *
- */
-template< class TScalar, class TRGBPixel >
-class ITK_EXPORT GreenColormapFunctor
-: public ColormapFunctor<TScalar, TRGBPixel>
-{
-public:
-
-  typedef GreenColormapFunctor                   Self;
-  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
-  typedef SmartPointer<Self>                     Pointer;
-  typedef SmartPointer<const Self>               ConstPointer;
-
-  /** Method for creation through the object factory. */
-  itkNewMacro( Self );
-
-  typedef typename Superclass::RGBPixelType      RGBPixelType;
-  typedef typename Superclass::ScalarType        ScalarType;  
-  typedef typename Superclass::RealType          RealType;
-  
-  virtual RGBPixelType operator()( const TScalar & ) const;
-
-protected:
[…]

### 2011-02-11 Temporary/itkGreyColormapFunctor.h

@@ -1,78 +1,0 @@
-/*=========================================================================
-
-  Program:   Advanced Normalization Tools
-  Module:    $RCSfile: itkGreyColormapFunctor.h,v $
-  Language:  C++
-  Date:      $Date: 2009-05-15 19:22:31 $
-  Version:   $Revision: 1.2 $
-
-  Copyright (c) ConsortiumOfANTS. All rights reserved.
-  See accompanying COPYING.txt or 
- http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
-
-     This software is distributed WITHOUT ANY WARRANTY; without even 
-     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
-     PURPOSE.  See the above copyright notices for more information.
-
-=========================================================================*/
-#ifndef __itkGreyColormapFunctor_h
-#define __itkGreyColormapFunctor_h
-
-#include "itkColormapFunctor.h"
-
-namespace itk {
-
-namespace Functor {  
-
-/**
- * \class GreyColormapFunctor
- * \brief Function object which maps a scalar value into an RGB colormap value.
- *
- * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
- * 
- * This code was contributed in the Insight Journal paper:
- *
- * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
- * http://www.insight-journal.org/browse/publication/285
- * http://hdl.handle.net/1926/1452
- *
- */
-template< class TScalar, class TRGBPixel >
-class ITK_EXPORT GreyColormapFunctor
-: public ColormapFunctor<TScalar, TRGBPixel>
-{
-public:
-
-  typedef GreyColormapFunctor                    Self;
-  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
-  typedef SmartPointer<Self>                     Pointer;
-  typedef SmartPointer<const Self>               ConstPointer;
-
-  /** Method for creation through the object factory. */
-  itkNewMacro( Self );
-
-  typedef typename Superclass::RGBPixelType      RGBPixelType;
-  typedef typename Superclass::ScalarType        ScalarType;  
-  typedef typename Superclass::RealType          RealType;
-  
-  virtual RGBPixelType operator()( const TScalar & ) const;
-
-protected:
[…]

### 2011-02-11 Temporary/itkHSVColormapFunctor.h

@@ -1,78 +1,0 @@
-/*=========================================================================
-
-  Program:   Advanced Normalization Tools
-  Module:    $RCSfile: itkHSVColormapFunctor.h,v $
-  Language:  C++
-  Date:      $Date: 2009-05-15 19:22:31 $
-  Version:   $Revision: 1.2 $
-
-  Copyright (c) ConsortiumOfANTS. All rights reserved.
-  See accompanying COPYING.txt or 
- http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
-
-     This software is distributed WITHOUT ANY WARRANTY; without even 
-     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
-     PURPOSE.  See the above copyright notices for more information.
-
-=========================================================================*/
-#ifndef __itkHSVColormapFunctor_h
-#define __itkHSVColormapFunctor_h
-
-#include "itkColormapFunctor.h"
-
-namespace itk {
-
-namespace Functor {  
-
-/**
- * \class HSVColormapFunctor
- * \brief Function object which maps a scalar value into an RGB colormap value.
- *
- * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
- * 
- * This code was contributed in the Insight Journal paper:
- *
- * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
- * http://www.insight-journal.org/browse/publication/285
- * http://hdl.handle.net/1926/1452
- *
- */
-template< class TScalar, class TRGBPixel >
-class ITK_EXPORT HSVColormapFunctor
-: public ColormapFunctor<TScalar, TRGBPixel>
-{
-public:
-
-  typedef HSVColormapFunctor                     Self;
-  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
-  typedef SmartPointer<Self>                     Pointer;
-  typedef SmartPointer<const Self>               ConstPointer;
-
-  /** Method for creation through the object factory. */
-  itkNewMacro( Self );
-
-  typedef typename Superclass::RGBPixelType      RGBPixelType;
-  typedef typename Superclass::ScalarType        ScalarType;  
-  typedef typename Superclass::RealType          RealType;
-  
-  virtual RGBPixelType operator()( const TScalar & ) const;
-
-protected:
[…]

### 2011-02-11 Temporary/itkHotColormapFunctor.h

@@ -1,78 +1,0 @@
-/*=========================================================================
-
-  Program:   Advanced Normalization Tools
-  Module:    $RCSfile: itkHotColormapFunctor.h,v $
-  Language:  C++
-  Date:      $Date: 2009-05-15 19:22:31 $
-  Version:   $Revision: 1.2 $
-
-  Copyright (c) ConsortiumOfANTS. All rights reserved.
-  See accompanying COPYING.txt or 
- http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
-
-     This software is distributed WITHOUT ANY WARRANTY; without even 
-     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
-     PURPOSE.  See the above copyright notices for more information.
-
-=========================================================================*/
-#ifndef __itkHotColormapFunctor_h
-#define __itkHotColormapFunctor_h
-
-#include "itkColormapFunctor.h"
-
-namespace itk {
-
-namespace Functor {  
-
-/**
- * \class HotColormapFunctor
- * \brief Function object which maps a scalar value into an RGB colormap value.
- *
- * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
- * 
- * This code was contributed in the Insight Journal paper:
- *
- * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
- * http://www.insight-journal.org/browse/publication/285
- * http://hdl.handle.net/1926/1452
- *
- */
-template< class TScalar, class TRGBPixel >
-class ITK_EXPORT HotColormapFunctor
-: public ColormapFunctor<TScalar, TRGBPixel>
-{
-public:
-
-  typedef HotColormapFunctor                     Self;
-  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
-  typedef SmartPointer<Self>                     Pointer;
-  typedef SmartPointer<const Self>               ConstPointer;
-
-  /** Method for creation through the object factory. */
-  itkNewMacro( Self );
-
-  typedef typename Superclass::RGBPixelType      RGBPixelType;
-  typedef typename Superclass::ScalarType        ScalarType;  
-  typedef typename Superclass::RealType          RealType;
-  
-  virtual RGBPixelType operator()( const TScalar & ) const;
-
-protected:
[…]

### 2011-02-11 Temporary/itkJetColormapFunctor.h

@@ -1,78 +1,0 @@
-/*=========================================================================
-
-  Program:   Advanced Normalization Tools
-  Module:    $RCSfile: itkJetColormapFunctor.h,v $
-  Language:  C++
-  Date:      $Date: 2009-05-15 19:22:31 $
-  Version:   $Revision: 1.2 $
-
-  Copyright (c) ConsortiumOfANTS. All rights reserved.
-  See accompanying COPYING.txt or 
- http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
-
-     This software is distributed WITHOUT ANY WARRANTY; without even 
-     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
-     PURPOSE.  See the above copyright notices for more information.
-
-=========================================================================*/
-#ifndef __itkJetColormapFunctor_h
-#define __itkJetColormapFunctor_h
-
-#include "itkColormapFunctor.h"
-
-namespace itk {
-
-namespace Functor {  
-
-/**
- * \class JetColormapFunctor
- * \brief Function object which maps a scalar value into an RGB colormap value.
- *
- * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
- * 
- * This code was contributed in the Insight Journal paper:
- *
- * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
- * http://www.insight-journal.org/browse/publication/285
- * http://hdl.handle.net/1926/1452
- *
- */
-template< class TScalar, class TRGBPixel >
-class ITK_EXPORT JetColormapFunctor
-: public ColormapFunctor<TScalar, TRGBPixel>
-{
-public:
-
-  typedef JetColormapFunctor                     Self;
-  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
-  typedef SmartPointer<Self>                     Pointer;
-  typedef SmartPointer<const Self>               ConstPointer;
-
-  /** Method for creation through the object factory. */
-  itkNewMacro( Self );
-
-  typedef typename Superclass::RGBPixelType      RGBPixelType;
-  typedef typename Superclass::ScalarType        ScalarType;  
-  typedef typename Superclass::RealType          RealType;
-  
-  virtual RGBPixelType operator()( const TScalar & ) const;
-
-protected:
[…]

### 2011-02-11 Temporary/itkOverUnderColormapFunctor.h

@@ -1,78 +1,0 @@
-/*=========================================================================
-
-  Program:   Advanced Normalization Tools
-  Module:    $RCSfile: itkOverUnderColormapFunctor.h,v $
-  Language:  C++
-  Date:      $Date: 2009-05-15 19:22:31 $
-  Version:   $Revision: 1.2 $
-
-  Copyright (c) ConsortiumOfANTS. All rights reserved.
-  See accompanying COPYING.txt or 
- http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
-
-     This software is distributed WITHOUT ANY WARRANTY; without even
-     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
-     PURPOSE.  See the above copyright notices for more information.
-
-=========================================================================*/
-#ifndef __itkOverUnderColormapFunctor_h
-#define __itkOverUnderColormapFunctor_h
-
-#include "itkColormapFunctor.h"
-
-namespace itk {
-
-namespace Functor {
-
-/**
- * \class OverUnderColormapFunctor
- * \brief Function object which maps a scalar value into an RGB colormap value.
- *
- * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
- * 
- * This code was contributed in the Insight Journal paper:
- *
- * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
- * http://www.insight-journal.org/browse/publication/285
- * http://hdl.handle.net/1926/1452
- *
- */
-template< class TScalar, class TRGBPixel >
-class ITK_EXPORT OverUnderColormapFunctor
-: public ColormapFunctor<TScalar, TRGBPixel>
-{
-public:
-
-  typedef OverUnderColormapFunctor               Self;
-  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
-  typedef SmartPointer<Self>                     Pointer;
-  typedef SmartPointer<const Self>               ConstPointer;
-
-  /** Method for creation through the object factory. */
-  itkNewMacro( Self );
-
-  typedef typename Superclass::RGBPixelType      RGBPixelType;
-  typedef typename Superclass::ScalarType        ScalarType;
-  typedef typename Superclass::RealType          RealType;
-
-  virtual RGBPixelType operator()( const TScalar & ) const;
-
[…]

### 2011-02-11 Temporary/itkRedColormapFunctor.h

@@ -1,78 +1,0 @@
-/*=========================================================================
-
-  Program:   Advanced Normalization Tools
-  Module:    $RCSfile: itkRedColormapFunctor.h,v $
-  Language:  C++
-  Date:      $Date: 2009-05-15 19:22:31 $
-  Version:   $Revision: 1.2 $
-
-  Copyright (c) ConsortiumOfANTS. All rights reserved.
-  See accompanying COPYING.txt or 
- http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
-
-     This software is distributed WITHOUT ANY WARRANTY; without even 
-     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
-     PURPOSE.  See the above copyright notices for more information.
-
-=========================================================================*/
-#ifndef __itkRedColormapFunctor_h
-#define __itkRedColormapFunctor_h
-
-#include "itkColormapFunctor.h"
-
-namespace itk {
-
-namespace Functor {  
-
-/**
- * \class RedColormapFunctor
- * \brief Function object which maps a scalar value into an RGB colormap value.
- *
- * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
- * 
- * This code was contributed in the Insight Journal paper:
- *
- * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
- * http://www.insight-journal.org/browse/publication/285
- * http://hdl.handle.net/1926/1452
- *
- */
-template< class TScalar, class TRGBPixel >
-class ITK_EXPORT RedColormapFunctor
-: public ColormapFunctor<TScalar, TRGBPixel>
-{
-public:
-
-  typedef RedColormapFunctor                     Self;
-  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
-  typedef SmartPointer<Self>                     Pointer;
-  typedef SmartPointer<const Self>               ConstPointer;
-
-  /** Method for creation through the object factory. */
-  itkNewMacro( Self );
-
-  typedef typename Superclass::RGBPixelType      RGBPixelType;
-  typedef typename Superclass::ScalarType        ScalarType;  
-  typedef typename Superclass::RealType          RealType;
-  
-  virtual RGBPixelType operator()( const TScalar & ) const;
-
-protected:
[…]

### 2011-02-11 Temporary/itkScalarToRGBColormapImageFilter.h

@@ -1,139 +1,0 @@
-/*=========================================================================
-
-  Program:   Advanced Normalization Tools
-  Module:    $RCSfile: itkScalarToRGBColormapImageFilter.h,v $
-  Language:  C++
-  Date:      $Date: 2009-05-15 12:51:13 $
-  Version:   $Revision: 1.2 $
-
-  Copyright (c) ConsortiumOfANTS. All rights reserved.
-  See accompanying COPYING.txt or 
- http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
-
-     This software is distributed WITHOUT ANY WARRANTY; without even
-     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
-     PURPOSE.  See the above copyright notices for more information.
-
-=========================================================================*/
-#ifndef __itkScalarToRGBColormapImageFilter_h
-#define __itkScalarToRGBColormapImageFilter_h
-
-#include "itkImageToImageFilter.h"
-
-#include "itkColormapFunctor.h"
-
-namespace itk
-{
-
-/** \class ScalarToRGBColormapImageFilter
- * \brief Implements pixel-wise intensity->rgb mapping operation on one image.
- *
- * This class is parameterized over the type of the input image and
- * the type of the output image.
- *
- * ScalarToRGBColormapImageFilter
- *
- * \sa BinaryFunctorImageFilter TernaryFunctorImageFilter
- *
- * \ingroup   IntensityImageFilters     Multithreaded
- */
-template <class TInputImage, class TOutputImage>
-class ITK_EXPORT ScalarToRGBColormapImageFilter
-: public ImageToImageFilter<TInputImage, TOutputImage>
-{
-public:
-  /** Standard class typedefs. */
-  typedef ScalarToRGBColormapImageFilter                 Self;
-  typedef ImageToImageFilter<TInputImage, TOutputImage>  Superclass;
-  typedef SmartPointer<Self>                             Pointer;
-  typedef SmartPointer<const Self>                       ConstPointer;
-
-  /** Method for creation through the object factory. */
-  itkNewMacro( Self );
-
-  /** Run-time type information (and related methods). */
-  itkTypeMacro( ScalarToRGBColormapImageFilter, ImageToImageFilter );
-
-  /** Some typedefs. */
-  typedef TInputImage                                    InputImageType;
-  typedef typename InputImageType::ConstPointer          InputImagePointer;
[…]

### 2011-02-11 Temporary/itkSpringColormapFunctor.h

@@ -1,78 +1,0 @@
-/*=========================================================================
-
-  Program:   Advanced Normalization Tools
-  Module:    $RCSfile: itkSpringColormapFunctor.h,v $
-  Language:  C++
-  Date:      $Date: 2009-05-15 19:22:31 $
-  Version:   $Revision: 1.2 $
-
-  Copyright (c) ConsortiumOfANTS. All rights reserved.
-  See accompanying COPYING.txt or 
- http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
-
-     This software is distributed WITHOUT ANY WARRANTY; without even 
-     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
-     PURPOSE.  See the above copyright notices for more information.
-
-=========================================================================*/
-#ifndef __itkSpringColormapFunctor_h
-#define __itkSpringColormapFunctor_h
-
-#include "itkColormapFunctor.h"
-
-namespace itk {
-
-namespace Functor {  
-
-/**
- * \class SpringColormapFunctor
- * \brief Function object which maps a scalar value into an RGB colormap value.
- *
- * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
- * 
- * This code was contributed in the Insight Journal paper:
- *
- * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
- * http://www.insight-journal.org/browse/publication/285
- * http://hdl.handle.net/1926/1452
- *
- */
-template< class TScalar, class TRGBPixel >
-class ITK_EXPORT SpringColormapFunctor
-: public ColormapFunctor<TScalar, TRGBPixel>
-{
-public:
-
-  typedef SpringColormapFunctor                  Self;
-  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
-  typedef SmartPointer<Self>                     Pointer;
-  typedef SmartPointer<const Self>               ConstPointer;
-
-  /** Method for creation through the object factory. */
-  itkNewMacro( Self );
-
-  typedef typename Superclass::RGBPixelType      RGBPixelType;
-  typedef typename Superclass::ScalarType        ScalarType;  
-  typedef typename Superclass::RealType          RealType;
-  
-  virtual RGBPixelType operator()( const TScalar & ) const;
-
-protected:
[…]

### 2011-02-11 Temporary/itkSummerColormapFunctor.h

@@ -1,78 +1,0 @@
-/*=========================================================================
-
-  Program:   Advanced Normalization Tools
-  Module:    $RCSfile: itkSummerColormapFunctor.h,v $
-  Language:  C++
-  Date:      $Date: 2009-05-15 19:22:31 $
-  Version:   $Revision: 1.2 $
-
-  Copyright (c) ConsortiumOfANTS. All rights reserved.
-  See accompanying COPYING.txt or 
- http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
-
-     This software is distributed WITHOUT ANY WARRANTY; without even 
-     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
-     PURPOSE.  See the above copyright notices for more information.
-
-=========================================================================*/
-#ifndef __itkSummerColormapFunctor_h
-#define __itkSummerColormapFunctor_h
-
-#include "itkColormapFunctor.h"
-
-namespace itk {
-
-namespace Functor {  
-
-/**
- * \class SummerColormapFunctor
- * \brief Function object which maps a scalar value into an RGB colormap value.
- *
- * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
- * 
- * This code was contributed in the Insight Journal paper:
- *
- * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
- * http://www.insight-journal.org/browse/publication/285
- * http://hdl.handle.net/1926/1452
- *
- */
-template< class TScalar, class TRGBPixel >
-class ITK_EXPORT SummerColormapFunctor
-: public ColormapFunctor<TScalar, TRGBPixel>
-{
-public:
-
-  typedef SummerColormapFunctor                  Self;
-  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
-  typedef SmartPointer<Self>                     Pointer;
-  typedef SmartPointer<const Self>               ConstPointer;
-
-  /** Method for creation through the object factory. */
-  itkNewMacro( Self );
-
-  typedef typename Superclass::RGBPixelType      RGBPixelType;
-  typedef typename Superclass::ScalarType        ScalarType;  
-  typedef typename Superclass::RealType          RealType;
-  
-  virtual RGBPixelType operator()( const TScalar & ) const;
-
-protected:
[…]

### 2011-02-11 Temporary/itkWinterColormapFunctor.h

@@ -1,78 +1,0 @@
-/*=========================================================================
-
-  Program:   Advanced Normalization Tools
-  Module:    $RCSfile: itkWinterColormapFunctor.h,v $
-  Language:  C++
-  Date:      $Date: 2009-05-15 19:22:31 $
-  Version:   $Revision: 1.2 $
-
-  Copyright (c) ConsortiumOfANTS. All rights reserved.
-  See accompanying COPYING.txt or 
- http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
-
-     This software is distributed WITHOUT ANY WARRANTY; without even 
-     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
-     PURPOSE.  See the above copyright notices for more information.
-
-=========================================================================*/
-#ifndef __itkWinterColormapFunctor_h
-#define __itkWinterColormapFunctor_h
-
-#include "itkColormapFunctor.h"
-
-namespace itk {
-
-namespace Functor {  
-
-/**
- * \class WinterColormapFunctor
- * \brief Function object which maps a scalar value into an RGB colormap value.
- *
- * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
- * 
- * This code was contributed in the Insight Journal paper:
- *
- * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
- * http://www.insight-journal.org/browse/publication/285
- * http://hdl.handle.net/1926/1452
- *
- */
-template< class TScalar, class TRGBPixel >
-class ITK_EXPORT WinterColormapFunctor
-: public ColormapFunctor<TScalar, TRGBPixel>
-{
-public:
-
-  typedef WinterColormapFunctor                  Self;
-  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
-  typedef SmartPointer<Self>                     Pointer;
-  typedef SmartPointer<const Self>               ConstPointer;
-
-  /** Method for creation through the object factory. */
-  itkNewMacro( Self );
-
-  typedef typename Superclass::RGBPixelType      RGBPixelType;
-  typedef typename Superclass::ScalarType        ScalarType;  
-  typedef typename Superclass::RealType          RealType;
-  
-  virtual RGBPixelType operator()( const TScalar & ) const;
-
-protected:
[…]

### 2011-03-18 Tensor/TensorFunctions.h
BUG: fixing bad bug that set negative eigenvalues to positive one even when using log-tensors
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@807 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -125,9 +125,7 @@ TensorType TensorLogAndExp( TensorType dtv, bool takelog , bool success=true)
   double e1 = D(0,0);
   double e2 = D(1,1);
   double e3 = D(2,2);
-  if ( e1 < 0 ) e1=e2;
-  if ( e3 < 0 ) e3=e2;
-  //float peigeps=1.e-12;   
+  // float peigeps=1.e-12;   
 
   if ( fabs(e3) < eps ) { success=false; std::cout << "-4" << std::flush; return dtv; }
   
@@ -135,6 +133,8 @@ TensorType TensorLogAndExp( TensorType dtv, bool takelog , bool success=true)
   eigmat.Fill(0);
   if (takelog)
 	{
+          if ( e1 < 0 ) e1=e2;
+          if ( e3 < 0 ) e3=e2;
 	  eigmat(0,0)=log(fabs(e1));
 	  eigmat(1,1)=log(fabs(e2));
 	  eigmat(2,2)=log(fabs(e3));
@@ -219,8 +219,6 @@ float  GetMetricTensorCost(  TVectorType dpath,  TTensorType dtv , unsigned int 
   double e1 = (eig.D(0,0));
   double e2 = (eig.D(1,1));
   double e3 = (eig.D(2,2));
-  if ( e1 < 0 ) e1=e2;
-  if ( e3 < 0 ) e3=e2;
   double etot=e1+e2+e3;
   if (etot==0) etot=1;
  
@@ -258,8 +256,6 @@ TVectorType ChangeTensorByVector(  TVectorType dpath,  TTensorType dtv, float ep
   double e3 = (eig.D(0,0));
   double e2 = (eig.D(1,1));
   double e1 = (eig.D(2,2));  
-  if ( e1 < 0 ) e1=e2;
-  if ( e3 < 0 ) e3=e2;
   double etot=e1+e2+e3;
   if (etot==0) etot=1;
  
@@ -369,8 +365,6 @@ float  GetTensorADC( TTensorType dtv,  unsigned int opt = 0)
   double e1 = (eig.D(0,0));
   double e2 = (eig.D(1,1));
   double e3 = (eig.D(2,2));
-  if ( e1 < 0 ) e1=e2;
-  if ( e3 < 0 ) e3=e2;
   if (opt <= 1 )  return (e1+e1+e3)/3.0;
   //  else if (opt == 4 ) return e2; 
[…]

### 2011-03-20 Tensor/TensorFunctions.h
ENH:  Added separate calculation of FA numerator and denominator.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@808 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -23,7 +23,7 @@ MatrixType Vector2Matrix( TensorType dtv )
   dtm(1,1) = dtv[3];
   dtm(1,2) = dtm(2,1) = dtv[4];
   dtm(2,2) = dtv[5];
-  
+
   return dtm;
 }
 
@@ -42,14 +42,14 @@ template<class TensorType, class MatrixType>
 TensorType Matrix2Vector( MatrixType dtm )
 {
   TensorType dtv;
-  
+
   dtv[0] = dtm(0,0);
   dtv[1] = dtm(0,1);
   dtv[2] = dtm(0,2);
   dtv[3] = dtm(1,1);
   dtv[4] = dtm(1,2);
   dtv[5] = dtm(2,2);
-  
+
   return dtv;
 }
 
@@ -58,7 +58,7 @@ void EigenAnalysis(TensorType dtv,  MatrixType &evals, MatrixType &evecs)
 {
   MatrixType dtm = Vector2Matrix<TensorType, MatrixType>(dtv);
   itk::DecomposeTensorFunction2<MatrixType,typename MatrixType::ValueType,MatrixType> decomposer;
-  decomposer.EvaluateSymmetricEigenDecomposition( dtm, evals, evecs ); 
+  decomposer.EvaluateSymmetricEigenDecomposition( dtm, evals, evecs );
 }
 
 template <class TensorType, class VectorType>
@@ -77,27 +77,27 @@ float DiffusionCoefficient( TensorType dtv, VectorType direction, bool normalize
   tangentmat(0,0) = direction[0];
   tangentmat(1,0) = direction[1];
   tangentmat(2,0) = direction[2];
-  
+
   vnl_matrix<double> fddmat = tangentmat.transpose() * tensor * tangentmat;
   float fdd = (float) fddmat(0,0);
 
   if (normalized > 0) {
 	  fdd = fdd/(tensor[0][0]+tensor[1][1]+tensor[2][2]);
 	}
-  
+
   return fdd;
 }
 
 template <class TensorType>
-TensorType TensorLogAndExp( TensorType dtv, bool takelog , bool success=true)        
+TensorType TensorLogAndExp( TensorType dtv, bool takelog , bool success=true)
 {
 
   float eps=1.e-12,mag=0;
-  for (unsigned int jj=0; jj<6; jj++) 
[…]

### 2011-03-20 ImageRegistration/itkExpectationBasedPointSetRegistrationFunction.h
ENH: minor changes for modularized itk compatibility
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@812 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -275,7 +275,7 @@ private:
   mutable double                  m_SumOfSquaredChange;
 
   /** Mutex lock to protect modification to metric. */
-  mutable SimpleFastMutexLock     m_MetricCalculationLock;
+  //  mutable SimpleFastMutexLock     m_MetricCalculationLock;
 
   DeformationFieldTypePointer m_DerivativeFixedField;
   DeformationFieldTypePointer m_DerivativeMovingField;

### 2011-03-20 Documentation/Tex/ants.tex
ENH: documentation on metric weights was unclear.  this should help some.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@813 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 
-\usepackage[dvips]{graphicx}
+% \usepackage[dvips]{graphicx}
 %\usepackage{listing}						
@@
 \subsubsection{Deformation Field File}
-ANTS writes the deformation field using one file for the deformation in each dimension. 
-For example, if the deformation field is named ABCD.nii.gz and exists in three dimensions, the corresponding files are:
-ABCDxvec.nii.gz, 
-ABCDyvec.nii.gz, 
-ABCDzvec.nii.gz (for 3D).  Note that ANTS programs {\em assume} the $*$vec part of the name will exist.  Thus, 
-one refers to the warp file as  ABCD.nii.gz , dropping the $*$vec from the name. 
-The displacement for each voxel is defined in physical image space.  If you want to 
+ANTS writes the deformation field into a multi-component image (latest
+release).  Thus, 
+one refers to the warp file as ABCD.nii.gz which, in 3D, contains 3 components.  If you want to 
 see the details of how a warp is achieved, look at the code in itkWarpImageMultiTransform.txx. 
@@
 Warping deformation field: 
-RegA2BWarpxvec.nii.gz, RegA2BWarpyvec.nii.gz, RegA2BWarpzvec.nii.gz
+RegA2BWarp.nii.gz.
 Inverse warping deformation field:
-RegA2BInverseWarpxvec.nii.gz, RegA2BInverseWarpyvec.nii.gz, RegA2BInverseWarpzvec.nii.gz.  
+RegA2BInverseWarp.nii.gz.  
 To apply the affine transform and the deformation field to the moving image, use WarpImageMultiTransform. 
@@
 transformations (InvId), to compute the fractional anisotropy of an
-image (TensorFA) and to perform Markov Random Field segmentation with
-or without priors (Segment).   Many other operations are available.
+image (TensorFA) and to perform many other feats of image processing strength.   
 Like all other ANTS programs, one may call ImageMath from the command
@@
 combinations of basic ANTS functions.  These scripts are called as
-\verb sh ~ \verb antsscriptname.sh ~ and provide usage when called
-from the command line.  For instance, try \verb sh ~ \verb ants.sh .
+\verb bash ~ \verb antsscriptname.sh ~ and provide usage when called
+from the command line.  For instance, try \verb bash ~ \verb ants.sh .
 Additionally, if you run ants in tcsh shells, you will need to call it
@@
 \end{verbatim}
-because tcsh does no interpret the brackets as intended.  Using the
-bash shell script interface also avoids this.  The key change involves using the slash to force tcsh to 
-interpet the bracket literally. 
+because tcsh does not interpret the brackets as intended.  Using the
+bash shell script interface also avoids this.  The key change involves using the slash to force tcsh to interpet the bracket literally. 
 \newpage
@@
   ImageMath ImageDimension 
-where ImageDimenion is 2 or 3.}
+where ImageDimenion is 2, 3 or 4.}
 \end{minipage}}
@@
 ANTS uses the nifti standard defintions of an image space.  So, the
-image orientation (direction), origin and spacing are important!  You
-may ``check'' for consistency between two image header definitions by
-using: ImageMath 3 Image2repaired.nii.gz CompareHeadersAndImages
-Image1.nii.gz Image2.nii.gz \newline This will compare Image2 to
-Image1 and ``fix'' the header of Image2, writing out to
-Image2repaired.
+image orientation (direction), origin and spacing are important!  The
+program 'PrintHeader' will give you information about your image's header.
 
@@
 chosen here -- \verb SyN[0.25] -- may be replaced with other 
-diffeomorphic transformation models.  The most general is global-in-time SyN via \verb SyN[0.25,2,0.01]   , 
+diffeomorphic transformation models.  The step-size (0.25) impacts
+accuracy.  Smaller is generally more accurate but takes more time to
+compute and may not capture as much deformation if the optimization is
+caught in a local minimum.  The most general is global-in-time SyN via \verb SyN[0.25,2,0.01]   , 
 where the time step for integration is 0.01 (lower is more accurate).  A fast approximate 
@@
 
-\subsection{ANTS Similarity Terms}
+\subsection{ANTS Similarity Terms and Other Functions}
 Here is a script that will let you experiment with different similarity term 
@@
 \begin{verbatim}
-II=r16slice.nii # change to your own images 
-JJ=r64slice.nii 
-DIM=2 # two-dimensional images 
+# The user should define two images.  here we use ANTS reference data
+II=r16slice.nii.gz 
+JJ=r64slice.nii.gz
+# 
+OUTPUTNAME=OUT  # define the output prefix 
 ITS=" -i 100x100x30 " # 3 optimization levels
-TSYNWITHTIME=" -t SyN[0.5,5,0.01]  -r Gauss[3,0] "
-TGREEDYSYN=" -t SyN[0.25]  -r Gauss[3,0] "
-TELAST=" -t Elast[1] -r Gauss[0.5,3] " 
-TEXP="  -t Exp[0.5,10] -r Gauss[0.5,3] " 
-LABELGUIDED=" -m PSE[${II},${JJ},${II},${JJ},0.75,0.1,25,0,10] "
+
+# different transformation models you can choose
+TSYNWITHTIME=" -t SyN[0.25,5,0.01] -r Gauss[3,0] " # spatiotemporal (full) diffeomorphism
+TGREEDYSYN=" -t SyN[0.15] -r Gauss[3,0] "          # fast symmetric normalization
+TELAST=" -t Elast[1] -r Gauss[0.5,3] "             # elastic 
+TEXP=" -t Exp[0.5,10] -r Gauss[0.5,3] "            # exponential
+
+# different metric choices for the user 
 INTMSQ=" -m MSQ[${II},${JJ},1,0] "
@@
 INTCC=" -m CC[${II},${JJ},1,4] " 
-NAME=TEST
-INT=$INTPR
-TRAN=$TSYN
-ANTS $DIM -o $NAME $ITS  $TRAN  $INT
-INVW=" -i ${NAME}Affine.txt ${NAME}InverseWarp.nii.gz "
-FWDW=" ${NAME}Warp.nii.gz ${NAME}Affine.txt "
-WarpImageMultiTransform  2 ${II} a.nii.gz -R ${JJ} $INVW
-WarpImageMultiTransform  2 ${JJ} b.nii.gz -R ${II} $FWDW
+#
+#
+# these are the forward and backward warps. 
+#
+INVW=" -i ${OUTPUTNAME}Affine.txt ${OUTPUTNAME}InverseWarp.nii.gz " 
+FWDW=" ${OUTPUTNAME}Warp.nii.gz ${OUTPUTNAME}Affine.txt " 
+#
+INT=$INTCC  # choose a  metric ( here , cross-correlation )
+TRAN=$TGREEDYSYN  # choose a transformation 
+if [[ $OPT == "Reg" ]] ; then
+
+# run the registration
+${ANTSPATH}/ANTS $DIM -o $OUTPUTNAME $ITS $TRAN $INT 
+# this is how you apply the output transformation 
+${ANTSPATH}/WarpImageMultiTransform $DIM ${II} ${OUTPUTNAME}IItoJJ.nii.gz -R ${JJ} $INVW 
+${ANTSPATH}/WarpImageMultiTransform $DIM ${JJ} ${OUTPUTNAME}JJtoII.nii.gz -R ${II} $FWDW
+
+elif [[ $OPT == "RegSeg" ]] ; then
+
+# try something different -- segment the images and use them in ANTS 
+# to run multivariate registration.
+IISEG=${OUTPUTNAME}A_seg.nii.gz
+JJSEG=${OUTPUTNAME}B_seg.nii.gz
+#
+# get masks for atropos 
+#
+${ANTSPATH}ThresholdImage $DIM $II $IISEG 1 1.e9
+${ANTSPATH}ThresholdImage $DIM $JJ $JJSEG 1 1.e9
+AtroposParams=" -d $DIM -m [0.1,1x1] -c [5,0] -i kmeans[3] "
+${ANTSPATH}/Atropos $AtroposParams -a $II -o $IISEG  -x $IISEG 
+${ANTSPATH}/Atropos $AtroposParams -a $JJ -o $JJSEG  -x $JJSEG 
+# compute some segmentations and use them in labelguided mapping 
+LABELGUIDED=" -m PSE[${II},${JJ},${IISEG},${JJSEG},0.75,0.1,25,0,10] "  
+#
+${ANTSPATH}/ANTS $DIM -o ${OUTPUTNAME} $ITS $TRAN $INT $LABELGUIDED 
+${ANTSPATH}/WarpImageMultiTransform $DIM ${II} ${OUTPUTNAME}IItoJJ.nii.gz -R ${JJ} $INVW 
+${ANTSPATH}/WarpImageMultiTransform $DIM ${JJ} ${OUTPUTNAME}JJtoII.nii.gz -R ${II} $FWDW
+
+elif [[ $OPT == "BTP" ]] ; then 
+#
+# now build a template from the two input images .
+# in a real application, you would modify the call to btp 
+# such that it's more appropriate for your data.  
+# call buildtemplateparallel.sh -h to get usage.
+# 
+${ANTSPATH}buildtemplateparallel.sh -d 2  -o ${OUTPUTNAME}BTP -c 0 $II $JJ
+fi 
 \end{verbatim}
@@
 of the change in the optimization objective is negative or very small.
-{\em Note: ANTS does not like spaces between brackets in the call to the metrics.}
 
 \subsection{Choosing a Metric}
-\input{metric.tex}
+ANTS supports both volumetric registration and point set registration.
+The image / point set similarity metrics in ANTS are unified in the form of a function on the images or the point sets:       $$\textbf{Similarity}[\textbf{fixedImage}, \textbf{movingImage}, \textbf{weight}, \textbf{parameters}].$$
+
+The similarity type for the deformation transformation is specified by
+the \verb -m  option, which contains two parts: simarity type and the parameters inside the brackets.
+(Note: no white spaces exist between parameters.) The possible similarity metrics for volumetric images are:
+\begin{itemize}
+    \item Cross correlation estimate:  \verb " -m  CC[fixedImage,movingImage,weight,radius]". This metric works
+      well for intra-modality and some inter-modality image registration. For example, \verb"-m CC[fixed.nii,moving.nii,1,5]" specifies:
+    \begin{itemize}
+        \item the fixed image: fixed.nii
+        \item the moving image: moving.nii             
+        \item weight for this metric is 1 (the weights are relative to
+          the weights on the $N$ other metrics passed to ANTS --- $N$
+          is unlimited).  So, the weight, $w_i$ on the $i^{th}$ metric
+          will be  $w_i=w_i/ ( \sum_i  w_i )$.
+        \item the region radius for computing cross correlation is 5
+    \end{itemize}
+
+    \item Mutual information: \verb"-m MI[fixedImage,movingImage,weight,number-of-histogram-bins]".
+This metric works both well for intra-modality and inter-modality image registration.
+For example, the first three parameters in \verb"-m MI[fixed.nii,moving.nii,1,32]" is similar to the example above in cross correlation, except that the last parameter means that the number of bins in computing mutual information is 32.
+
+    \item PR: \verb"-m PR[fixedImage,movingImage,weight,radius]". This
+      metric is deprecated.  It still ``works'' but CC is preferred.
+
+    \item Mean square difference: \verb"-m MSQ[fixedImage,movingImage,weight,0]".
+This metric works for intra-modality image registration.
+The last parameter \verb"0" is a padding value of no real meaning.
+For example, \verb"-m MSQ[fixed.nii,moving.nii,1,0]".
+
+\end{itemize}
+
+ANTS also support registration of point sets. The supported formats for point sets can be found in I/O section. The similarity metrics for point sets are:
+\begin{itemize}
+    \item Point set expectation: 
+     \begin{verbatim}
+-m PSE [fixedImage,movingImage,fixedPoints,movingPoints,weight,
+pointSetPercentage,pointSetSigma,boundaryPointsOnly,
+kNeighborhood,PartialMatchingIterations=100000]
+    \end{verbatim}
+    \begin{itemize}
+        \item \verb"fixedImage": defines the space domain of the fixed point set.
+        \item \verb"movingImage": defines the space domain of the moving point set.
+        \item \verb"fixedPoints/Image": defines the coordinates of the fixed point set or label image.
+It can be an image with discrete positive labels, a VTK format point set file, or a text file.
+Details can be found in I/O section (TODO).
+        \item \verb"movingPoints/Image": defines the coordinates of the moving point set or label image. 
+        \item \verb"weight": weight for this metric. \verb"1" weights are relative to the weights on the $N$ other metrics passed to ANTS --- $N$ is unlimited.
+        \item \verb"pointSetPercentage": the percentage of points to be randomly sampled used in the registration.
+        \item \verb"pointSetSigma": the standard deviation of the Parzen window used to estimate the expectation. 
+        \item \verb"boundaryPointsOnly":  \verb"1" (or ``true'') means only the boundary points in the label image is used to drive registration.
+        \item \verb"kNeighborhood" is a positive discrete number. The first $k$ neighbors are used to compute the deformation during the registration. 
+        \item \verb"PartialMatchingIterations" controls the symmtry in the matching.
+This option assumes the complete labeling is in the first set of label parameters ...
+more iterations leads to more symmetry in the matching  - 0 iterations means full asymmetry
+    \end{itemize}
+
+    \item Jensen-Tsallis BSpline
+    \begin{verbatim}
+-m JTB[fixedImage,movingImage,fixedPoints,movingPoints,weight,
+pointSetPercentage,pointSetSigma,boundaryPointsOnly,kNeighborhood,alpha,
+meshResolution,splineOrder,numberOfLevels,useAnisotropicCovariances] 
+    \end{verbatim}
+    \begin{itemize}
+        \item \verb"fixedImage": defines the space domain of the fixed point set.
+        \item \verb"movingImage": defines the space domain of the moving point set.
+        \item \verb"fixedPoints": defines the coordinates of the fixed point set.
+It can be an image with discrete positive labels, a VTK format point set file, or a text file.
+Details can be found in I/O section (TODO).
+        \item \verb"movingPoints": defines the coordinates of the moving point set.
+        \item \verb"weight": weight for this metric.  The weights are relative to the weights on the $N$ other metrics passed to ANTS --- $N$ is unlimited.
+        \item \verb"pointSetPercentage": the percentage of points to be randomly sampled used in the registration.
+        \item \verb"pointSetSigma": the sigma for the Parzen window used to estimate probabilities.  
+        \item \verb"boundaryPointsOnly": [TODO] \verb"1" (or ``true'') means only the boundary points in the point sets are used to drive registration.
+        \item \verb"kNeighborhood" is a positive discrete number. The first $k$ neighbors are used to compute the deformation during the registration. 
+        \item \verb"alpha"
+        \item \verb"meshResolution"
+        \item \verb"splineOrder"
+        \item \verb"numberOfLevels"
+        \item \verb"useAnisotropicCovariances"
+    \end{itemize}
+
+\end{itemize}
+
+ANTS has three affine registration similarity metrics for volumetric images, which are specified using \verb"--affine-metric-type":
+\begin{itemize}
+    \item Mutual information, specified as \verb"--affine-metric-type MI".
+This usually works for both inter and intra-modalities in 3D.
+Also, the options used in computing mututal information in affine registration can be specified as \verb"--MI-option N1xN2".
+The first parameter \verb"N1" specifies the number of bins.
+The second parameter \verb"N2" specifies the number of samples.
+For example: \verb"--MI-option 32x8000".
+    \item Cross-correlation  \verb"--affine-metric-type CC".  CC has a
+      wide capture range.  
+    \item Mean square difference, specified using \verb"--affine-metric-type MSE". There is no options for this metric.
+\end{itemize}
+
+Fig. \ref{fig:metric_example} shows the registration result using intensity difference with the following command:
+\begin{verbatim}
+ANTS 2 -m MSQ[r16slice.nii,r64slice.nii,1,0] -r Gauss[3,0] -t SyN[0.5] -i 50x50x30
+\end{verbatim}
+
+%        \begin{wrapfigure}{r}{0.5\textwidth}
+\begin{figure}
+    \label{fig:metric_example}
+    \centering
+    \begin{tabular}[h]{c|c|c}
+%        \includegraphics[width=0.16\textwidth]{Figures/r16slice.jpg} &
+%        \includegraphics[width=0.16\textwidth]{Figures/r64slice.jpg} &
+ %       \includegraphics[width=0.16\textwidth]{Figures/resMSQ.jpg} \\
+        fixed image &
+        moving image & 
+        MSE  \\
+    \end{tabular} 
+    \itkcaption{registration using mean square intensity difference}
+\end{figure}
+    %    \end{wrapfigure}
+Here is also an example script to register a pair of images using mean square intensity difference and computing the metrics of the registration image.
+\begin{verbatim}
+#use intensity difference with radius 0 -- radius no effect on intensity difference
+ANTS 2 -m MSQ[r16slice.nii,r64slice.nii,1,0] -r Gauss[3,0] -t SyN[0.5] -i 50x50x30
+WarpImageMultiTransform 2 r64slice.nii resMSQ.nii Warp.nii Affine.txt
+MeasureImageSimilarity 2 0 r16slice.nii resMSQ.nii metricexamplelog.txt
+MeasureImageSimilarity 2 1 r16slice.nii resMSQ.nii metricexamplelog.txt
+MeasureImageSimilarity 2 2 r16slice.nii resMSQ.nii metricexamplelog.txt
+ConvertToJpg resMSQ.nii resMSQ.jpg
+\end{verbatim}
 
@@
 \item   Make a directory for this example.   Copy or link all the B*tiff images into it.
-\item Copy the file ANTS/Scripts/buildtemplate.sh into that directory.  
+\item Copy the file ANTS/Scripts/buildtemplateparallel.sh into that directory.  
 \item  On the command line, within that directory, run the following command: 
-\verb  bash ~ \verb buildtemplateparallel.sh -h ~ to get usage. 
+\verb  bash ~ \verb buildtemplateparallel.sh ~ \verb -h ~ to get usage. 
 \item Following the usage, you can write a script that contains a call like this: 
@@
 \end{enumerate}
-The script offers an option ( -c 1  for qsub , -c 2 for voxbo )  so that you can send the whole thing to distributed computing. 
+The script offers an option ( -c 1  for qsub , -c 2 to attempt to use
+multiple cores on your machine )  so that you can send the whole thing to distributed computing. 
 Typically, one would use voxbo or the qsub program available in the Sun Grid Engine.  
@@
 A larger deformation mapping was required and so we increased the maximum number of iterations (the "-i" parameter).
-One might also increase the gradient-descent step-size (SyN[0.25] => SyN[0.5]).
+One might also increase the gradient-descent step-size (SyN[0.25] $\rightarrow$ SyN[0.5]).
 Large deformation mapping is challenging not only because of the amount of deformation, but also because the "true" solution is difficult to find.
@@
 at left.
+
 \newpage
@@
 An advanced MRF segmentation algorithm is available in ANTS via the
-\verb Atropos tool, which is still under development.  We apply 
+\verb Atropos ~tool, which is still under development.  We apply 
 the segmentation to the example data in: ANTS/Examples/Data/r16slice.nii.  
@@
   Atropos -d 2 -a r16slice.nii -o [r16slice_seg.nii.gz,r16slice_prob_%02d.nii.gz] \
-      -c [10,0.001] -m [0.05,1] -i kmeans[3] -x mask.nii.gz
+      -c [10,0.001] -m [0.1,1x1x1] -i kmeans[3] -x mask.nii.gz
 \end{verbatim}
@@
 in figure~\ref{fig:seg} and -- with priors -- in figure~\ref{fig:seg2}.
+The document atropos.pdf in the Documentation directory provides much
+more explanation.
 \begin{figure}
@@
 \includegraphics[width=1\textwidth]{Figures/segmentation2.pdf}
-\itkcaption[ANTS Segmentation]{ANTS extends basic Markov Random Field
-  regularized Gaussian model tissue segmentation to include priors,
+\itkcaption[ANTS Segmentation]{The ANTs program {\em Atropos} extends basic Markov Random Field regularized Gaussian model tissue segmentation to include priors,
   which allow spatially varying tissue models to determine
-  segmentation.  The ANTS program N3BiasFieldCorrection is less
+  segmentation.  The ANTS program N4BiasFieldCorrection is 
   critical as preprocessing when using this approach to segmentation.
@@
 \begin{enumerate}
-\item Preprocess images -- bias correction (e.g. via the \verb
-  N4BiasFieldCorrection program), segmentation (via \verb Atropos ), etc.  Potentially construct an optimal template. 
+\item Preprocess images -- bias correction (e.g. via the \verb N4BiasFieldCorrection ~program), 
+segmentation (via \verb Atropos ), etc.  Potentially construct an optimal template. 
 \item Normalize images -- run \verb ants.sh ~ to map a population of images to a template and store the deformations. 

### 2011-04-14 Documentation/Tex/ants.tex
ENH: a simple implementation of COMPCOR based on input ROIs = CompCorRestingStateConnectivity in ImageMath
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@829 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@
 
+\subsection{Statistical Mapping with ANTS: rsfMRI}
+We are developing methods within ANTs to enable rsfMRI processing.  
+The current strategy is within an ImageMath function called
+CompCorRestingStateConnectivity.   This function will use pre-labeled
+regions of the rsfMRI image to estimate the impact of physiological
+confounds on the rsfMRI signal.  We implement the following steps 
+\begin{enumerate}
+\item Define matrices ${\bf G} , {\bf C},  {\bf R}$ which cover the
+  gray matter region of interest, the confound region of interest and
+  the reference region against which to compare the gray matter. 
+\item Perform PCA on the ${\bf C}$ matrix to get its first N (e.g. 5)
+  eigenvectors.  Use ordinary least squares to factor the effect of
+  these out from the matrices ${\bf G} ,  {\bf R}$.
+\item Get the first and second principal eigenvector from ${\bf R}$,
+  the reference region.  
+\item Correlate these against ${\bf G}$ on a voxel-by-voxel basis. 
+\end{enumerate}
+This has not been validated yet except by performing the same analysis
+on a test-retest dataset with anterior cingulate as the reference
+region.  This highlighted the thalamus, posterior cingulate and
+cerebellum as part of the default mode network in both test and retest data.
+
 \subsection{Statistics with ANTS and {\bf R}}

### 2011-04-14 Utilities/antsMatrixUtilities.h

@@ -1,0 +1,244 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: antsMatrixUtilities.h,v $
+  Language:  C++
+  Date:      $Date: $
+  Version:   $Revision: $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or
+  http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt
+  for details.
+
+  This software is distributed WITHOUT ANY WARRANTY; without even
+  the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
+  PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __antsMatrixUtilities_h
+#define __antsMatrixUtilities_h
+#include <vnl/algo/vnl_matrix_inverse.h>
+#include <vnl/algo/vnl_cholesky.h>
+#include "itkImageToImageFilter.h"
+namespace itk {
+namespace ants {
+
+template<class TInputImage, class TRealType = double>
+class ITK_EXPORT antsMatrixUtilities :
+    public ImageToImageFilter<TInputImage, TInputImage>
+{
+public:
+  /** Standard class typdedefs. */
+  typedef antsMatrixUtilities                     Self;
+  typedef ImageToImageFilter<TInputImage, TInputImage>  Superclass;
+  typedef SmartPointer<Self>                                 Pointer;
+  typedef SmartPointer<const Self>                           ConstPointer;
+
+
+  /** Method for creation through the object factory. */
+  itkNewMacro( Self );
+
+  /** Run-time type information (and related methods). */
+  itkTypeMacro( antsMatrixUtilities, ImageToImageFilter );
+
+  /** Dimension of the images. */
+  itkStaticConstMacro( ImageDimension, unsigned int,
+                       TInputImage::ImageDimension );
+
+  itkStaticConstMacro( MatrixDimension, unsigned int, 2 );
+
+  /** Typedef support of input types. */
[…]

### 2011-04-15 Utilities/antsMatrixUtilities.h
BUG: fix to a simple implementation of COMPCOR based on input ROIs = CompCorRestingStateConnectivity in ImageMath
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@830 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -144,6 +144,14 @@ public:
 
   VectorType GetCovMatEigenvector( MatrixType p, unsigned int evec );
 
+  VectorType AverageColumns( MatrixType p ){
+    unsigned int ncol=p.columns();
+    VectorType v=p.get_column(0);
+    v.fill(0);
+    for (unsigned int i=0; i<ncol; i++) v=v+p.get_column(i);
+    return v/(RealType)ncol;
+  }
+
   MatrixType WhitenMatrix(MatrixType p, RealType regularization=1.e-2 ) {
     MatrixType invcov=this->CovarianceMatrix(p,regularization);
     invcov=this->PseudoInverse( invcov, true );

### 2011-04-16 ImageRegistration/itkANTSImageRegistrationOptimizer.h
ENH: Paul's commit for using all metrics in convergence.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@831 fd2da0a4-d968-48df-9cb5-d8c496287cbf

@@ -1070,6 +1070,13 @@ PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer mo
          * the desired window will start at the user-specified origin and
          * end at the current iteration).
          */
+
+
+        // Added by Paul: allow weighted metric to be used here
+        typename ParserType::OptionType::Pointer regularizationOption
+          = this->m_Parser->GetOption( "use-all-metrics-for-convergence" );
+        bool use_all_metrics = (atoi(regularizationOption->GetValue().c_str()) > 0);
+
 	unsigned int domtar=12;
 	if( this->m_CurrentIteration > domtar )
           {
@@ -1099,7 +1106,22 @@ PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer mo
             point[0] = qq;
             ProfilePointDataType energy;
             energy.Fill( 0 );
-            energyProfiles[0]->GetPointData( qq, &energy );
+            if( use_all_metrics )
+              {
+              for (unsigned int im = 0; im < numberOfMetrics; im++)
+                {
+                ProfilePointType pm;
+                pm[0] = qq;
+                ProfilePointDataType em;
+                energyProfiles[im]->GetPointData( qq, &em );
+                RealType weight = this->m_SimilarityMetrics[im]->GetWeightScalar();
+                energy[0] += weight * em[0];
+                }
+              }
+            else
+              {
+              energyProfiles[0]->GetPointData( qq, &energy );
+              }
             totale+=energy[0];
             energyProfileWindow->SetPoint( qq-windowBegin, point );
             energyProfileWindow->SetPointData( qq-windowBegin, energy );
@@ -1109,7 +1131,22 @@ PointSetPointer  WarpMultiTransform(ImagePointer referenceimage, ImagePointer mo
           for( unsigned int qq = windowBegin; qq < this->m_CurrentIteration; qq++ )
             {
             ProfilePointDataType energy;  energy.Fill(0);
-            energyProfiles[0]->GetPointData( qq, &energy );
+            if( use_all_metrics )
+              {
+              for (unsigned int im = 0; im < numberOfMetrics; im++)
+                {
+                ProfilePointType pm;
+                pm[0] = qq;
+                ProfilePointDataType em;
+                energyProfiles[im]->GetPointData( qq, &em );
+                RealType weight = this->m_SimilarityMetrics[im]->GetWeightScalar();
+                energy[0] += weight * em[0];
+                }
+              }
[…]
